### 2015-03-16 Kokkos_PG_Compilation.tex
Initial commit.

\chapter{Compiling}\label{C:build}

This chapter explains how to compile Kokkos, and how to link your
application against Kokkos.  Kokkos supports two build systems:
\begin{itemize}
\item Using the embedded Makefile
\item Trilinos' CMake build system
\end{itemize}
Note that the two explicitly supported build methods should not be
mixed.  For example, do not include the embedded Makefile in your
application build process, while explicitly linking against a
pre-compiled Kokkos library in Trilinos.  We also include specific
advice for building for NVIDIA GPUs, and for Intel Xeon Phi.

\section{General Information}\label{S:build:gen}

Kokkos consists mainly of header files. 
Only a few functions have to be compiled into object files outside of the application's source code.
Those functions are contained in \verb!.cpp! files inside the \lstinline|kokkos/core/src| directory and its subdirectories.
The files are internally protected with macros to prevent compilation if the related execution space is not enabled. 
Thus, it is not necessary to create a list of included object files specific to your compilation target.
One may simply compile all \verb!.cpp! files. 
The enabled features are controlled via macros which have to be provided in the compilation line or in the \lstinline|KokkosCore_config.h| include file. 
A list of macros can be found in Table \ref{TBL:CompileMacros}.
\begin{table}
\caption{Table of configuration Macros}
\label{TBL:CompileMacros}
\begin{small}
\begin{tabular}[t]{p{0.25\textwidth}p{0.3\textwidth}p{0.42\textwidth}}
\hline\hline
Macro & Effect & Comment \\
\hline
{\tiny KOKKOS\_HAVE\_CUDA} & Enable the CUDA execution space. & Requires a compiler capable of understanding CUDA-C. See Section \ref{S:build:CUDA}. \\
{\tiny KOKKOS\_HAVE\_OPENMP} & Enable the OpenMP execution space. & Requires the compiler to support OpenMP (e.g., \verb!-fopenmp!). \\
{\tiny KOKKOS\_HAVE\_PTHREADS} & Enable the Threads execution space. & Requires linking with libpthread.\\
{\tiny KOKKOS\_HAVE\_CXX11} & Enable internal usage of C++11 features. & The code needs to be compile with the C++11 standard. Most compilers accept the \verb!-std=c++11! flag for this.\\
{\tiny KOKKOS\_HAVE\_HWLOC} & Enable thread and memory pinning via hwloc. & Requires linking with libhwloc.\\
{\tiny KOKKOS\_DONT\_INCLUDE \_KOKKOSCORE\_CONFIG\_H} & Do not include the KokkosCore\_config.h file. & Useful when providing all macros via the compile line instead of auto-generating that file. 
This option is set for the embedded Makefile. \\
\hline\hline
\end{tabular}
\end{small}
\end{table}

\section{Using Kokkos' Makefile system}\label{S:build:make}

Kokkos provides an embedded Makefile (\lstinline|kokkos/Makefile.kokkos|) for inclusion in an application Makefile. 
This embedded Makefile generates a \lstinline|KOKKOS_INC| and a \newline
\lstinline|KOKKOS_LINK| variable which need to be appended to the compile and the link line respectively in the application Makefile. 
If you want to include dependencies (i.e. trigger a rebuild of the application object files when Kokkos files change) you can include \lstinline|KOKKOS_HEADERS| in you dependency list.
Note that you cannot compile and link at the same time!
Makefile.kokkos uses a number of variables which can be set either in the parent Makefile prior to including it, or on the command line.
These variables take the form of \lstinline|SOME_VAR=yes/no| or \lstinline|SOME_PATH=/path/to/library| and control build options such as the target architecture. 
A list can be found in Table \ref{TBL:MakefileOptions}.
Example application Makefiles can be found in the tutorial examples under \verb!kokkos/example/tutorial!.

\begin{table}
\caption{Table of Makefile options}
\label{TBL:MakefileOptions}
\begin{small}
\begin{tabular}[t]{lp{0.7\textwidth}}
\hline\hline
Option & Description \\
\hline
CUDA & Enable the CUDA execution space. \\
OMP & Enable the OpenMP execution space. \\
PTHREADS & Enable the Threads execution space.\\
MIC & Enable compilation for Intel Xeon Phi. \\
CXX11 & Enable internal usage of C++11 features.\\
HWLOC & Enable hwloc usage.\\
LIBRT & Enable librt usage for the \lstinline|Kokkos::Impl::Timer| class.\\
CUDA\_PATH & Set the path to the CUDA toolkit directory.\\
HWLOC\_PATH & Set the path to the hwloc library.\\
\hline\hline
\end{tabular}
\end{small}
\end{table}

\section{Using Trilinos' CMake build system}\label{S:build:Trilinos}

The Trilinos project (see \url{trilinos.org}) is an effort to develop
algorithms and enabling technologies within an object-oriented
software framework for the solution of large-scale, complex
multiphysics engineering and scientific problems.  Trilinos is
organized into packages.  Even though Kokkos is a stand-alone software
project, Trilinos uses Kokkos extensively.  Thus, Trilinos' source
code includes Kokkos' source code, and builds Kokkos as part of its
build process.

Trilinos' build system uses CMake.  Thus, in order to build Kokkos as
part of Trilinos, you must first install CMake (version
\texttt{2.8.12} or newer; CMake \texttt{3.x} works).  
To enable Kokkos when building Trilinos, set the CMake option \verb!Trilinos_ENABLE_Kokkos!.
Trilinos' build system lets packages express dependencies on other packages or external libraries.
If you enable any Trilinos package (e.g., Tpetra) that has a required dependency on Kokkos, 
Trilinos will enable Kokkos automatically.
Configuration macros are automatically inferred from Trilinos settings. 
For example, if the CMake option \lstinline|Trilinos_ENABLE_OpenMP| is ON, Trilinos will define the macro \lstinline|KOKKOS_HAVE_OPENMP|.
Trilinos' build system will autogenerate the previously mentioned \lstinline|KokkosCore_config.h| file that contains those macros.

We refer readers to Trilinos' documentation for details.  Also, the
\texttt{kokkos/config} directory includes examples of Trilinos
configuration scripts.

\section{Building for CUDA}\label{S:build:CUDA}

Any Kokkos application compiled for CUDA embeds CUDA code via template metaprogramming.
Thus, the whole application must be built with a CUDA-capable compiler.
(At the moment, the only such compiler is NVIDIA's NVCC.)
More precisely, every compilation unit containing a Kokkos kernel or a function called from a Kokkos kernel has to be compiled with a CUDA-capable compiler. 
This includes files containing \lstinline|Kokkos::View| allocations, which call an initialization kernel.

The current version of NVCC has some shortcomings when used as the main compiler for a project, in particular when part of a complex build system.
For example, it does not understand most GCC command-line options, which must be prepended by the \lstinline|-Xcompiler| flag when calling NVCC. 
Kokkos comes with a shell script, called \lstinline|nvcc_wrapper|, that wraps NVCC to address these issues.
We intend this as a drop-in replacement for a normal GCC-compatible compiler (e.g., GCC or Intel) in your build system.
It analyzes the provided command-line options and prepends them correctly. 
It also adds the correct flags for compiling generic C++ files containing CUDA code (e.g., \verb!*.cpp!, \verb!*.cxx!, or \verb!*.CC!).
By default \lstinline|nvcc_wrapper| calls \verb!g++! as the host compiler.
You may override this by providing NVCC's '\lstinline|-ccbin|' option as a compiler flag.
The default can be reset by editing the script itself.

Many people use a system like Environment Modules (see \\ \url{http://modules.sourceforge.net/}) to manage their shell environment.
When using a module system, it can be useful to provide different versions for different back-end compiler types (e.g., \verb!icpc!, \verb!pgc++!, \verb!g++!, and \verb!clang!).
To use the \lstinline|nvcc_wrapper| in conjunction with MPI wrappers, simply overwrite which C++ compiler is called by the MPI wrapper. 
For example, you can reset OpenMPI's C++ compiler by setting the \lstinline|OMPI_CXX| environment variable.
Make sure that \lstinline|nvcc_wrapper| calls the host compiler with which the MPI library was compiled.

\section{Building for Intel Xeon Phi}\label{S:build:Phi}

TODO

### 2015-03-16 Kokkos_PG_HierarchicalParallelism.tex
Initial commit.

\chapter{Hierarchical Parallelism}\label{C:Hierarchical}
Node architectures on modern high-performance computers are characterized by ever more \emph{hierarchical parallelism}. 
A level in the hierachy is determined by the hardware resources which are shared between compute units at that level. 
Higher levels in the hierarchy also have access to all resources in its branch at lower levels of the hierarchy.
This concept is orthogonal to the concept of heterogeneity. 
For example, a node in a typical CPU-based cluster consists of a number of multicore CPUs.  Each core supports one or more hyperthreads, and each hyperthread can execute vector instructions.
This means there are 4 levels in the hierarchy of parallelism: 
\begin{enumerate}
\item CPU sockets share access to the same memory and network resources,
\item cores within a socket typically have a shared last level cache (LLC), 
\item hyperthreads on the same core have access to a shared L1 (and L2) cache and they submit instructions to the same execution units, and
\item vector units execute a shared instruction on multiple data items.
\end{enumerate}
GPU-based systems also have a hierarchy of 4 levels:
\begin{enumerate}
\item Multiple GPUs in the same node share access to the same host memory and network resources, 
\item core clusters (e.g. the SMs on an NVIDIA GPU) have a shared cache and access to the same high bandwidth memory on a single GPU, 
\item threads running on the same core cluster have access to the same L1 cache and scratch memory and they are 
\item grouped in so called Warps or Wave Fronts within which threads are always synchronous and can collaborate more closely for example via direct register swapping. 
\end{enumerate}
Kokkos provides a number of abstract levels of parallelism which at compile time or run time are mapped to the appropriate hardware features. 
This mapping is not necessarily static or predefined; it may differ for each kernel.
Furthermore, some mapping decisions happen at run time. 
This enables adaptive kernels which map work to different hardware resources depending on the work set size.
While Kokkos provides defaults and suggestions, the optimal mapping can be algorithm dependent. 
Hierarchical parallelism is accessible through execution policies.

\section{Thread Teams Basics}

Kokkos' most basic hierarchical parallelism concept is a thread team.
A \emph{thread team} is a collection of threads which can synchronize,
and which share a ``scratch pad'' memory
(see Section\ref {S:Hierarchical:Scratch}).

Instead of mapping a 1-D range of indices to hardware resources,
Kokkos' thread teams map a 2-D index range.
The first index is the \emph{league}, the index of the team.
The second index is the thread index within a team.
In CUDA this is equivalent to launching a 1-D grid of 1-D blocks.
The league size is arbitrary -- that is, it is only limited by the integer size type -- while the team size must fit in the hardware constraints.
As in CUDA, only a limited number of teams are actually active at the same time,
and they must run to completion before new ones are executed. 
Consequently it is not valid to use inter thread-team synchronization mechanisms
such as waits for events initiated by other thread teams.

\subsection{Creating a Policy Instance}

Kokkos exposes use of thread teams with the \lstinline!Kokkos::TeamPolicy! execution policy.
To use thread teams you need to create a \lstinline|Kokkos::TeamPolicy| instance.
It can be created inline for the parallel dispatch call.
The constructors requires two arguments: a league size and a team size. 
As with the  \lstinline|Kokkos::RangePolicy| a specific execution tag and a specific execution space can be given as optional template arguments.
\begin{lstlisting}
// Using default execution space and launching 
// a league with league_size teams with team_size threads each
Kokkos::TeamPolicy<> 
        policy( league_size, team_size );

// Using  a specific execution space to 
// run a n_worksets x team_size parallelism
Kokkos::TeamPolicy<ExecutionSpace> 
        policy( league_size, team_size );

// Using a specific execution space and an execution tag 
Kokkos::TeamPolicy<SomeTag, ExecutionSpace> 
        policy( league_size, team_size ); 
\end{lstlisting}

\subsection{Basic Kernels}
Within the kernel the team policy member type provides the necessary functionality to use teams. 
It allows access to thread identifiers such as league rank and size and team rank and size, as well as team synchronous actions such as team barriers, reductions and scans.
\begin{lstlisting}
using Kokkos::TeamPolicy;
using Kokkos::parallel_for;

typedef TeamPolicy<ExecutionSpace>::member_type member_type;
// Create an instance of the policy
TeamPolicy<ExecutionSpace> policy (league_size, team_size);
// Launch a kernel
parallel_for (policy, KOKKOS_LAMBDA (member_type team_member) {
    // Calculate a global thread id
    int k = team_member.league_rank () * team_member.team_size () +
            team_member.team_rank ();
    // Calculate the sum of the global thread ids of this team
    int team_sum = team_member.reduce (k);
    // Atomicly add the value to a global value
    a() += team_sum;
  });
\end{lstlisting}

This approach of writing hierarchical parallelism algorithms makes it very explicit that a kernel using a \lstinline!TeamPolicy! constitutes a parallel region with respect to the team.

\section{Team Scratch Pads}\label{S:Hierarchical:Scratch}

Each Kokkos team has a ``scratch pad.''
This is an instance of a memory space accessible only by threads in that team.
Scratch pads let an algorithm load a workset into a shared space
and then collaboratively work on it with all members of a team.
The lifetime of data in a scratch pad is the lifetime of the team.
In particular, scratch pads are recycled by all logical teams running on the same physical set of cores.
During the lifetime of the team all operations allowed on global memory are allowed on the scratch memory.
This includes taking addresses and performing atomic operations on elements located in scratch space. 
Team-level scratch pads correspond to the per-block shared memory in Cuda,
or to the ``local store'' memory on the Cell processor.

Kokkos exposes scratch pads through a special memory space associated with the execution space:
\lstinline|execution_space::scratch_memory_space|.
You may allocate a chunk of scratch memory through the \lstinline|TeamPolicy| member type.
You may request multiple allocations from scratch, up to a user-provided maximum. 
The maximum is provided either through a function in the functor which returns a potentially team-size dependent value, 
or it can be specified as an argument to the constructor of the TeamPolicy. 
It is not valid to provide both values at the same time. 
The argument to the TeamPolicy can be used to set the shared memory size when using functors. 
One restriction on shared memory allocations is that they can not be freed during the lifetime of the team. 
This avoids the complexity of a memory pool,
and reduces the time it takes to obtain an allocation
(which currently is a few operations to calculate the offset).

\begin{lstlisting}
template<class ExecutionSpace>
struct functor {
  typedef ExecutionSpace execution_space;
  typedef execution_space::member_type member_type;

  KOKKOS_INLINE_FUNCTION
  void operator() (member_type team_member) const {
    size_t double_size = 5*team_member.team_size()*sizeof(double);

    // Get a shared team allocation on the scratch pad
    double* team_shared_a = (double*)
      team_member.team_shmem().get_shmem(double_size);

    // Get another allocation on the scratch pad
    int* team_shared_b = (int*)
      team_member.team_shmem().get_shmem(160*sizeof(int));

    // ... use the scratch allocations ...
  }

  // Provide the shared memory capacity.
  // This function takes the team_size as an argument,
  // which allows team_size dependent allocations.
  size_t team_shmem_size (int team_size) const {
    return sizeof(double)*5*team_size +
           sizeof(int)*160;
  }
};
\end{lstlisting}

Instead of simply getting raw allocations in memory, users can also allocate Views directly in scratch memory. 
This is achieved by providing the shared memory handle as the first argument of the View constructor.
Views also have a static member function which return their shared memory size requirements. 
The function expects the run-time dimensions as arguments, corresponding to View's constructor.

\begin{lstlisting}
tyepdef Kokkos::DefaultExecutionSpace::scratch_memory_space
  ScratchSpace;
// Define a view type in ScratchSpace
typedef Kokkos::View<int*[4],ScratchSpace> shared_int_2d;

// Get the size of the shared memory allocation
size_t shared_size = shared_int_2d::shmem_size(team_size);
Kokkos::parallel_for(Kokkos::TeamPolicy<>(league_size,team_size),
                     KOKKOS_LAMBDA ( member_type team_member) {
  // Get a view allocated in team shared memory.
  // The constructor takes the shared memory handle and the 
  // runtime dimensions
  shared_int_2d A(team_member.team_shmem(), team_member.team_size());
  ...

});
\end{lstlisting}

\section{Nested Parallelism}

Instead of writing code which explicitly uses league and team rank indices, one can use nested parallelism to implement hierarchical algorithms.
Kokkos lets the user have up to three nested layers of parallelism.
The team and thread levels are the first two levels.
The third level is \emph{vector} parallelism.

You may use any of the three parallel patterns -- for, reduce, or scan -- at each level.
You may nest them and use them in conjunction with code that is aware of the league and team rank.
The different layers are accessible via special execution policies:
\lstinline|TeamThreadLoop| and \lstinline|ThreadVectorLoop|.

\subsection{Team and Vector Loops}

The first nested level of parallel loops splits an index range over the threads of a team. 
This motivates the policy name \lstinline|TeamThreadRange|, 
which indicates that the loop is executed once by the team with the index range split over threads.
The loop count is not limited to the number of threads in a team, and how the index range is mapped to threads is architecture dependent.
It is not legal to nest multiple parallel loops using the \lstinline!TeamThreadRange! policy.
However, it is valid to have multiple parallel loops using the \lstinline!TeamThreadRange! policy follow each other in sequence, in the same kernel.  
Note that it is not legal to make a write access to POD data outside of the closure of a nested parallel layer. 
This is a conscious choice to prevent difficult to debug issues related to thread private, team shared and globally shared variables. 
A simple way to enforce this is by using the ``capture by value'' clause with lambdas. 
With the lambda being considered as \lstinline|const| inside the \lstinline!TeamThreadRange! loop,
the compiler will catch illegal accesses at compile time as a \lstinline|const| violation.

The most simple use case is to have another \lstinline|parallel_for| nested inside a kernel. 
\begin{lstlisting}
Kokkos::parallel_for(Kokkos::TeamPolicy<>(league_size,team_size), 
                     KOKKOS_LAMBDA ( member_type team_member) {
  Scalar tmp;
  Kokkos::parallel_for(
    Kokkos::TeamThreadRange(team_member,loop_count), 
    [=] (int& i) {
    ....
    // tmp += i; // This would be an illegal access
  });
});
\end{lstlisting}

The \lstinline|parallel_reduce| construct can be used to perform optimized team-level reductions:

\begin{lstlisting}
using Kokkos::parallel_reduce;
using Kokkos::TeamPolicy;
using Kokkos::TeamThreadRange;
parallel_for (TeamPolicy<> (league_size, team_size),
                 KOKKOS_LAMBDA (member_type team_member) {
    // The default reduction uses Scalar's += operator
    // to combine thread contributions.
    Scalar sum;
    parallel_reduce (TeamThreadRange (team_member, loop_count), 
      [=] (int& i, Scalar& lsum) {
        // ... 
        lsum += ...;
      }, sum);

    // You may provide a custom reduction as another
    // lambda together with an initialization value.
    Scalar product;
    Scalar init_value = 1;
    parallel_reduce (TeamThreadRange (team_member, loop_count), 
      [=] (int& i, Scalar& lsum) {
        // ...
        lsum *= ...;
      }, product, [=] (Scalar& lsum, Scalar& update) {
        lsum *= update; 
      }, init_value);
  });
\end{lstlisting}

The third pattern is \lstinline|parallel_scan| which can be used to perform prefix scans.

\subsection{Execution Restriction with \lstinline|single|}

As stated in the earlier subsections, a kernel has to be treated as a parallel region with respect to threads (and vector lanes) within a team.
This means that global memory accesses outside of the respective nested levels potentially have to be protected against repetitive execution. 
A common example is the case where a team performs some calculation but only one result per team has to be written back to global memory. 
Kokkos provides the \lstinline|Kokkkos::single(Policy,Lambda)| function for this case.
The two policies \lstinline|Kokkos::PerTeam| and \lstinline|Kokkos::PerThread| restrict execution to once per thread or once per team.

\begin{lstlisting}
using Kokkos::parallel_for;
using Kokkos::parallel_reduce;
using Kokkos::TeamThreadRange;
using Kokkos::ThreadVectorRange;
using Kokkos::PerThread;
// ...
parallel_for (TeamThreadRange (thread, 100), [=] (int i) {
    double sum = 0;
    // Perform a vector reduction with a thread
    parallel_reduce (ThreadVectorRange (thread, 100), 
      [=] (int i, double& lsum) {
        // ...
        lsum += ...;
      }, sum);
    // Add the result value into a team shared array.
    // Make sure it is only added once per thread.
    Kokkos::single (PerThread (), [=] () {
        shared_array(i) += sum;
      });
  });

double sum;
parallel_reduce (TeamThreadRange (thread, 99), 
  [=] (int i, double& lsum) {
    // Add the result value into a team shared array
    // Make sure its only added once per thread
    Kokkos::single (PerThread (thread), [=] () {
        lsum += someFunction (shared_array(i), 
                              shared_array(i+1));
      });
  }, sum);

// Add the per team contribution to global memory
Kokkos::single (PerTeam (thread), [=] () {
    global_array(thread.league_rank()) = sum;
  });

// ...
\end{lstlisting}

### 2015-03-16 Kokkos_PG_Initialization.tex
Initial commit.

\chapter{Initialization}\label{C:init}

In order to use Kokkos an initialization call is required. 
That call is responsible for aquiring hardware resources such as threads. 
Typically this call should be placed right at the start of a program.
If you use both MPI and Kokkos, your program should initialize Kokkos right after calling \lstinline!MPI_Init!.
That way, if MPI sets up process binding masks,
Kokkos will get that information and use it for best performance.
Your program must also \emph{finalize} Kokkos when done using it,
in order to free hardware resources.

\section{Initialization by command-line arguments}\label{S:init:cmdline}

The simplest way to initialize Kokkos is by calling the following function:
\begin{lstlisting}
Kokkos::initialize(int& argc, char* argv[]); 
\end{lstlisting}
Just like \lstinline!MPI_Init!, this function interprets command-line arguments to determine the requested settings.
Also like \lstinline!MPI_Init!, it reserves the right to remove command-line arguments from the input list.
This is why it takes \lstinline!argc! by reference, rather than by value; it may change the value on output.

This function will initialize the default execution space
\begin{lstlisting}
Kokkos::DefaultExecutionSpace;
\end{lstlisting}
and its default host execution space
\begin{lstlisting}
Kokkos::DefaultHostExecutionSpace;
\end{lstlisting}
if applicable.
These defaults depend on the Kokkos configuration.
Kokkos chooses the two spaces using the following list, ordered from low to high:
\begin{enumerate}
\item \lstinline|Kokkos::Serial| 
\item \lstinline|Kokkos::Threads|
\item \lstinline|Kokkos::OpenMP|
\item \lstinline|Kokkos::Cuda| 
\end{enumerate}
The highest execution space in the list which is actually enabled is Kokkos' default execution space,
and the highest enabled host execution space is Kokkos' default host execution space.
(Currently, the only non-host execution space is \lstinline!Cuda!.)
For example, if  \lstinline|Kokkos::Cuda|, \lstinline|Kokkos::OpenMP|, and 
\lstinline|Kokkos::Serial| are enabled, then \lstinline|Kokkos::Cuda| is the
default execution space and \lstinline|Kokkos::OpenMP| is the default host execution 
space.\footnote{This is the preferred set of defaults when CUDA and OpenMP are enabled.
If you use a thread-parallel host execution space, we prefer Kokkos' OpenMP back-end, 
as this ensures compatibility of Kokkos' threads with the application's direct use of OpenMP threads.
Kokkos cannot promise that its Threads back-end will not conflict with the application's direct use of operating system threads.}

Command-line arguments come in ``prefixed'' and ``non-prefixed'' versions.
Prefixed versions start with the string \verb!--kokkos-!.
\lstinline!Kokkos::initialize! will remove prefixed options from the input list,
but will preserve non-prefixed options.
Argument options are given with an equals (\verb!=!) sign.
If the same argument occurs more than once, the last one counts. 
Furthermore, prefixed versions of the command line arguments take precedence over the non-prefixed ones.
For example, the arguments
\begin{verbatim}
--kokkos-threads=4 --threads=2
\end{verbatim}
set the number of threads to 4, while
\begin{verbatim}
--kokkos-threads=4 --threads=2 --kokkos-threads=3
\end{verbatim}
set the number of threads to 3.
Table \ref{TBL:CommandLineOptions} gives a full list of command-line options.

\begin{table}
\caption{Command-line options for \lstinline|Kokkos::initialize|}
\label{TBL:CommandLineOptions}
\begin{small}
\begin{tabular}[t]{lp{0.5\textwidth}}
\hline
Argument & Description \\\hline
\lstinline|--kokkos-help| & print this message \\
\lstinline|--kokkos-threads=INT| & 
specify total number of threads or number of threads per NUMA region if used in conjunction with '--numa' option. \\
\lstinline|--kokkos-numa=INT| & specify number of NUMA regions used by process. \\
\lstinline|--kokkos-device=INT| & specify device id to be used by Kokkos. \\
\lstinline|--kokkos-ndevices=INT[,INT]| & used when running MPI jobs. Specify number of
devices per node to be used. Process to device
mapping happens by obtaining the local MPI rank
and assigning devices round-robin. The optional
second argument allows for an existing device
to be ignored. This is most useful on workstations
with multiple GPUs, of which one is used to drive
screen output.\\
\hline
\end{tabular}
\end{small}
\end{table}

\section{Initialization by struct}\label{S:init:struct}

Instead of giving \lstinline|Kokkos::initialize()| command-line arguments,
one may directly pass in initialization parameters, using the following struct:
\begin{lstlisting}
struct Kokkos::InitArguments {
  int num_threads;
  int num_numa;
  int device_id;
  // ... the struct may have more members ...
};
\end{lstlisting}
The \lstinline|num_threads| field corresponds to the \verb!--kokkos-threads! command-line argument,
\lstinline|num_numa| to \verb!--kokkos-numa!, and \lstinline|device_id| to \verb!--kokkos-device!.
(See Table \ref{TBL:CommandLineOptions} for details.)
Not all parameters are observed by all execution spaces, and the struct might expand in the future if needed.

If you set \lstinline|num_threads| or \lstinline|num_numa| to zero or less, 
Kokkos will \emph{not} initialize the default host execution space.
Note that the default value of each parameter is -1.
Thus, you must set both \lstinline|num_threads| and \lstinline|num_numa|
in order for Kokkos to initialize the default host execution space.

Here is an example of how to use the struct.
\begin{lstlisting}
Kokkos::InitArguments args;
// 8 (CPU) threads per NUMA region
args.num_threads = 8;
// 2 (CPU) NUMA regions per process
args.num_numa = 2;
// If Kokkos was built with CUDA enabled,
// use the GPU with device ID 1.
args.device_id = 1;

Kokkos::initialize(args);
\end{lstlisting}

\section{Initializing non-default execution spaces}\label{S:init:specific}

Instead of calling the generic initialization, one can call initialization for each execution space on its own. 
If the associated host execution space of an execution space is not identical to the latter, it has to be initialized first.
For example when compiling with support for pthreads and Cuda, \lstinline|Kokkos::Threads| has to be initialized before \lstinline|Kokkos::Cuda|.
The initialization calls take different arguments for each of the execution spaces.

If you want to initialize an execution space other than those that Kokkos initializes by default,
you \emph{must} initialize it explicitly in code, by calling its \lstinline!initialize()! method and passing it the appropriate arguments.
Furthermore, if the associated host execution space of an execution space is not identical to the latter, the host execution space must be initialized first.
For example, if you have set the default execution space to \lstinline!Kokkos::Serial!, 
but want to use \lstinline!Kokkos::Cuda! and \lstinline!Kokkos::OpenMP!, 
you must initialize \lstinline!Kokkos::OpenMP! before \lstinline!Kokkos::Cuda!.
The initialization calls take different arguments for each of the execution spaces.

We do \emph{not} recommend initializing multiple host execution spaces,
because the different spaces may fight over threads or other hardware resources.
This may have a big effect on performance.
The only exception is the \lstinline!Serial! execution space.
You may safely initialize \lstinline!Serial! along with another host execution space.

\section{Finalization}\label{S:init:finalize}

At the end of each program Kokkos needs to be shut down in order to free resources. 
Do this by calling \lstinline!Kokkos::finalize()!.
You may wish to set this to be called automatically at program exit,
either by setting an \lstinline!atexit! hook
or by attaching the function to \lstinline!MPI_COMM_SELF!,
so that it is called automatically at \lstinline!MPI_Finalize!.

### 2015-03-16 Kokkos_PG_Introduction.tex
Initial commit.

\chapter{Introduction}\label{C:intro}

The field of High Performance Computing (HPC) is entering a new era.
Since the mid 1990's, HPC applications have had great success 
with distributed-memory parallel computing using the Message-Passing Interface (MPI).
However, after two decades of MPI dominance, 
increasing core counts on standard processors and the emergence of new throughput-oriented processors, such as GPUs, make a new approach necessary.
MPI will not necessarily go away,
but application developers will need to expose finer-grained shared-memory parallelism
\emph{as well as} coarser-grained MPI-level parallelism,
in order to get their money's worth on new architectures.

Developers have many shared-memory parallel programming model options today.
Those include OpenMP, OpenACC, CUDA, TBB, Cilk, OpenCL, and C++AMP, among others.
Some of these programming models are vendor specific, while others have varying degrees of support on different hardware. 
Most of the models lack a concept to handle complex memory hierarchies, 
which in addition to threading is one of the defining upcoming trends in HPC.

The Kokkos programming model addresses the increasing complexity and variety of hardware architectures
through a rigorous introduction of parallel execution and memory handling abstractions.
These abstraction layers isolate a software developer from hardware details and allow the implementation of performance-portable applications.

This Guide covers both the programming model, and its implementation as an embedded C++ library.
It describes the abstraction concepts employed in Kokkos and their exposure as a C++ interface.
We intend it as an introduction for new users as well as a reference for application developers already using Kokkos.
We have also supplemented the Guide with tutorial examples in the \verb!example! subdirectory of Kokkos' source code.

### 2015-03-16 Kokkos_PG_MachineModel.tex
Initial commit.

\chapter{Machine Model}

The abstract machine model of Kokkos allows for very complex heterogenous node architectures containing N execution spaces and M memory spaces.

### 2015-03-16 Kokkos_PG_ParallelDispatch.tex
Initial commit.

\chapter{Parallel dispatch}

You probably started reading this Guide because you wanted to learn
how Kokkos can parallelize your code.  This chapter will teach you
different kinds of parallel operations that Kokkos can execute.  We
call these operations collectively \emph{parallel dispatch}, because
Kokkos ``dispatches'' them for execution by a particular execution
space.  Kokkos provides three different parallel operations:
\begin{itemize}
\item \lstinline!parallel_for! implements a ``for loop'' with
  independent iterations.
\item \lstinline!parallel_reduce! implements a reduction.
\item \lstinline!parallel_scan! implements a prefix scan.
\end{itemize}
Kokkos gives users two options for defining the body of a parallel
loop: functors and lambdas.  It also lets users control how the
parallel operation executes, by specifying an \emph{execution policy}.

\section{Specifying the parallel loop body}

\subsection{Functors}

A \emph{functor} is one way to define the body of a parallel loop.
It is a class or struct\footnote{A ``struct'' in C++ is just a class, all of whose members are public by default.} with a public \lstinline!operator()! instance method.
That method's arguments depend on both which parallel operation you want to execute (for, reduce, or scan), and on the loop's execution policy (e.g., range or team).
For an example of a functor, see the section in this chapter for each type of parallel operation.
In the most common case of a \lstinline!parallel_for!, it takes an integer argument which is the for loop's index.
Other arguments are possible; see Chapter \ref{C:Hierarchical} on ``hierarchical parallelism.''

The \lstinline|operator()| method must be const, and must be marked with the \\ \lstinline!KOKKOS_INLINE_FUNCTION! macro.
If building with CUDA, this macro will mark your method as suitable for running on the CUDA device (as well as on the host).
If not building with CUDA, the macro is unnecessary but harmless.  Here is an example of the signature of such a method:
\begin{lstlisting}
KOKKOS_INLINE_FUNCTION void operator() (...) const;
\end{lstlisting}

The entire parallel operation (for, reduce, or scan) shares the same instance of the functor.
However, any variables declared inside the \lstinline!operator()! method are local to that iteration of the parallel loop.
Kokkos may pass the functor instance by ``copy,'' not by pointer or reference, to the execution space that executes the code. 
In particular, the functor might need to be copied to a different execution space than the host.  For this reason, it is generally not valid to have any pointer or reference members in the functor.
Pass in Kokkos Views by copy as well; this works by shallow copy.
The functor is also passed as a const object, so it is not valid to change members of the functors.
(However, it is valid for the functor to change the contents of, for example, a View or a raw array which is a member of the functor.)

\subsection{Lambdas}

The 2011 version of the C++ standard (``C++11'') provides a new language construct, the \emph{lambda}, also called ``anonymous function'' or ``closure.''
Kokkos lets users supply parallel loop bodies as either functors (see above) or lambdas.
Lambdas work like automatically generated functors.
Just like a class, a lambda may have state.  The only difference is that with a lambda, the state comes in from the environment.
(The name ``closure'' means that the function ``closes over'' state from the environment.)
Just like with functors, lambdas must bring in state by ``value'' (copy), not by reference or pointer.

By default, lambdas capture nothing (as the default capture specifier \lstinline![]! specifies).
This is not likely to be useful, since parallel for generally works by its side effects.
Thus, we recommend using the ``capture by value'' specifier \lstinline![=]! by default.
You may also explicitly specify variables to capture, but they must be captured by value.

\subsection{Should I use a functor or a lambda?}

Kokkos lets users choose whether to use a functor or a lambda.
Lambdas are convenient for short loop bodies.
For a much more complicated loop body,
you might find it easier for testing to separate it out and name it as a functor.
Lambdas by definition are ``anonymous functions,'' meaning that they have no name.
This makes it harder to test them.
Furthermore, if you would like to use lambdas with CUDA,
you must have a sufficiently new version of CUDA.
Finally, the ``execution tag'' feature, which lets you put together several different parallel loop bodies into a single functor,
only works with functors.  (See Chapter \ref{C:Hierarchical} for details.)

\subsection{Specifying the execution space}

If a functor has an \lstinline|execution_space| public typedef, a parallel dispatch will only run the functor in that execution space.
That's a good way to mark a functor as specific to an execution space.  
If the functor lacks this typedef, \lstinline|parallel_for| will run it in the default execution space, unless you tell it otherwise (that's an advanced topic; see ``execution policies'').
Lambdas do not have typedefs, so they run on the default execution space, unless you tell Kokkos otherwise.

\section{Parallel for}

The most common parallel dispatch operation is a \lstinline|parallel_for| call.
It corresponds to the OpenMP construct \lstinline!#pragma omp parallel for!.
Parallel for splits the index range over the available hardware resources and executes the loop body in parallel.
Each iteration is executed independently.
Kokkos promises nothing about the loop order or the amount of work which actually runs concurrently. 
This means in particular that not all loop iterations are active at the same time.  Consequently, it is not legal to use wait constructs (e.g., wait for a prior iteration to finish). 
Kokkos also doesn't guarantee that it will use all available parallelism. 
For example, it can decide to execute in serial if the loop count is very small, and it would typically be faster to run in serial instead of introducing parallelization overhead.

The lambda or the \lstinline!operator()! method of the functor takes one argument.  That argument is the parallel loop ``index.''
The type of the index depends on the execution policy used for the \lstinline!parallel_for!.  It is an integer type for the implicit or explicit \lstinline|RangePolicy|.
The former is used if the first argument to \lstinline|parallel_for| is an integer.

\section{Parallel reduce}

Kokkos' \lstinline!parallel_reduce! operation implements a reduction.
It is like \lstinline!parallel_for!, except that each iteration
produces a value, and the values are accumulated into a single value
with a user-specified associative binary operation.  It corresponds to
the OpenMP construct \lstinline!#pragma omp parallel reduction!, but
with fewer restrictions on the reduction operation.

The lambda or the \lstinline!operator()! method of the functor takes two arguments.
The first argument is the parallel loop ``index.''
The type of the index depends on the execution policy used for the \lstinline!parallel_reduce!.
If you give \lstinline!parallel_reduce! an integer range as its first argument,
or use \lstinline!RangePolicy! explicitly,
then the first argument of the lambda or \lstinline!operator()! method is an integer index.
Its second argument is a nonconst reference to the type of the reduction result.

\subsection{Example using lambda}

Here is an example reduction using a lambda,
where the reduction result is a \lstinline!double!.
\begin{lstlisting}
const size_t N = ...;
View<double*> x ("x", N);
// ... fill x with some numbers ... 
double sum = 0.0;
parallel_reduce (N, [=] (const int i, double& update) {
    update += x(i); }, sum);
\end{lstlisting}
This version of \verb!parallel_reduce! is easy to use,
but it imposes some assumptions on the reduction.
For example, it assumes that it is correct for threads to join their intermediate reduction results using binary \lstinline!operator+!.
If you want to change this, you must either implement your own reduction result type with a custom binary \lstinline!operator+!,
or define the reduction using a functor instead of a lambda.

\subsection{Example using functor}

The following example shows a reduction using the \emph{max-plus semiring},
where \lstinline!max(a,b)! corresponds to addition and ordinary addition corresponds to multiplication:
\begin{lstlisting}
class MaxPlus {
public:
  // Kokkos reduction functors need the value_type typedef.
  // This is the type of the result of the reduction.
  typedef double value_type;

  // Just like with parallel_for functors, you may specify 
  // an execution_space typedef.  If not provided, Kokkos
  // will use the default execution space by default.

  // Since we're using a functor instead of a lambda,
  // the functor's constructor must do the work of capturing
  // the Views needed for the reduction.
  MaxPlus (const View<double*>& x) : x_ (x) {}

  // This is helpful for determining the right index type,
  // especially if you expect to need a 64-bit index.
  typedef View<double*>::size_type size_type;

  KOKKOS_INLINE_FUNCTION void
  operator() (const size_type i, value_type& update) const
  { // max-plus semiring equivalent of "plus"
    if (update < x_(i)) {
      update = x_(i);
    }
  }

  // "Join" intermediate results from different threads.
  // This should normally implement the same reduction 
  // operation as operator() above.  Note that both input
  // arguments MUST be declared volatile.
  KOKKOS_INLINE_FUNCTION void
  join (volatile value_type& dst,
        const volatile value_type& src) const
  { // max-plus semiring equivalent of "plus"
    if (dst < src) {
      dst = src;
    }
  }

  // Tell each thread how to initialize its reduction result.
  KOKKOS_INLINE_FUNCTION void
  init (value_type& dst) const
  { // The identity under max is -Inf.
    // Kokkos does not come with a portable way to access 
    // floating-point Inf and NaN.  Trilinos does, however; 
    // see Kokkos::ArithTraits in the Tpetra package.
#ifdef __CUDA_ARCH__
    return -CUDART_INF;
#else
    return strtod ("-Inf", (char**) NULL);
#endif // __CUDA_ARCH__
  }

private:
  View<double*> x_;
};
\end{lstlisting}

This example shows how to use the above functor:
\begin{lstlisting}
const size_t N = ...;
View<double*> x ("x", N);
// ... fill x with some numbers ...

//  Trivial reduction in max-plus semiring is -Inf.
double result = strtod ("-Inf", (char**) NULL);
parallel_reduce (N, MaxPlus (x), result);
\end{lstlisting}

\subsection{Example using functor with default join and init}

If your functor does not supply a \lstinline!join! method with the correct signature,
Kokkos will supply a default \lstinline!join! that uses binary \lstinline!operator+!.
Likewise, if your functor does not supply an \lstinline!init! method with the correct signature,
Kokkos will supply a default \lstinline!init! that sets the reduction result to zero.

Here is an example of a reduction functor that computes the sum of squares
of the entries of a View.
Since it does not implement the \lstinline!join! and \lstinline!init! methods,
Kokkos will supply defaults.
\begin{lstlisting}
struct SquareSum {
  // Specify the type of the reduction value with a "value_type"
  // typedef.  In this case, the reduction value has type int.
  typedef int value_type;

  // The reduction functor's operator() looks a little different than
  // the parallel_for functor's operator().  For the reduction, we
  // pass in both the loop index i, and the intermediate reduction
  // value lsum.  The latter MUST be passed in by nonconst reference.
  // (If the reduction type is an array like int[], indicating an
  // array reduction result, then the second argument is just int[].)
  KOKKOS_INLINE_FUNCTION
  void operator () (const int i, value_type& lsum) const {
    lsum += i*i; // compute the sum of squares
  }
};

// Use the above functor to compute the sum of squares from 0 to N-1.
const size_t N = ...;
sum = 0;
// parallel_reduce needs an instance of SquareSum, 
// so we invoke its constructor with ().
parallel_reduce (N, SquareSum ());
\end{lstlisting}
This example has a short enough loop body that it would be better to use a lambda:
\begin{lstlisting}
// Compute the sum of squares from 0 to N-1.
int sum = 0;
parallel_reduce (N, [=] (const int i, int& lsum) { lsum += i*i; });
\end{lstlisting}

\subsection{Reductions with an array of results}

Kokkos lets you compute reductions with an array of reduction results,
as long as that array has a (run-time) constant number of entries.
This currently only works with functors.  Here is an example functor
that computes column sums of a 2-D View.
\begin{lstlisting}
struct ColumnSums {
  // In this case, the reduction result is an array of float.
  // Note the C++ notation for an array typedef.
  typedef float value_type[];

  typedef View<float**>::size_type size_type;

  // Tell Kokkos the result array's number of entries.
  // This must be a public value in the functor.
  size_type value_count;

  View<float**> X_;

  // As with the above examples, you may supply an
  // execution_space typedef.  If not supplied, Kokkos
  // will use the default execution space for this functor.

  // Be sure to set value_count in the constructor.
  ColumnSums (const View<float**>& X) : 
    value_count (X.dimension_1 ()), // # columns in X
    X_ (X)
  {}

  // value_type here is already a "reference" type,
  // so we don't pass it in by reference here.
  KOKKOS_INLINE_FUNCTION void
  operator() (const size_type i, value_type sum) const {
    // You may find it helpful to put pragmas above
    // this loop to convince the compiler to vectorize it.
    // This is probably only helpful if the View type
    // has LayoutRight.
    for (size_type j = 0; j < value_count; ++j) {
      sum[j] += X_(i, j);
    }
  }

  // value_type here is already a "reference" type,
  // so we don't pass it in by reference here.
  KOKKOS_INLINE_FUNCTION void
  join (volatile value_type dst,
        volatile value_type src) const {
    for (size_type j = 0; j < value_count; ++j) {
      dst[j] += src[j];
    }
  }

  KOKKOS_INLINE_FUNCTION void init () const {
    for (size_type j = 0; j < value_count; ++j) {
      sum[j] += 0.0;
    }
  }
};
\end{lstlisting}
We show how to use this functor here:
\begin{lstlisting}
  const size_t numRows = 10000;
  const size_t numCols = 10;

  View<float**> X ("X", numRows, numCols);
  // ... fill X before the following ...
  ColumnSums cs (X);
  float sums[10];
  parallel_reduce (X.dimension_0 (), cs, sums);
\end{lstlisting}

\section{Parallel scan}

Kokkos' \lstinline!parallel_scan! operation implements a \emph{prefix scan}.
A prefix scan is like a reduction over a 1-D array,
but it also stores all intermediate results (``partial sums'').
It can use any associative binary operator.
The default is \lstinline!operator+!, 
and we call a scan with that operator a ``sum scan''
if we need to distinguish it from scans with other operators.
The scan operation comes in two variants.
An \emph{exclusive scan} excludes (hence the name) the first entry of the array,
and an \emph{inclusive scan} includes that entry.
Given an example array $(1, 2, 3, 4, 5)$, 
an exclusive sum scan overwrites the array with $(0, 1, 3, 6, 10)$,
and an inclusive sum scan overwrites the array with $(1, 3, 6, 10, 15)$.

Many operations that ``look'' sequential can be parallelized with a
scan.  To learn more, see e.g., ``Vector Models for Data-Parallel
Computing,'' Guy Blelloch, 1990 (the book version of Prof.\ Blelloch's
PhD dissertation).

Kokkos lets users specify a scan by either a functor or a lambda.
Both look like their \lstinline!parallel_reduce! equivalents,
except that the \lstinline!operator()! method or lambda takes three arguments:
the loop index, the ``update'' value by nonconst reference, and a \lstinline!bool!.
Here is a lambda example where the intermediate results have type \lstinline!float!.
\begin{lstlisting}
View<float*> x = ...; // assume filled with input values
const size_t N = x.dimension_0 ();
parallel_scan (N, [=] (const int i, float& upd, const bool final) {
    if (final) {
      x(i) = upd; // only update array on final pass
    }
    // For exclusive scan, change the update value after 
    // updating array, like we do here.  For inclusive scan, 
    // change the update value before updating array.
    upd += x(i); 
  });
\end{lstlisting}
Kokkos may use a multiple-pass algorithm to implement scan.
This means that it may call your \lstinline!operator()! or lambda multiple times per loop index value.
The \lstinline!final! Boolean argument tells you whether Kokkos is on the final pass.
You must only update the array on the final pass.

For an exclusive scan, change the \lstinline!update! value after
updating the array, as in the above example.  For an inclusive scan,
change \lstinline!update! \emph{before} updating the array.  Just as
with reductions, your functor may need to specify a nondefault
\lstinline!join! or \lstinline!init! method if the defaults do not do
what you want.

\section{Functors}
\section{Function Name Tags}

### 2015-03-16 Kokkos_PG_ProgrammingModel.tex
Initial commit.

\chapter{Programming Model}

The programming model Kokkos is characterized by 6 core abstractions: Execution Spaces, Execution Patterns, Execution Policies, Memory Spaces, Memory Layout and Memory Traits. 
These abstraction concepts allow the formulation of generic algorithms and data structures which can then be mapped to different types of architectures. 
Effectively they allow for compile time transformation of algorithms to allow for adaptions of varying degrees of hardware parallelism as well as of the memory hierarchy.

\section{Execution Spaces}

An Execution Space is the place {\it Where} code can actually be executed. 
For example on current Hybrid GPU / CPU systems there are two types of execution spaces: the GPU cores and the CPU cores. 
In the future this could include Processing in Memory (PIM) modules or different core types on a heterogenous CPU.
In principle this can also be used to introduce remote memory spaces, e.g. the capability of sending work to a different node.
Execution Spaces thus give an application developer the means to target different parts of a heterogenous hardware architecture.

\section{Execution Patterns}

Execution Patterns are the fundamental parallel algorithms in which an application has to be expressed.
Examples are \lstinline|parallel_for|: execute a function in undeterminded order a specified amount of times.,
\lstinline|parallel_reduce|: which combines a \lstinline|parallel_for| execution with a reduction operation,
\lstinline|parallel_scan|: which combines a \lstinline|parallel_for| operation with a prefix or postfix scan on output values of each operation, and
\lstinline|task|: which executes a single function with dependencies on other functions.
Expressing an application in these patterns allows the underlying implementation or the used compiler to reason about valid transformations.
For example all \lstinline|parallel_***| patterns allow unspecified execution order, and only promise deterministic results of the reductions themselves.
This enables different mapping patterns on different hardware such as assignment of iterations to threads or vector lanes.

\section{Execution Policies}

An Execution Policy determines together with an Execution Pattern {\it How} a function is executed.

\section{Memory Spaces}

Memory Spaces are the places {\it Where} data resides.
They specify physical location of data as well as certain access characteristics. 
Different physical locations correspond to things such as high bandwidth memory, on die scratch memory or non-volatile bulk storage.
Different logical memory spaces allow for concepts such as UVM memory in the CUDA programming model, which is accessible from Host and GPU. 
Memory Spaces also can be used to express remote memory locations.
Furthermore they encapsulate functionality such as consistency control and persistence scopes.

\section{Memory Layout}

Layouts express the mapping from logical (or algorithmical) indicies to address offset for a data allocation. 
By adopting appropriate layouts for memort structures an application can optimise data access patterns in a given algorithm.
If an implementation provides polymorphic layouts (i.e. a data structure can be instantiated at compile or runtime with different layouts) an architecture dependent optimisation can be performed.

\section{Memory Traits}

Memory Traits specify how a data structure is accessed in an algorithm. 
Traits express usage scenarios such as atomic access, random access and streaming loads or stores.
By putting such attributes on data structures, an implementation of the programming model can insert optimal load and store operations.
If a compiler implements the programming model, it could reason about the access modes and use that to inform code transformations.

### 2015-03-16 Kokkos_PG_Subviews.tex
Initial commit.

\chapter{Subviews}\label{C:Subviews}

After reading this chapter, you should understand the following:
\begin{itemize}
\item A \emph{slice} of a multidimensional array behaves as an array,
  and is a view of a structured subset of that array
\item A \emph{subview} is a slice of an existing Kokkos View
\item A subview has the same reference count as their parent View
\item Use C++11 \texttt{auto} to let Kokkos pick the subview's type
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{A subview is a slice of a View}\label{S:Subviews:Slice}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

A \emph{slice} of a multidimensional array behaves as an array, and is
a view of a structured subset of the original array.  ``Behaves as an
array'' means that the slice has the same syntax as an array should;
one can access its entries using array indexing notation.  ``View''
means that the slice and the original array point to the same data.
That is, the slice sees changes to the original array, and vice versa.
``Structured subset'' means a cross product of indices along each
dimension, as for example a plane or face of a cube.  If the original
array has dimensions $(N_0, N_1, \dots, N_{k-1})$, then a slice views
all entries whose indices are $(\alpha_0, \alpha_1, \dots,
\alpha_{k-1})$, where $\alpha_j$ is a subset of $[0, 1, \dots,
N_j-1]$.

Array slices are handy for encapsulation.  A slice looks and acts like
an array, so you can pass it into functions that expect an array.  For
example, you can write a function for processing boundaries (as
slices) of a structured grid, without needing to tell that function
properties of the entire grid.

Programming languages like Fortran 90, Matlab, and Python have a
special ``colon'' notation for representing slices.  For example, if
\texttt{A} is an $M \times N$ array, then
\begin{itemize}
\item \texttt{A(:, :)} represents the whole array,
\item \texttt{A(:, 3)} represents the fourth column (if the language
  has zero-based indices, or the third column, if the language has
  one-based indices),
\item \texttt{A(4, :)} represents the fifth row, 
\item \texttt{A(2:4, 3:7)} represent the sub-array of rows 3-4 and
  columns 4--7 (different languages differ on whether the ranges are
  inclusive or exclusive of the last index -- Kokkos, like Python, is
  exclusive), and
\item \texttt{A(3, 4)} represents a ``zero-dimensional'' slice which
  views the entry in the fourth row and fifth column of the matrix.
\end{itemize}
These languages may have more elaborate notation for expressing sets
of indices other than contiguous ranges.  These may include
``strided'' subsets of indices, like \texttt{3:2:9} = $\{ 3, 5, 7, 9
\}$, or even arbitrary sets of indices.

In Kokkos, a \emph{subview} is a slice of a View.  A subview has the
same reference count as its parent View, so the parent View won't be
deallocated before all subviews go away.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{How to take a subview}\label{S:Subviews:How}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

To take a subview of a View, use the \texttt{subview} function.  This
function is overloaded for all different kinds of Views and index
ranges.  For example, the following is equivalent to the above example
\texttt{A(2:4, 3:7)}:

\begin{lstlisting}
const size_t N0 = ...;
const size_t N1 = ...;
View<double**> A ("A", N0, N1);

auto A_sub = 
  subview (A, std::make_pair (2, 4), std::make_pair (3, 7));
\end{lstlisting}

The Kokkos equivalent of a contiguous index range \texttt{3:7} is
\\ \lstinline!std::pair<size_t, size_t>(3, 7)!.  The Kokkos equivalent of
\texttt{:} (a colon by itself; the whole index range for that
dimension) is \texttt{ALL()} (an instance of the \texttt{Kokkos::ALL}
class).  Kokkos does not currently have equivalents of the strided or
arbitrary index sets.

Every subview is also a View.  This means that you may take a subview
of a subview.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{C++11 type deduction}\label{SS:Subviews:How:auto}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Note the use of the C++11 keyword \texttt{auto} in the above example.
A subview may have a different type than its parent View.  For
instance, if \lstinline!A! has \lstinline!LayoutRight!,
\lstinline!A_sub!  has \lstinline!LayoutStride!.  A subview of an
entire row of \lstinline!A! has \lstinline!LayoutRight! as well, but a
subview of an entire column of \lstinline!A! has
\lstinline!LayoutStride!.  If you assign the result of the
\texttt{subview} function to the wrong type, Kokkos will emit a
compile-time error.  The best way to avoid this is to use the C++11
\texttt{auto} keyword to let the compiler deduce the correct type for
you.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Dimension of a subview}\label{SS:Subviews:How:dim}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Suppose that a View has $k$ dimensions.  Then, when calling
\texttt{subview} on that View, you must pass in $k$ arguments.  Every
argument that is a single index -- that is, not a pair or
\lstinline!ALL()! -- reduces the dimension of the resulting View by 1.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Degenerate Views}\label{SS:Subviews:How:degen}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Given a View with $k$ dimensions, we call that View \emph{degenerate}
if any of those dimensions is zero.  Degenerate Views are useful for
keeping code simple, by avoiding special cases.  For example, in an
MPI (Message-Passing Interface) distributed-memory parallel code that
uses Kokkos to represent local (per-process) data structures,

### 2015-03-16 Kokkos_PG_Views.tex
Initial commit.

\chapter{View: Multidimensional array}

After reading this chapter, you should understand the following:
\begin{itemize}
\item A Kokkos View is an array of zero or more dimensions
\item How to use View's first template parameter to specify the type
  of entries, the number of dimensions, and whether the dimensions are
  determined at run time or compile time
\item Kokkos handles array deallocation automatically
\item Kokkos chooses array layout at compile time for best overall
  performance, as a function of the computer architecture
\item How to set optional template parameters of View for low-level
  control of execution space, layout, and how Kokkos accesses array
  entries
\end{itemize}
In all code examples in this chapter, we assume that all classes in
the \lstinline!Kokkos! namespace have been imported into the working
namespace.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Why Kokkos needs multidimensional arrays}\label{S:Views:Why}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Many scientific and engineering codes spend a lot of time computing
with arrays of data.  Programmers invest a lot of effort making array
computations as fast as possible.  This effort is often intimately
bound to details of the computer architecture, run-time environment,
language, and programming model.  For example, optimal array layout
may differ based on the architecture, with a large integer factor
penalty if wrong.  Low-level issues like pointer alignment, array
layout, indexing overhead, and initialization all affect performance.
This is true even with sequential codes.  Thread parallelism adds even
more pitfalls, like first-touch allocation and false sharing.

For best performance, coders need to tie details of how they manage
arrays, to details of how parallel code accesses and manages those
arrays.  Programmers who write architecture-specific code then need to
mix low-level features of the architecture and programming model into
high-level algorithms.  This makes it hard to port codes between
architectures, or to write a code that still performs well as
architectures evolve.

Kokkos aims to relieve some of this burden, by optimizing array
management and access for the specific architecture.  Tying arrays to
shared-memory parallelism lets Kokkos optimize the former to the
latter.  For example, Kokkos can easily do first-touch allocation,
because it controls threads that it can use to initialize arrays.
Kokkos' architecture awareness lets it pick optimal layout and pad
allocations for good alignment.  Expert coders can also use Kokkos to
access low-level or more architecture-specific optimizations in a more
user-friendly way.  For instance, Kokkos makes it easy to experiment
with different array layouts.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Creating and using a View}\label{S:Views:CreateUse}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Constructing a View}

A View is an array of zero or more dimensions.  Programmers set both
the type of entries, and the number of dimensions, at compile time, as
part of the type of the View.  For example, the following specifies
and allocates a View with four dimensions, whose entries have type
\texttt{int}:

\begin{lstlisting}
const size_t N0 = ...;
const size_t N1 = ...;
const size_t N2 = ...;
const size_t N3 = ...;
View<int****> a ("some label", N0, N1, N2, N3);
\end{lstlisting}

The string argument is a label, which Kokkos uses for debugging.
Different Views may have the same label.  The ellipses indicate some
integer dimensions specified at run time.  Users may also set some
dimensions at compile time.  For example, the following View has two
dimensions.  The first one (represented by the asterisk) is a run-time
dimension, and the second (represented by [3]) is a compile-time
dimension.  Thus, the View is an N by 3 array of type double, where N
is specified at run time in the View's constructor.

\begin{lstlisting}
const size_t N = ...;
View<double*[3]> b ("another label", N);
\end{lstlisting}

Views may have up to (at least) 8 dimensions, and any number of these
may be run-time or compile-time.  The only limitation is that the
run-time dimensions (if any) must appear first, followed by all the
compile-time dimensions (if any).  For example, the following are
valid three-dimensional View types:
\begin{itemize}
\item \lstinline!View<int***>!  (3 run-time dimensions)
\item \lstinline!View<int**[8]>!  (2 run-time, 1 compile-time)
\item \lstinline!View<int*[3][8]>!  (1 run-time, 2 compile-time)
\item \lstinline!View<int[4][3][8]>!  (3 compile-time)
\end{itemize}
and the following are \emph{not} valid three-dimensional View types:
\begin{itemize}
\item \lstinline!View<int[4]**>!
\item \lstinline!View<int[4][3]*>!
\item \lstinline!View<int[4]*[8]>!
\item \lstinline!View<int*[3]*>!
\item \lstinline!View<int*[3][8]>!
\end{itemize}
This limitation comes from the implementation of View using C++
templates.  View's first template parameter must be a valid C++ type.

\subsection{What types of data may a View contain?}

C++ lets users construct data types that may ``look like'' numbers in
terms of syntax, but do arbitrarily complicated things inside.  Some
of those things may not be thread safe, like unprotected updates to
global state.  Others may perform badly in parallel, like fine-grained
dynamic memory allocation.  As a result, Kokkos restricts the set of
data types \texttt{T} which may be the entries of a View.

Users may always construct a View whose entries are 
\begin{itemize}
\item built-in data types (``plain old data''), like \texttt{int} or
  \texttt{double}, or
\item structs of built-in data types.
\end{itemize}
The latter specifically forbids nontrivial destructors, constructors
with dynamic memory allocation, or virtual methods of any kind.  In
particular, entries must have size determined at compile time by the
\texttt{sizeof} operator.\footnote{Kokkos has a customization facility
  for users to add support for types with size determined at run time,
  but this is for expert users only.  For examples, see the Stokhos
  software package in the Trilinos project.}

\subsection{Accessing entries (indexing)}

You may access an entry of a View using parentheses enclosing a
comma-delimited list of integer indices.  This looks just like a
Fortran multidimensional array access.  For example:
\begin{lstlisting}
const size_t N = ...;
View<double*[3][4]> a ("some label", N);
parallel_for (N, [=] (const ptrdiff_t i) {
    const size_t j = ...;
    const size_t k = ...;
    const double a_ijk = a(i,j,k);
    /* rest of the loop body */
  });
\end{lstlisting}
Note how in the above example, we only access the View's entries in a
parallel loop body.  In general, you may only access a View's entries
in an execution space which is allowed to access that View's memory
space.  For example, if the default memory space is \lstinline!Cuda!
(CUDA device memory), you may not access it in host code, unless CUDA
UVM is enabled.  Furthermore, access costs (e.g., latency and
bandwidth) may vary, depending on the View's ``native'' memory and
execution spaces, and the execution space from which you access it.
CUDA UVM may work, but it may also be slow, depending on your access
pattern and performance requirements.  Thus, best practice is to
access the View only in a Kokkos parallel for, reduce, or scan, using
the same execution space as the View.  This also ensures that access
to the View's entries respect first-touch allocation.  The first
(leftmost) dimension of the View is the \emph{parallel dimension},
over which it is most efficient to do parallel array access.

\subsection{Memory management and copying}

\subsubsection{Reference counting}

Kokkos automatically manages deallocation of Views, through a
reference-counting mechanism.  Otherwise, Views behave like raw
pointers.  Copying or assigning a View does a shallow copy, and
changes the reference count.  (The View copied has its reference count
incremented, and the assigned-to View has its reference count
decremented.)  A View's destructor (called when the View falls out of
scope or during a stack unwind due to an exception) decrements the
reference count.  Once the reference count reaches zero, Kokkos may
deallocate the View.

For example, the following code allocates two Views, then assigns one
to the other.  That assignment may deallocate the first View, since it
reduces its reference count to zero.  It then increases the reference
count of the second View, since now both Views point to it.
\begin{lstlisting}
View<int*> a ("a", 10);
View<int*> b ("b", 10);
a = b; // assignment does shallow copy
\end{lstlisting}

For efficiency, View allocation and reference counting turn off inside
of Kokkos' parallel for, reduce, and scan operations.  This affects
what you can do with Views inside of Kokkos' parallel operations.

\subsubsection{Deep copy}

If you want to make a deep copy of a View's contents, you must do so
explicitly, using the \lstinline!deep_copy! function.  For example:
\begin{lstlisting}
View<int*> a ("a", 10);
View<int*> b ("b", 10);
deep_copy (a, b); // copy contents of b into a
\end{lstlisting}

\subsection{Where and how is the View allocated?}

\subsubsection{Execution and memory spaces}

Views are allocated by default, in the default execution space's
default memory space.  You may access the View's execution space via
its \lstinline!execution_space! typedef, and its memory space via its
\lstinline!memory_space! typedef.  You may also specify the execution space
explicitly as a template parameter.  For example, the following
allocates a View in CUDA device memory:
\begin{lstlisting}
View<int*, Cuda> a ("a", 100000);
\end{lstlisting}
and the following allocates a View in ``host'' memory, using OpenMP
for first-touch allocation:
\begin{lstlisting}
View<int*, OpenMP> a ("a", 100000);
\end{lstlisting}

\subsubsection{Initialization}

A View's entries are initialized to zero by default.  Initialization
happens in parallel, for first-touch allocation over the first
(leftmost) dimension of the View.

You may allocate a View without initializing.  For example:
\begin{lstlisting}
View<int*> x (ViewAllocateWithoutInitializing (label), 100000);
\end{lstlisting}

\subsection{How do I get the raw pointer?}

We discourage access to a View's ``raw'' pointer.  This circumvents
reference counting.  That is, the memory may be deallocated once the
View's reference count goes to zero, so holding on to a raw pointer
may result in invalid memory access.  Furthermore, it may not even be
possible to access the View's memory from a given execution space.
For example, a View in the \lstinline!Cuda! space points to CUDA device
memory.  Nevertheless, sometimes you really need access to the
pointer.  For such cases, we provide the \lstinline!ptr_on_device()!
method.  For example:
\begin{lstlisting}
// Legacy function that takes a raw pointer.
extern void legacyFunction (double* x_raw, const size_t len);

// Your function that takes a View.
void myFunction (const View<double*>& x) {
  // DON'T DO THIS UNLESS YOU MUST
  double* a_raw = a.ptr_on_device ();
  const size_t N = x.dimension_0 ();
  legacyFunction (a_raw, N);
}
\end{lstlisting}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Layout}\label{S:Views:Layout}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Strides and dimensions}\label{SS:Views:Layout:Strides}

\emph{Layout} refers to the mapping from a logical multidimensional
index $(i, j, k, \dots)$ to a physical memory offset.  Different
programming languages may have different layout conventions.  For
example, Fortran uses \emph{column-major} or ``left'' layout, where
consecutive entries in the same column of a 2-D array are contiguous
in memory.  Kokkos calls this \lstinline!LayoutLeft!.  C, C++, and Java use
\emph{row-major} or ``right'' layout, where consecutive entries in the
same row of a 2-D array are contiguous in memory.  Kokkos calls this
\lstinline!LayoutRight!.

The generalization of both left and right layouts is ``strided.''  For
a strided layout, each dimension has a \emph{stride}.  The stride for
that dimension determines how far apart in memory two array entries
are, whose indices in that dimension differ only by one, and whose
other indices are all the same.  For example, with a 3-D strided view
with strides $(s_1, s_2, s_3)$, entries $(i, j, k)$ and $(i, j+1, k)$
are $s_2$ entries (not bytes) apart in memory.  Kokkos calls this
\lstinline!LayoutStride!.

Strides may differ from dimensions.  For example, Kokkos reserves the
right to pad each dimension for cache or vector alignment.  You may
access all the dimensions of a View using the \lstinline!dimension! method,
and all the strides using the \lstinline!stride! method.  Both take a raw
integer array, and only fill in as many entries as the number of
dimensions.  For example:
\begin{lstlisting}
const size_t N0 = ...;
const size_t N1 = ...;
const size_t N2 = ...;
View<int***> a ("a", N0, N1, N2);
size_t dims[3]
a.dimension (dims); // fill 'dims' with dimensions
size_t strides[3]
a.strides (dims); // fill 'strides' with strides
\end{lstlisting}
You may also refer to specific dimensions:
\begin{lstlisting}
const size_t n0 = a.dimension_0 ();
const size_t n2 = a.dimension_2 ();
\end{lstlisting}

Users of the BLAS and LAPACK libraries may be familiar with the ideas
of layout and stride.  These libraries only accept matrices in
column-major format.  The stride between consecutive entries in the
same column is 1, and the stride between consecutive entries in the
same row is \lstinline!LDA! (``leading dimension of the matrix A'').  The
number of rows may be less than \lstinline!LDA!, but may not be greater.

\subsection{Other layouts}\label{SS:Views::Layout::Other}

Other layouts are possible.  For example, Kokkos has a ``tiled''
layout, where a tile's entries are stored contiguously (in either row-
or column-major order) and tiles have compile-time dimensions.  One
may also use Kokkos to implement Morton ordering or variants thereof.

\subsection{Default layout depends on execution space}\label{SS:Views:Layout:Default}

Kokkos selects a View's default layout for optimal parallel access
over the leftmost dimension, based on its execution space.  For
example, \lstinline!View<int**, Cuda>! has \lstinline!LayoutLeft!, so that
consecutive threads in the same warp access consecutive entries in
memory.  This \emph{coalesced access} gives the code better memory
bandwidth.  In contrast, \lstinline!View<int**, OpenMP>! has
\lstinline!LayoutRight!, so that a single thread accesses contiguous
entries of the array.  This avoids wasting cache lines and helps
prevent false sharing of a cache line between threads.

\subsection{Explicitly specifying layout}\label{SS:Views:Layout:Explicit}

We prefer that users let Kokkos determine a View's layout, based on
its execution space.  However, sometimes you really need to specify
the layout.  For example, the BLAS and LAPACK libraries only accept
column-major arrays.  If you want to give a View to the BLAS or
LAPACK, that View must be \lstinline!LayoutLeft!.  You may specify the
layout as a template parameter of View.  For example:
\begin{lstlisting}
const size_t N0 = ...;
const size_t N1 = ...;
View<double**, LayoutLeft> A ("A", N0, N1);

// Get 'LDA' for BLAS / LAPACK
int strides[2]; // any integer type works in stride()
A.stride (strides);
const int LDA = strides[1];
\end{lstlisting}
You may ask a View for its layout via its \lstinline!array_layout! typedef.
This can be helpful for C++ template metaprogramming.  For example
(C++11 only):
\begin{lstlisting}
template<class ViewType>
void callBlas (const ViewType& A) {
  typedef typename ViewType::array_layout array_layout;
  if (std::is_same<array_layout, LayoutLeft>::value) {
    callSomeBlasFunction (A.ptr_on_device (), ...);
  } else {
    throw std::invalid_argument ("A is not LayoutLeft");
  }
}
\end{lstlisting}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Managing Data Placement}\label{S:Views:Placement}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

TODO: Hey Christian, what did you mean by this?

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Access traits}\label{S:Views:AccessTraits}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Const Views}

Different optional template parameters of View let you control how
Kokkos accesses entries of the array.  The simplest example involves
making a View \lstinline!const!, that is, making its entries read-only.  It
is a compile-time error to assign to an entry of a const View.  It
does not make sense to allocate a const View, but it does make sense
to assign a non-const View to a const View.  Here is an example:
\begin{lstlisting}
const size_t N0 = ...;
View<double*> a_nonconst ("a_nonconst", N0);

// Assign a nonconst View to a const View
View<const double*> a_const = a_nonconst;
// Pass the const View to some read-only function.
const double result = readOnlyFunction (a_const);
\end{lstlisting}
Note that a ``const View'' means the \emph{entries} are const.  You
may still assign to a const View.  \lstinline!View<const double*>!
corresponds exactly to \lstinline!const double*!, and
\lstinline!const View<double*>! to \lstinline!double* const!.

\subsection{Random read Views}

Const Views are useful for declaring intent and helping you debug at
compile time.  They also open the door to other, optional
optimizations.  In particular, Kokkos lets you declare that you plan
to access a read-only array irregularly (less spatial locality), using
the optional \lstinline!RandomAccess! template parameter.  You may only use
this with a const View.  For example:
\begin{lstlisting}
const size_t N0 = ...;
View<int*> a_nonconst ("a", N0); // allocate nonconst View
// Assign to const, RandomAccess View
View<const int*, RandomAccess> a_ra = a_nonconst;
\end{lstlisting}
For a View whose execution space is \lstinline!Cuda!, access to a
\lstinline!RandomAccess! View may use CUDA texture fetches.  Texture
fetches are not cache coherent with respect to writes, which is why
you must use read-only access.  The texture cache is optimized for
noncontiguous access, since it has a shorter cache line than the
regular cache.

\lstinline!RandomAccess! is valid for other execution spaces, and Kokkos
also requires a const View for those execution spaces.  Kokkos
reserves the right to perform analogous optimizations on other
execution spaces.

\subsection{Standard idiom for specifying access traits}

The standard idiom for View is to pass it around using as few template
parameters as possible.  Then, assign to a View with the desired
access traits only at the ``last moment,'' when those access traits
are needed just before entering a computational kernel.  This lets you
template C++ classes and functions on the View type, without
proliferating instantiations.  Here is an example:

\begin{lstlisting}
// Compute a sparse matrix-vector product, for a sparse
// matrix stored in compressed sparse row (CSR) format.
void
spmatvec (const View<double*>& y, 
  const View<const size_t*>& ptr,
  const View<const int*>& ind,
  const View<const double*>& val,
  const View<const double*>& x)
{
  // Access to x has less locality than access to y.
  View<const double*, RandomAccess> x_ra;
  typedef View<const size_t*>::size_type size_type;

  parallel_for (y.dimension_0 (), [=] (const size_type i) {
      double y_i = y(i);
      for (size_t k = ptr(i); k < ptr(i+1); ++k) {
        y_i += val(k) * x_ra(ind(k));
      }
      y(i) = y_i;
    });
}
\end{lstlisting}

\subsection{Unmanaged Views}

It's always better to let Kokkos control memory allocation, but
sometimes you don't have a choice.  You might have to work with an
application or an interface that returns a raw pointer, for example.
Kokkos lets you wrap raw pointers in an \emph{unmanaged View}.
``Unmanaged'' means that Kokkos does \emph{not} do reference counting
or automatic deallocation for those Views.  The following example
shows how to create an unmanaged View of host memory.  You may do this
for CUDA device memory too, or indeed for memory allocated in any
memory space, by specifying the View's execution or memory space
accordingly.
\begin{lstlisting}
// Sometimes other code gives you a raw pointer, ...
const size_t N0 = ...;
double* x_raw = malloc (N0 * sizeof (double));
{
  // ... but you want to access it with Kokkos.
  //
  // malloc() returns host memory, so we use the 
  // host memory space HostSpace.  Unmanaged
  // Views have no label, because labels work with
  // the reference counting system.
  View<double*, HostSpace, Unmanaged> x_view (x_raw, N0);

  functionThatTakesKokkosView (x_view);

  // It's safest for unmanaged Views to fall out of
  // scope, before freeing their memory.
}
free (x_raw);
\end{lstlisting}

\subsection{Atomic access}

The \lstinline!Atomic! access trait lets you create a View, such that every
read or write to any entry uses an atomic update.  This only works for
certain data types, for which the hardware and programming model
support atomic updates.  Currently, Kokkos limits atomic updates to
32-bit or 64-bit built-in types.

### 2015-03-16 Kokkos_PG_HierarchicalParallelism.tex
Added intro to Hierarchical Parallelism chapter.

@@
 \chapter{Hierarchical Parallelism}\label{C:Hierarchical}
+
+This chapter explains how to use Kokkos to exploit multiple levels of shared-memory parallelism.
+These levels include thread teams, threads within a team, and vector lanes.
+You may nest these levels of parallelism,
+and execute \lstinline!parallel_for!, \lstinline!parallel_reduce!, or \lstinline!parallel_scan! at each level.
+The syntax differs only by the execution policy,
+which is the first argument to the \verb!parallel_*! operation.
+Kokkos also exposes a ``scratch pad'' memory which all threads within a team may share.
+
+\section{Motivation}\label{S:Hierarchical:Motivation}
+
 Node architectures on modern high-performance computers are characterized by ever more \emph{hierarchical parallelism}. 
@@
 \end{enumerate}
-Kokkos provides a number of abstract levels of parallelism which at compile time or run time are mapped to the appropriate hardware features. 
+Kokkos provides a number of abstract levels of parallelism,
+which it maps to the appropriate hardware features.
 This mapping is not necessarily static or predefined; it may differ for each kernel.
@@
 
-\section{Thread Teams Basics}
+\section{Thread teams}\label{S:Hierarchical:Teams}
 
@@
 
-\subsection{Creating a Policy Instance}
+\subsection{Creating a Policy Instance}\label{SS:Hierarchical:Teams:Policy}
 
@@
 
-\subsection{Basic Kernels}
+\subsection{Basic Kernels}\label{SS:Hierarchical:Teams:Kernels}
+
 Within the kernel the team policy member type provides the necessary functionality to use teams. 
@@
 
-\section{Team Scratch Pads}\label{S:Hierarchical:Scratch}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Team scratch pad memory}\label{S:Hierarchical:Scratch}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\section{Nested Parallelism}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Nested parallelism}\label{S:Hierarchical:Nested}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\subsection{Team and Vector Loops}
+\subsection{Team and vector loops}\label{SS:Hierarchical:Nested:Loops}
 

### 2015-03-17 Kokkos_PG_Views.tex
Added intro to Hierarchical Parallelism chapter.
Minor changes to section labels.

@@
-\chapter{View: Multidimensional array}
+\chapter{View: Multidimensional array}\label{C:View}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Why Kokkos needs multidimensional arrays}\label{S:Views:Why}
+\section{Why Kokkos needs multidimensional arrays}\label{S:View:Why}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Creating and using a View}\label{S:Views:CreateUse}
+\section{Creating and using a View}\label{S:View:CreateUse}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Layout}\label{S:Views:Layout}
+\section{Layout}\label{S:View:Layout}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\subsection{Strides and dimensions}\label{SS:Views:Layout:Strides}
+\subsection{Strides and dimensions}\label{SS:View:Layout:Strides}
 
@@
 
-\subsection{Other layouts}\label{SS:Views::Layout::Other}
+\subsection{Other layouts}\label{SS:View::Layout::Other}
 
@@
 
-\subsection{Default layout depends on execution space}\label{SS:Views:Layout:Default}
+\subsection{Default layout depends on execution space}\label{SS:View:Layout:Default}
 
@@
 
-\subsection{Explicitly specifying layout}\label{SS:Views:Layout:Explicit}
+\subsection{Explicitly specifying layout}\label{SS:View:Layout:Explicit}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Managing Data Placement}\label{S:Views:Placement}
+\section{Managing Data Placement}\label{S:View:Placement}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+Christian says: \lstinline!CudaSpace!, \lstinline!HostSpace!, etc.
+Also, please emphasize memory space over execution space, and talk
+about the UVM and pinned variants of \lstinline!CudaSpace!.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Access traits}\label{S:Views:AccessTraits}
+\section{Access traits}\label{S:View:AccessTraits}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2015-03-17 Kokkos_PG_MachineModel.tex
Filled out Machine Models chapter (NOT DONE).
Carter and Christian, please take a look to ensure that my idea of
Kokkos' semantics corresponds with your idea.  Thanks!

@@
-\chapter{Machine Model}
+\chapter{Machine model}\label{C:Model}
 
-The abstract machine model of Kokkos allows for very complex heterogenous node architectures containing N execution spaces and M memory spaces.
+After reading this chapter, you will understand the abstract model of
+a shared-memory parallel computer which underlies Kokkos.  
+Respecting the abstract machine model ensures performance portability.
+The model has two components:
+\begin{itemize}
+\item \emph{Memory spaces}, in which data live
+\item \emph{Execution spaces}, which execute parallel operations using
+  data in one or more memory spaces
+\end{itemize}
+All code that uses Kokkos involves relations between memory and execution spaces.
+% A given execution space may or may not be able to access data in a
+% given memory space.  Users can ask Kokkos if it can.  Users may also
+% copy data between any two different memory spaces by using
+% \lstinline!deep_copy!.
+
+\section{Motivation}\label{S:Model:Motivation}
+
+Kokkos is a programming model which happens to have a C++ implementation.
+As you are first learning Kokkos,
+you may find yourself focusing on details of the implementation,
+especially if you have less experience with C++ syntax.
+However, as the syntax becomes more familiar to you,
+you will find it more and more helpful to understand the underlying abstractions.
+
+Kokkos assumes an \emph{abstract machine model},
+which represents a shared-memory parallel computer architecture.
+The model may represent either a stand-alone computer,
+like a laptop or workstation,
+or a single ``parallel process'' in a distributed-memory parallel programming model such as MPI (the Message-Passing Interface).
+In general, we write \emph{node} to describe the hardware that Kokkos governs.
+Kokkos' machine model lets programmers reason about very complicated heterogenous node architectures.
+It also lets them start with simple, even sequential architectures,
+and gradually port code from there to perform well on more complicated computer architectures.
+The abstract machine model might look complicated,
+but it reflects computers that companies are actually designing and building.
+Many chapters in this Guide show how to use the more complex aspects of the model.
+For example, see Chapter \ref {C:Hierarchical},
+which shows how to use Kokkos to expose multiple levels of parallelism.
+
+The point of this chapter is that \emph{respecting the abstract machine model ensures performance portability}.
+The phrase \emph{performance portability} means that the same code performs reasonably well on many different computer architectures,
+with no or minimal\footnote{For example, users might change a single C++ typedef or a single configure-time option.} changes to code.
+Kokkos depends on this abstract machine model for performance portability.
+\emph{Implementations} of Kokkos may depend on other things for performance portability.
+For example, efficiency of \lstinline!parallel_for! in the C++ implementation relies in part on compilers' ability to inline code.
+One could imagine a different Kokkos implementation based on source-to-source translation,
+that inlines parallel loop bodies into \lstinline!parallel_for! before they reach the compiler.
+Kokkos' OpenMP and Pthreads back-ends may rely on the Portable Hardware Locality (hwloc) library
+to control assignment of software threads to hardware cores or hyperthreads.
+However, Kokkos as an abstract programming model does not depend on these things.
+
+\section{Memory spaces and execution spaces}\label{S:Model:Spaces}
+
+A \emph{memory space} contains data.
+Users access those data as k-dimensional typed arrays, where k may be any nonnegative integer.
+(Zero-dimensional arrays are single values.  See Chapter \ref{C:View} for details.)
+% In the C++ implementation of the Kokkos programming model,
+% these memory addresses are pointers,
+% which point to contiguous address ranges,
+% each of which represents an allocation of data.
+The memory space controls dynamic allocation and deallocation of those arrays,
+if it allows them at all.
+
+An \emph{execution space} runs Kokkos' parallel operations,
+like parallel for, reduce, and scan.  (See Chapter \ref{C:Dispatch}.)
+It may access data in some available memory spaces.
+Whether an execution space may access data in a memory space
+depends on \emph{both} the execution space and the memory space.
+
+Here is an analogy: Many operating systems have ``processes.''
+A process is like an execution space, in that code may execute in one or more concurrent threads.
+Those threads share access to a common stack and heap, which comprise the process' ``memory space.''
+Two different processes may not access each other's memory spaces by default.
+However, some operating systems have a ``shared memory'' mechanism.\footnote{See e.g., the POSIX \lstinline!shm_open! and related functions.}
+This lets different processes share access to a specific region of memory,
+which in Kokkos terms is like a ``memory space,'' separate from each process' normal memory space.
+Some operating systems also let you open a file,
+and then ``memory map'' that file into a contiguous memory address range.\footnote{See e.g., the POSIX \lstinline!mmap! function.}
+This makes the file behave like yet another memory space.
+Memory-mapped files may have very different performance characteristics than the usual memory,
+even though programmers access both ``memory spaces'' through the same interface.
+Whether or not a process may access a file depends on both the process and the file.
+
+Kokkos distinguishes between \emph{types} of spaces, and \emph{instances} of spaces.
+For example, a node might have multiple NVIDIA GPUs.
+All of these GPUs' memory spaces have the type \lstinline!Kokkos::CudaSpace!,
+but each space is a different instance of \lstinline!Kokkos::CudaSpace!.
+The common case of using Kokkos presumes a single instance of a given execution or memory space.
+In that case, one uses the type of the space as a synonym for the default (single) instance of that space.
+When we speak of a space by its type,
+we mean the default single instance of that space,
+unless we say otherwise.
+
+% Kokkos' abstract machine model may contain $N$ types of execution spaces
+% and $M$ types of memory spaces.
+% It is possible that $N \neq M$.
+
+\section{Memory consistency}\label{S:Model:Consistency}
+
+A \emph{memory consistency model} describes the behavior of a memory space $M$,
+with respect to multiple threads running in an execution space $E$
+that access the same data in $M$ concurrently.
+One may also encounter this idea as the phrase ``cache coherency.''
+We prefer ``memory consistency,'' because Kokkos does not require caches.
+Consistency models are a complicated subject which we cannot cover in sufficient detail here.
+We encourage readers to refer to standard computer hardware textbooks,
+like ``Computer Architecture: A Quantitative Approach'' (Hennessy and Patterson, 2011).
+
+Kokkos imposes a very weak consistency model.
+For example, it does \emph{not} assume sequential consistency.
+Much like MPI 3 one-sided communication
+or other partitioned global address space (PGAS) programming models,
+Kokkos requires an execution space to issue a \emph{memory fence}
+to a given memory space, 
+in order to bring that memory space into a consistent state.
+The fence forces completion of all outstanding memory writes,
+and reorders all loads to occur after all those writes complete.
+
+\section{Asynchronous parallel dispatch}\label{S:Model:Async}
+
+Kokkos lets users tell execution spaces to execute parallel operations.
+These include parallel for, reduce, and scan (see Chapter \ref{C:Dispatch})
+as well as View initialization (see Chapter \ref{C:View}).
+We name the class of all such operations \emph{parallel dispatch}.
+
+In Kokkos, parallel dispatch executes \emph{asynchronously}.  
+This means that they may return ``early,'' before they have actually completed.
+However, despite this asynchrony,
+they execute in sequence with respect to other Kokkos operations.
+This matters for things like timing.
+For example, a \lstinline!parallel_for! may return ``right away,''
+so if you want to measure how long it takes,
+you must first call \lstinline!fence()! on that execution space.
+This forces all functors to complete before \lstinline!fence()! returns.
+
+\section{Atomic updates}\label{S:Model:Atomic}
+
+Some execution spaces may implement \emph{atomic updates} with respect to some memory spaces.
+An atomic update runs on an execution space, and changes a word of data in a memory space.
+The atomic update behaves atomically with respect to other atomic updates happening concurrently by that execution space on that memory space.
+``Behaves atomically'' means that all of the atomic updates will complete,
+and the word of data will be changed in a way that respects some order of those update operations.
+Kokkos does not promise any particular order.
+Furthermore, for a nonatomic memory read or write to ``see'' the results of atomic updates,
+the execution space must first execute a \lstinline!fence()! operation to that memory space.

### 2015-03-17 Kokkos_PG_MachineModel.tex
Machine model: Filled out the section on "Program execution."

@@
 
+\subsection{Types and instances of spaces}\label{SS:Model:Spaces:Types}
+
+Kokkos distinguishes between \emph{types} of spaces, and \emph{instances} of spaces.
+For example, a node might have multiple NVIDIA GPUs.
+All of these GPUs' memory spaces have the type \lstinline!Kokkos::CudaSpace!,
+but each space is a different instance of \lstinline!Kokkos::CudaSpace!.
+The common case of using Kokkos presumes a single instance of a given execution or memory space.
+In that case, one uses the type of the space as a synonym for the default (single) instance of that space.
+When we speak of a space by its type,
+we mean the default single instance of that space,
+unless we say otherwise.
+
+\subsection{Analogy with operating system processes}\label{SS:Model:Spaces:Analogies}
+
 Here is an analogy: Many operating systems have ``processes.''
@@
 
-Kokkos distinguishes between \emph{types} of spaces, and \emph{instances} of spaces.
-For example, a node might have multiple NVIDIA GPUs.
-All of these GPUs' memory spaces have the type \lstinline!Kokkos::CudaSpace!,
-but each space is a different instance of \lstinline!Kokkos::CudaSpace!.
-The common case of using Kokkos presumes a single instance of a given execution or memory space.
-In that case, one uses the type of the space as a synonym for the default (single) instance of that space.
-When we speak of a space by its type,
-we mean the default single instance of that space,
-unless we say otherwise.
-
 % Kokkos' abstract machine model may contain $N$ types of execution spaces
@@
 % It is possible that $N \neq M$.
+
+\section{Thread teams}\label{S:Model:Teams}
+
+Kokkos supports multiple levels of parallelism.
+In particular, it groups threads into \emph{teams}.
+A \emph{thread team} is a collection of one or more parallel ``threads'' of execution.
+Kokkos allows an arbitrary number of teams -- the \emph{league size}.
+Hardware constrains the number of threads in a team -- the \emph{team size}.
+
+Threads in a team can synchronize -- they have a ``barrier'' primitive -- 
+and share a ``scratch pad'' memory which they may use for temporary storage.
+Scratch pad memory exists only during parallel operations;
+allocations in it do not persist across kernels.
+Teams themselves may run in any order,
+and may not necessarily run all in parallel.
+For example, if the user asks for $T$ teams,
+the hardware may choose to run them one after another in sequence,
+or in groups of up to $G$ teams at a time in parallel.
+
+Users may \emph{nest} parallel operations.
+Teams may perform one parallel operation (for, reduce, or scan),
+and threads within each team may perform another, possibly different parallel operation.
+Different teams may do entirely different things.
+For example, all the threads in one team may execute a \lstinline!parallel_for!,
+and all the threads in a different team may execute a \lstinline!parallel_scan!.
+Different threads within a team may also do different things.
+However, performance may vary if threads in a team ``diverge'' in their behavior
+(e.g., take different sides of a branch).
+Chapter \ref{C:Hierarchical} shows how the C++ implementation of Kokkos exposes thread teams.
+
+NVIDIA's CUDA programming model inspired Kokkos' thread team model.
+The scratch pad memory corresponds with CUDA's per-team ``shared memory.''
+The ``league / team'' vocabulary comes from OpenMP 4.0,
+and has many aspects in common with our thread team model.
+We have found that programming to this model results in good performance,
+even on computer architectures which only implement parts of the full model.
+For example, most multicore processors in common use for high-performance computing lack ``scratch pad'' hardware.
+However, if users request a scratch pad size that fits comfortably in the largest cache shared by the threads in a team,
+programming as if a scratch pad exists forces users to address locality in their algorithms.
+This also reflects the common experience that rewriting a code for more restrictive hardware,
+then porting the code \emph{back} to conventional hardware,
+tends to improve performance relative to an unoptimized code.
+
+\section{Program execution}\label{S:Model:Exec}
+
+It is tempting to try to define formally what it means for a processor to execute code.
+None of us authors have a background in logic or what computer scientists call ``formal methods,''
+so our attempt might not go very far!
+We will stick with informal definitions and rely on Kokkos' C++ implementation
+as an existence proof that the definitions make sense.
+
+Kokkos lets users tell execution spaces to execute parallel operations.
+These include parallel for, reduce, and scan (see Chapter \ref{C:Dispatch})
+as well as View allocation and initialization (see Chapter \ref{C:View}).
+We name the class of all such operations \emph{parallel dispatch}.
+
+From our perspective, there are three kinds of code:
+\begin{enumerate}
+\item Code executing inside of a Kokkos parallel operation
+\item Code outside of a Kokkos parallel operation that asks Kokkos to
+  do something (e.g., parallel dispatch itself)
+\item Code that has nothing to do with Kokkos
+\end{enumerate}
+The first category is the most restrictive.
+Section \ref{S:Model:Teams} above explains restrictions on inter-team synchronization.
+In general, we limit the ability of Kokkos-parallel code to invoke Kokkos operations
+(other than for nested parallelism; 
+see Section \ref{S:Model:Teams} above and Chapter \ref{C:Hierarchical}).
+We also forbid dynamic memory allocation (other than from the team's scratch pad) in parallel operations.
+Whether Kokkos-parallel code may invoke operating system routines or third-party libraries
+depends on the execution and memory spaces being used.
+Regardless, restrictions on inter-team synchronization have implications for things like filesystem access.
+
+\emph{Kokkos threads are for computing in parallel,}
+not for overlapping I/O and computation,
+and not for making graphical user interfaces responsive.
+Use other kinds of threads (e.g., operating system threads) for the latter two purposes.
+You may be able to mix Kokkos' parallelism with other kinds of threads;
+see Section \ref{SS:Model:Exec:ThreadSafety}.
+Kokkos' developers are also working on a task parallelism model
+that will work with Kokkos' existing data-parallel constructs.
+
+\subsection{Reproducible reductions and scans}\label{SS:Model:Exec:Repro}
+
+Kokkos promises \emph{nothing} about the order in which the iterations of a parallel loop occur.
+However, it \emph{does} promise that if you execute the same parallel reduction or scan,
+using the same hardware resources and run-time settings,
+then you will get the same results each time you run the operation.
+``Same results'' even means ``with respect to floating-point rounding error.''
+
+\subsection{Asynchronous parallel dispatch}\label{SS:Model:Exec:Async}
+
+This concerns the second category of code, that calls Kokkos operations.
+In Kokkos, parallel dispatch executes \emph{asynchronously}.  
+This means that it may return ``early,'' before it has actually completed.
+Nevertheless, it executes \emph{in sequence} with respect to other Kokkos operations on the same execution or memory space.
+This matters for things like timing.
+For example, a \lstinline!parallel_for! may return ``right away,''
+so if you want to measure how long it takes,
+you must first call \lstinline!fence()! on that execution space.
+This forces all functors to complete before \lstinline!fence()! returns.
+
+\subsection{Thread safety?}\label{SS:Model:Exec:ThreadSafety}
+
+Users may wonder about ``thread safety,'' that is,
+whether multiple operating system threads may safely call into Kokkos concurrently.
+Kokkos' thread safety depends on both its implementation, 
+and on the execution and memory spaces that the implementation uses.
+The C++ implementation has made great progress towards (non-Kokkos) thread safety of View memory management.
+For now, however, the most portable approach is for only one (non-Kokkos) thread of execution to control Kokkos.
+Also, be aware that operating system threads might interfere with Kokkos' performance,
+depending on the execution space that you use.
+
+Thread safety with respect to Kokkos' threads is a different matter.
+View (see Chapter \ref{C:View}) promises safety of its memory management.
+That is, you may safely access Views inside of Kokkos' parallel operations.
+Kokkos provides some synchronization constructs as well; see the following section.
 
@@
 
-Kokkos imposes a very weak consistency model.
+Kokkos imposes a very weak consistency model across an entire execution space.
 For example, it does \emph{not} assume sequential consistency.
@@
 
-\section{Asynchronous parallel dispatch}\label{S:Model:Async}
-
-Kokkos lets users tell execution spaces to execute parallel operations.
-These include parallel for, reduce, and scan (see Chapter \ref{C:Dispatch})
-as well as View initialization (see Chapter \ref{C:View}).
-We name the class of all such operations \emph{parallel dispatch}.
-
-In Kokkos, parallel dispatch executes \emph{asynchronously}.  
-This means that they may return ``early,'' before they have actually completed.
-However, despite this asynchrony,
-they execute in sequence with respect to other Kokkos operations.
-This matters for things like timing.
-For example, a \lstinline!parallel_for! may return ``right away,''
-so if you want to measure how long it takes,
-you must first call \lstinline!fence()! on that execution space.
-This forces all functors to complete before \lstinline!fence()! returns.
-
-\section{Atomic updates}\label{S:Model:Atomic}
+\subsection{Atomic updates}\label{SS:Model:Consistency:Atomic}
 

### 2015-03-17 Kokkos_PG_Subviews.tex
Subviews: Minor formatting and wording changes

@@
 \item A \emph{subview} is a slice of an existing Kokkos View
-\item A subview has the same reference count as their parent View
-\item Use C++11 \texttt{auto} to let Kokkos pick the subview's type
+\item A subview has the same reference count as its parent View
+\item Use C++11 type inference (\lstinline!auto!) to let Kokkos pick the subview's type
 \end{itemize}
@@
 
-A \emph{slice} of a multidimensional array behaves as an array, and is
-a view of a structured subset of the original array.  ``Behaves as an
-array'' means that the slice has the same syntax as an array should;
-one can access its entries using array indexing notation.  ``View''
-means that the slice and the original array point to the same data.
-That is, the slice sees changes to the original array, and vice versa.
+In Kokkos, a \emph{subview} is a slice of a View.  A \emph{slice} of a
+multidimensional array behaves as an array, and is a view of a
+structured subset of the original array.  ``Behaves as an array''
+means that the slice has the same syntax as an array should; one can
+access its entries using array indexing notation.  ``View'' means that
+the slice and the original array point to the same data.  That is, the
+slice sees changes to the original array, and vice versa.
 ``Structured subset'' means a cross product of indices along each
@@
 all entries whose indices are $(\alpha_0, \alpha_1, \dots,
-\alpha_{k-1})$, where $\alpha_j$ is a subset of $[0, 1, \dots,
-N_j-1]$.
+\alpha_{k-1})$, where $\alpha_j$ is an ordered subset of $\{0, 1,
+\dots, N_j-1\}$.
 
@@
 
-In Kokkos, a \emph{subview} is a slice of a View.  A subview has the
-same reference count as its parent View, so the parent View won't be
-deallocated before all subviews go away.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-To take a subview of a View, use the \texttt{subview} function.  This
-function is overloaded for all different kinds of Views and index
-ranges.  For example, the following is equivalent to the above example
-\texttt{A(2:4, 3:7)}:
-
+To take a subview of a View, use the \texttt{Kokkos::subview}
+function.  This function is overloaded for all different kinds of
+Views and index ranges.  For example, the following code is equivalent
+to the above example \texttt{A(2:4, 3:7)}:
+%%
 \begin{lstlisting}
@@
 
-auto A_sub = 
-  subview (A, std::make_pair (2, 4), std::make_pair (3, 7));
+auto A_sub = subview (A, make_pair (2, 4), make_pair (3, 7));
 \end{lstlisting}
+%
+In the above example and those that follow in this chapter, 
+we assume that \lstinline!Kokkos::View!, \lstinline!Kokkos::subview!,
+\lstinline!Kokkos::ALL!, \lstinline!std::make_pair!, and \lstinline!std::pair!
+have been imported into the working C++ namespace.
 
-The Kokkos equivalent of a contiguous index range \texttt{3:7} is
-\\ \lstinline!std::pair<size_t, size_t>(3, 7)!.  The Kokkos equivalent of
+The Kokkos equivalent of a contiguous index range \texttt{3:7} is \\
+\lstinline!pair<size_t, size_t>(3, 7)!.  The Kokkos equivalent of
 \texttt{:} (a colon by itself; the whole index range for that
-dimension) is \texttt{ALL()} (an instance of the \texttt{Kokkos::ALL}
-class).  Kokkos does not currently have equivalents of the strided or
-arbitrary index sets.
+dimension) is \texttt{ALL()} (an instance of the \texttt{ALL} class,
+which Kokkos uses only for this purpose).  Kokkos does not currently
+have equivalents of the strided or arbitrary index sets.
 
-Every subview is also a View.  This means that you may take a subview
-of a subview.
+A subview has the same reference count as its parent View, so the
+parent View won't be deallocated before all subviews go away.  Every
+subview is also a View.  This means that you may take a subview of a
+subview.
 
@@
 if any of those dimensions is zero.  Degenerate Views are useful for
-keeping code simple, by avoiding special cases.  For example, in an
-MPI (Message-Passing Interface) distributed-memory parallel code that
-uses Kokkos to represent local (per-process) data structures,
+keeping code simple, by avoiding special cases.  For example, consider
+a MPI (Message-Passing Interface) distributed-memory parallel code
+that uses Kokkos to represent local (per-process) data structures.
+Suppose that the code distributes a dense matrix (2-D array) in block
+row fashion over the MPI processes in a communicator.  It could be
+that some processes own zero rows of the matrix.  This may not be
+efficient, since those processes do no work yet participate in
+collectives, but it might be possible.  In this case, allowing Views
+with zero rows would reduce the number of special cases in the code.

### 2015-03-20 Kokkos_PG_HierarchicalParallelism.tex
Replaced [=] with KOKKOS_LAMBDA; explained Kokkos::single better.
I added an explanation of the KOKKOS_LAMBDA macro.  If CUDA is OFF, it
just turns into [=].  If CUDA is ON, it may do something special to make
the lambda work with CUDA.

I also improved the explanation of Kokkos::single.  In particular,
Kokkos::single's lambda is NOT allowed to modify any captured variable
by reference.

@@
 
-\subsection{Creating a Policy Instance}\label{SS:Hierarchical:Teams:Policy}
+\subsection{Creating a Policy instance}\label{SS:Hierarchical:Teams:Policy}
 
@@
 
-\subsection{Basic Kernels}\label{SS:Hierarchical:Teams:Kernels}
+\subsection{Basic kernels}\label{SS:Hierarchical:Teams:Kernels}
 
-Within the kernel the team policy member type provides the necessary functionality to use teams. 
-It allows access to thread identifiers such as league rank and size and team rank and size, as well as team synchronous actions such as team barriers, reductions and scans.
+The team policy's \lstinline!member_type! provides the necessary functionality to use teams within a parallel kernel.
+It allows access to thread identifiers such as the league rank and size, and the team rank and size.
+It also provides team-synchronous actions such as team barriers, reductions and scans.
 \begin{lstlisting}
@@
 
-This approach of writing hierarchical parallelism algorithms makes it very explicit that a kernel using a \lstinline!TeamPolicy! constitutes a parallel region with respect to the team.
+The name ``\lstinline!TeamPolicy!'' makes it explicit that a kernel
+using it constitutes a parallel region with respect to the team.
 
@@
 
-The most simple use case is to have another \lstinline|parallel_for| nested inside a kernel. 
+The simplest use case is to have another \lstinline|parallel_for| nested inside a kernel. 
 \begin{lstlisting}
-Kokkos::parallel_for(Kokkos::TeamPolicy<>(league_size,team_size), 
-                     KOKKOS_LAMBDA ( member_type team_member) {
+using Kokkos::parallel_for;
+using Kokkos::TeamPolicy;
+using Kokkos::TeamThreadRange;
+
+parallel_for (TeamPolicy<> (league_size, team_size), 
+                    KOKKOS_LAMBDA (member_type team_member)
+{
   Scalar tmp;
-  Kokkos::parallel_for(
-    Kokkos::TeamThreadRange(team_member,loop_count), 
+  parallel_for (TeamThreadRange (team_member, loop_count), 
     [=] (int& i) {
-    ....
+      // ...
       // tmp += i; // This would be an illegal access
@@
 
-\subsection{Execution Restriction with \lstinline|single|}
+\subsection{Restrict execution with \lstinline|single|}
 
-As stated in the earlier subsections, a kernel has to be treated as a parallel region with respect to threads (and vector lanes) within a team.
+As stated above, a kernel is a parallel region with respect to threads (and vector lanes) within a team.
 This means that global memory accesses outside of the respective nested levels potentially have to be protected against repetitive execution. 
 A common example is the case where a team performs some calculation but only one result per team has to be written back to global memory.
+
 Kokkos provides the \lstinline|Kokkkos::single(Policy,Lambda)| function for this case.
-The two policies \lstinline|Kokkos::PerTeam| and \lstinline|Kokkos::PerThread| restrict execution to once per thread or once per team.
+It currently accepts two policies:
+\begin{itemize}
+\item \lstinline|Kokkos::PerTeam| restricts execution of the lambda's
+  body to once per team
+\item \lstinline|Kokkos::PerThread| restricts execution of the
+  lambda's body to once per thread (that is, to only one vector lane
+  in a thread)
+\end{itemize}
+The \lstinline!single! function takes a lambda as its second argument.
+That lambda takes zero arguments, 
+so its body must perform side effects in order to have an effect.
+It must always be correct for the lambda to capture variables by value
+(\lstinline![=]!, not \lstinline![&]!).
+Thus, if the lambda captures by reference,
+it must \emph{not} modify variables that it has captured by reference.
 
@@
 using Kokkos::PerThread;
+
+TeamPolicy<...> policy (...);
+typedef TeamPolicy<...>::member_type team_member;
+
+parallel_for (policy, KOKKOS_LAMBDA (const team_member& thread) {
   // ...
-parallel_for (TeamThreadRange (thread, 100), [=] (int i) {
+
+  parallel_for (TeamThreadRange (thread, 100), 
+    KOKKOS_LAMBDA (const int& i) {
       double sum = 0;
@@
   parallel_reduce (TeamThreadRange (thread, 99), 
-  [=] (int i, double& lsum) {
-    // Add the result value into a team shared array
-    // Make sure its only added once per thread
+    KOKKOS_LAMBDA (int i, double& lsum) {
+      // Add the result value into a team shared array.
+      // Make sure it is only added once per thread.
       Kokkos::single (PerThread (thread), [=] () {
@@
 
-// Add the per team contribution to global memory
+  // Add the per team contribution to global memory.
   Kokkos::single (PerTeam (thread), [=] () {
@@
     });
-
-// ...
+  });
 \end{lstlisting}

### 2015-03-20 Kokkos_PG_ParallelDispatch.tex
Minor changes to section labels.
Replaced [=] with KOKKOS_LAMBDA; explained Kokkos::single better.
I added an explanation of the KOKKOS_LAMBDA macro.  If CUDA is OFF, it
just turns into [=].  If CUDA is ON, it may do something special to make
the lambda work with CUDA.

I also improved the explanation of Kokkos::single.  In particular,
Kokkos::single's lambda is NOT allowed to modify any captured variable
by reference.

@@
-\chapter{Parallel dispatch}
+\chapter{Parallel dispatch}\label{C:Dispatch}
 
@@
 parallel operation executes, by specifying an \emph{execution policy}.
+Later chapters will cover more advanced execution policies that allow
+nested parallelism.
+
+Important notes on syntax:
+\begin{itemize}
+\item Use the \lstinline!KOKKOS_INLINE_FUNCTION! macro to mark a
+  functor's methods that Kokkos will call in parallel
+\item Use the \lstinline!KOKKOS_LAMBDA! macro to replace a lambda's
+  capture clause when giving the lambda to Kokkos for parallel
+  execution
+\end{itemize}
 
@@
 You may also explicitly specify variables to capture, but they must be captured by value.
+We prefer that for the outermost level of parallelism (see Chapter \ref{C:Hierarchical}),
+you use the \lstinline!KOKKOS_LAMBDA! macro instead of the capture clause.
+If CUDA is disabled, this just turns into the usual capture-by-value clause \lstinline![=]!.
+That captures variables from the surrounding scope by value.
+Do NOT capture them by reference!
+If CUDA is enabled, this macro may have a special definition 
+that makes the lambda work correctly with CUDA.
+Compare to the \lstinline!KOKKOS_INLINE_FUNCTION! macro,
+which has a special meaning if CUDA is enabled.
+If you do not plan to build with CUDA, you may use \lstinline![=]! explicitly,
+but we find using the macro easier than remembering the capture clause syntax.
 
@@
 double sum = 0.0;
-parallel_reduce (N, [=] (const int i, double& update) {
+// KOKKOS_LAMBDA macro includes capture-by-value specifier [=].
+parallel_reduce (N, KOKKOS_LAMBDA (const int i, double& update) {
     update += x(i); }, sum);
@@
 int sum = 0;
-parallel_reduce (N, [=] (const int i, int& lsum) { lsum += i*i; });
+parallel_reduce (N, KOKKOS_LAMBDA (const int i, int& lsum) { lsum += i*i; });
 \end{lstlisting}
@@
 const size_t N = x.dimension_0 ();
-parallel_scan (N, [=] (const int i, float& upd, const bool final) {
+parallel_scan (N, KOKKOS_LAMBDA (const int i, float& upd, const bool final) {
     if (final) {

### 2015-03-20 Kokkos_PG_Views.tex
Replaced [=] with KOKKOS_LAMBDA; explained Kokkos::single better.
I added an explanation of the KOKKOS_LAMBDA macro.  If CUDA is OFF, it
just turns into [=].  If CUDA is ON, it may do something special to make
the lambda work with CUDA.

I also improved the explanation of Kokkos::single.  In particular,
Kokkos::single's lambda is NOT allowed to modify any captured variable
by reference.

@@
 View<double*[3][4]> a ("some label", N);
-parallel_for (N, [=] (const ptrdiff_t i) {
+// KOKKOS_LAMBDA macro includes capture-by-value specifier [=].
+parallel_for (N, KOKKOS_LAMBDA (const ptrdiff_t i) {
     const size_t j = ...;
@@
 
-  parallel_for (y.dimension_0 (), [=] (const size_type i) {
+  parallel_for (y.dimension_0 (), KOKKOS_LAMBDA (const size_type i) {
       double y_i = y(i);

### 2015-03-23 Kokkos_PG_Views.tex
Some corrections.

@@
 \item \lstinline!View<int*[3]*>!
-\item \lstinline!View<int*[3][8]>!
 \end{itemize}
@@
 global state.  Others may perform badly in parallel, like fine-grained
-dynamic memory allocation.  As a result, Kokkos restricts the set of
-data types \texttt{T} which may be the entries of a View.
+dynamic memory allocation.  Therefore it is strongly advised to use 
+only simple data types inside Kokkos Views.
 
@@
 \end{itemize}
-The latter specifically forbids nontrivial destructors, constructors
-with dynamic memory allocation, or virtual methods of any kind.  In
-particular, entries must have size determined at compile time by the
-\texttt{sizeof} operator.\footnote{Kokkos has a customization facility
-  for users to add support for types with size determined at run time,
-  but this is for expert users only.  For examples, see the Stokhos
-  software package in the Trilinos project.}
+
+While it is in principal possible to have Kokkos Views of arbitrary 
+objects there are a number of restrictions. The \lstinline|T| type 
+must have a default constructor and destructor. For portability reasons
+\lstinline|T| is not allowed to have virtual functions and one is not 
+allowed to call functions with dynamic memory allocations inside of 
+kernel code. Furthermore assignment operators as well as default constructor
+and deconstructor must be marked with \lstinline|KOKKOS_[INLINE_]FUNCTION|. 
+Keep in mind that the semantics of the resulting View are a combination of the 
+Views 'view' semantics and the behaviour of the element type.
 
@@
 in an execution space which is allowed to access that View's memory
-space.  For example, if the default memory space is \lstinline!Cuda!
-(CUDA device memory), you may not access it in host code, unless CUDA
-UVM is enabled.  Furthermore, access costs (e.g., latency and
+space.  For example, if the default execution space is \lstinline!Cuda!, 
+a View for which no specific Memory Space was fiven may not be accessed 
+in host code \footnote{An exemption is if you specified for CUDA compilation
+that the default memory space is CudaUVMSpace, which can be accessed from 
+the host.}
+Furthermore, access costs (e.g., latency and
 bandwidth) may vary, depending on the View's ``native'' memory and
@@
 (leftmost) dimension of the View is the \emph{parallel dimension},
-over which it is most efficient to do parallel array access.
+over which it is most efficient to do parallel array access if the 
+default memory layout is used (e.g. if no specific memory layout is 
+specified).
 
@@
 its \lstinline!execution_space! typedef, and its memory space via its
-\lstinline!memory_space! typedef.  You may also specify the execution space
+\lstinline!memory_space! typedef.  You may also specify the memory space
 explicitly as a template parameter.  For example, the following
@@
 \begin{lstlisting}
-View<int*, Cuda> a ("a", 100000);
+View<int*, CudaSpace> a ("a", 100000);
 \end{lstlisting}
-and the following allocates a View in ``host'' memory, using OpenMP
-for first-touch allocation:
+and the following allocates a View in ``host'' memory, using the default 
+host execution space for first-touch initialization:
 \begin{lstlisting}
-View<int*, OpenMP> a ("a", 100000);
+View<int*, HostSpace> a ("a", 100000);
 \end{lstlisting}
+
+Since there is no bijective association between execution spaces and memory 
+spaces, Kokkos provides a way to explicitly provide both to a View as a 
+\lstinline|Device|. 
+\begin{lstlisting}
+View<int*, Device<Cuda,CudaUVMSpace> > a ("a", 100000);
+View<int*, Device<OpenMP,CudaUVMSpace> > b ("b", 100000);
+\end{lstlisting}
+In this cas \lstinline|a| and \lstinline|b| will live in the same memory space,
+but \lstinline|a| will be initialized on the GPU while \lstinline|b| will be 
+initialized on the host.
 
@@
 \end{lstlisting}
+This is useful in situations where your dominant use of the View
+exhibits a complicated access pattern. In this case it is best
+to run the most costly kernel directly after initialization to 
+execute the first touch pattern and get optimal memory affinity.
 
@@
 For example, a View in the \lstinline!Cuda! space points to CUDA device
-memory.  Nevertheless, sometimes you really need access to the
+memory. Also using raw pointers would normally defeat the usability 
+of polymorpic layouts and automatic padding. 
+Nevertheless, sometimes you really need access to the
 pointer.  For such cases, we provide the \lstinline!ptr_on_device()!
@@
 \end{lstlisting}
+A user is in most cases also allowed to obtain a pointer to a specific 
+element via the usual \lstinline|&| operator. For example
+
+\begin{lstlisting}
+// Legacy function that takes a raw pointer.
+void someLibraryFunction (double* x_raw);
+
+KOKKOS_INLINE_FUNCTION
+void foo(const View<double*>& x) {
+  someLibraryFunction(&x[3]);
+}
+\end{lstlisting}
+
+This is only valid if a Views reference type is an \lstinline|lvalue|.
+That property can be queried statically at compile time from the view through
+its \lstinline|reference_is_lvalue| member.
 
@@
 right to pad each dimension for cache or vector alignment.  You may
-access all the dimensions of a View using the \lstinline!dimension! method,
-and all the strides using the \lstinline!stride! method.  Both take a raw
-integer array, and only fill in as many entries as the number of
-dimensions.  For example:
+access the dimensions of a View using the \lstinline!dimension! method,
+which takes the index of the dimension.
+
+Strides are accessed using the \lstinline!stride! method.  It takes a raw
+integer array, and only fills in as many entries as the rank of the View.
+For example:
 \begin{lstlisting}
@@
 View<int***> a ("a", N0, N1, N2);
-size_t dims[3]
-a.dimension (dims); // fill 'dims' with dimensions
+
+int dim1 = a.dimension (1); // returns dimension 1
 size_t strides[3]
@@
 \end{lstlisting}
-You may also refer to specific dimensions:
+You may also refer to specific dimensions without a runtime parameter:
 \begin{lstlisting}
@@
 column-major arrays.  If you want to give a View to the BLAS or
-LAPACK, that View must be \lstinline!LayoutLeft!.  You may specify the
+LAPACK library, that View must be \lstinline!LayoutLeft!.  You may specify the
 layout as a template parameter of View.  For example:
@@
 You may ask a View for its layout via its \lstinline!array_layout! typedef.
-This can be helpful for C++ template metaprogramming.  For example
-(C++11 only):
+This can be helpful for C++ template metaprogramming.  For example:
 \begin{lstlisting}
@@
 
-\subsection{Random read Views}
+Const Views often allow the compiler to optimize more agressively by allowing it
+to reason about possible write conflicts and data aliasing. For example in a vector
+update \lstinline|a(i+1)+=b(i) with skewed indexing it is safe to vectorize if
+\lstinline|b| is a View of const data.
 
-Const Views are useful for declaring intent and helping you debug at
-compile time.  They also open the door to other, optional
-optimizations.  In particular, Kokkos lets you declare that you plan
+\subsection{Memory Traits}
+
+Another way to get optimized data accesses is to specify a memory trait. 
+These traits are used to declare intended use of the particular view of 
+an allocation. For example a particular kernel might use a view only for 
+streaming writes. By declaring that intention Kokkos can insert the 
+appropriate store intrinsics on each architecture if available.
+
+Currently 
+In particular, Kokkos lets you declare that you plan
 to access a read-only array irregularly (less spatial locality), using

### 2015-03-25 Kokkos_PG_Compilation.tex
Upgraded the embedded Makefile description for the new Makefile.kokkos

@@
 {\tiny KOKKOS\_HAVE\_PTHREADS} & Enable the Threads execution space. & Requires linking with libpthread.\\
+{\tiny KOKKOS\_HAVE\_Serial} & Enable the Serial execution space. & \\
 {\tiny KOKKOS\_HAVE\_CXX11} & Enable internal usage of C++11 features. & The code needs to be compile with the C++11 standard. Most compilers accept the \verb!-std=c++11! flag for this.\\
 {\tiny KOKKOS\_HAVE\_HWLOC} & Enable thread and memory pinning via hwloc. & Requires linking with libhwloc.\\
-{\tiny KOKKOS\_DONT\_INCLUDE \_KOKKOSCORE\_CONFIG\_H} & Do not include the KokkosCore\_config.h file. & Useful when providing all macros via the compile line instead of auto-generating that file. 
-This option is set for the embedded Makefile. \\
 \hline\hline
@@
 
-Kokkos provides an embedded Makefile (\lstinline|kokkos/Makefile.kokkos|) for inclusion in an application Makefile. 
-This embedded Makefile generates a \lstinline|KOKKOS_INC| and a \newline
-\lstinline|KOKKOS_LINK| variable which need to be appended to the compile and the link line respectively in the application Makefile. 
-If you want to include dependencies (i.e. trigger a rebuild of the application object files when Kokkos files change) you can include \lstinline|KOKKOS_HEADERS| in you dependency list.
-Note that you cannot compile and link at the same time!
-Makefile.kokkos uses a number of variables which can be set either in the parent Makefile prior to including it, or on the command line.
-These variables take the form of \lstinline|SOME_VAR=yes/no| or \lstinline|SOME_PATH=/path/to/library| and control build options such as the target architecture. 
-A list can be found in Table \ref{TBL:MakefileOptions}.
-Example application Makefiles can be found in the tutorial examples under \verb!kokkos/example/tutorial!.
+The base of the build system is the Makefile.kokkos, which is designed to be included by application Makefiles.
+It contains logic to (re)generate the KokkosCore\_config.h file if necessary, build the Kokkos library, and provide updated compiler and linker flags.
+
+The system can digest a number of variables which are used to configure Kokkos settings. 
+Generally the variables are parsed for Keywords. 
+This allows for multiple options being given for each variable. 
+The separator doesn't matter as long as it doesn't interact with the Make system.
+A list of variables, their meaning and options is given in table \ref{TBL:MakefileOptions}.
 
 \begin{table}
-\caption{Table of Makefile options}
+\caption{Variables for the embedded Makefile}
 \label{TBL:MakefileOptions}
 \begin{small}
-\begin{tabular}[t]{lp{0.7\textwidth}}
+\begin{tabular}[t]{p{0.3\textwidth}p{0.65\textwidth}}
 \hline\hline
-Option & Description \\
+Variable & Description \\
 \hline
-CUDA & Enable the CUDA execution space. \\
-OMP & Enable the OpenMP execution space. \\
-PTHREADS & Enable the Threads execution space.\\
-MIC & Enable compilation for Intel Xeon Phi. \\
-CXX11 & Enable internal usage of C++11 features.\\
-HWLOC & Enable hwloc usage.\\
-LIBRT & Enable librt usage for the \lstinline|Kokkos::Impl::Timer| class.\\
-CUDA\_PATH & Set the path to the CUDA toolkit directory.\\
-HWLOC\_PATH & Set the path to the hwloc library.\\
+   KOKKOS\_PATH (IN): & Path to the Kokkos root or install directory. 
+                   One can either build against an existing install 
+                   of Kokkos or use its source directly for an 
+                   embedded build. In the former case the "Input variables"
+                   are set inside the embedded Makefile.kokkos and it 
+                   is not valid to set them differently in the including
+                   Makefile.\\
+
+\hline
+   CUDA\_PATH (IN): & Path to the Cuda toolkit root directory.\\
+
+\hline
+   HWLOC\_PATH (IN): & Path to the hardware locality library. \\
+
+\hline
+   KOKKOS\_DEVICES (IN): & What Execution and Memory Spaces should be enabled. \\
+      \hspace{0.5cm}Options: & OpenMP,Serial,Pthreads,Cuda \\
+      \hspace{0.5cm}Default: & OpenMP \\
+
+\hline    
+    KOKKOS\_ARCH (IN): & What backend Architecture to build for. \\
+      \hspace{0.5cm}Options: & KNC,SNB,HSW,Kepler,Kepler30,Kepler35,Kepler37,Maxwell,
+               Maxwell50,ARMv8,BGQ,Power7,Power8 \\
+      \hspace{0.5cm}Default: & None -- this means no particular architecture flags
+               are set. \\
+
+\hline
+    KOKKOS\_DEBUG (IN): & Enable debugging. \\
+      \hspace{0.5cm}Options: & yes,no \\
+      \hspace{0.5cm}Default: & no \\
+
+\hline
+    KOKKOS\_CXX\_STANDARD (IN): & Set the C++ standard to be used. \\
+      \hspace{0.5cm}Options: & c++11 \\
+      \hspace{0.5cm}Default: & c++11 \\
+
+\hline\hline
+
+    KOKKOS\_CPPFLAGS (OUT): & Pre processor flags (include directories and defines). 
+                              Add this to applications compiler and preprocessor flags. \\
+
+\hline
+    KOKKOS\_CXXFLAGS (OUT): & Compiler flags. Add this to the applications compiler flags. \\
+
+\hline
+    KOKKOS\_LDFLAGS (OUT): & Linker flags. Add this to the applications linker flags. \\
+
+\hline
+    KOKKOS\_LIBS (OUT): & Libraries required by Kokkos. Add this to the link line after the linker flags. \\
+
+\hline
+    KOKKOS\_CPP\_DEPENDS (OUT): & Dependencies for compilation units which include any Kokkos header files.
+                                  Add this as a dependency to compilation targets including any Kokkos code. \\
+
+\hline
+    KOKKOS\_LINK\_DEPENDS (OUT): & Dependencies of an application linking in the Kokkos library.
+                                   Add this to the dependency list of link targets. \\
+
+\hline\hline
+    CXXFLAGS (IN): & User provided compiler flags which will be used to compile the Kokkos library. \\
+
+\hline
+    CXX (IN): & The compiler used to compile the Kokkos library. \\
 \hline\hline
@@
 \end{table}
+
+    A word of caution on where to include the Makefile.kokkos:
+    since the embedded Makefiles defines targets it is usually
+    better to include it after the first application target
+    has been defined. Since that target can't use the flags
+    from the embedded Makefiles it should be a meta target:
+
+\begin{lstlisting}    
+    CXX=g++
+
+    default: main
+
+    include Makefile.kokkos
+
+    main: $(KOKKOS_LINK_DEPENDS) $(KOKKOS_CPP_DEPENDS) main.cpp
+        $(CXX) $(KOKKOS_CPPFLAGS) $(KOKKOS_CXXFLAGS) \
+        $(KOKKOS_LDFLAGS) $(KOKKOS_LIBS) main.cpp -o main
+\end{lstlisting}
+
+More example application Makefiles can be found in the tutorial examples under \verb!kokkos/example/tutorial!.
 

### 2015-03-25 Kokkos_PG_Views.tex
Fixed missing character
Upgraded the embedded Makefile description for the new Makefile.kokkos

@@
 only simple data types inside Kokkos Views.
-
 Users may always construct a View whose entries are 
@@
 space.  For example, if the default execution space is \lstinline!Cuda!, 
-a View for which no specific Memory Space was fiven may not be accessed 
+a View for which no specific Memory Space was given may not be accessed 
 in host code \footnote{An exemption is if you specified for CUDA compilation
 that the default memory space is CudaUVMSpace, which can be accessed from 
-the host.}
+the host.}.
 Furthermore, access costs (e.g., latency and
@@
 to reason about possible write conflicts and data aliasing. For example in a vector
-update \lstinline|a(i+1)+=b(i) with skewed indexing it is safe to vectorize if
+update \lstinline|a(i+1)+=b(i)| with skewed indexing it is safe to vectorize if
 \lstinline|b| is a View of const data.

### 2015-03-27 Kokkos_PG_LegacyCodeIntegration.tex
Added new chapter which will discuss legacy code transisition and interoperability

\chapter{Interoperability and Legacy Codes}
\section{Legacy data structures}

There are two principal mechanism to facilitate interoperability with legacy data 
structures: (i) Kokkos allocates data and raw pointers are extracted to create 
legacy data structures and (ii) unmanaged views can be used to view externally 
allocated data. In both cases it is mandatory to fix the Layout of the Kokkos view
to the actual layout used in the legacy data structure.

\subsection{Kokkos manages memory}

\subsubsection{Raw allocations}
\subsubsection{Views as the fundamental data owning structure}

\subsection{External memory management}

\subsection{\lstinline|std::vector|}

\section{Calling non-Kokkos libraries}

\section{Keeping track of data transfers}

### 2015-03-27 Kokkos_PG_Views.tex
Worked on view section
this cleans up and reorders some sections to make the flow clearer.
I also expanded a few subsections and corrected errenous statements.

@@
 
+\subsection{Const Views}
+
+A view can have const data semantics (i.e. its entries are read-only) by 
+specifying a `const` data type.  It is a compile-time error to assign to an 
+entry of a ``const View''.  Assignment semantics are equivalent to a pointer to 
+const data.  
+A const View means the \emph{entries} are const.  You
+may still assign to a const View.  \lstinline!View<const double*>!
+corresponds exactly to \lstinline!const double*!, and
+\lstinline!const View<double*>! to \lstinline!double* const!.
+Therefore it does not make sense to allocate a const View,
+since you could not obtain a non-const view of the same data and you 
+can not assign to it. You can however assign a non-const view to a const
+view.  Here is an example:
+\begin{lstlisting}
+const size_t N0 = ...;
+View<double*> a_nonconst ("a_nonconst", N0);
+
+// Assign a nonconst View to a const View
+View<const double*> a_const = a_nonconst;
+// Pass the const View to some read-only function.
+const double result = readOnlyFunction (a_const);
+\end{lstlisting}
+
+Const Views often enables the compiler to optimize more aggressively by allowing it
+to reason about possible write conflicts and data aliasing. For example in a vector
+update \lstinline|a(i+1)+=b(i)| with skewed indexing it is safe to vectorize if
+\lstinline|b| is a View of const data.
+
 \subsection{Accessing entries (indexing)}
@@
 
-\subsection{Memory management and copying}
-
-\subsubsection{Reference counting}
+\subsection{Reference counting}
 
@@
 
-\subsubsection{Deep copy}
+\subsection{Resizing}
 
-If you want to make a deep copy of a View's contents, you must do so
-explicitly, using the \lstinline!deep_copy! function.  For example:
-\begin{lstlisting}
-View<int*> a ("a", 10);
-View<int*> b ("b", 10);
-deep_copy (a, b); // copy contents of b into a
-\end{lstlisting}
-
-\subsection{Where and how is the View allocated?}
-
-\subsubsection{Execution and memory spaces}
-
-Views are allocated by default, in the default execution space's
-default memory space.  You may access the View's execution space via
-its \lstinline!execution_space! typedef, and its memory space via its
-\lstinline!memory_space! typedef.  You may also specify the memory space
-explicitly as a template parameter.  For example, the following
-allocates a View in CUDA device memory:
-\begin{lstlisting}
-View<int*, CudaSpace> a ("a", 100000);
-\end{lstlisting}
-and the following allocates a View in ``host'' memory, using the default 
-host execution space for first-touch initialization:
-\begin{lstlisting}
-View<int*, HostSpace> a ("a", 100000);
-\end{lstlisting}
-
-Since there is no bijective association between execution spaces and memory 
-spaces, Kokkos provides a way to explicitly provide both to a View as a 
-\lstinline|Device|. 
-\begin{lstlisting}
-View<int*, Device<Cuda,CudaUVMSpace> > a ("a", 100000);
-View<int*, Device<OpenMP,CudaUVMSpace> > b ("b", 100000);
-\end{lstlisting}
-In this cas \lstinline|a| and \lstinline|b| will live in the same memory space,
-but \lstinline|a| will be initialized on the GPU while \lstinline|b| will be 
-initialized on the host.
-
-\subsubsection{Initialization}
-
-A View's entries are initialized to zero by default.  Initialization
-happens in parallel, for first-touch allocation over the first
-(leftmost) dimension of the View.
-
-You may allocate a View without initializing.  For example:
-\begin{lstlisting}
-View<int*> x (ViewAllocateWithoutInitializing (label), 100000);
-\end{lstlisting}
-This is useful in situations where your dominant use of the View
-exhibits a complicated access pattern. In this case it is best
-to run the most costly kernel directly after initialization to 
-execute the first touch pattern and get optimal memory affinity.
-
-\subsection{How do I get the raw pointer?}
-
-We discourage access to a View's ``raw'' pointer.  This circumvents
-reference counting.  That is, the memory may be deallocated once the
-View's reference count goes to zero, so holding on to a raw pointer
-may result in invalid memory access.  Furthermore, it may not even be
-possible to access the View's memory from a given execution space.
-For example, a View in the \lstinline!Cuda! space points to CUDA device
-memory. Also using raw pointers would normally defeat the usability 
-of polymorpic layouts and automatic padding. 
-Nevertheless, sometimes you really need access to the
-pointer.  For such cases, we provide the \lstinline!ptr_on_device()!
-method.  For example:
-\begin{lstlisting}
-// Legacy function that takes a raw pointer.
-extern void legacyFunction (double* x_raw, const size_t len);
-
-// Your function that takes a View.
-void myFunction (const View<double*>& x) {
-  // DON'T DO THIS UNLESS YOU MUST
-  double* a_raw = a.ptr_on_device ();
-  const size_t N = x.dimension_0 ();
-  legacyFunction (a_raw, N);
-}
-\end{lstlisting}
-A user is in most cases also allowed to obtain a pointer to a specific 
-element via the usual \lstinline|&| operator. For example
+Kokkos Views can be resized using the \lstinline|resize| non-member function.
+It takes an existing view as its input by reference and the new dimension information
+corresponding to the constructor arguments. A new view with the new dimensions will
+be created and a kernel will be run in the views execution space to copy the data 
+element by element from the old view to the new one. Note that the old allocation is 
+only deleted if the view to be resized was the {\it only} view referencing the underlying 
+allocation.
 
 \begin{lstlisting}
-// Legacy function that takes a raw pointer.
-void someLibraryFunction (double* x_raw);
+// Allocate a view with 100x50x4 elements 
+View<int**[4]> a( "a", 100,50);
 
-KOKKOS_INLINE_FUNCTION
-void foo(const View<double*>& x) {
-  someLibraryFunction(&x[3]);
-}
+// Resize a to 200x50x4 elements; the original allocation is freed
+resize(a, 200,50);
+
+// Create a second view b viewing the same data as a
+View<int**[4]> b = a;
+// Resize a again to 300x60x4 elements; b is still 200x50x4
+resize(a,300,60);
 \end{lstlisting}
-
-This is only valid if a Views reference type is an \lstinline|lvalue|.
-That property can be queried statically at compile time from the view through
-its \lstinline|reference_is_lvalue| member.
 
@@
 prevent false sharing of a cache line between threads. 
+In section \ref{S:View:Placement} more details will be discussed.
 
@@
 
-TODO: Hey Christian, what did you mean by this?
+\subsection{Memory spaces}
 
-Christian says: \lstinline!CudaSpace!, \lstinline!HostSpace!, etc.
-Also, please emphasize memory space over execution space, and talk
-about the UVM and pinned variants of \lstinline!CudaSpace!.
+Views are allocated by default, in the default execution space's
+default memory space.  You may access the View's execution space via
+its \lstinline!execution_space! typedef, and its memory space via its
+\lstinline!memory_space! typedef.  You may also specify the memory space
+explicitly as a template parameter.  For example, the following
+allocates a View in CUDA device memory:
+\begin{lstlisting}
+View<int*, CudaSpace> a ("a", 100000);
+\end{lstlisting}
+and the following allocates a View in ``host'' memory, using the default 
+host execution space for first-touch initialization:
+\begin{lstlisting}
+View<int*, HostSpace> a ("a", 100000);
+\end{lstlisting}
+
+Since there is no bijective association between execution spaces and memory 
+spaces, Kokkos provides a way to explicitly provide both to a View as a 
+\lstinline|Device|. 
+\begin{lstlisting}
+View<int*, Device<Cuda,CudaUVMSpace> > a ("a", 100000);
+View<int*, Device<OpenMP,CudaUVMSpace> > b ("b", 100000);
+\end{lstlisting}
+In this case \lstinline|a| and \lstinline|b| will live in the same memory space,
+but \lstinline|a| will be initialized on the GPU while \lstinline|b| will be 
+initialized on the host. 
+The \lstinline|Device| type can be accessed as a views \lstinline|device_type|
+typedef. A \lstinline|Device| has only three typedef members 
+\lstinline|device_type|, \lstinline|execution_space|, and \lstinline|memory_space|.
+The \lstinline|execution_space|, and \lstinline|memory_space| typedefs are
+the same for a view and the \lstinline|device_type| typedef.
+
+It is important to understand that accessibility of a View does not 
+depend on its execution space directly. It is only determined by its
+memory space. Therefore both \lstinline|a| and \lstinline|b| have the 
+same access properties. They differ only in how they are initialized, 
+as well as in where parallel kernels associated with operations such 
+as resizing or deep copies are run.
+
+\subsection{Initialization}
+
+A View's entries are initialized to zero by default.  Initialization
+happens in parallel, for first-touch allocation over the first
+(leftmost) dimension of the View using the execution space
+of the View.
+
+You may allocate a View without initializing.  For example:
+\begin{lstlisting}
+View<int*> x (ViewAllocateWithoutInitializing (label), 100000);
+\end{lstlisting}
+This is useful in situations where your dominant use of the View
+exhibits a complicated access pattern. In this case it is best
+to run the most costly kernel directly after initialization to 
+execute the first touch pattern and get optimal memory affinity.
+
+\subsection{Deep copy and HostMirror}
+
+Copying data from one view to another, in particular between views 
+in different memory spaces, is called deep copy. 
+Kokkos never performs a hidden deep copy. To do so a user has to 
+call the \lstinline!deep_copy! function.  For example:
+\begin{lstlisting}
+View<int*> a ("a", 10);
+View<int*> b ("b", 10);
+deep_copy (a, b); // copy contents of b into a
+\end{lstlisting}
+
+Deep copies can only be performed between views with an identical 
+memory layout and padding. For example the following two operations
+are not valid:
+\begin{lstlisting}
+View<int*[3]> a ("a", CudaSpace, 10);
+View<int*[3]> b ("b", HostSpace, 10);
+deep_copy (a, b); // This will give a compiler error
+
+View<int*[3], LayoutLeft> c ("c", CudaSpace, 10);
+View<int*[3], LayoutLeft> d ("d", HostSpace, 10);
+deep_copy (c, d); // This might give a runtime error
+\end{lstlisting}
+
+The first one will not work because the default layouts of \lstinline|CudaSpace|
+and \lstinline|HostSpace| are different. The compiler will catch that since 
+no overload of the \lstinline|deep_copy| function exists to copy view from
+one layout to another. The second case will fail at runtime if padding settings
+are different for the two memory spaces. This would result in different 
+allocation sizes and thus prevent a direct memcopy.
+
+The reasoning for allowing only direct bitwise copies is that a deep copy
+between different memory spaces would otherwise require a temporary 
+copy of the data to which a bitwise copy is performed followed by a parallel
+kernel to transfer the data element by element.
+
+Kokkos provides the following way to work around those limitations. 
+First views have a \lstinline|HostMirror| typedef which is a view type with 
+compatible layout inside the \lstinline|HostSpace|. Additionally there is a
+\lstinline|create_mirror| and \lstinline|create_mirror_view| function which 
+allocate views of the \lstinline|HostMirror| type of a view. 
+The difference between the two is that \lstinline|create_mirror| will always
+allocate a new view, while \lstinline|create_mirror_view| will only create a 
+new view if the original one is not in \lstinline|HostSpace|.
+
+\begin{lstlisting}
+View<int*[3]> a ("a", MemorySpace, 10);
+// Allocate a view in HostSpace with the 
+// layout and padding of a
+typename View<int*[3]>::HostMirror b =
+    create_mirror(a);
+// This is always a memcopy
+deep_copy (a, b);
+
+typename View<int*[3]>::HostMirror c =
+    create_mirror_view(a);
+// This is a no-op if MemorySpace is HostSpace
+deep_copy (a,c)
+\end{lstlisting}
+
+\subsection{How do I get the raw pointer?}
+
+We discourage access to a View's ``raw'' pointer.  This circumvents
+reference counting.  That is, the memory may be deallocated once the
+View's reference count goes to zero, so holding on to a raw pointer
+may result in invalid memory access.  Furthermore, it may not even be
+possible to access the View's memory from a given execution space.
+For example, a View in the \lstinline!Cuda! space points to CUDA device
+memory. Also using raw pointers would normally defeat the usability 
+of polymorpic layouts and automatic padding. 
+Nevertheless, sometimes you really need access to the
+pointer.  For such cases, we provide the \lstinline!ptr_on_device()!
+method.  For example:
+\begin{lstlisting}
+// Legacy function that takes a raw pointer.
+extern void legacyFunction (double* x_raw, const size_t len);
+
+// Your function that takes a View.
+void myFunction (const View<double*>& x) {
+  // DON'T DO THIS UNLESS YOU MUST
+  double* a_raw = a.ptr_on_device ();
+  const size_t N = x.dimension_0 ();
+  legacyFunction (a_raw, N);
+}
+\end{lstlisting}
+A user is in most cases also allowed to obtain a pointer to a specific 
+element via the usual \lstinline|&| operator. For example
+
+\begin{lstlisting}
+// Legacy function that takes a raw pointer.
+void someLibraryFunction (double* x_raw);
+
+KOKKOS_INLINE_FUNCTION
+void foo(const View<double*>& x) {
+  someLibraryFunction(&x[3]);
+}
+\end{lstlisting}
+
+This is only valid if a Views reference type is an \lstinline|lvalue|.
+That property can be queried statically at compile time from the view through
+its \lstinline|reference_is_lvalue| member.
 
@@
 
-\subsection{Const Views}
-
-Different optional template parameters of View let you control how
-Kokkos accesses entries of the array.  The simplest example involves
-making a View \lstinline!const!, that is, making its entries read-only.  It
-is a compile-time error to assign to an entry of a const View.  It
-does not make sense to allocate a const View, but it does make sense
-to assign a non-const View to a const View.  Here is an example:
-\begin{lstlisting}
-const size_t N0 = ...;
-View<double*> a_nonconst ("a_nonconst", N0);
-
-// Assign a nonconst View to a const View
-View<const double*> a_const = a_nonconst;
-// Pass the const View to some read-only function.
-const double result = readOnlyFunction (a_const);
-\end{lstlisting}
-Note that a ``const View'' means the \emph{entries} are const.  You
-may still assign to a const View.  \lstinline!View<const double*>!
-corresponds exactly to \lstinline!const double*!, and
-\lstinline!const View<double*>! to \lstinline!double* const!.
-
-Const Views often allow the compiler to optimize more agressively by allowing it
-to reason about possible write conflicts and data aliasing. For example in a vector
-update \lstinline|a(i+1)+=b(i)| with skewed indexing it is safe to vectorize if
-\lstinline|b| is a View of const data.
-
-\subsection{Memory Traits}
-
 Another way to get optimized data accesses is to specify a memory trait. 
@@
 streaming writes. By declaring that intention Kokkos can insert the 
-appropriate store intrinsics on each architecture if available.
+appropriate store intrinsics on each architecture if available. Access traits
+are specified through an optional template parameter which comes last in the list
+of parameters. Multiple traits can be combined with binary ``or'' operators:
 
-Currently 
-In particular, Kokkos lets you declare that you plan
-to access a read-only array irregularly (less spatial locality), using
-the optional \lstinline!RandomAccess! template parameter.  You may only use
-this with a const View.  For example:
+\begin{lstlisting}
+View<double*, MemoryTraits<SomeTrait> > a;
+View<const double*, MemoryTraits<SomeTrait | SomeOtherTrait> > b;
+View<double*, LayoutLeft, MemoryTraits<SomeTrait | SomeOtherTrait> > c;
+View<double*, MemorySpace, MemoryTraits<SomeTrait | SomeOtherTrait> > d;
+View<double*, LayoutLeft, MemorySpace, MemoryTraits<SomeTrait> > e;
+\end{lstlisting}
+
+\subsection{Atomic access}
+
+The \lstinline!Atomic! access trait lets you create a View of data , such that every
+read or write to any entry uses an atomic update.  This only works for
+certain data types, for which the hardware and programming model
+support atomic updates.  Currently, Kokkos limits atomic updates to
+32-bit or 64-bit types\footnote{On X86 architectures Kokkos also supports 
+128-bit atomics. Types for which atomic access are performed must support the 
+necessary operators such as =,+=,-=,+,- etc.}.
+
+Note that it is a violation of the semantics to write \emph{or} read from a 
+non-atomic view of data if at the same time an atomic view of the data is used.
+
+\begin{lstlisting}
+View<int*> a("a" , 100);
+View<int*, MemoryTraits<Atomic> > a_atomic = a;
+
+a_atomic(1) += 1; // This access will do an atomic addition
+\end{lstlisting}
+
+\subsection{Random Access}
+
+The \lstinline|RandomAccess| trait declares the intent to access a View
+irregularly (in particular non consecutively). If used for a const View 
+in the \lstinline|CudaSpace| or \lstinline|CudaUVMSpace|, Kokkos will use 
+texture fetches for accesses when executing in the \lstinline|Cuda| execution
+space.
+For example:
 \begin{lstlisting}
@@
 \end{lstlisting}
-For a View whose execution space is \lstinline!Cuda!, access to a
+If the default execution space is \lstinline!Cuda!, access to a
 \lstinline!RandomAccess! View may use CUDA texture fetches.  Texture
@@
 
-\lstinline!RandomAccess! is valid for other execution spaces, and Kokkos
-also requires a const View for those execution spaces.  Kokkos
-reserves the right to perform analogous optimizations on other
-execution spaces.
+While \lstinline!RandomAccess! is valid for other execution spaces, 
+currently no specific optimizations are performed. But in the future a view
+allocated with the \lstinline!RandomAccess! attribute might for example
+use a larger page size, and thus reducing page faults in the memory 
+system.
 
@@
   // the reference counting system.
-  View<double*, HostSpace, Unmanaged> x_view (x_raw, N0);
+  View<double*, HostSpace, MemoryTraits<Unmanaged> > x_view (x_raw, N0);
 
@@
 \end{lstlisting}
-
-\subsection{Atomic access}
-
-The \lstinline!Atomic! access trait lets you create a View, such that every
-read or write to any entry uses an atomic update.  This only works for
-certain data types, for which the hardware and programming model
-support atomic updates.  Currently, Kokkos limits atomic updates to
-32-bit or 64-bit built-in types.

### 2015-03-28 Kokkos_PG_Introduction.tex
Updated introduction with some attempt to start at motivations for the reader.

@@
 
-The field of High Performance Computing (HPC) is entering a new era.
-Since the mid 1990's, HPC applications have had great success 
-with distributed-memory parallel computing using the Message-Passing Interface (MPI).
-However, after two decades of MPI dominance, 
-increasing core counts on standard processors and the emergence of new throughput-oriented processors, such as GPUs, make a new approach necessary.
-MPI will not necessarily go away,
-but application developers will need to expose finer-grained shared-memory parallelism
-\emph{as well as} coarser-grained MPI-level parallelism,
-in order to get their money's worth on new architectures.
+\noindent
+The field of High Performance Computing (HPC) is on the verge of entering a new era. The need
+for a fundamental change comes from many angles including the growing acceptance that rates of
+pure computation (often called FLOP/s) are a poor single optimization goal for scientific
+workloads as well as practical challenges in the form of producing energy efficient and cost
+efficient processors. Since the convergence on the Message-Passing Interface (MPI) standard
+in the mid 1990s, application developers have enjoyed a fairly static view of the machine --
+that of a distributed collection of homogeneous nodes executing in collaboration. However, 
+after almost two decades of domaince, the sole use of MPI to derive parallelism is acting as
+a potential limiter to greater future performance. While we expect MPI to continue to function
+as the basic mechanism for communication between compute nodes, additional parallelism is
+likely to be required on the computing node itself if high performance and efficiency are to be
+realized.
 
-Developers have many shared-memory parallel programming model options today.
-Those include OpenMP, OpenACC, CUDA, TBB, Cilk, OpenCL, and C++AMP, among others.
-Some of these programming models are vendor specific, while others have varying degrees of support on different hardware. 
-Most of the models lack a concept to handle complex memory hierarchies, 
-which in addition to threading is one of the defining upcoming trends in HPC.
+In reviewing potential options for the computing nodes of the future the reader might fall
+upon three broad categories of computing device: (1) the multi-core processor with powerful
+serial performance, optimized to reduce operation latency; (2) many-cores processors with
+low to medium powered compute cores that are designed to extent the multi-core concept toward
+throughput based computation and, finally, (3), the general-purpose graphics processor unit
+(GP-GPU, or often, GPU) which is a much more numerous collection of low powered cores designed
+to tolerate long latencies but provide performance through a much higher degree of 
+parallelism. Any combination of these options might also be combined in the future.
 
-The Kokkos programming model addresses the increasing complexity and variety of hardware architectures
-through a rigorous introduction of parallel execution and memory handling abstractions.
-These abstraction layers isolate a software developer from hardware details and allow the implementation of performance-portable applications.
+The diversity of the options available for processor selection raises as interested question
+as to how these should be programmed. Partly due to their heritage but also due to their
+optimized designs, each of these hardware types offers a different programming solution and
+a different set of guidelines to write applications to for highest performance. Options available
+today include a number of shared memory approaches such as OpenMP, Cilk+, Thread Building Blocks
+(TBB) as well as Linux p-threads. To target both contemporary multi/many-core processors and
+GPUs technologies such as OpenMP, OpenACC and OpenCL might be used. Finally, for GPUs a 
+programming model such as CUDA may be selected. This poses a problem to the application
+developer of today which is reminiscent of the challenges before MPI became the default
+communication library -- which model should be selected to provide portability across hardware
+solutions as well as provide reasonable to high performance across each class of processor and
+protect algorithm investment into the future. So far none of these models have been able to 
+provide practical solutions to these questions.
 
-This Guide covers both the programming model, and its implementation as an embedded C++ library.
-It describes the abstraction concepts employed in Kokkos and their exposure as a C++ interface.
-We intend it as an introduction for new users as well as a reference for application developers already using Kokkos.
-We have also supplemented the Guide with tutorial examples in the \verb!example! subdirectory of Kokkos' source code.
+The Kokkos programming model described in this programming guide seeks to address these
+concerns by providing an abstraction of both computation and application data allocation and
+layout. These abstraction layers are designed specifically to isolate software developers for
+the fluctuation in hardware details yet provide portability and high levels of performance
+across architectures.
+
+This guide provides an introduction to the motivations of developing such an abstraction library,
+a coverage of the programming model and its implementation as an embedded C++ library requiring
+no additional modifications to the base C++ language. As such it should be seen as an introduction
+for new users and as a reference for application developers who are already employing Kokkos
+in their applications. Finally, supplementary tutorial examples are included as part of the
+Kokkos software release to help users experiment and explore the programming model through a 
+gradual series of steps.

### 2015-03-28 Kokkos_PG_MachineModel.tex
Added Kokkos model image and some new descriptions of memory spaces/execution spaces.

@@
-\chapter{Machine model}\label{C:Model}
+\chapter{Machine model}
+\label{C:Model}
 
@@
 
-\section{Motivation}\label{S:Model:Motivation}
+\section{Motivation}\label{S:Model:Motivations}
 
-Kokkos is a programming model which happens to have a C++ implementation.
-As you are first learning Kokkos,
-you may find yourself focusing on details of the implementation,
-especially if you have less experience with C++ syntax.
-However, as the syntax becomes more familiar to you,
-you will find it more and more helpful to understand the underlying abstractions.
+Kokkos is comprised of two orthogonal aspects. The first of these is 
+an underlying {\em abstract machine model} which describes fundamental
+concepts required for the development of future portable and performant
+high performance computing applications; the second is a concrete instantiation
+of the programming this model written in C++, which allows programmers
+to write to the concept machine model. It is important to treat these
+two aspects to Kokkos as distinct entities because the underlying model
+being used by Kokkos could, in the future, be instantiated in additional
+languages beyond C++ yet the algorithmic specification would remain
+valid.
 
-Kokkos assumes an \emph{abstract machine model},
-which represents a shared-memory parallel computer architecture.
-The model may represent either a stand-alone computer,
-like a laptop or workstation,
-or a single ``parallel process'' in a distributed-memory parallel programming model such as MPI (the Message-Passing Interface).
-In general, we write \emph{node} to describe the hardware that Kokkos governs.
-Kokkos' machine model lets programmers reason about very complicated heterogenous node architectures.
-It also lets them start with simple, even sequential architectures,
-and gradually port code from there to perform well on more complicated computer architectures.
-The abstract machine model might look complicated,
-but it reflects computers that companies are actually designing and building.
-Many chapters in this Guide show how to use the more complex aspects of the model.
-For example, see Chapter \ref {C:Hierarchical},
-which shows how to use Kokkos to expose multiple levels of parallelism.
+\subsection{Kokkos Abstract Machine Model}
 
-The point of this chapter is that \emph{respecting the abstract machine model ensures performance portability}.
-The phrase \emph{performance portability} means that the same code performs reasonably well on many different computer architectures,
-with no or minimal\footnote{For example, users might change a single C++ typedef or a single configure-time option.} changes to code.
-Kokkos depends on this abstract machine model for performance portability.
-\emph{Implementations} of Kokkos may depend on other things for performance portability.
-For example, efficiency of \lstinline!parallel_for! in the C++ implementation relies in part on compilers' ability to inline code.
-One could imagine a different Kokkos implementation based on source-to-source translation,
-that inlines parallel loop bodies into \lstinline!parallel_for! before they reach the compiler.
-Kokkos' OpenMP and Pthreads back-ends may rely on the Portable Hardware Locality (hwloc) library
-to control assignment of software threads to hardware cores or hyperthreads.
-However, Kokkos as an abstract programming model does not depend on these things.
+\begin{figure}
+\includegraphics[width=2in]{figures/kokkos-node.pdf}
+\caption{Conceptual Model of a Future High Performance Computing Node}
+\label{fig:kokkosnode}
+\end{figure}
 
-\section{Memory spaces and execution spaces}\label{S:Model:Spaces}
+Kokkos assumes an {\em abstract machine model} for the design of
+future shared-memory computing architectures. The model (shown in
+Figure~\ref{fig:kokkosnode}) assumes that there may be multiple
+execution units in a compute node. In the Figure we have elected to
+show two different types of compute units -- one which represents
+multiple latency-optimized cores, similar to contemporary processor
+cores, and a second source of compute in the form of an off die
+accelerator. Of note is that the processor and accelerator each 
+have distinct memories, each with unique performance properties,
+that are accessible across the node ({\em i.e.} the memory is
+reachable or {\em shared} by all execution units. The specific
+layout shown in Figure~\ref{fig:kokkosnode} is an instantiation
+of the Kokkos abstract machine model used to describe the potential
+for multiple types of compute engines and memories within a single
+node. In future systems there may be a range of execution engines
+which are used in the node ranging from a single type of core
+as in many/multi-core processors found today through to an range
+of execution units where many-core processors may be joined to
+numerous types of accelerator cores. In order to ensure portability
+to the potential range of nodes an abstraction of the
+compute engines and available memories are required.
 
-A \emph{memory space} contains data.
-Users access those data as k-dimensional typed arrays, where k may be any nonnegative integer.
-(Zero-dimensional arrays are single values.  See Chapter \ref{C:View} for details.)
-% In the C++ implementation of the Kokkos programming model,
-% these memory addresses are pointers,
-% which point to contiguous address ranges,
-% each of which represents an allocation of data.
-The memory space controls dynamic allocation and deallocation of those arrays,
-if it allows them at all.
+\subsection{Kokkos Execution Spaces}
 
-An \emph{execution space} runs Kokkos' parallel operations,
-like parallel for, reduce, and scan.  (See Chapter \ref{C:Dispatch}.)
-It may access data in some available memory spaces.
-Whether an execution space may access data in a memory space
-depends on \emph{both} the execution space and the memory space.
+Kokkos uses the term {\em execution spaces} to describe a logical
+grouping of computation units which share an indentical set of
+performance properties. An execution space provides a set of 
+parallel execution resources which can be utilized by the
+programmer using several types of fundamental parallel
+operation. For a list of the operations available see
+Chapter~\ref{C:Dispatch}.
 
-\subsection{Types and instances of spaces}\label{SS:Model:Spaces:Types}
+\subsubsection{Execution Space Instances}
 
-Kokkos distinguishes between \emph{types} of spaces, and \emph{instances} of spaces.
-For example, a node might have multiple NVIDIA GPUs.
-All of these GPUs' memory spaces have the type \lstinline!Kokkos::CudaSpace!,
-but each space is a different instance of \lstinline!Kokkos::CudaSpace!.
-The common case of using Kokkos presumes a single instance of a given execution or memory space.
-In that case, one uses the type of the space as a synonym for the default (single) instance of that space.
-When we speak of a space by its type,
-we mean the default single instance of that space,
-unless we say otherwise.
+An {\em instance} of an execution space is a specific instantiation
+of an execution space to which a programmer can target parallel
+work. By means of example, an execution space might be used
+to describe a multi-core processor. In this example, the
+execution space contains several homogeneous cores which share
+some logical grouping. In a program written to the Kokkos model,
+an instance of this execution space would be made available
+on which parallel kernels could be executed. As a second example,
+if we were to add a GPU to the multi-core processor a second
+execution space type is available in the system, the 
+application programmer would then have two 
+execution space instances available to select from. The
+important consideration here is that the method of
+compiling code for different execution spaces and the
+dispatch of kernels to instances is abstracted by the Kokkos
+framework allowed application programmers to be free from 
+written algorithms in hardware specific languages or
+requiring the programmer to understand the restrictions
+required for parallel execution in the execution space.
 
-\subsection{Analogy with operating system processes}\label{SS:Model:Spaces:Analogies}
+\subsection{Kokkos Memory Spaces}
 
-Here is an analogy: Many operating systems have ``processes.''
-A process is like an execution space, in that code may execute in one or more concurrent threads.
-Those threads share access to a common stack and heap, which comprise the process' ``memory space.''
-Two different processes may not access each other's memory spaces by default.
-However, some operating systems have a ``shared memory'' mechanism.\footnote{See e.g., the POSIX \lstinline!shm_open! and related functions.}
-This lets different processes share access to a specific region of memory,
-which in Kokkos terms is like a ``memory space,'' separate from each process' normal memory space.
-Some operating systems also let you open a file,
-and then ``memory map'' that file into a contiguous memory address range.\footnote{See e.g., the POSIX \lstinline!mmap! function.}
-This makes the file behave like yet another memory space.
-Memory-mapped files may have very different performance characteristics than the usual memory,
-even though programmers access both ``memory spaces'' through the same interface.
-Whether or not a process may access a file depends on both the process and the file.
+The multiple types of memory which will become available in future
+computing nodes are abstracted by Kokkos through {\em memory
+spaces}. Each memory space provides a finite storage capacity
+at which data structures can be allocated and accessed. In
+order to provide portable allocations and performant access,
+Kokkos requires data to be allocated as a $k$-dimensioned array,
+where $k$ must be greater than equal to zero (a zero array being
+a scalar). Each dimension of the array can be either fixed in
+size at compile time or dynamically sized with size resolution
+performed during execution.
 
-% Kokkos' abstract machine model may contain $N$ types of execution spaces
-% and $M$ types of memory spaces.
-% It is possible that $N \neq M$.
+\subsubsection{Instances of Kokkos Memory Spaces}
+
+In much the same way execution spaces have specific instaniations through
+the availablilty of an {\em instance} so do memory spaces. An instance
+of a memory space provides a concrete method for the
+application programmer to request data storage allocatons. Returning
+to the examples provided for execution spaces, the multi-core
+processor may have multiple memory spaces available including
+on-package memory, slower DRAM and additional a set of non-volatile
+memories. The GPU may also provide an additional memory space
+through its local on-package memory. The programmer is free
+to decide where each data structure may be allocated by requesting
+these from the specific instance associated with that memory space.
+Kokkos provides the appropriate abstraction of the allocation
+routines and any associated data management operations including
+releasing the memory and returning it for future use, as well as
+copy operations.
 

### 2015-03-28 Kokkos_PG_MachineModel.tex
Adds a bibliography to the reference guide and a citation to the CAL-AMM document from Sandia National Laboratories.

@@
 \begin{figure}
-\includegraphics[width=2in]{figures/kokkos-node.pdf}
+\begin{center}
+\includegraphics[width=4in]{figures/kokkos-node.pdf}
 \caption{Conceptual Model of a Future High Performance Computing Node}
 \label{fig:kokkosnode}
+\end{center}
 \end{figure}
@@
 Figure~\ref{fig:kokkosnode}) assumes that there may be multiple
-execution units in a compute node. In the Figure we have elected to
+execution units in a compute node. For a more general discussion
+of abstract machine models for Exascale computing the reader should
+consult reference~\cite{cal_amm}. In the figure shown here,
+we have elected to
 show two different types of compute units -- one which represents

### 2015-03-28 Kokkos_PG_MachineModel.tex
Adds Hennessy and Paterson reference and a section for atomic memory references.

@@
 copy operations.
+
+\subsubsection{Memory Consistency in Kokkos}
+
+Memory consistency models are a complex topic in and of themselves
+and usually rely on complex operations associated with hardware
+caches (for more information see reference~\cite{handp_hardware}. 
+Kokkos itself does not {\em require} caches to be present
+in hardware and so assumes an extremely weak memory consistency
+model. In this model, the programmer should not assume any
+specific ordering of memory operations being isused by a
+kernel. This has the potential to create race conditions between
+memory operations if these are not appropriately protected. In
+order to provide a guaranteed that memory are completed, Kokkos
+provides a {\em fence} operation which forces the compute 
+engine to complete all outstanding memory operations before
+any new ones can be issued. With appropriate use of fences,
+programmers are thereby able to ensure that guarantees can
+be made as to when data will {\em definitely} have been 
+written to memory.
+
+\subsubsection{Atomic accesses to Memory in Kokkos}
 

### 2015-03-28 Kokkos_PG_MachineModel.tex
Adds figures for execution spaces and memory spaces in Kokkos.

@@
 
+\begin{figure}
+\begin{center}
+\includegraphics[width=4in]{figures/kokkos-execution-space.pdf}
+\caption{Example Execution Spaces in a Future Computing Node}
+\label{fig:kokkos_execution_spaces}
+\end{center}
+\end{figure}
+
 Kokkos uses the term {\em execution spaces} to describe a logical
@@
 
+\begin{figure}
+\begin{center}
+\includegraphics[width=4in]{figures/kokkos-memory-space.pdf}
+\caption{Example Memory Spaces in a Future Computing Node}
+\label{fig:kokkos_memory_spaces}
+\end{center}
+\end{figure}
+
 The multiple types of memory which will become available in future
@@
 
+\subsubsection{Atomic accesses to Memory in Kokkos}
+
+In cases where multiple executing threads attempt to read
+a memory address, complete a computation on the item and
+write it back to same address in memory, an ordering
+collision may occur. These situations, known as
+{\em race conditions} (because the data value
+stored in memory are the threads complete is dependent
+on which thread completes its memory operation last),
+are often the cause of non-determinism in parallel
+programs. A number of methods can be employed to ensure
+that race conditions do not occur in parallel programs
+including the use of locks (which allow only a single
+thread to gain access to data structure at a time),
+critical regions (which allow only one thread to execute
+a code sequence at any point in time) and {\em atomic}
+operations. Memory operations which are atomic guarantee
+that a read, simple computation, and write to memory
+are completed a single unit. This might allow application
+programmers to safely increment a memory value for instance,
+or, more commonly, to safely accumulate values from multiple
+threads into a single memory locations.
+
+Although atomic operations are logically an operation
+on memory, they are performed by compute units and so
+are a property of execution spaces in the Kokkos framework.
+Execution spaces are not guaranteed to provide atomic
+operations support but attempts to use them on spaces
+which do not support atomic access will result in 
+a compile time error.
+
 \subsubsection{Memory Consistency in Kokkos}
@@
 and usually rely on complex operations associated with hardware
-caches (for more information see reference~\cite{handp_hardware}. 
-Kokkos itself does not {\em require} caches to be present
+caches or memory access coherency (for more information 
+see reference~\cite{handp_hardware}. 
+Kokkos, does not {\em require} caches to be present
 in hardware and so assumes an extremely weak memory consistency
-model. In this model, the programmer should not assume any
+model. In the Kokkos model, the programmer should not assume any
 specific ordering of memory operations being isused by a
@@
 written to memory.
-
-\subsubsection{Atomic accesses to Memory in Kokkos}
 
@@
 Kokkos provides some synchronization constructs as well; see the following section.
-
-\section{Memory consistency}\label{S:Model:Consistency}
-
-A \emph{memory consistency model} describes the behavior of a memory space $M$,
-with respect to multiple threads running in an execution space $E$
-that access the same data in $M$ concurrently.
-One may also encounter this idea as the phrase ``cache coherency.''
-We prefer ``memory consistency,'' because Kokkos does not require caches.
-Consistency models are a complicated subject which we cannot cover in sufficient detail here.
-We encourage readers to refer to standard computer hardware textbooks,
-like ``Computer Architecture: A Quantitative Approach'' (Hennessy and Patterson, 2011).
-
-Kokkos imposes a very weak consistency model across an entire execution space.
-For example, it does \emph{not} assume sequential consistency.
-Much like MPI 3 one-sided communication
-or other partitioned global address space (PGAS) programming models,
-Kokkos requires an execution space to issue a \emph{memory fence}
-to a given memory space, 
-in order to bring that memory space into a consistent state.
-The fence forces completion of all outstanding memory writes,
-and reorders all loads to occur after all those writes complete.
-
-\subsection{Atomic updates}\label{SS:Model:Consistency:Atomic}
-
-Some execution spaces may implement \emph{atomic updates} with respect to some memory spaces.
-An atomic update runs on an execution space, and changes a word of data in a memory space.
-The atomic update behaves atomically with respect to other atomic updates happening concurrently by that execution space on that memory space.
-``Behaves atomically'' means that all of the atomic updates will complete,
-and the word of data will be changed in a way that respects some order of those update operations.
-Kokkos does not promise any particular order.
-Furthermore, for a nonatomic memory read or write to ``see'' the results of atomic updates,
-the execution space must first execute a \lstinline!fence()! operation to that memory space.

### 2015-03-28 Kokkos_PG_LegacyCodeIntegration.tex
Added stuff to legacy code section.

@@
 \chapter{Interoperability and Legacy Codes}
+
+One goal of Kokkos is to support incremental adoption in legacy 
+applications. This facilitates a step by step conversion allowing
+for continuous testing of functionality (and in certain bounds) of 
+performance. One feature of this is full interoperability with the 
+underlying backend programming models. This also allows for 
+target specific optimizations written directly in the backend model
+in order to achieve maximal performance.
+
+After reading this chapter, you should understand the following:
+\begin{itemize}
+\item Restriction on interoperability with raw OpenMP and Cuda.
+\item How to handle external data structures.
+\item How to incrementally convert legacy data structures.
+\item How to call non-Kokkos third party libraries safely.
+\end{itemize}
+In all code examples in this chapter, we assume that all classes in
+the \lstinline!Kokkos! namespace have been imported into the working
+namespace.
+
+\section{OpenMP, Pthreads and CUDA interoperability}
+
+Since the implementation of Kokkos is achieved with a C++ library
+it provides full interoperability with the underlying backend programming
+models. In particular it allows for mixing of OpenMP, CUDA and Kokkos 
+code in the same compilation unit. This is true for both the parallel 
+execution layers of Kokkos as for the data layer.
+
+It is important to recognize that this does not lift certain restrictions. 
+For example it is not valid to allocate a view inside of an OpenMP 
+parallel region, the same way as it is not valid to allocate a View 
+inside a \lstinline|parallel_for| kernel. Indeed there are things which 
+are slightly more cumbersome when mixing the models. Assigning 
+one view to another inside a Cuda kernel or an OpenMP parallel 
+region is only possible if the destination view is unmanaged. 
+During dispatch of kernels with parallel for  all Views referenced 
+in the functor or lambda are automatically switched into unmanaged 
+mode. This would not happen when simply entering an OpenMP 
+parallel region.
+
+\subsection{Cuda interoperabiltiy}
+
+The most important thing to know for Cuda interoperability is that the
+\lstinline|KOKKOS_INLINE_FUNCTION| macro evaluates to 
+\lstinline|__host__ __device__ inline|. This means that calling a pure
+\lstinline|__device__| function (for example Cuda intrinsics or device 
+functions of libraries) must be protected with the \lstinline|__CUDA_ARCH__|
+pragma.
+
+\begin{lstlisting}
+__device__ SomeFunction(double* x) {
+  ...
+}
+
+struct Functor {
+  typedef Cuda execution_space;
+  View<double*,Cuda> a;
+  KOKKOS_INLINE_FUNCTION
+  void operator(const int& i) {
+    #ifdef __CUDA_ARCH__
+    int block = blockIdx.x;
+    int thread = threadIdx.x;
+    if (thread == 0)
+      SomeFunction(&a(block));
+    #endif
+  }
+}
+\end{lstlisting}
+
+The \lstinline|RangePolicy| start a 1D grid of 1D thread blocks so that the index 
+\lstinline|i| is calculated as \lstinline|blockIdx.x * blockDim.x + threadIdx.x|. 
+For the \lstinline|TeamPolicy| the number of teams is the grid dimension, while
+the number of threads per team is mapped to the Y-dimension of the Cuda 
+thread-block. The optional vector length is mapped to the X-dimension. 
+For example \lstinline|TeamPolicy<Cuda>(100,12,16)| would start a 1D grid of size 
+100 with block-dimensions (16,12,1) while \lstinline|TeamPolicy<Cuda>(100,96)| would 
+result in a grid size of 100 with block-dimensions of (1,96,1). The restrictions on the 
+vector length (power of two and smaller than 32 for the Cuda execution space) 
+guarantee that vector loops are performed by threads which are part of a single 
+warp.
+
+\subsection{OpenMP}
+
+One restriction on OpenMP interoperability is that it is not valid to increase
+the number of threads via \lstinline|omp_set_num_threads()| after initializing 
+Kokkos. This restriction is caused due to Kokkos allocating internal per thread
+book keeping data structures. It is however valid to ask 
+for the thread ID inside a Kokkos parallel kernel compiled for the OpenMP 
+execution space. It is also valid to use OpenMP constructs such as OpenMP
+atomics inside a parallel kernel or functions called by it. However it is undefined
+what happens when mixing OpenMP and Kokkos atomics, since those will not 
+necessarily map to the the same underlying mechanism.
+
 \section{Legacy data structures}
@@
 allocated data. In both cases it is mandatory to fix the Layout of the Kokkos view
-to the actual layout used in the legacy data structure.
+to the actual layout used in the legacy data structure. Note that the user is responsible
+for insuring proper access capabilities. For example a pointer obtained from a view
+in the \lstinline|CudaSpace| may only be accessed from Cuda kernels, and a View
+constructed from memory acquired through a call to \lstinline|new| will typically only 
+be accessible from Execution spaces which can access the \lstinline|HostSpace|.
 
@@
 \subsubsection{Raw allocations}
-\subsubsection{Views as the fundamental data owning structure}
+
+A simple way to add support for multiple memory spaces to a legacy app is to 
+use \lstinline|kokkos_malloc|, \lstinline|kokkos_free| and \lstinline|kokkos_realloc|.
+The functions are templated on the memory space and thus allow targeted 
+placement of data structures:
+
+\begin{lstlisting}
+// Allocate an array of 100 doubles in the default memory space
+double* a = (double*) kokkos_malloc<>(100*sizeof(double));
+
+// Allocate an array of 150 int* in the Cuda UVM memory space
+// This allocation is accessible from the host
+int** 2d_array = (int**) kokkos_malloc<CudaUVMSpace>(150*sizeof(int*));
+
+// Fill the pointer array with pointers to data in the Cuda Space
+// Since it is not the UVM space you can access 2d_array[i][j] only inside a Cuda Kernel
+for(int i=0;i<150;i++)
+  2d_array[i] = (int*) kokkos_malloc<CudaSpace>(200*sizeof(int));
+\end{lstlisting}
+
+A common usage scenario of this capability is to allocate all memory in the CudaUVMSpace 
+when compiling for GPUs. This allows all allocations to be accessible from legacy code sections
+as well as from parallel kernels written with Kokkos.
 
@@
 
-\subsection{\lstinline|std::vector|}
+When memory is managed externally, for example because Kokkos is used
+in a library which is given pointers to data allocations as input, it can be necessary 
+or convenient to wrap the data into Kokkos views. If the library anyway receives the 
+data to create a copy it is straight forward to allocate the internal data structure as a
+view and copy the data in a parallel kernel element by element. Note that you might
+need to first copy into a host view before copying to the actual destination memory 
+space:
+
+\begin{lstlisting}
+template<class ExecutionSpace>
+void MyKokkosFunction(double* a, const double** b, int n, int m) {
+  // Define the host execution space and the view types
+  typedef HostSpace::execution_space host_space;
+  typedef View<double*,ExecutionSpace> t_1d_device_view;
+  typedef View<double**,ExecutionSpace> t_2d_device_view;
+
+  // Allocate the view in the memory space of ExecutionSpace
+  t_1d_device_view d_a("a",n);
+  // Create a host copy of that view
+  typename t_1d_device_view::HostMirror h_a = create_mirror_view(a);
+  // Copy the data from the external allocation into the host view
+  parallel_for(RangePolicy<host_space>(0,n),
+    KOKKOS_LAMBDA (const int& i) {
+    h_a(i) = a[i];
+  });
+  // Copy the data from the host view to the device view
+  deep_copy(d_a,h_a);
+
+  // Allocate a 2D view in the memory space of ExecutionSpace
+  t_2d_device_view d_b("b",n,m);
+  // Create a host copy of that view
+  typename t_2d_device_view::HostMirror h_b = create_mirror_view(b);
+
+  // Get the member_type of the team policy
+  typedef TeamPolicy<host_space>::member_type t_team;
+  // Run a 2D copy kernel using a TeamPolicy
+  parallel_for(TeamPolicy<host_space>(n,m),KOKKOS_LAMBDA (const t_team& t) {
+    const int i = t.team_rank();
+    parallel_for(TeamThreadRange(t,m), [&] (const int& j) {
+      h_b(i,j) = b[i][j];
+    });
+  });
+  // Copy the data from the host to the device
+  deep_copy(d_b,h_b);
+}
+\end{lstlisting}
+
+Alternatively one can create a view which directly references the external allocation.
+If that data is a multi dimensional view it is important to specify the Layout explicitly.
+Furthermore all data must be part of the same allocation.
+
+ \begin{lstlisting}
+void MyKokkosFunction(double* a, const double* b, int n, int m) {
+  // Define the host execution space and the view types
+  typedef View<double*,DefaultHostExecutionSpace, Unmanaged> t_1d_view;
+  typedef View<double**[3],LayoutRight, DefaultHostExecutionSpace, Unmanaged> 
+     t_3d_view;
+
+  // Create a 1D view of the external allocation
+  t_1d_view d_a(a,n);
+
+  // Create a 3D view of the second external allocation
+  // This assumes that the data had a row major layout (i.e. the third index is stride 1)
+  t_3d_view d_b(b,n,m);
+}
+\end{lstlisting}
+
+\subsection{Views as the fundamental data owning structure}
+
+An other option is to let Kokkos handle the basic allocations using Views, 
+and then constructing the legacy data structures around them. Again it is
+important to fix the Layout of the Views to whatever the layout of the legacy 
+data was.
+
+\begin{lstlisting}
+// Allocate a 2D view with row major layout
+View<double**,LayoutRight,HostSpace> a("A",n,m);
+
+// Allocate an array of pioneers
+double** a_old = new double*[n];
+
+// Fill the array with pointers to the rows of a
+for(int i=0; i<n; i++)
+  a_old[i] = &a(i,0);
+\end{lstlisting}
+
+\subsection{std::vector}
+
+One of the most common data objects in C++ codes is \lstinline|std::vector|. 
+Its semantics are unfortunately not compatible with Kokkos requirements 
+and it is thus not well supported. A major problem is that functors and lambdas
+are passed as const objects to the parallel execution. This design choice was made
+to (i) prevent a common cause of race conditions and (ii) allow the underlying 
+implementation more flexibility in how to share the functor and where to put it.
+In particular this leaves the choice open for the implementation to give each thread 
+an individual copy of the functor or place it in read only cache structures.
+
+The semantics of \lstinline|std::vector| would in this case prevent a kernel from modifying its 
+entries since a const \lstinline|std::vector| is read only. Furthermore creating multiple copies
+of the functor would in deed replicate the vector data, since it has copy semantics.
+
+Other issues with \lstinline|std::vector| are its unrestricted support for resize as well as push 
+functionality. In a threaded environment support for those capabilities would bring
+massive performance penalties. In particular access to the \lstinline|std::vector| would require
+locks in order to prevent one thread to deallocate the view while another accesses its content.
+And last but not least \lstinline|std::vector| is not supported on GPUs and thus would prevent
+portability.
+
+Kokkos provides a drop in replacement for |std::vector| in the form of \lstinline|Kokkos::vector|.
+Outside of parallel kernels its semantics are mostly the same as that of \lstinline|std::vector|, for
+example assignments performs deep copies and resize and push functionality are provided.
+One important difference is that it is valid to assign values to the elements of a const vector.
 

### 2015-03-29 Kokkos_PG_MachineModel.tex
Rewrite chapter heading text to clean up.

@@
-\chapter{Machine model}
+\chapter{Machine Model}
 \label{C:Model}
 
-After reading this chapter, you will understand the abstract model of
-a shared-memory parallel computer which underlies Kokkos.  
-Respecting the abstract machine model ensures performance portability.
-The model has two components:
+After reading this chapter you will understand the abstract model
+of a parallel computing node which underlies the design choices
+and structure of the Kokkos framework. The machine model ensures
+the applications written using Kokkos will have portability across
+architectures while being performant on a range of hardware.\\
+\\
+\noindent
+The machine model has two important components:
 \begin{itemize}
-\item \emph{Memory spaces}, in which data live
+\item \emph{Memory spaces}, in which data structures can be allocated
 \item \emph{Execution spaces}, which execute parallel operations using
-  data in one or more memory spaces
+  data from one or more memory spaces
 \end{itemize}
-All code that uses Kokkos involves relations between memory and execution spaces.
-% A given execution space may or may not be able to access data in a
-% given memory space.  Users can ask Kokkos if it can.  Users may also
-% copy data between any two different memory spaces by using
-% \lstinline!deep_copy!.
 
-\section{Motivation}\label{S:Model:Motivations}
+\section{Motivations}\label{S:Model:Motivations}
 

### 2015-03-30 Kokkos_PG_LegacyCodeIntegration.tex
More updates in legacy section.

@@
-\chapter{Interoperability and Legacy Codes}
+\section{Interoperability and Legacy Codes}
 
@@
 
-\section{OpenMP, Pthreads and CUDA interoperability}
+\subsection{OpenMP, Pthreads and CUDA interoperability}
 
@@
 region is only possible if the destination view is unmanaged. 
-During dispatch of kernels with parallel for  all Views referenced 
+During dispatch of kernels with \lstinline|parallel_for|  all Views referenced 
 in the functor or lambda are automatically switched into unmanaged 
@@
 
-\subsection{Cuda interoperabiltiy}
+\subsubsection{Cuda interoperabiltiy}
 
@@
 
-\subsection{OpenMP}
+\subsubsection{OpenMP}
 
@@
 
-\section{Legacy data structures}
+\subsection{Legacy data structures}
 
@@
 
-\subsection{Kokkos manages memory}
-
-\subsubsection{Raw allocations}
+\subsection{Raw allocations through Kokkos}
 
@@
 
-\subsection{External memory management}
+\subsubsection{External memory management}
 
@@
 
-\subsection{Views as the fundamental data owning structure}
+\subsubsection{Views as the fundamental data owning structure}
 
@@
 
-\subsection{std::vector}
+\subsubsection{std::vector}
 
@@
 
+Inside of parallel sections \lstinline|Kokkos::vector| switches to view semantics. That means
+in particular that assignments are shallow copies. Certain functions will also throw runtime 
+errors when called inside a parallel kernel. This includes resize and push.
+
 \section{Calling non-Kokkos libraries}
 
-\section{Keeping track of data transfers}
+\subsection{Keeping track of data transfers}

### 2015-03-31 Kokkos_PG_Compilation.tex
Convert to SAND format.

@@
-\chapter{Compiling}\label{C:build}
+\section{Compiling}\label{C:build}
 
@@
 
-\section{General Information}\label{S:build:gen}
+\subsection{General Information}\label{S:build:gen}
 
@@
 
-\section{Using Kokkos' Makefile system}\label{S:build:make}
+\subsection{Using Kokkos' Makefile system}\label{S:build:make}
 
@@
 
-\section{Using Trilinos' CMake build system}\label{S:build:Trilinos}
+\subsection{Using Trilinos' CMake build system}\label{S:build:Trilinos}
 
@@
 
-\section{Building for CUDA}\label{S:build:CUDA}
+\subsection{Building for CUDA}\label{S:build:CUDA}
 
@@
 
-\section{Building for Intel Xeon Phi}\label{S:build:Phi}
+\subsection{Building for Intel Xeon Phi}\label{S:build:Phi}
 

### 2015-03-31 Kokkos_PG_HierarchicalParallelism.tex
Convert to SAND format.

@@
-\chapter{Hierarchical Parallelism}\label{C:Hierarchical}
+\section{Hierarchical Parallelism}\label{C:Hierarchical}
 
@@
 
-\section{Motivation}\label{S:Hierarchical:Motivation}
+\subsection{Motivation}\label{S:Hierarchical:Motivation}
 
@@
 
-\section{Thread teams}\label{S:Hierarchical:Teams}
+\subsection{Thread teams}\label{S:Hierarchical:Teams}
 
@@
 
-\subsection{Creating a Policy instance}\label{SS:Hierarchical:Teams:Policy}
+\subsubsection{Creating a Policy instance}\label{SS:Hierarchical:Teams:Policy}
 
@@
 
-\subsection{Basic kernels}\label{SS:Hierarchical:Teams:Kernels}
+\subsubsection{Basic kernels}\label{SS:Hierarchical:Teams:Kernels}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Team scratch pad memory}\label{S:Hierarchical:Scratch}
+\subsection{Team scratch pad memory}\label{S:Hierarchical:Scratch}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Nested parallelism}\label{S:Hierarchical:Nested}
+\subsection{Nested parallelism}\label{S:Hierarchical:Nested}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\subsection{Team and vector loops}\label{SS:Hierarchical:Nested:Loops}
+\subsubsection{Team and vector loops}\label{SS:Hierarchical:Nested:Loops}
 
@@
 
-\subsection{Restrict execution with \lstinline|single|}
+\subsubsection{Restrict execution with \lstinline|single|}
 

### 2015-03-31 Kokkos_PG_Initialization.tex
Convert to SAND format.

@@
-\chapter{Initialization}\label{C:init}
+\section{Initialization}\label{C:init}
 
@@
 
-\section{Initialization by command-line arguments}\label{S:init:cmdline}
+\subsection{Initialization by command-line arguments}\label{S:init:cmdline}
 
@@
 
-\section{Initialization by struct}\label{S:init:struct}
+\subsection{Initialization by struct}\label{S:init:struct}
 
@@
 If you set \lstinline|num_threads| or \lstinline|num_numa| to zero or less, 
-Kokkos will \emph{not} initialize the default host execution space.
+Kokkos will try to determine default values if possible or otherwise set them to 1.
+In particular Kokkos can use the hwloc library to determine default settings, using the assumption that the process binding mask is unique, i.e.
+that this process does not share any cores with another process.
 Note that the default value of each parameter is -1.
-Thus, you must set both \lstinline|num_threads| and \lstinline|num_numa|
-in order for Kokkos to initialize the default host execution space.
 
@@
 
-\section{Initializing non-default execution spaces}\label{S:init:specific}
+\subsection{Initializing non-default execution spaces}\label{S:init:specific}
 
@@
 
-\section{Finalization}\label{S:init:finalize}
+\subsection{Finalization}\label{S:init:finalize}
 

### 2015-03-31 Kokkos_PG_Introduction.tex
Convert to SAND format.

@@
-\chapter{Introduction}\label{C:intro}
+\section{Introduction}
+\label{C:intro}
 
@@
 pure computation (often called FLOP/s) are a poor single optimization goal for scientific
-workloads as well as practical challenges in the form of producing energy efficient and cost
+workloads, as well as practical challenges in the form of producing energy efficient and cost
 efficient processors. Since the convergence on the Message-Passing Interface (MPI) standard
-in the mid 1990s, application developers have enjoyed a fairly static view of the machine --
+in the mid 1990s, application developers have enjoyed a seemingly static view of the 
+underlying machine --
 that of a distributed collection of homogeneous nodes executing in collaboration. However, 
@@
 a potential limiter to greater future performance. While we expect MPI to continue to function
-as the basic mechanism for communication between compute nodes, additional parallelism is
-likely to be required on the computing node itself if high performance and efficiency are to be
-realized.
+as the basic mechanism for communication between compute nodes for the immediate future, 
+additional parallelism is
+likely to be required on the computing node itself if high performance and efficiency goal 
+are to be realized.
 
@@
 upon three broad categories of computing device: (1) the multi-core processor with powerful
-serial performance, optimized to reduce operation latency; (2) many-cores processors with
-low to medium powered compute cores that are designed to extent the multi-core concept toward
+serial performance, optimized to reduce operation latency; (2) many-core processors with
+low to medium powered compute cores that are designed to extend the multi-core concept toward
 throughput based computation and, finally, (3), the general-purpose graphics processor unit
@@
 to tolerate long latencies but provide performance through a much higher degree of 
-parallelism. Any combination of these options might also be combined in the future.
+parallelism and computational throughput. Any combination of these options might also be 
+combined in the future.
 
 The diversity of the options available for processor selection raises as interested question
-as to how these should be programmed. Partly due to their heritage but also due to their
+as to how these should be programmed. In part due to their heritage, but also due to their
 optimized designs, each of these hardware types offers a different programming solution and
@@
 (TBB) as well as Linux p-threads. To target both contemporary multi/many-core processors and
-GPUs technologies such as OpenMP, OpenACC and OpenCL might be used. Finally, for GPUs a 
-programming model such as CUDA may be selected. This poses a problem to the application
+GPUs technologies such as OpenMP, OpenACC and OpenCL might be used. Finally, for highest
+performance on GPUs a 
+programming model such as CUDA may be selected. Such a variety of options poses a problem 
+to the application
 developer of today which is reminiscent of the challenges before MPI became the default
 communication library -- which model should be selected to provide portability across hardware
-solutions as well as provide reasonable to high performance across each class of processor and
-protect algorithm investment into the future. So far none of these models have been able to 
+solutions, as well as provide high performance across each class of processor and
+protect algorithm investment into the future. None of the models listed above have been able to 
 provide practical solutions to these questions.
@@
 concerns by providing an abstraction of both computation and application data allocation and
-layout. These abstraction layers are designed specifically to isolate software developers for
-the fluctuation in hardware details yet provide portability and high levels of performance
-across architectures.
+layout. These abstraction layers are designed specifically to isolate software developers from
+fluctuation and diversity in hardware details yet provide portability and high levels of performance
+across many architectures.
 

### 2015-03-31 Kokkos_PG_LegacyCodeIntegration.tex
Convert to SAND format.

@@
 region is only possible if the destination view is unmanaged. 
-During dispatch of kernels with \lstinline|parallel_for|  all Views referenced 
+During dispatch of kernels with parallel for  all Views referenced 
 in the functor or lambda are automatically switched into unmanaged 
@@
 
-\subsection{Raw allocations through Kokkos}
+\subsubsection{Kokkos manages memory}
+
+\paragraph{Raw allocations}
 
@@
 
-Inside of parallel sections \lstinline|Kokkos::vector| switches to view semantics. That means
-in particular that assignments are shallow copies. Certain functions will also throw runtime 
-errors when called inside a parallel kernel. This includes resize and push.
-
-\section{Calling non-Kokkos libraries}
+\subsection{Calling non-Kokkos libraries}
 

### 2015-03-31 Kokkos_PG_MachineModel.tex
Convert to SAND format.

@@
-\chapter{Machine Model}
+\section{Machine Model}
 \label{C:Model}
@@
 
-\section{Motivations}\label{S:Model:Motivations}
+\subsection{Motivations}\label{S:Model:Motivations}
 
@@
 
-\subsection{Kokkos Abstract Machine Model}
+\subsubsection{Kokkos Abstract Machine Model}
 
@@
 
-\subsection{Kokkos Memory Spaces}
+\subsubsection{Kokkos Memory Spaces}
 
@@
 
-\subsubsection{Atomic accesses to Memory in Kokkos}
+\paragraph{Atomic accesses to Memory in Kokkos}
 
@@
 
-\subsubsection{Memory Consistency in Kokkos}
+\paragraph{Memory Consistency in Kokkos}
 
@@
 
-\section{Thread teams}\label{S:Model:Teams}
+\subsection{Thread teams}\label{S:Model:Teams}
 
@@
 
-\section{Program execution}\label{S:Model:Exec}
+\subsection{Program execution}\label{S:Model:Exec}
 
@@
 
-\subsection{Reproducible reductions and scans}\label{SS:Model:Exec:Repro}
+\paragraph{Reproducible reductions and scans}\label{SS:Model:Exec:Repro}
 
@@
 
-\subsection{Asynchronous parallel dispatch}\label{SS:Model:Exec:Async}
+\paragraph{Asynchronous parallel dispatch}\label{SS:Model:Exec:Async}
 
@@
 
-\subsection{Thread safety?}\label{SS:Model:Exec:ThreadSafety}
+\subsubsection{Thread safety?}\label{SS:Model:Exec:ThreadSafety}
 

### 2015-03-31 Kokkos_PG_Subviews.tex
Convert to SAND format.

@@
-\chapter{Subviews}\label{C:Subviews}
+\section{Subviews}\label{C:Subviews}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{A subview is a slice of a View}\label{S:Subviews:Slice}
+\subsection{A subview is a slice of a View}\label{S:Subviews:Slice}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{How to take a subview}\label{S:Subviews:How}
+\subsection{How to take a subview}\label{S:Subviews:How}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{C++11 type deduction}\label{SS:Subviews:How:auto}
+\subsubsection{C++11 type deduction}\label{SS:Subviews:How:auto}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Dimension of a subview}\label{SS:Subviews:How:dim}
+\subsubsection{Dimension of a subview}\label{SS:Subviews:How:dim}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Degenerate Views}\label{SS:Subviews:How:degen}
+\subsubsection{Degenerate Views}\label{SS:Subviews:How:degen}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2015-03-31 Kokkos_PG_Views.tex
Convert to SAND format.

@@
-\chapter{View: Multidimensional array}\label{C:View}
+\section{View: Multidimensional array}\label{C:View}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Why Kokkos needs multidimensional arrays}\label{S:View:Why}
+\subsection{Why Kokkos needs multidimensional arrays}\label{S:View:Why}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Creating and using a View}\label{S:View:CreateUse}
+\subsection{Creating and using a View}\label{S:View:CreateUse}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\subsection{Constructing a View}
+\subsubsection{Constructing a View}
 
@@
 
-\subsection{What types of data may a View contain?}
+\subsubsection{What types of data may a View contain?}
 
@@
 
-\subsection{Const Views}
+\subsubsection{Const Views}
 
@@
 
-\subsection{Accessing entries (indexing)}
+\subsubsection{Accessing entries (indexing)}
 
@@
 
-\subsection{Reference counting}
+\subsubsection{Reference counting}
 
@@
 
-\subsection{Resizing}
+\subsubsection{Resizing}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Layout}\label{S:View:Layout}
+\subsection{Layout}\label{S:View:Layout}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\subsection{Strides and dimensions}\label{SS:View:Layout:Strides}
+\subsubsection{Strides and dimensions}\label{SS:View:Layout:Strides}
 
@@
 
-\subsection{Other layouts}\label{SS:View::Layout::Other}
+\subsubsection{Other layouts}\label{SS:View::Layout::Other}
 
@@
 
-\subsection{Default layout depends on execution space}\label{SS:View:Layout:Default}
+\subsubsection{Default layout depends on execution space}\label{SS:View:Layout:Default}
 
@@
 
-\subsection{Explicitly specifying layout}\label{SS:View:Layout:Explicit}
+\subsubsection{Explicitly specifying layout}\label{SS:View:Layout:Explicit}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Managing Data Placement}\label{S:View:Placement}
+\subsection{Managing Data Placement}\label{S:View:Placement}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\subsection{Memory spaces}
+\subsubsection{Memory spaces}
 
@@
 
-\subsection{Initialization}
+\subsubsection{Initialization}
 
@@
 
-\subsection{Deep copy and HostMirror}
+\subsubsection{Deep copy and HostMirror}
 
@@
 
-\subsection{How do I get the raw pointer?}
+\subsubsection{How do I get the raw pointer?}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Access traits}\label{S:View:AccessTraits}
+\subsection{Access traits}\label{S:View:AccessTraits}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\subsection{Atomic access}
+\subsubsection{Atomic access}
 
@@
 
-\subsection{Random Access}
+\subsubsection{Random Access}
 
@@
 
-\subsection{Standard idiom for specifying access traits}
+\subsubsection{Standard idiom for specifying access traits}
 
@@
 
-\subsection{Unmanaged Views}
+\subsubsection{Unmanaged Views}
 

### 2015-04-07 Kokkos_PG_LegacyCodeIntegration.tex
More updates to LegacyCodeIntegration

--- 2015-03-30 Kokkos_PG_LegacyCodeIntegration.tex
@@
 
+\begin{lstlisting}
+// Create a vector of 1000 double elements
+Kokkos::vector<double> v(1000);
+// Create another vector as a copy of v; This allocates another 1000 doubles
+Kokkos::vector<double> x = v;
+
+parallel_for(1000, KOKKOS_LAMBDA (const int& i) {
+   // Create a view of x; m and x will reference the same data.
+   Kokkos::vector<double> m = x; 
+   x[i] = 2*i+1;
+   v[i] = m[i] - 1;
+});
+
+// Now x contains the first 1000 uneven numbers
+// v contains the first 1000 even numbers
+\end{lstlisting}
+
 \section{Calling non-Kokkos libraries}
 
-\subsection{Keeping track of data transfers}
+There are no restrictions on calling non-Kokkos libraries outside of parallel kernels. 
+However due to the polymorphic layouts of Kokkos views it is often required to test
+layouts for compatibility with third party libraries. The usual Blas interface for example,
+expects matrixes to be laid out in column major format (i.e. LayoutLeft in Kokkos). 
+Furthermore it is necessary to test that the library can access the memory space 
+of the view.
+
+\begin{lstlisting}
+template<class Scalar, class DeviceType>
+Scalar dot(View<const Scalar* , DeviceType> a, View<const Scalar*, DeviceType> b) {
+// Check for Cuda memory and call cublas if true
+#ifdef KOKKOS_HAVE_CUDA
+  if(Impl::is_same<typename DeviceType::memory_space,CudaSpace> ::value ||
+     Impl::is_same<typename DeviceType::memory_space,CudaUVMSpace> ::value) {
+    return call_cublas_dot(a.ptr_on_device(),b.ptr_on_device(),a.dimension_0() );
+  }
+#endif
+
+// Call CBlas on the host otherwise
+  if(Impl::is_same<typename DeviceType::memory_space,HostSpace> ::value) {
+    return call_cblas_dot(a.ptr_on_device(),b.ptr_on_device(),a.dimension_0() );
+  }
+}
+\end{lstlisting}
+
+Note the \lstinline|Impl::is_same| construct can be replaced

### 2015-04-15 Kokkos_PG_Compilation.tex
Updated style of the Guide . This comments out the SAND Report style
The stuff is still in if we really need to its not that hard to go back.

--- 2015-03-25 Kokkos_PG_Compilation.tex
@@
 \begin{small}
-\begin{tabular}[t]{p{0.25\textwidth}p{0.3\textwidth}p{0.42\textwidth}}
+\begin{tabular}[t]{p{0.18\textwidth}p{0.3\textwidth}p{0.42\textwidth}}
 \hline\hline
@@
 \hline
-   HWLOC\_PATH (IN): & Path to the hardware locality library. \\
-
-\hline
    KOKKOS\_DEVICES (IN): & What Execution and Memory Spaces should be enabled. \\
@@
                are set. \\
+
+\hline
+   KOKKOS\_USE\_TPLS (IN): & Enable optional third party libraries. \\
+      \hspace{0.5cm}Options: & hwloc,librt \\
+      \hspace{0.5cm}Default: & \\
+
+\hline
+   HWLOC\_PATH (IN): & Path to the hardware locality library if enabled. \\
 
@@
 
-Any Kokkos application compiled for CUDA embeds CUDA code via template metaprogramming.
+Any Kokkos application compiled for CUDA embeds CUDA code via template meta-programming.
 Thus, the whole application must be built with a CUDA-capable compiler.

### 2015-04-15 Kokkos_PG_LegacyCodeIntegration.tex
Updated style of the Guide . This comments out the SAND Report style
The stuff is still in if we really need to its not that hard to go back.

--- 2015-03-30 Kokkos_PG_LegacyCodeIntegration.tex
@@
-\section{Interoperability and Legacy Codes}
+\chapter{Interoperability and Legacy Codes}
 
@@
 
-\subsection{OpenMP, Pthreads and CUDA interoperability}
+\section{OpenMP, Pthreads and CUDA interoperability}
 
@@
 
-\subsubsection{Cuda interoperabiltiy}
+\subsection{Cuda interoperabiltiy}
 
-The most important thing to know for Cuda interoperability is that the
-\lstinline|KOKKOS_INLINE_FUNCTION| macro evaluates to 
+The most important thing to know for Cuda interoperability is that the provided macro
+\lstinline|KOKKOS_INLINE_FUNCTION| evaluates to 
 \lstinline|__host__ __device__ inline|. This means that calling a pure
@@
 
-\subsubsection{OpenMP}
+\subsection{OpenMP}
 
@@
 
-\subsection{Legacy data structures}
+\section{Legacy data structures}
 
@@
 
-\subsection{Raw allocations through Kokkos}
+\section{Raw allocations through Kokkos}
 
@@
 // This allocation is accessible from the host
-int** 2d_array = (int**) kokkos_malloc<CudaUVMSpace>(150*sizeof(int*));
+int** 2d_array = (int**) kokkos_malloc<CudaUVMSpace>
+                         (150*sizeof(int*));
 
 // Fill the pointer array with pointers to data in the Cuda Space
-// Since it is not the UVM space you can access 2d_array[i][j] only inside a Cuda Kernel
+// Since it is not the UVM space you can access 2d_array[i][j] 
+// only inside a Cuda Kernel
 for(int i=0;i<150;i++)
@@
 
-\subsubsection{External memory management}
+\subsection{External memory management}
 
@@
   // Run a 2D copy kernel using a TeamPolicy
-  parallel_for(TeamPolicy<host_space>(n,m),KOKKOS_LAMBDA (const t_team& t) {
+  parallel_for(TeamPolicy<host_space>(n,m),
+    KOKKOS_LAMBDA (const t_team& t) {
     const int i = t.team_rank();
@@
  \begin{lstlisting}
-void MyKokkosFunction(double* a, const double* b, int n, int m) {
+void MyKokkosFunction(int* a, const double* b, int n, int m) {
   // Define the host execution space and the view types
-  typedef View<double*,DefaultHostExecutionSpace, Unmanaged> t_1d_view;
-  typedef View<double**[3],LayoutRight, DefaultHostExecutionSpace, Unmanaged> 
-     t_3d_view;
+  typedef View<int*, DefaultHostExecutionSpace, Unmanaged> t_1d_view;
+  typedef View<double**[3],LayoutRight, DefaultHostExecutionSpace, 
+               Unmanaged> t_3d_view;
 
@@
   // Create a 3D view of the second external allocation
-  // This assumes that the data had a row major layout (i.e. the third index is stride 1)
+  // This assumes that the data had a row major layout 
+  // (i.e. the third index is stride 1)
   t_3d_view d_b(b,n,m);
@@
 
-\subsubsection{Views as the fundamental data owning structure}
+\subsection{Views as the fundamental data owning structure}
 
@@
 
-\subsubsection{std::vector}
+\subsection{std::vector}
 
@@
 
-Kokkos provides a drop in replacement for |std::vector| in the form of \lstinline|Kokkos::vector|.
+Kokkos provides a drop in replacement for \lstinline|std::vector| with \lstinline|Kokkos::vector|.
 Outside of parallel kernels its semantics are mostly the same as that of \lstinline|std::vector|, for
@@
 
+\begin{lstlisting}
+// Create a vector of 1000 double elements
+Kokkos::vector<double> v(1000);
+// Create another vector as a copy of v; 
+// This allocates another 1000 doubles
+Kokkos::vector<double> x = v;
+
+parallel_for(1000, KOKKOS_LAMBDA (const int& i) {
+   // Create a view of x; m and x will reference the same data.
+   Kokkos::vector<double> m = x; 
+   x[i] = 2*i+1;
+   v[i] = m[i] - 1;
+});
+
+// Now x contains the first 1000 uneven numbers
+// v contains the first 1000 even numbers
+\end{lstlisting}
+
 \section{Calling non-Kokkos libraries}
 
-\subsection{Keeping track of data transfers}
+There are no restrictions on calling non-Kokkos libraries outside of parallel kernels. 
+However due to the polymorphic layouts of Kokkos views it is often required to test
+layouts for compatibility with third party libraries. The usual Blas interface for example,
+expects matrixes to be laid out in column major format (i.e. LayoutLeft in Kokkos). 
+Furthermore it is necessary to test that the library can access the memory space 
+of the view.
+
+\begin{lstlisting}
+template<class Scalar, class Device>
+Scalar dot(View<const Scalar* , Device> a, 
+           View<const Scalar*, Device> b) {
+// Check for Cuda memory and call cublas if true
+#ifdef KOKKOS_HAVE_CUDA
+  if(Impl::is_same<typename Device::memory_space,
+                            CudaSpace>::value ||
+     Impl::is_same<typename Device::memory_space,
+                            CudaUVMSpace>::value) {
+    return call_cublas_dot(a.ptr_on_device(), b.ptr_on_device(),
+                           a.dimension_0() );
+  }
+#endif
+
+// Call CBlas on the host otherwise
+  if(Impl::is_same<typename Device::memory_space,HostSpace>::value) {
+    return call_cblas_dot(a.ptr_on_device(), b.ptr_on_device(),
+                          a.dimension_0() );
+  }
+}
+\end{lstlisting}
+
+Note the \lstinline|Impl::is_same| construct can be replaced

### 2015-04-15 Kokkos_PG_MachineModel.tex
Updated style of the Guide . This comments out the SAND Report style
The stuff is still in if we really need to its not that hard to go back.

--- 2015-03-29 Kokkos_PG_MachineModel.tex
@@
 
-\subsection{Kokkos Execution Spaces}
+\section{Kokkos Execution Spaces}
 
@@
 
-\subsubsection{Execution Space Instances}
+\subsection{Execution Space Instances}
 
@@
 
-\subsubsection{Instances of Kokkos Memory Spaces}
+\subsection{Instances of Kokkos Memory Spaces}
 
@@
 
-\section{Thread teams}\label{S:Model:Teams}
-
-Kokkos supports multiple levels of parallelism.
-In particular, it groups threads into \emph{teams}.
-A \emph{thread team} is a collection of one or more parallel ``threads'' of execution.
-Kokkos allows an arbitrary number of teams -- the \emph{league size}.
-Hardware constrains the number of threads in a team -- the \emph{team size}.
-
-Threads in a team can synchronize -- they have a ``barrier'' primitive -- 
-and share a ``scratch pad'' memory which they may use for temporary storage.
-Scratch pad memory exists only during parallel operations;
-allocations in it do not persist across kernels.
-Teams themselves may run in any order,
-and may not necessarily run all in parallel.
-For example, if the user asks for $T$ teams,
-the hardware may choose to run them one after another in sequence,
-or in groups of up to $G$ teams at a time in parallel.
-
-Users may \emph{nest} parallel operations.
-Teams may perform one parallel operation (for, reduce, or scan),
-and threads within each team may perform another, possibly different parallel operation.
-Different teams may do entirely different things.
-For example, all the threads in one team may execute a \lstinline!parallel_for!,
-and all the threads in a different team may execute a \lstinline!parallel_scan!.
-Different threads within a team may also do different things.
-However, performance may vary if threads in a team ``diverge'' in their behavior
-(e.g., take different sides of a branch).
-Chapter \ref{C:Hierarchical} shows how the C++ implementation of Kokkos exposes thread teams.
-
-NVIDIA's CUDA programming model inspired Kokkos' thread team model.
-The scratch pad memory corresponds with CUDA's per-team ``shared memory.''
-The ``league / team'' vocabulary comes from OpenMP 4.0,
-and has many aspects in common with our thread team model.
-We have found that programming to this model results in good performance,
-even on computer architectures which only implement parts of the full model.
-For example, most multicore processors in common use for high-performance computing lack ``scratch pad'' hardware.
-However, if users request a scratch pad size that fits comfortably in the largest cache shared by the threads in a team,
-programming as if a scratch pad exists forces users to address locality in their algorithms.
-This also reflects the common experience that rewriting a code for more restrictive hardware,
-then porting the code \emph{back} to conventional hardware,
-tends to improve performance relative to an unoptimized code.
+%\section{Thread teams}\label{S:Model:Teams}
+%
+%Kokkos supports multiple levels of parallelism.
+%In particular, it groups threads into \emph{teams}.
+%A \emph{thread team} is a collection of one or more parallel ``threads'' of execution.
+%Kokkos allows an arbitrary number of teams -- the \emph{league size}.
+%Hardware constrains the number of threads in a team -- the \emph{team size}.
+%
+%Threads in a team can synchronize -- they have a ``barrier'' primitive -- 
+%and share a ``scratch pad'' memory which they may use for temporary storage.
+%Scratch pad memory exists only during parallel operations;
+%allocations in it do not persist across kernels.
+%Teams themselves may run in any order,
+%and may not necessarily run all in parallel.
+%For example, if the user asks for $T$ teams,
+%the hardware may choose to run them one after another in sequence,
+%or in groups of up to $G$ teams at a time in parallel.
+%
+%Users may \emph{nest} parallel operations.
+%Teams may perform one parallel operation (for, reduce, or scan),
+%and threads within each team may perform another, possibly different parallel operation.
+%Different teams may do entirely different things.
+%For example, all the threads in one team may execute a \lstinline!parallel_for!,
+%and all the threads in a different team may execute a \lstinline!parallel_scan!.
+%Different threads within a team may also do different things.
+%However, performance may vary if threads in a team ``diverge'' in their behavior
+%(e.g., take different sides of a branch).
+%Chapter \ref{C:Hierarchical} shows how the C++ implementation of Kokkos exposes thread teams.
+%
+%NVIDIA's CUDA programming model inspired Kokkos' thread team model.
+%The scratch pad memory corresponds with CUDA's per-team ``shared memory.''
+%The ``league / team'' vocabulary comes from OpenMP 4.0,
+%and has many aspects in common with our thread team model.
+%We have found that programming to this model results in good performance,
+%even on computer architectures which only implement parts of the full model.
+%For example, most multicore processors in common use for high-performance computing lack ``scratch pad'' hardware.
+%However, if users request a scratch pad size that fits comfortably in the largest cache shared by the threads in a team,
+%programming as if a scratch pad exists forces users to address locality in their algorithms.
+%This also reflects the common experience that rewriting a code for more restrictive hardware,
+%then porting the code \emph{back} to conventional hardware,
+%tends to improve performance relative to an unoptimized code.
 

### 2015-04-15 Kokkos_PG_ParallelDispatch.tex
Convert to SAND format.
Updated style of the Guide . This comments out the SAND Report style
The stuff is still in if we really need to its not that hard to go back.

@@
 int sum = 0;
-parallel_reduce (N, KOKKOS_LAMBDA (const int i, int& lsum) { lsum += i*i; });
+parallel_reduce (N, KOKKOS_LAMBDA (const int i, int& lsum) { 
+  lsum += i*i; 
+});
 \end{lstlisting}
@@
 const size_t N = x.dimension_0 ();
-parallel_scan (N, KOKKOS_LAMBDA (const int i, float& upd, const bool final) {
+parallel_scan (N, KOKKOS_LAMBDA (const int& i, 
+          float& upd, const bool& final) {
     if (final) {
@@
 
-\section{Functors}
 \section{Function Name Tags}
+
+When writing class based applications it often is useful to make the classes themselves functors.
+Using that approach allows the kernels to access all other class members, both data and functions. 
+An issue coming up in that case is the necessity for multiple parallel kernels in the same class.
+Kokkos supports that through function name tags. An application can use optional (unused) first arguments
+to differentiate multiple operators in the same class. Execution policies can take the type of that argument as
+an optional template parameter. The same applies to init, join and final functions.
+
+\begin{lstlisting}
+class Foo {
+  struct BarTag {};
+  struct RabTag {};
+
+  void compute() {
+     Kokkos::parallel_for(RangePolicy<BarTag>(0,100), *this);
+     Kokkos::parallel_for(RangePolicy<RabTag>(0,1000), *this);
+  }
+
+  KOKKOS_INLINE_FUNCTION
+  void operator() (const BarTag&, const int& i) const {
+    ...
+    foobar();
+    ...
+  }
+
+  KOKKOS_INLINE_FUNCTION
+  void operator() (const RabTag&, const int& i) const {
+    ...
+    foobar();
+    ...
+  }
+
+  void foobar() {
+    ...
+  }
+};
+\end{lstlisting}

### 2015-04-15 Kokkos_PG_ProgrammingModel.tex
Convert to SAND format.
Updated style of the Guide . This comments out the SAND Report style
The stuff is still in if we really need to its not that hard to go back.

@@
 Execution Spaces thus give an application developer the means to target different parts of a heterogenous hardware architecture.
+This corresponds directly to the previously described machine model.
 
@@
 An Execution Policy determines together with an Execution Pattern {\it How} a function is executed.
+Some policies can be nested in others.
+
+\subsection{Range Policies}
+
+The most simple form of execution policies are {\it Range Policies}. 
+They are used to execute an operation once for each element in a range. 
+There are no prescriptions of order of execution or concurrency, which means that it is not legal
+to synchronize different iterations.
+
+\subsection{Team Policies}
+
+Team policies are used to implement hierarchical parallelism. 
+For that purpose Kokkos groups threads into \emph{teams}.
+A \emph{thread team} is a collection of one or more parallel ``threads'' of execution.
+Kokkos allows an arbitrary number of teams -- the \emph{league size}.
+Hardware constrains the number of threads in a team -- the \emph{team size}.
+All threads in a team are guaranteed to run concurrently.
+
+Threads in a team can synchronize -- they have a ``barrier'' primitive -- 
+and share a ``scratch pad'' memory which they may use for temporary storage.
+Note that the not all forms of synchronization mechanism are legal in Kokkos, in particular
+implementing ``spin-locks'' for threads in a team may result in dead locks. 
+In Kokkos there is no forward progress guarantee for threads, that means that a single thread spinning
+on a lock acquired by another thread of the team may use 100\% of the compute engines resources.
+Calling the explicit Kokkos team barrier is the only safe way to synchronize threads in a team.
+
+Scratch pad memory exists only during parallel operations;
+allocations in it do not persist across kernels.
+Teams themselves may run in any order,
+and may not necessarily run all in parallel.
+For example, if the user asks for $T$ teams,
+the hardware may choose to run them one after another in sequence,
+or in groups of up to $G$ teams at a time in parallel.
+
+Users may \emph{nest} parallel operations.
+Teams may perform one parallel operation (for, reduce, or scan),
+and threads within each team may perform another, possibly different parallel operation.
+Different teams may do entirely different things.
+For example, all the threads in one team may execute a \lstinline!parallel_for!,
+and all the threads in a different team may execute a \lstinline!parallel_scan!.
+Different threads within a team may also do different things.
+However, performance may vary if threads in a team ``diverge'' in their behavior
+(e.g., take different sides of a branch).
+Chapter \ref{C:Hierarchical} shows how the C++ implementation of Kokkos exposes thread teams.
+
+NVIDIA's CUDA programming model inspired Kokkos' thread team model.
+The scratch pad memory corresponds with CUDA's per-team ``shared memory.''
+The ``league / team'' vocabulary comes from OpenMP 4.0,
+and has many aspects in common with our thread team model.
+We have found that programming to this model results in good performance,
+even on computer architectures which only implement parts of the full model.
+For example, most multicore processors in common use for high-performance computing lack ``scratch pad'' hardware.
+However, if users request a scratch pad size that fits comfortably in the largest cache shared by the threads in a team,
+programming as if a scratch pad exists forces users to address locality in their algorithms.
+This also reflects the common experience that rewriting a code for more restrictive hardware,
+then porting the code \emph{back} to conventional hardware,
+tends to improve performance relative to an unoptimized code.
 

### 2015-04-15 Kokkos_PG_Views.tex
Updated style of the Guide . This comments out the SAND Report style
The stuff is still in if we really need to its not that hard to go back.

--- 2015-03-27 Kokkos_PG_Views.tex
@@
 memory.  This \emph{coalesced access} gives the code better memory
-bandwidth.  In contrast, \lstinline!View<int**, OpenMP>! has
+bandwidth.
+
+In contrast, \lstinline!View<int**, OpenMP>! has
 \lstinline!LayoutRight!, so that a single thread accesses contiguous
@@
 View<const double*, MemoryTraits<SomeTrait | SomeOtherTrait> > b;
-View<double*, LayoutLeft, MemoryTraits<SomeTrait | SomeOtherTrait> > c;
-View<double*, MemorySpace, MemoryTraits<SomeTrait | SomeOtherTrait> > d;
-View<double*, LayoutLeft, MemorySpace, MemoryTraits<SomeTrait> > e;
+View<int*, LayoutLeft, MemoryTraits<SomeTrait | SomeOtherTrait> > c;
+View<int*, MemorySpace, MemoryTraits<SomeTrait | SomeOtherTrait> > d;
+View<int*, LayoutLeft, MemorySpace, MemoryTraits<SomeTrait> > e;
 \end{lstlisting}
@@
   // the reference counting system.
-  View<double*, HostSpace, MemoryTraits<Unmanaged> > x_view (x_raw, N0);
+  View<double*, HostSpace, MemoryTraits<Unmanaged> > 
+    x_view (x_raw, N0);
 

### 2015-04-15 Kokkos_QuickStartGuide.tex
Updated style of the Guide . This comments out the SAND Report style
The stuff is still in if we really need to its not that hard to go back.

\documentclass[a4paper,11pt]{book}
\usepackage[utf8x]{inputenc}

\usepackage{graphicx}

%% or use the epsfig package if you prefer to use the old commands
%% \usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols
%% \usepackage{amssymb}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers after \end{frontmatter}.
%% \usepackage{lineno}

%% \usepackage{placeins}

\usepackage{listings}
\usepackage{subfigure}
\usepackage{courier}
\usepackage{tikz}
\usetikzlibrary{shapes,positioning}
\usepackage{hyperref}
\hypersetup{
    colorlinks,
    citecolor=black,
    filecolor=black,
    linkcolor=black,
    urlcolor=black
}
\usepackage{url}

\usepackage{color}

\definecolor{lightlightgray}{gray}{0.97}
\definecolor{lightlightgreen}{rgb}{0.8,1,0.8}
\definecolor{darkblue}{rgb}{0,0,0.8}
\definecolor{darkgreen}{rgb}{0,0.5,0}
\definecolor{LightSkyBlue}{rgb}{0.8,1,0.8}
\definecolor{MidnightBlue}{rgb}{0.8,1,0.8}
\definecolor{SandiaGray}{RGB}{130, 120, 111}
\definecolor{SandiaBlue}{RGB}{0, 50, 90}
\definecolor{SandiaLightBlue}{RGB}{0,150,180}
\definecolor{SandiaLightLightBlue}{RGB}{0,200,220}
\definecolor{SandiaLightLightLightBlue}{RGB}{0,228,238}
\definecolor{SandiaRed}{RGB}{130, 36, 51}
\lstset{language=C++,
           frame=single,
           basicstyle=\ttfamily\scriptsize,
           %%basicstyle=\ttfamily,
           %%basicstyle=\sffamily\normalsize,
           keywordstyle=\color{darkblue},
           backgroundcolor=\color{lightlightgray},
           commentstyle=\color{darkgreen}
           }

\tikzset{HighLightBox/.style = {draw=SandiaBlue, fill=SandiaLightLightLightBlue, line width=2pt,
                          rectangle, rounded corners, inner sep=10pt, inner ysep=20pt},
         fancytitle/.style ={fill=SandiaBlue, text=white, rounded corners}
         }

\begin{document}
\begin{minipage}[h]{\paperwidth}
  {\begin{tikzpicture}[remember picture,overlay]
    \node[] at (current page.north west)
      {\begin{tikzpicture}[remember picture, overlay, scale=0.5]
          \draw[fill=SandiaLightLightBlue,color=SandiaLightLightBlue] (0,-10.5cm) rectangle (8cm,5.5cm);
          \draw[fill=SandiaLightBlue,color=SandiaLightBlue] (0,-8.5cm) rectangle (6cm,3.5cm);
          \draw[fill=SandiaBlue,color=SandiaBlue] (0,-6.5cm) rectangle (4cm,1.5cm);          
          \draw[fill=SandiaLightBlue,color=SandiaLightBlue] (4cm,-6.5cm) circle [radius=4cm];
          \draw[fill=SandiaLightLightBlue,color=SandiaLightLightBlue] (6cm,-8.5cm) circle [radius=6cm];
          \draw[fill=white,color=white] (8cm,-10.5cm) circle [radius=8cm];
          \draw[fill=SandiaGray,color=SandiaGray] (0,0cm) rectangle (2\paperwidth,-0.5cm);
          \draw[fill=SandiaBlue,color=SandiaBlue] (0,-0.5cm) rectangle (2\paperwidth,-2.5cm);
       \end{tikzpicture}
      };
   \end{tikzpicture}
   }

  {\begin{tikzpicture}[remember picture,overlay]
      \node[] at (current page.south east)
      {\begin{tikzpicture}[remember picture, overlay,scale=0.5]
          \draw[fill=SandiaLightLightBlue,color=SandiaLightLightBlue] (-8cm,11.5cm) rectangle (8cm,5.5cm);
          \draw[fill=SandiaLightBlue,color=SandiaLightBlue] (-6cm,9.5cm) rectangle (6cm,3.5cm);
          \draw[fill=SandiaBlue,color=SandiaBlue] (0,7.5cm) rectangle (-4cm,1.5cm);          
          \draw[fill=SandiaLightBlue,color=SandiaLightBlue] (-4cm,7.5cm) circle [radius=4cm];
          \draw[fill=SandiaLightLightBlue,color=SandiaLightLightBlue] (-6cm,9.3cm) circle [radius=6cm];
          \draw[fill=white,color=white] (-8cm,11.5cm) circle [radius=8cm];
        \draw[fill=SandiaBlue,color=SandiaBlue] (-2\paperwidth,0.5cm) rectangle
          (\paperwidth,3.5cm);
        \draw[fill=SandiaGray,color=SandiaGray] (-2\paperwidth,0) rectangle
          (\paperwidth,0.5cm);
       \end{tikzpicture}
      };                           
   \end{tikzpicture}
   }

  {\begin{tikzpicture}[remember picture,overlay]
     \node[HighLightBox,yshift=-7cm,xshift=0.25\paperwidth] at (current page.north west) (ParallelPatterns)
     {\begin{minipage}[h]{0.4\paperwidth}
Algorithms are expressed in parallel patterns:
\begin{lstlisting}
\\ Execute Functor Operator 
parallel_for( ExecutionPolicy , Functor);
\end{lstlisting}
\begin{lstlisting}
\\ Execute Functor Operator and perform Reduction 
parallel_reduce( ExecutionPolicy , Functor);
parallel_reduce( ExecutionPolicy , Functor, Result);
\end{lstlisting}
\begin{lstlisting}
\\ Execute Functor Operator and perform Scan 
parallel_scan( ExecutionPolicy , Functor);
\end{lstlisting}

      \end{minipage}
      };
      \node[fancytitle, right=10pt] at (ParallelPatterns.north west) {Parallel Patterns};   
      \end{tikzpicture}
  }
  {\begin{tikzpicture}[remember picture,overlay]
     \node[HighLightBox,yshift=-15cm,xshift=0.25\paperwidth] at (current page.north west) (ExecutionPolicies)
     {\begin{minipage}[h]{0.4\paperwidth}
Algorithms are expressed in parallel patterns:
\begin{lstlisting}
\\ Execute Functor Operator 
parallel_for( ExecutionPolicy , Functor);
\end{lstlisting}
\begin{lstlisting}
\\ Execute Functor Operator and perform Reduction 
parallel_reduce( ExecutionPolicy , Functor);
parallel_reduce( ExecutionPolicy , Functor, Result);
\end{lstlisting}
\begin{lstlisting}
\\ Execute Functor Operator and perform Scan 
parallel_scan( ExecutionPolicy , Functor);
\end{lstlisting}

      \end{minipage}
      };
      \node[fancytitle, right=10pt] at (ExecutionPolicies.north west) {Execution Policies};   
      \end{tikzpicture}
  }
\end{minipage}

\end{document}

### 2015-04-20 Kokkos_QuickStartGuide.tex
Updated Kokkos Quick Reference

@@
 \usetikzlibrary{shapes,positioning}
+\usepackage{mdwlist}
 \usepackage{hyperref}
@@
       };
+     \node[anchor=north west,xshift=-.05\paperwidth,rectangle,
+               yshift=3.75cm, rounded corners=10pt,inner sep=5pt,
+              fill=SandiaBlue]
+              {\color{white} \LARGE Kokkos: Quick Reference};
    \end{tikzpicture}
@@
       };                
+      \node[xshift=0.75\paperwidth,rectangle, yshift=2.65cm] at (current page.south west)
+              {\includegraphics[height=0.3in]{figures/SNL_Stacked_Black_Blue.jpg}};
+      \node[xshift=0.6\paperwidth,rectangle, yshift=2.65cm] at (current page.south west)
+              {\includegraphics[height=0.45in]{figures/DOEbwlogo.pdf}};
+
    \end{tikzpicture}
@@
 
-  {\begin{tikzpicture}[remember picture,overlay]
-     \node[HighLightBox,yshift=-7cm,xshift=0.25\paperwidth] at (current page.north west) (ParallelPatterns)
+  {\begin{tikzpicture}[remember picture,overlay,trim left]
+     \node[HighLightBox,yshift=-12cm,xshift=0.265\paperwidth] at (current page.north west) (ParallelPatterns)
      {\begin{minipage}[h]{0.4\paperwidth}
-Algorithms are expressed in parallel patterns:
+     \begin{small}
+Algorithms are expressed in parallel patterns for all which applies it:
+\begin{itemize}\itemsep1pt \parskip0pt \parsep0pt
+\item execute every element of index set
+\item order and concurrency are undefined
+\item \lstinline|MemberType| is used for \lstinline|ExecPolicy::member_type|
+\item \lstinline|TagType| is used for \lstinline|ExecPolicy::tag_type|
+\item \lstinline|TagType| is used if it is not \lstinline|void|
+\end{itemize}
+Using appropriate Exececution Policies parallel patterns can be nested.
 \begin{lstlisting}
-\\ Execute Functor Operator 
-parallel_for( ExecutionPolicy , Functor);
+parallel_for( ExecPolicy , Functor);
 \end{lstlisting}
+\vspace{-0.4cm}
+\begin{itemize}\itemsep1pt \parskip0pt \parsep0pt
+\item calls \lstinline|Functor::operator([TagType,] MemberType)|
+\end{itemize}
 \begin{lstlisting}
-\\ Execute Functor Operator and perform Reduction 
-parallel_reduce( ExecutionPolicy , Functor);
-parallel_reduce( ExecutionPolicy , Functor, Result);
+parallel_reduce( ExecPolicy , Functor);
+parallel_reduce( ExecPolicy , Functor, Result);
 \end{lstlisting}
+\vspace{-0.4cm}
+\begin{itemize}\itemsep1pt \parskip0pt \parsep0pt
+\item calls \lstinline|Functor::operator([TagType,] MemberType,| \lstinline|ValueType&)|
+\item performs a deterministic reduction over \lstinline|ValueType|
+\item \lstinline|ValueType| is inferred from \lstinline|Functor::operator|
+\item \lstinline|Result| must be a 0D \lstinline|View| of \lstinline|ValueType| or
+an lvalue of \lstinline|ValueType|
+\item thread contributions are combined by \lstinline|Functor::join([TagType, ]volatile ValueType&,| 
+ \lstinline|volatile ValueType&)| or \lstinline|ValueType::operator+=|  \lstinline|(volatile ValueType&)|
+if \lstinline|join| is not defined
+\item thread contributions are initialized by \lstinline|Functor::init([TagType, ] ValueType& )| 
+or \lstinline|ValueType()| if \lstinline|init| is not defined
+\item \lstinline|Functor::finalize([TagType, ] ValueType&)| is called for the final reduction
+value if defined 
+\end{itemize}
 \begin{lstlisting}
-\\ Execute Functor Operator and perform Scan 
-parallel_scan( ExecutionPolicy , Functor);
+parallel_scan( ExecPolicy , Functor);
 \end{lstlisting}
-
+\vspace{-0.4cm}
+\begin{itemize}\itemsep1pt \parskip0pt \parsep0pt
+\item calls \lstinline|Functor::operator([TagType,] MemberType,| \lstinline|ValueType&, bool final)|
+\item if \lstinline|final == true| \lstinline|ValueType| is the prefix sum
+\item the contributions have to be added to \lstinline|ValueType| irrespective of \lstinline|final|  
+\end{itemize}
+     \end{small}     
       \end{minipage}
@@
   }
+
   {\begin{tikzpicture}[remember picture,overlay]
-     \node[HighLightBox,yshift=-15cm,xshift=0.25\paperwidth] at (current page.north west) (ExecutionPolicies)
+     \node[HighLightBox,yshift=-14cm,xshift=0.75\paperwidth] at (current page.north west) (ExecutionPolicies)
      {\begin{minipage}[h]{0.4\paperwidth}
-Algorithms are expressed in parallel patterns:
+\begin{small}
+Execution Policies govern how a function is executed and what the index space is over which
+the parallel iteration is performed. 
 \begin{lstlisting}
-\\ Execute Functor Operator 
-parallel_for( ExecutionPolicy , Functor);
+RangePolicy<[ExecSpace,] [TagType]>
+      (begin, end)
+RangePolicy<[ExecSpace,] [TagType,] INT>
+      (begin, end)
+RangePolicy<[ExecSpace,] [TagType,] IntType>
+      (begin, end)
 \end{lstlisting}
+\vspace{-0.4cm}
+\begin{itemize}\itemsep1pt \parskip0pt \parsep0pt
+\item top level policy (i.e. not nested)
+\item statically splits \lstinline|(begin, end]| interval over compute resources
+\item \lstinline|ExecSpace| defaults to \lstinline|DefaultExecutionSpace|
+\item \lstinline|TagType| defaults to \lstinline|void|
+\item providing an integer number specifies chunk size for static splitting
+\item providing an integer type sets the type of the loop index
+\end{itemize}
 \begin{lstlisting}
-\\ Execute Functor Operator and perform Reduction 
-parallel_reduce( ExecutionPolicy , Functor);
-parallel_reduce( ExecutionPolicy , Functor, Result);
+TeamPolicy<[ExecSpace,] [TagType]>
+      (league_size, team_size[, vector_length])
+TeamPolicy<[ExecSpace,] [TagType,] INT>
+      (league_size, team_size[, vector_length])
+TeamPolicy<[ExecSpace,] [TagType,] IntType>
+      (league_size, team_size[, vector_length])
 \end{lstlisting}
+\vspace{-0.4cm}
+\begin{itemize}\itemsep1pt \parskip0pt \parsep0pt
+\item top level policy (i.e. not nested)
+\item statically splits \lstinline|league_size| range over interval over compute resources
+\item \lstinline|ExecSpace| defaults to \lstinline|DefaultExecutionSpace|
+\item \lstinline|TagType| defaults to \lstinline|void|
+\item providing an integer number specifies chunk size for static splitting
+\item providing an integer type sets the type of the loop index
+\item starts 2D or 3D parallelism: a league of thread teams with vector length
+\end{itemize}
 \begin{lstlisting}
-\\ Execute Functor Operator and perform Scan 
-parallel_scan( ExecutionPolicy , Functor);
+TeamThreadRange(TeamPolicy<...>::member_type, 
+                begin, end)
 \end{lstlisting}
-
+\vspace{-0.4cm}
+\begin{itemize}\itemsep1pt \parskip0pt \parsep0pt
+\item for each calling team split the range \lstinline|[begin,end)| over threads in the team
+\item must be nested in call with \lstinline|TeamPolicy|
+\item team collective
+\end{itemize}        
+\begin{lstlisting}
+ThreadVectorRange(TeamPolicy<...>::member_type, 
+                begin, end)
+\end{lstlisting}
+\vspace{-0.4cm}
+\begin{itemize}\itemsep1pt \parskip0pt \parsep0pt
+\item for each calling thread split the range \lstinline|[begin,end)| over vector lanes
+\item must be nested in call with \lstinline|TeamPolicy|
+\end{itemize}        
+\end{small}
       \end{minipage}
@@
   }
+
+  {
+  \begin{tikzpicture}[remember picture,overlay,trim left]
+     \node[HighLightBox,yshift=-24.75cm,xshift=0.265\paperwidth] at (current page.north west) (ExecutionSpace)
+     {\begin{minipage}[h]{0.4\paperwidth}
+     \begin{small}
+\begin{lstlisting}
+Serial
+\end{lstlisting}
+\vspace{-0.4cm}
+\begin{itemize}\itemsep1pt \parskip0pt \parsep0pt
+\item basic serial execution
+\end{itemize}
+\vspace{-0.4cm}
+\begin{lstlisting}
+OpenMP
+\end{lstlisting}
+\vspace{-0.4cm}
+\begin{itemize}\itemsep1pt \parskip0pt \parsep0pt
+\item using OpenMP 3.1 as backend
+\end{itemize}
+\vspace{-0.4cm}
+\begin{lstlisting}
+Threads
+\end{lstlisting}
+\vspace{-0.4cm}
+\begin{itemize}\itemsep1pt \parskip0pt \parsep0pt
+\item using Pthreads on Unix based systems
+\end{itemize}
+\vspace{-0.4cm}
+\begin{lstlisting}
+Cuda
+\end{lstlisting}
+\vspace{-0.4cm}
+\begin{itemize}\itemsep1pt \parskip0pt \parsep0pt
+\item using NVIDIAs Cuda 6.5 or higher as backend
+\end{itemize}
+\vspace{-0.4cm}
+\end{small}
+      \end{minipage}
+      };
+      \node[fancytitle, right=10pt] at (ExecutionSpace.north west) {Execution Spaces};   
+   \end{tikzpicture}
+  }
+
 \end{minipage}

### 2015-05-27 Kokkos_PG_Compilation.tex
Updated to the document. This version will now be send to R&A.

@@
 \hline
+   KOKKOS\_CUDA\_OPTIONS (IN): & Enable optional settings specific to CUDA. \\
+      \hspace{0.5cm}Options: & force\_uvm,use\_ldg,rdc \\
+      \hspace{0.5cm}Default: & \\
+
+\hline
    HWLOC\_PATH (IN): & Path to the hardware locality library if enabled. \\
@@
 
+Kokkos provides a script \lstinline|generate_makefile.bash| which can generate a Makefile for building and installing the library
+as well as building and running the tests. Please run \lstinline|generate_makefile.bash --help| for options. 
+Note that paths given to the script must be absolute paths, and the script must be run with the \lstinline|bash| shell (the script will
+do it if it is run directly i.e. as \lstinline|./generate_makefile.bash|.
+
 \section{Using Trilinos' CMake build system}\label{S:build:Trilinos}
@@
 You may override this by providing NVCC's '\lstinline|-ccbin|' option as a compiler flag.
-The default can be reset by editing the script itself.
+The default can be set by editing the script itself or by setting the environment variable \lstinline|NVCC_WRAPPER_DEFAULT_COMPILER|.
 
@@
 Make sure that \lstinline|nvcc_wrapper| calls the host compiler with which the MPI library was compiled.
-
-\section{Building for Intel Xeon Phi}\label{S:build:Phi}
-
-TODO

### 2015-05-27 Kokkos_PG_HierarchicalParallelism.tex
Updated style of the Guide . This comments out the SAND Report style
The stuff is still in if we really need to its not that hard to go back.
Updated to the document. This version will now be send to R&A.

--- 2015-03-20 Kokkos_PG_HierarchicalParallelism.tex
@@
 and reduces the time it takes to obtain an allocation
-(which currently is a few operations to calculate the offset).
+(which currently is a few tens of integer operations to calculate the offset).
 
@@
 The function expects the run-time dimensions as arguments, corresponding to View's constructor. 
+Note that the view must be unmanaged (i.e. it must have the \lstinline|Unmanaged| memory trait).
 
@@
 // Define a view type in ScratchSpace
-typedef Kokkos::View<int*[4],ScratchSpace> shared_int_2d;
+typedef Kokkos::View<int*[4],ScratchSpace,Kokkos::MemoryTraits<Kokkos::Unmanaged>> shared_int_2d;
 
@@
 
-\subsection{Restrict execution with \lstinline|single|}
+\subsection{Restricting execution to a single executor}
 

### 2015-05-27 Kokkos_PG_LegacyCodeIntegration.tex
Updated to the document. This version will now be send to R&A.

@@
 #ifdef KOKKOS_HAVE_CUDA
-  if(Impl::is_same<typename Device::memory_space,
+  if(std::is_same<typename Device::memory_space,
                            CudaSpace>::value ||
-     Impl::is_same<typename Device::memory_space,
+     std::is_same<typename Device::memory_space,
                            CudaUVMSpace>::value) {
@@
 // Call CBlas on the host otherwise
-  if(Impl::is_same<typename Device::memory_space,HostSpace>::value) {
+  if(std::is_same<typename Device::memory_space,HostSpace>::value) {
     return call_cblas_dot(a.ptr_on_device(), b.ptr_on_device(),
@@
 \end{lstlisting}
-
-Note the \lstinline|Impl::is_same| construct can be replaced

### 2015-05-27 Kokkos_PG_ParallelDispatch.tex
Updated to the document. This version will now be send to R&A.

@@
 
+It is a violation of Kokkos semantics to capture by reference \lstinline|[&]| for two reasons.
+First Kokkos might give the lambda to an execution space which can not access the stack of 
+the dispatching thread. Secondly, capturing by reference allows the programmer to violate the 
+const semantics of the lambda. For correctness and portability reasons lambdas and functors are
+treated as const objects inside the parallel code section. Capturing by reference allows a circumvention
+of that const property, and enables many more possibilities of writing non-threads-safe code.
+
+When using lambdas for nested parallelism (see Chapter \ref{C:Hierarchical} for details) using
+capture by reference can be useful for performance reasons, but the code is only valid Kokkos
+code if it also works with capturing by copy.
+
 \subsection{Should I use a functor or a lambda?}
@@
 Furthermore, if you would like to use lambdas with CUDA,
-you must have a sufficiently new version of CUDA.
+you must have a sufficiently new version of CUDA (at the time of writing Lambdas are not supported yet supported by the release 7.0).
 Finally, the ``execution tag'' feature, which lets you put together several different parallel loop bodies into a single functor,
@@
 For example, it can decide to execute in serial if the loop count is very small, and it would typically be faster to run in serial instead of introducing parallelization overhead.
+The \lstinline|RangePolicy| allows you to specify minimal chunk sizes in order to control potential concurrency for low trip count loops.
 
@@
 \end{lstlisting}
+
+This approach can also be used to template the operators by templating the tag classes which is useful to enable compile time evaluation of appropriate conditionals.

### 2015-05-27 Kokkos_PG_Views.tex
Updated to the document. This version will now be send to R&A.

@@
 templates.  View's first template parameter must be a valid C++ type.
+
+Note that the above used constructor is not necessarily available
+for all view types. Specific Layouts or Memory Spaces may require more 
+specialized allocators. This will be discussed later.
 
@@
 
+Note the return type of \lstinline|dimension_N()| is the \lstinline|size_type| of the 
+views memory space. This causes some issues if warning free compilation
+should be achieved since it will typically be necessary to cast the return value.
+In particular in cases where the \lstinline|size_type| is more conservative than 
+required it can be beneficial to cast the value to \lstinline|int|, since signed 32 bit
+integers typically give the best performance when used as index types. In index heavy 
+codes this performance difference can be significant compared to using \lstinline|size_t| 
+since the vector length on many architectures is twice as long for 32 bit values as for
+64 bit values, and signed integers have less stringent overflow testing requirements than
+unsigned integers.
+
 Users of the BLAS and LAPACK libraries may be familiar with the ideas
@@
 \begin{lstlisting}
-View<int*[3]> a ("a", CudaSpace, 10);
-View<int*[3]> b ("b", HostSpace, 10);
+View<int*[3], CudaSpace> a ("a", 10);
+View<int*[3], HostSpace> b ("b", 10);
 deep_copy (a, b); // This will give a compiler error
 
-View<int*[3], LayoutLeft> c ("c", CudaSpace, 10);
-View<int*[3], LayoutLeft> d ("d", HostSpace, 10);
+View<int*[3], LayoutLeft, CudaSpace> c ("c", 10);
+View<int*[3], LayoutLeft, HostSpace> d ("d", 10);
 deep_copy (c, d); // This might give a runtime error
@@
 \begin{lstlisting}
-View<int*[3]> a ("a", MemorySpace, 10);
+View<int*[3], MemorySpace> a ("a", 10);
 // Allocate a view in HostSpace with the 
 // layout and padding of a
-typename View<int*[3]>::HostMirror b =
+typename View<int*[3], MemorySpace>::HostMirror b =
     create_mirror(a);
@@
 The \lstinline!Atomic! access trait lets you create a View of data , such that every
-read or write to any entry uses an atomic update.  This only works for
-certain data types, for which the hardware and programming model
-support atomic updates.  Currently, Kokkos limits atomic updates to
-32-bit or 64-bit types\footnote{On X86 architectures Kokkos also supports 
-128-bit atomics. Types for which atomic access are performed must support the 
-necessary operators such as =,+=,-=,+,- etc.}.
+read or write to any entry uses an atomic update. Kokkos supports atomics for 
+all data types independent of size. Restrictions are that you are (i) not allowed to alias data
+for which atomic operations are performed, and the results of non atomic accesses (
+including read) to data which is at the same time atomically accessed is not defined.
 
-Note that it is a violation of the semantics to write \emph{or} read from a 
-non-atomic view of data if at the same time an atomic view of the data is used.
+Performance characteristics of atomic operation depend on the data type. 
+Some types (in particular integer types) are natively supported and might even provide 
+asynchronous atomic operations. Others (such as 32 bit and 64 bit atomics for non 
+integer types) are often implemented using CAS loops of integers. 
+Everything else is implemented with a locking approach, where an atomic operations 
+aquires a lock based on a hash of the pointer value of the data element.
+
+Types for which atomic access are performed must support the 
+necessary operators such as =,+=,-=,+,- etc. as well as have a number of \lstinline|volatile|
+overloads of functions such as assign and copy constructors defined.
 
@@
   // Access to x has less locality than access to y.
-  View<const double*, RandomAccess> x_ra;
+  View<const double*, RandomAccess> x_ra = x;
   typedef View<const size_t*>::size_type size_type;

### 2015-10-20 Kokkos_PG_ParallelDispatch.tex
Fixed minor omission of "const"
...  in parallel_reduce functor example.

@@
   join (volatile value_type dst,
-        volatile value_type src) const {
+        const volatile value_type src) const {
     for (size_type j = 0; j < value_count; ++j) {

### 2016-06-21 Kokkos_PG_Compilation.tex
Some updates

@@
 
+In order to compile Kokkos a C++11 compliant compiler is needed. 
+For an up to date list of compilers which are tested on a nightly basis, please refer to the README on the github repository. 
+At the time of writing supported compilers include: 
+\begin{lstlisting}
+  GCC   4.7.2, 4.8.x, 4.9.x, 5.1.x, 5.3.x
+  Intel 14.x, 15.x, 16.x, 17.x 
+  Clang 3.5.1, 3.6.x, 3.7.x, 3.8.x, 3.9.x; 
+  Cuda  6.5, 7.0, 7.5, 8.0; 
+  XL 13.3;
+\end{lstlisting}
+
 \section{Using Kokkos' Makefile system}\label{S:build:make}
@@
     KOKKOS\_ARCH (IN): & What backend Architecture to build for. \\
-      \hspace{0.5cm}Options: & KNC,SNB,HSW,Kepler,Kepler30,Kepler35,Kepler37,Maxwell,
+      \hspace{0.5cm}Options: & KNL,KNC,SNB,HSW,Kepler,Kepler30,Kepler35,Kepler37,Maxwell,
                Maxwell50,ARMv8,BGQ,Power7,Power8 \\
@@
 \hline
+   KOKKOS\_OPTIONS (IN): & Enable optional settings \\
+      \hspace{0.5cm}Options: & aggressive\_vectorization \\
+      \hspace{0.5cm}Default: & \\
+
+\hline
    KOKKOS\_CUDA\_OPTIONS (IN): & Enable optional settings specific to CUDA. \\
-      \hspace{0.5cm}Options: & force\_uvm,use\_ldg,rdc \\
+      \hspace{0.5cm}Options: & force\_uvm,use\_ldg,rdc,enable\_lambda \\
       \hspace{0.5cm}Default: & \\

### 2016-06-21 Kokkos_PG_HierarchicalParallelism.tex
Some updates

@@
 Kokkos' thread teams map a 2-D index range.
-The first index is the \emph{league}, the index of the team.
-The second index is the thread index within a team.
+The first index is the \emph{league rank}, the index of the team.
+The second index is the \emph{team rank}, the thread index within a team.
 In CUDA this is equivalent to launching a 1-D grid of 1-D blocks.
@@
 The constructors requires two arguments: a league size and a team size. 
+In place of the team size a user can utilize \lstinline|Kokkos::AUTO| to let Kokkos guess a good team size for a given architecture.
+Doing that is the recommend way for most developers to utilize the \lstinline|TeamPolicy|.
 As with the  \lstinline|Kokkos::RangePolicy| a specific execution tag and a specific execution space can be given as optional template arguments.
@@
 // Using  a specific execution space to 
-// run a n_worksets x team_size parallelism
+// run a n_worksets parallelism with Kokkos choosing the team size
 Kokkos::TeamPolicy<ExecutionSpace> 
-        policy( league_size, team_size );
+        policy( league_size, Kokkos::AUTO() );
 
@@
 // Create an instance of the policy
-TeamPolicy<ExecutionSpace> policy (league_size, team_size);
+TeamPolicy<ExecutionSpace> policy (league_size, Kokkos::AUTO() );
 // Launch a kernel
@@
 You may allocate a chunk of scratch memory through the \lstinline|TeamPolicy| member type.
-You may request multiple allocations from scratch, up to a user-provided maximum. 
-The maximum is provided either through a function in the functor which returns a potentially team-size dependent value, 
-or it can be specified as an argument to the constructor of the TeamPolicy. 
+You may request multiple allocations from scratch, up to a user-provided maximum aggregate size. 
+The maximum is provided either through a \lstinline|team_shmem_size| function in the functor which returns a potentially team-size dependent value, 
+or it can be specified through a setting of the TeamPolicy \lstinline|set_scratch_size|. 
 It is not valid to provide both values at the same time. 
@@
 
+The following is an example of using the functor interface:
 \begin{lstlisting}
@@
 };
+\end{lstlisting}
+
+The \lstinline|set_scratch_size| function of the \lstinline|TeamPolicy| takes two or three arguments.
+The first argument specifies the level in the scratch hierarchy for which a specific size is requested. 
+Different levels have different restrictions. 
+Generally the first level is restricted to a few tenths of kilobyte roughly corresponding to L1 cache size. 
+The second level can be used to get an aggregate over all teams of a few gigabyte, corresponding to available 
+space in high-bandwidth memory. 
+The third level generally falls back to capacity memory in the node. 
+The second and third argument are either per-thread or per-team sizes for scratch memory.
+Note like previously discussed, the setter function does not modify the instance it is called on, but returns 
+a copy of the policy object with adjusted scratch size request.
+
+Here are some examples:
+\begin{lstlisting}
+TeamPolicy<> policy_1 = TeamPolicy<>(league_size, team_size).
+                          set_scratch_size(1, PerTeam(1024), PerThread(32));
+TeamPolicy<> policy_2 = TeamPolicy<>(league_size, team_size).
+                          set_scratch_size(1, PerThread(32));
+TeamPolicy<> policy_3 = TeamPolicy<>(league_size, team_size).
+                          set_scratch_size(1, PerTeam(1024));
+\end{lstlisting}
+
+The total amount of scratch space available for each team will be the per-team value plus the per-thread value multiplied by the team-size. 
+The interface allows users to specify those settings inline:
+\begin{lstlisting}
+parallel_for(TeamPolicy<>(league_size, team_size).set_scratch_size(1, PerTeam(1024)),
+  KOKKOS_LAMBDA (const TeamPolicy<>::member_type& team) {
+    ...
+});
 \end{lstlisting}
@@
 
-You may use any of the three parallel patterns -- for, reduce, or scan -- at each level.
+You may use any of the three parallel patterns -- for, reduce, or scan -- at each level\footnote{The parallel scan operation is not implemented for all execution spaces on the thread level, and it doesn't support a TeamPolicy on the top level.}.
 You may nest them and use them in conjunction with code that is aware of the league and team rank.
@@
 
-\subsection{Team and vector loops}\label{SS:Hierarchical:Nested:Loops}
+\subsection{Team loops}\label{SS:Hierarchical:Nested:Loops:Team}
 
@@
 This is a conscious choice to prevent difficult to debug issues related to thread private, team shared and globally shared variables. 
-A simple way to enforce this is by using the ``capture by value'' clause with lambdas. 
+A simple way to enforce this is by using the ``capture by value'' clause with lambdas, 
+but ``capture by reference'' is recommended for release builds since it typically results in better performance.
 With the lambda being considered as \lstinline|const| inside the \lstinline!TeamThreadRange! loop,
@@
 
+\subsection{Vector loops}\label{SS:Hierarchical:Nested:Loops:Vector}
+
+At the innermost level of nesting parallel loops in a kernel is comprised of the \emph{vector}-loop.
+Vector level parallelism works identical to the team level loops using the execution policy \lstinline|ThreadVectorRange|.
+In contrast to the team-level, there is no legal way to exploit the vector level outside of a parallel pattern using the \lstinline|ThreadVectorRange|. 
+However one can use such a parallel construct in- and outside of a \lstinline|TeamThreadRange| parallel operation.
+
+\begin{lstlisting}
+using Kokkos::parallel_reduce;
+using Kokkos::TeamPolicy;
+using Kokkos::TeamThreadRange;
+using Kokkos::ThreadVectorRange;
+parallel_for (TeamPolicy<> (league_size, team_size),
+                 KOKKOS_LAMBDA (member_type team_member) {
+
+    int k = team_member.team_rank();
+    // The default reduction uses Scalar's += operator
+    // to combine thread contributions.
+    Scalar sum;
+    parallel_reduce (ThreadVectorRange (team_member, loop_count),
+      [=] (int& i, Scalar& lsum) {
+        // ...
+        lsum += ...;
+      }, sum);
+
+    parallel_for (TeamThreadRange (team_member, workset_size),
+      [&] (int& j) {
+      // You may provide a custom reduction as another
+      // lambda together with an initialization value.
+      Scalar product;
+      Scalar init_value = 1;
+      parallel_reduce (ThreadVectorRange (team_member, loop_count),
+        [=] (int& i, Scalar& lsum) {
+          // ...
+          lsum *= ...;
+        }, product, [=] (Scalar& lsum, Scalar& update) {
+          lsum *= update;
+        }, init_value);
+      });
+  });
+\end{lstlisting}
+
+As the name indicates the vector-level must be vectorizable. 
+The parallel patterns will exploit available mechanisms to encourage vectorization by the compiler. 
+When using the Intel compiler for example, the vector level loop will be internally decorated with 
+\lstinline|#pragma ivdep|, telling the compiler to ignore assumed vector dependencies.
+
 \subsection{Restricting execution to a single executor}
@@
 \end{itemize}
-The \lstinline!single! function takes a lambda as its second argument.
+The \lstinline|single| function takes a lambda as its second argument.
 That lambda takes zero arguments, 
@@
 It must always be correct for the lambda to capture variables by value
-(\lstinline![=]!, not \lstinline![&]!).
+(\lstinline|[=]|, not \lstinline|[&]|).
 Thus, if the lambda captures by reference,

### 2016-06-21 Kokkos_PG_ParallelDispatch.tex
Some updates

@@
 Furthermore, if you would like to use lambdas with CUDA,
-you must have a sufficiently new version of CUDA (at the time of writing Lambdas are not supported yet supported by the release 7.0).
+you must have a sufficiently new version of CUDA. 
+At the time of writing CUDA 7.5 supports lambdas using a special flag. 
+To enable this support, use the \lstinline|KOKKOS_CUDA_OPTIONS=enable_lambda| option.
+
 Finally, the ``execution tag'' feature, which lets you put together several different parallel loop bodies into a single functor,

### 2016-06-21 Kokkos_PG_Views.tex
Some updates

@@
 specialized allocators. This will be discussed later.
+
+Another important thing to keep in mind is that a \lstinline|View| handle is a 
+stateful object. It is not legal to create a \lstinline|View| handle from
+raw memory by typecasting a pointer. To call any operator on a \lstinline|View|
+its constructor must have been called before. This includes the assignment
+operator. If it is necessary to initialize raw memory with a \lstinline|View| 
+handle, one can legally do so using a move constructor (``placement new'').
+The above has nothing to do with the data a \lstinline|View| is referencing.
+It is completely legal to give a typecast pointer to the constructor of an
+unmanaged \lstinline|View|.
+
+\begin{lstlisting}
+View<int*> * a_ptr = (View<int*>*) malloc(10*sizeof(View<int*);
+a_ptr[0] = View<int*>("A0",1000); // This is illegal
+new(&a_ptr[1]) View<int*>("A1",10000); // This is legal 
+\end{lstlisting}
 
@@
 may also use Kokkos to implement Morton ordering or variants thereof.
+In order to write a custom layout one has to define a new layout class and
+specialise the \lstinline|ViewMapping| class for that layout. 
+The \lstinline|ViewMapping| class implements the offset operator as well
+as stride calculation for regular layouts. 
+A good way to start such a customization is by copying the implementation of 
+\lstinline|LayoutLeft| and its associated \lstinline|ViewMapping| specialization,
+renaming the layout and then change the offset operator.
 
@@
   if (std::is_same<array_layout, LayoutLeft>::value) {
-    callSomeBlasFunction (A.ptr_on_device (), ...);
+    callSomeBlasFunction (A.data(), ...);
   } else {
@@
 Nevertheless, sometimes you really need access to the
-pointer.  For such cases, we provide the \lstinline!ptr_on_device()!
+pointer.  For such cases, we provide the \lstinline!data()!
 method.  For example:
@@
   // DON'T DO THIS UNLESS YOU MUST
-  double* a_raw = a.ptr_on_device ();
+  double* a_raw = a.data ();
   const size_t N = x.dimension_0 ();
@@
 \end{lstlisting}
+
+\subsection{Conversion Rules and Function Specialization}
+
+Not all view types can be assigned to each other. In summary
+the data type and dimension have to match, the layout must be
+compatible and the memory space has to match. Examples illustrating
+the rules are:
+
+\begin{enumerate}
+\item Memory Spaces must match
+\subitem \lstinline{ Kokkos::View<int*> -> Kokkos::View<int*,HostSpace> // ok if default memory space is HostSpace}
+\item Data Type and Rank has to Match
+\subitem \lstinline{int*  -> int*          //ok}
+\subitem \lstinline{int*  -> const int*    //ok}
+\subitem \lstinline{const int*  -> int*    //not ok, const violation}
+\subitem \lstinline{int** -> int*          //not ok, rank mismatch}
+\subitem \lstinline{int*[3] -> int**       //ok}
+\subitem \lstinline{int** -> int*[3]       //ok if runtime dimension check matches}
+\subitem \lstinline{int*  -> long*         //not ok, type mismatch}
+\item Layouts must be compatible
+\subitem \lstinline{LayoutRight -> LayoutRight //ok}
+\subitem \lstinline{LayoutLeft -> LayoutRight  //not ok}
+\subitem \lstinline{LayoutLeft -> LayoutSride  //ok}
+\subitem \lstinline{LayoutStride -> LayoutLeft //ok if runtime dimensions allow assignment}
+\item Memory Traits 
+\end{enumerate}

### 2016-10-16 Kokkos_PG_HierarchicalParallelism.tex
Updates Hierarchical Parallelism

@@
 which is the first argument to the \verb!parallel_*! operation.
-Kokkos also exposes a ``scratch pad'' memory which all threads within a team may share.
+Kokkos also exposes a ``scratch pad'' memory which provides thread private and team private allocations..
 
@@
 
+You should use Hierarchical Parallelism in particular in a number of cases:
+\begin{enumerate}
+\item Non-tightly nested loops: Hierarchical Parallelism allows you to expose more parallelism.
+\item Data gather + reuse: If you gather data for a particular iteration of an outer loop, and then repeatiptly use it in an inner loop, Hierarchical Parallelism with scratch memory may match the usecase well.
+\item Force Cache Blocking: Using Hierarchical Parallelism forces a developer into algorithmic choices which are good for cache blocking.
+This can sometimes lead to better performing algorithms, than simple flat parallelism.
+\end{enumarate}
+
+On the other hand you should probably not use Hierarchical Parallelism if you have tightly nested loops. 
+For that usecase a multi dimensional RangePolicy is the better fit.
+
 \section{Thread teams}\label{S:Hierarchical:Teams}
@@
 Doing that is the recommend way for most developers to utilize the \lstinline|TeamPolicy|.
-As with the  \lstinline|Kokkos::RangePolicy| a specific execution tag and a specific execution space can be given as optional template arguments.
+As with the  \lstinline|Kokkos::RangePolicy| a specific execution tag, a specific execution space, a \lstinline|Kokkos::IndexType|, and a \lstinline|Kokkos::Schedule| can be given as optional template arguments.
 \begin{lstlisting}
@@
 
-\subsection{Basic kernels}\label{SS:Hierarchical:Teams:Kernels}
+\subsection{Basic kernels}\label{SS:Hierarchical:Teams:Kernels}:
 

### 2017-05-15 Kokkos_PG_Compilation.tex
SOme more updates

@@
   Clang 3.5.1, 3.6.x, 3.7.x, 3.8.x, 3.9.x; 
-  Cuda  6.5, 7.0, 7.5, 8.0; 
+  Cuda  7.0, 7.5, 8.0; 
   XL 13.3;
@@
     KOKKOS\_ARCH (IN): & What backend Architecture to build for. \\
-      \hspace{0.5cm}Options: & KNL,KNC,SNB,HSW,Kepler,Kepler30,Kepler35,Kepler37,Maxwell,
-               Maxwell50,ARMv8,BGQ,Power7,Power8 \\
+      \hspace{0.5cm}Options: & KNL,KNC,SNB,HSW,BDW,Kepler,Kepler30,Kepler35,Kepler37,Maxwell,
+               Maxwell50,Pascal60,Pascal61,ARMv8,ARMv81,ARMv8-ThunderX,BGQ,Power7,Power8 \\
       \hspace{0.5cm}Default: & None -- this means no particular architecture flags
@@
    KOKKOS\_USE\_TPLS (IN): & Enable optional third party libraries. \\
-      \hspace{0.5cm}Options: & hwloc,librt \\
+      \hspace{0.5cm}Options: & hwloc,librt,experimental\_memkind \\
       \hspace{0.5cm}Default: & \\
@@
 Thus, the whole application must be built with a CUDA-capable compiler.
-(At the moment, the only such compiler is NVIDIA's NVCC.)
+(At the moment, the only such compilers are NVIDIA's NVCC and Clang 4.0 [not released yet at time of writing].)
 More precisely, every compilation unit containing a Kokkos kernel or a function called from a Kokkos kernel has to be compiled with a CUDA-capable compiler. 

### 2017-05-15 Kokkos_PG_HierarchicalParallelism.tex
SOme more updates

@@
 This can sometimes lead to better performing algorithms, than simple flat parallelism.
-\end{enumarate}
+\end{enumerate}
 
@@
 
-\subsection{Basic kernels}\label{SS:Hierarchical:Teams:Kernels}:
+\subsection{Basic kernels}\label{SS:Hierarchical:Teams:Kernels}
 
@@
 // Define a view type in ScratchSpace
-typedef Kokkos::View<int*[4],ScratchSpace,Kokkos::MemoryTraits<Kokkos::Unmanaged>> shared_int_2d;
+typedef Kokkos::View<int*[4],ScratchSpace,
+          Kokkos::MemoryTraits<Kokkos::Unmanaged>> shared_int_2d;
 
@@
   ...  
-
 });
@@
 The \lstinline|single| function takes a lambda as its second argument.
-That lambda takes zero arguments, 
-so its body must perform side effects in order to have an effect.
+That lambda takes zero arguments or one argument by reference. 
+If it takes no argument, its body must perform side effects in order to have an effect.
+If it takes one argument, the final value of that argument is broadcast to every executor on the level: 
+i.e. every vectorlane of the thread, or every thread (and vector lane) of the team.  
 It must always be correct for the lambda to capture variables by value
@@
 \end{lstlisting}
+
+Here is an example of using the broadcast capabilities to determine the start offset for a team
+in a buffer:
+
+\begin{lstlisting}
+using Kokkos::parallel_for;
+using Kokkos::parallel_reduce;
+using Kokkos::TeamThreadRange;
+using Kokkos::ThreadVectorRange;
+using Kokkos::PerThread;
+
+TeamPolicy<...> policy (...);
+typedef TeamPolicy<...>::member_type team_member;
+
+Kokkos::View<int> offset("Offset");
+offset() = 0;
+
+parallel_for (policy, KOKKOS_LAMBDA (const team_member& thread) {
+  // ...
+
+  parallel_reduce (TeamThreadRange (thread, 100),
+    KOKKOS_LAMBDA (const int& i, int& lsum) {
+      if(...) lsum++;
+  });
+  Kokkos::single (PerTeam (thread), [=] (int& my_offset) {
+    my_offset = Kokkos::atomic_fetch_add(&offset(),lsum);
+  });
+  ...
+\end{lstlisting}
+
+To further illustrate the "parallel region" semantics of the team execution consider the following code:
+
+\begin{lstlisting}
+using Kokkos::parallel_reduce;
+using Kokkos::TeamThreadRange;
+using Kokkos::TeamPolicy;
+
+parallel_reduce(TeamPolicy<>(N,team_size),
+  KOKKOS_LAMBDA (const member_type& teamMember, int& lsum) {
+    int s = 0;
+    for(int i = 0; i<10; i++) s++;
+    lsum += s;
+},sum);
+\end{lstlisting}
+
+In this example \lstinline{sum} will contain the value \lstinline{N * team_size * 10}.
+Every thread in each team will compute \lstinline{s=10} and then contribute it to the sum.
+
+Lets go one step further and add a nested \lstinline{parallel_reduce}. 
+By choosing the loopbound to be \lstinline{team_size} every thread still only runs
+once through the inner loop.
+
+\begin{lstlisting}
+using Kokkos::parallel_reduce;
+using Kokkos::TeamThreadRange;
+using Kokkos::TeamPolicy;
+
+parallel_reduce(TeamPolicy<>(N,team_size),
+  KOKKOS_LAMBDA (const member_type& teamMember, int& lsum) {
+
+  int s = 0;
+  parallel_reduce(TeamThreadRange(teamMember, team_size),
+    [=] (const int k, int & inner_lsum) {
+    int inner_s = 0;
+    for(int i = 0; i<10; i++) inner_s++;
+    inner_lsum += inner_s;
+  },s);
+  lsum += s;
+},sum);
+\end{lstlisting}
+
+The answer in this case is neverless \lstinline{N * team_size * team_size * 10}.
+Each thread computes \lstinline{inner_s = 10}.
+But all threads in the team combine their results to compute a \lstinline{s} value of \lstinline{team_size * 10}.
+Since every thread in each team contributes that value to the global sum, we arrive at the final value of \lstinline{N * team_size * team_size * 10}. 
+If the intended goal was for each team to only contribute \lstinline{s} once to the global sum, 
+the contribution should have been protected with a \lstinline{single} clause.

### 2017-05-15 Kokkos_PG_MachineModel.tex
SOme more updates

@@
 high performance computing applications; the second is a concrete instantiation
-of the programming this model written in C++, which allows programmers
+of the programming model written in C++, which allows programmers
 to write to the concept machine model. It is important to treat these
@@
 have distinct memories, each with unique performance properties,
-that are accessible across the node ({\em i.e.} the memory is
-reachable or {\em shared} by all execution units. The specific
-layout shown in Figure~\ref{fig:kokkosnode} is an instantiation
+that may or may not be accessible across the node ({\em i.e.} the memory may be
+reachable or {\em shared} by all execution units, but specific 
+memory spaces may also be only accessible by specifc execution units). 
+The specific layout shown in Figure~\ref{fig:kokkosnode} is an instantiation
 of the Kokkos abstract machine model used to describe the potential
@@
 dispatch of kernels to instances is abstracted by the Kokkos
-framework allowed application programmers to be free from 
-written algorithms in hardware specific languages or
-requiring the programmer to understand the restrictions
-required for parallel execution in the execution space.
+model. 
+This allows application programmers to be free from 
+writing algorithms in hardware specific languages.
 
@@
 spaces}. Each memory space provides a finite storage capacity
-at which data structures can be allocated and accessed. In
-order to provide portable allocations and performant access,
-Kokkos requires data to be allocated as a $k$-dimensioned array,
-where $k$ must be greater than equal to zero (a zero array being
-a scalar). Each dimension of the array can be either fixed in
-size at compile time or dynamically sized with size resolution
-performed during execution.
+at which data structures can be allocated and accessed. 
+Differnt memory space types have different characteristics with respect
+to accesibility from execution spaces as well as their performance
+characteristics.
 
@@
 
-Although atomic operations are logically an operation
-on memory, they are performed by compute units and so
-are a property of execution spaces in the Kokkos framework.
-Execution spaces are not guaranteed to provide atomic
-operations support but attempts to use them on spaces
-which do not support atomic access will result in 
-a compile time error.
-
 \paragraph{Memory Consistency in Kokkos}
@@
 model. In the Kokkos model, the programmer should not assume any
-specific ordering of memory operations being isused by a
+specific ordering of memory operations being isued by a
 kernel. This has the potential to create race conditions between

### 2017-05-15 Kokkos_PG_ProgrammingModel.tex
SOme more updates

@@
 Effectively they allow for compile time transformation of algorithms to allow for adaptions of varying degrees of hardware parallelism as well as of the memory hierarchy.
+
+\begin{figure}[h]
+\begin{center}
+\includegraphics[clip, trim=1cm 5.5cm 12cm 2.5cm, width=\textwidth]{figures/kokkos-abstractions.pdf}
+\caption{The core abstractions of the Kokkos Programming Model}
+\label{fig:kokkosabstractions}
+\end{center}
+\end{figure}
 
@@
 and share a ``scratch pad'' memory which they may use for temporary storage.
-Note that the not all forms of synchronization mechanism are legal in Kokkos, in particular
+Note that not all forms of synchronization mechanism are legal in Kokkos, in particular
 implementing ``spin-locks'' for threads in a team may result in dead locks. 
@@
 Different logical memory spaces allow for concepts such as UVM memory in the CUDA programming model, which is accessible from Host and GPU. 
-Memory Spaces also can be used to express remote memory locations.
+Memory Spaces also could in the futrue also be used to express remote memory locations.
 Furthermore they encapsulate functionality such as consistency control and persistence scopes.
@@
 Layouts express the mapping from logical (or algorithmical) indicies to address offset for a data allocation. 
-By adopting appropriate layouts for memort structures an application can optimise data access patterns in a given algorithm.
+By adopting appropriate layouts for memory structures an application can optimise data access patterns in a given algorithm.
 If an implementation provides polymorphic layouts (i.e. a data structure can be instantiated at compile or runtime with different layouts) an architecture dependent optimisation can be performed.

### 2017-05-15 Kokkos_PG_Subviews.tex
SOme more updates

--- 2015-03-17 Kokkos_PG_Subviews.tex
@@
 
-To take a subview of a View, use the \texttt{Kokkos::subview}
+To take a subview of a View, you can use the \texttt{Kokkos::subview}
 function.  This function is overloaded for all different kinds of
@@
 
-A subview has the same reference count as its parent View, so the
-parent View won't be deallocated before all subviews go away.  Every
-subview is also a View.  This means that you may take a subview of a
+A subview has the same reference count as its parent \lstinline!View!, so the
+parent \lstinline!View! won't be deallocated before all subviews go away.  Every
+subview is also a \lstinline!View!.  This means that you may take a subview of a
 subview.
+
+Another way of getting a subview is throught the appropriate \lstinline{View} constructor.
+\begin{lstlisting}
+const size_t N0 = ...;
+const size_t N1 = ...;
+View<double**> A ("A", N0, N1);
+
+View<double**,LayoutStride> A_sub(A,make_pair(2,4),make_pair(3,7));
+\end{lstlisting}
+For this usage you must know the layouttype of the subview. 
+On the positive side, such a direct construction is generally a bit cheaper than
+through the \lstinline!Kokkos::subview! function.
 
@@
 \texttt{subview} function to the wrong type, Kokkos will emit a
-compile-time error.  The best way to avoid this is to use the C++11
+compile-time error.  The easiest way to avoid this is to use the C++11
 \texttt{auto} keyword to let the compiler deduce the correct type for
-you.
+you. That said \lstinline!auto! comes with its own cost. 
+It generally is more expansive for compilers to deal with \lstinline!auto! than with 
+explicitly known types.
 

### 2017-06-28 Kokkos_PG_HierarchicalParallelism.tex
Reformat lines overrunning margins

@@
 \begin{lstlisting}
-parallel_for(TeamPolicy<>(league_size, team_size).set_scratch_size(1, PerTeam(1024)),
+parallel_for(
+  TeamPolicy<>(league_size, team_size).set_scratch_size(1, PerTeam(1024)),
     KOKKOS_LAMBDA (const TeamPolicy<>::member_type& team) {

### 2017-12-17 Kokkos_PG_Views.tex
Reformat lines overrunning margins
Show how to make a View of Views; explain why this is generally unwise

@@
 \subsection{Constructing a View}
+\label{SS:View:CreateUse:Constructing}
 
@@
 \subsection{What types of data may a View contain?}
+\label{SS:View:CreateUse:ValidValueType}
 
@@
 global state.  Others may perform badly in parallel, like fine-grained
-dynamic memory allocation.  Therefore it is strongly advised to use 
-only simple data types inside Kokkos Views.
-Users may always construct a View whose entries are 
+dynamic memory allocation.  Therefore, we strongly advise that users
+only store simple data types inside Kokkos Views.
+
+Users may always construct a View whose entries are either
 \begin{itemize}
@@
 
-While it is in principal possible to have Kokkos Views of arbitrary 
-objects there are a number of restrictions. The \lstinline|T| type 
-must have a default constructor and destructor. For portability reasons
-\lstinline|T| is not allowed to have virtual functions and one is not 
-allowed to call functions with dynamic memory allocations inside of 
-kernel code. Furthermore assignment operators as well as default constructor
-and deconstructor must be marked with \lstinline|KOKKOS_[INLINE_]FUNCTION|. 
-Keep in mind that the semantics of the resulting View are a combination of the 
-Views 'view' semantics and the behaviour of the element type.
+It is possible to make Kokkos Views of other kinds of objects.
+However, Kokkos imposes restrictions on the set of types \lstinline|T|
+for which one can construct a \lstinline|View<T*>|.  For example:
+\begin{enumerate}
+\item \lstinline|T| must have a default constructor and destructor. 
+\item \lstinline|T| must not have virtual methods.
+\item \lstinline|T|'s default constructor and destructor must not
+  allocate data, and must be thread safe.
+\item \lstinline|T|'s assignment operators as well as its default
+  constructor and deconstructor must be marked with the
+  \lstinline|KOKKOS_INLINE_FUNCTION| or \lstinline|KOKKOS_FUNCTION|
+  macro.
+\end{enumerate}
+
+All restrictions but the first come from the requirement that
+\lstinline|View<T*>| work with every execution and memory space,
+including those that use CUDA.  The constructor of
+\lstinline|View<T*>| does not just allocate memory; it also
+initializes the allocation by iterating over it using a
+\lstinline|Kokkos::parallel_for|, and invoking \lstinline|T|'s default
+constructor for each entry.  If the View's execution space is
+\lstinline|Cuda|, then that means \lstinline|T|'s default constructor
+needs to be correct to call on device.
+
+\subsection{Can I make a View of Views?}
+\label{SS:View:CreateUse:ViewOfViews}
+
+A ``View of Views'' is a special case of View, where the type of each
+entry is itself a View.  It is possible to make this, but before you
+try, please see below.
+
+\subsubsection{You probably don't want this}
+
+If you really just want a multidimensional array, please don't do
+this.  Instead, see Section \ref{SS:View:CreateUse:Constructing} for
+the correct syntax.
+
+If you want to represent an array of arrays, and the inner arrays have
+fixed length or a fixed upper bound on length, consider instead using
+a ``compressed sparse row'' data structure.  Kokkos' Containers
+subpackage as a \lstinline|StaticCrsGraph| class that you may use for
+this purpose.
+
+If you want a hash table, Kokkos' Containers subpackage has an
+\lstinline|UnorderedMap| class that you may use for this purpose.
+
+One reason you might \emph{actually} want a View of Views is because
+you need a representation of a ``ragged'' array of arrays -- where the
+inner arrays have widely varying length -- and you need to be able to
+reallocate the inner arrays dynamically.
+
+You might also want a View of some class that itself contains Views.
+If you want this, first think about how to reorganize your data
+structures for better efficiency.
+
+\subsubsection{I really want a View of Views; what do I do?}
+
+Section \ref{SS:View:CreateUse:ValidValueType} above explains how the
+outer View's constructor works.  The outer View's constructor does not
+just allocate memory; it also initializes the allocation by iterating
+over it using a \lstinline|Kokkos::parallel_for|, and invoking the
+entry type's default constructor for each entry.  If the View's
+execution space is \lstinline|Cuda|, then that means the entry type's
+default constructor needs to be correct to call on device.  That is a
+problem, because the entry type in this case is itself View.  View's
+constructor wants to allocate memory, and thus does not work on
+device.  Kokkos parallel regions generally forbid memory allocation.
+
+One work-around is to invoke all inner Views' constructors on host.
+For example, one might allocate a host mirror version of the outer
+View, do the inner View allocations on host, then deep copy the outer
+View to device.  Alternately, one might use \lstinline!CudaUVMSpace!
+for the outer View, do the inner View allocations on host, then issue
+a \lstinline!Kokkos::fence! before using the outer View on device.
+The fence ensures that no device kernel will attempt to use the outer
+View before the host initialization has finished.
+
+A different approach defers constructing the inner Views until a
+separate Kokkos parallel region.  One first constructs the outer View
+using the \lstinline!WithoutInitializing! option, like this:
+\begin{lstlisting}
+using Kokkos::View;
+using Kokkos::view_alloc;
+using Kokkos::WithoutInitializing;
+
+// Need an std::string here, because the compiler gets confused
+// if you pass view_alloc a char* as its first argument.
+const std::string label ("v_outer");
+View<View<int*>> v_outer (view_alloc (label, WithoutInitializing));
+\end{lstlisting}
+This ensures that the inner Views' constructors don't get called on
+device.  However, that also means that the inner Views are not
+correctly constructed; they are not yet safe to use!  Thus, one must
+then use ``placement new'' in a Kokkos parallel region, in order to
+construct the inner Views.  This approach works best when the inner
+Views share a common memory pool for all their allocations.
+
+Here is a summary of different ways to make a View of Views:
+\begin{enumerate}
+\item Allocate a host mirror version of the outer View, do the inner
+  View allocations on host, then deep copy the outer View to device.
+\item Use \lstinline!CudaUVMSpace! for the outer View, do the inner
+  View allocations on host, then fence before using the outer View on
+  device.
+\item Allocate the outer View using the
+  \lstinline!WithoutInitializing! option, then use ``placement new''
+  in a Kokkos parallel region to construct the inner Views.
+\end{enumerate}
 
@@
 \item Memory Spaces must match
-\subitem \lstinline{ Kokkos::View<int*> -> Kokkos::View<int*,HostSpace> // ok if default memory space is HostSpace}
+  \subitem \lstinline{ Kokkos::View<int*> -> Kokkos::View<int*,HostSpace>}
+  \subitem \lstinline{                      //ok if default memory space is HostSpace}
 \item Data Type and Rank has to Match

### 2019-08-22 latex/access_policy_table.tex
initial commit

\begin{tabular}{|c|p{15em}|}
\hline
Expression & Meaning \\
\hline
    \texttt{A} & An \texttt{AccessPolicy} type. \\
    \texttt{a} & An instance of \texttt{A}. \\
    \texttt{A::element\_type} & The type of each element in the set of elements described by the associated \texttt{mdspan}. \\
    \texttt{A::pointer} & The pointer type through which a range of elements are accessed. \\
    \texttt{p} & An instance of \texttt{A::pointer}. \\
    \texttt{i} & Non-negative value of type \texttt{ptrdiff\_t}. \\
    \texttt{A::reference} & The type through which an element is accessed.  Must be convertible to \texttt{A::element\_type} \\
    \texttt{A::offset\_policy} & An \texttt{AccessPolicy} type convertible from \texttt{A}. \\
    \texttt{a.access(p, i)} & Returns an object that provides access to the \texttt{i}-th element in the range of elements that starts at \texttt{p}. \\
    \texttt{a.offset(p, i)} & An instance of \texttt{A::offset\_policy::pointer} for which \texttt{A::offset\_policy(a).access( a.offset(p, i), 0)} references the same element as \texttt{a.access(p, i)}. \\
    \texttt{a.decay(p)} & An instance of \texttt{A::element\_type*} for which \texttt{*a.decay(p)} references the same element as \texttt{A::element\_type(a.access(p, 0))} \\
\hline
\end{tabular}

### 2019-08-22 latex/layout_policy_table.tex
initial commit

\begin{tabular}{|c|p{15em}|}
\hline
Expression & Meaning \\
\hline
    \texttt{M} & A \texttt{LayoutMapping} type. \\
    \texttt{m} & An instance of \texttt{M}. \\
    \texttt{E} & A specialization of \texttt{std::extents}. \\
    \texttt{e} & An instance of \texttt{E}. \\
    \texttt{i...} and \texttt{j...} & Multidimensional indices in the multidimensional index space described by \texttt{e} \\
    \texttt{m.extents()} & The extents object \texttt{e} representing the multidimensional index domain of the mapping.  \\
    \texttt{m(i...)} & A nonnegative value representing the codomain offset corresponding to \texttt{i...}. \\
    \texttt{m.required\_span\_size()} & The maximum value of \texttt{m(i...)} plus \texttt{1} if all extents are non-zero, or \texttt{0} otherwise. \\
    \texttt{m.is\_unique()} & \texttt{true} only if for every \texttt{i... != j...}, \texttt{m(i...) != m(j...)}. \\
    \texttt{m.is\_contiguous()} & \texttt{true} only if the set defined by all \texttt{m(i...)} equals the set $\{0, ..., \mathtt{m.required\_span\_size()} - 1\}$ \\
    \texttt{m.is\_strided()} & \texttt{true} only if $\forall r \in [0, \mathtt{e.rank()}), \exists K_r$ such that $\forall \mathtt{i...}, \mathtt{j...} \in \mathtt{e}$, if all elements of \texttt{i...} and \texttt{j...} are equal except for the $r^{\mathrm{th}}$ element, with $j_r = i_r + 1$, then $K_r$ equals \texttt{m(j...) - m(i...)}. \\
    \texttt{m.stride(r)} & The integer $K_r$, as described above.  Only required if \texttt{m.is\_strided()} is \texttt{true}. \\
    \texttt{M::is\_always\_unique()} & \texttt{true} only if \texttt{m.is\_unique()} is \texttt{true} for all instances of \texttt{M}. \\
    \texttt{M::is\_always\_contiguous()} & \texttt{true} only if \texttt{m.is\_contiguous()} is \texttt{true} for all instances of \texttt{M}. \\
    \texttt{M::is\_always\_strided()} & \texttt{true} only if \texttt{m.is\_strided()} is \texttt{true} for all instances of \texttt{M}. \\
\hline
\end{tabular}

### 2019-08-22 latex/make_pdf.sh

@@ -1,0 +1,17 @@
+#!/usr/bin/env sh
+
+function section_to_tex() {
+  pandoc -f markdown-auto_identifiers+raw_tex -t latex ${mydir}/../$1.md > $mydir/$1.tex
+}
+
+
+mydir=$(cd $(dirname $0) && pwd)
+section_to_tex abstract
+section_to_tex design
+section_to_tex implementation
+section_to_tex benchmarks
+section_to_tex conclusions
+section_to_tex introduction
+cd $mydir
+pdflatex $mydir/paper.tex
+pdflatex $mydir/paper.tex

### 2019-08-22 plots/parse_gbench.py

@@ -1,0 +1,322 @@
+from os.path import split, join as path_join, abspath, dirname
+my_directory = dirname(abspath(__file__))
+
+import sys
+sys.path.insert(0, "/Users/dshollm/Projects/plotting_tools/source/head")
+import json
+import attr
+from glob import glob
+import re
+
+from plotting.plottable import *
+from plotting.plotters.bar_plot import bar_plot
+
+import matplotlib as mpl
+from matplotlib.figure import SubplotParams
+from matplotlib import pyplot as plt
+
+data_dir = "/Users/dshollm/Projects/mdspan/etc/p3hpc-19-paper/plots/data"
+
+class MicroBenchmark(DataPoint):
+    
+    def __init__(self, **kwargs):
+        for key, value in kwargs.items():
+            setattr(self, key, value)
+
+    @classmethod
+    def from_json(cls, entry, context):
+        kwargs = dict()
+        if context is not None:
+            kwargs.update(context)
+        kwargs.update(dict(entry))
+        return cls(**kwargs)
+
+    @classmethod
+    def load_all_from_json_output(cls, filename):
+        file = open(filename)
+        benchmark = json.load(file)
+        rv = []
+        assert("context" in benchmark)
+        assert("benchmarks" in benchmark)
+        for b in benchmark["benchmarks"]:
+            rv.append(cls.from_json(b, benchmark["context"]))
+        for m in rv:
+            setattr(m, "source_data_file", filename)
+        return rv
+    
+#===============================================================================
+
+class Variable:
+    pass
+
+class InputVariable(Variable):
+    pass
+
+class OutputVariable(Variable):
+    pass
+
+#===============================================================================
+
+@plottable_attribute
+class cpu_time(OutputVariable):
+    name = "Time"
+    value_type = float
+    
+@plottable_attribute
+class repeats(InputVariable):
+    name = "Repeats"
+    value_type = int
+    default_value = 1
+
+@plottable_derived_variable(
+    cpu_time / repeats / 1000
+)
+class execution_time(OutputVariable):
+    name = "Time (us)"
+    
+class Sum3DVariable:
+    def invalid_if(self, dp):
+        return "Sum_3D" not in run_name(dp)
+
+@plottable_attribute
+class host_name(InputVariable):
+    pass
+
+@plottable_attribute
+class source_data_file(InputVariable):
+    pass
+
+@plottable_attribute
+class run_name(InputVariable):
+    name = "Benchmark Name"
+
[…]

### 2019-08-27 plots/parse_gbench.py
updates

@@ -34,7 +34,11 @@ class MicroBenchmark(DataPoint):
     @classmethod
     def load_all_from_json_output(cls, filename):
         file = open(filename)
-        benchmark = json.load(file)
+        try:
+            benchmark = json.load(file)
+        except:
+            print(f"Error loading json file {filename}")
+            raise
         rv = []
         assert("context" in benchmark)
         assert("benchmarks" in benchmark)
@@ -84,6 +88,10 @@ class host_name(InputVariable):
 
 @plottable_attribute
 class source_data_file(InputVariable):
+    pass
+
+@plottable_attribute
+class run_type(InputVariable):
     pass
 
 @plottable_attribute
@@ -157,20 +165,67 @@ class sum_3d_iter_order(InputVariable, Sum3DVariable):
             return "Right"
         elif "_left/" in dp.run_name:
             return "Left"
-
+        
 @plottable_variable
-class sum_3d_type_and_shape(InputVariable, Sum3DVariable):
-    label_x_axis = False
-    name = "  "
+class data_structure_type(InputVariable):
+    name = "Data Structure"
+    value_type = str
     
     def invalid_if(self, dp):
-        return Sum3DVariable().invalid_if(dp) or (re.search(r'(d?)(\d+)_(d?)(\d+)_(d?)(\d+)', dp.run_name) is None)
+        val = dp.run_name
+        return "MDSpan" not in val and "Raw_" not in val
+    
+    def get_value(self, dp):
+        val = dp.run_name
+        if "MDSpan" in val:
+            return 'mdspan'
+        elif "Raw_" in val:
+            return 'raw pointer'
+        else:
+            return None
+
+@plottable_variable
+class sum_3d_shape(InputVariable):
+    label_x_axis = False
+    name = "  "
+
+    def invalid_if(self, dp):
+        return re.search(r'(d?)(\d+)_(d?)(\d+)_(d?)(\d+)', dp.run_name) is None
 
     def get_value(self, dp):
         val = dp.run_name
         m = re.search(r'(d?)(\d+)_(d?)(\d+)_(d?)(\d+)', val)
         x, y, z = map(int, (m[2], m[4], m[6]))
-        assert(x == y and y == z)
+        return f"{x}x{y}x{z}"
+
+@plottable_variable
+class sum_3d_dynamicness(InputVariable):
+    label_x_axis = False
[…]

### 2019-08-28 plots/parse_gbench.py
more text in benchmarks and implementation sections

@@ -1,4 +1,4 @@
-from os.path import split, join as path_join, abspath, dirname
+from os.path import split, join as path_join, abspath, dirname, basename
 my_directory = dirname(abspath(__file__))
 
 import sys
@@ -7,6 +7,7 @@ import json
 import attr
 from glob import glob
 import re
+import shutil
 
 from plotting.plottable import *
 from plotting.plotters.bar_plot import bar_plot
@@ -38,12 +39,14 @@ class MicroBenchmark(DataPoint):
             benchmark = json.load(file)
         except:
             print(f"Error loading json file {filename}")
-            raise
+            shutil.move(filename, f"{dirname(filename)}/broken/{basename(filename)}")
+            return None
         rv = []
         assert("context" in benchmark)
         assert("benchmarks" in benchmark)
         for b in benchmark["benchmarks"]:
-            rv.append(cls.from_json(b, benchmark["context"]))
+            if b["run_type"] != "aggregate":
+                rv.append(cls.from_json(b, benchmark["context"]))
         for m in rv:
             setattr(m, "source_data_file", filename)
         return rv
@@ -65,6 +68,12 @@ class OutputVariable(Variable):
 class cpu_time(OutputVariable):
     name = "Time"
     value_type = float
+
+
+@plottable_attribute
+class real_time(OutputVariable):
+    name = "Time"
+    value_type = float
     
 @plottable_attribute
 class repeats(InputVariable):
@@ -73,7 +82,7 @@ class repeats(InputVariable):
     default_value = 1
 
 @plottable_derived_variable(
-    cpu_time / repeats / 1000
+    real_time / repeats / 1000
 )
 class execution_time(OutputVariable):
     name = "Time (us)"
@@ -168,7 +177,7 @@ class sum_3d_iter_order(InputVariable, Sum3DVariable):
         
 @plottable_variable
 class data_structure_type(InputVariable):
-    name = "Data Structure"
+    name = ""
     value_type = str
     
     def invalid_if(self, dp):
@@ -265,7 +274,7 @@ class compiler(InputVariable):
 
 @plottable_variable
 class raw_vs_mdspan_plot_test_type(InputVariable):
-    name = "Benchmark"
+    name = ""
     value_type = str
 
     def get_value(self, dp):
@@ -277,42 +286,49 @@ class raw_vs_mdspan_plot_test_type(InputVariable):
                 if sum_3d_layout(dp) == "Right" or data_structure_type(dp) == "raw pointer":
                     bm = "Sum3D"
[…]

### 2019-08-29 latex/access_policy_table.tex
talk fully outlined

@@
     \texttt{A::reference} & The type through which an element is accessed.  Must be convertible to \texttt{A::element\_type} \\
-    \texttt{A::offset\_policy} & An \texttt{AccessPolicy} type convertible from \texttt{A}. \\
+    \texttt{A::offset\_policy} & (Optional) An \texttt{AccessPolicy} type convertible from \texttt{A}. Defaults to \texttt{A}. \\
     \texttt{a.access(p, i)} & Returns an object that provides access to the \texttt{i}-th element in the range of elements that starts at \texttt{p}. \\
@@
 \\
-    \texttt{a.decay(p)} & An instance of \texttt{A::element\_type*} for which \texttt{*a.decay(p)} references the same element as \texttt{A::element\_type(a.access(p, 0))} \\
 \hline

### 2019-08-29 plots/parse_gbench.py

@@ -18,6 +18,7 @@ from matplotlib import pyplot as plt
 
 data_dir = "/Users/dshollm/Projects/mdspan/etc/p3hpc-19-paper/plots/data"
 
+
 class MicroBenchmark(DataPoint):
     
     def __init__(self, **kwargs):
@@ -323,12 +324,21 @@ if __name__ == "__main__":
         file_data = MicroBenchmark.load_all_from_json_output(f)
         if file_data is not None:
             data.extend(file_data)
-        
 
+
+    use_paper_figure_style = True
     plot_sum_3d_cuda = False
     plot_sum_3d_left_right_apollo = False
     plot_sums = False
     make_overview_figure = True
+
+    #================================================================================
+    # paper figure style
+    if use_paper_figure_style:
+        plt.rc('font', family="serif")
+        plt.rc('legend', fontsize="small", framealpha=1.0)
+
+
     #================================================================================
     if plot_sums:
     #================================================================================
@@ -339,7 +349,7 @@ if __name__ == "__main__":
                   subplotpars=SubplotParams(
                       left=0.11,
                       right=0.87,
-                      bottom=0.2
+                      bottom=0.15
                   )
               )
               s = get_series(data,
@@ -440,9 +450,9 @@ if __name__ == "__main__":
             subplotpars=SubplotParams(
                 left=0.09,
                 right=0.95,
-                bottom=0.35
+                bottom=0.32
             ),
-            figsize=[7.0, 3.2]
+            figsize=[7.0, 3.0]
         )
         s = get_series(data,
             series_variable=data_structure_type,
@@ -465,7 +475,8 @@ if __name__ == "__main__":
             legend=True,
             error_bars=True,
             fig=fig, ax=ax,
-            xticklabels_keywords=dict(rotation=90, fontdict=dict(fontsize=8.5), family="serif")
+            xticklabels_keywords=dict(rotation=90, fontdict=dict(fontsize=7), family="serif"),
+            yticklabels_text_keywords=dict(fontsize=7, family="serif"),
         ).figure
         #ax.set_ylim(0.75, 1.25)
         fig.suptitle(f"Summary of Selected Benchmarks")

### 2019-08-30 plots/parse_gbench.py
more plots

@@ -3,10 +3,10 @@ my_directory = dirname(abspath(__file__))
 
 import sys
 sys.path.insert(0, "/Users/dshollm/Projects/plotting_tools/source/head")
+sys.path.insert(0, my_directory)
+
 import json
-import attr
 from glob import glob
-import re
 import shutil
 
 from plotting.plottable import *
@@ -15,15 +15,25 @@ from plotting.plotters.bar_plot import bar_plot
 import matplotlib as mpl
 from matplotlib.figure import SubplotParams
 from matplotlib import pyplot as plt
+from plotting.plottable.combined_data_point import *
 
-data_dir = "/Users/dshollm/Projects/mdspan/etc/p3hpc-19-paper/plots/data"
+from variables import *
 
 
 class MicroBenchmark(DataPoint):
     
     def __init__(self, **kwargs):
+        self._known_aspects = []
         for key, value in kwargs.items():
+            self._known_aspects.append(key)
             setattr(self, key, value)
+
+    def __repr__(self):
+        rv = f"{type(self).__name__}(\n"
+        for k in self._known_aspects:
+            rv += f"  {k}={repr(getattr(self, k))},\n"
+        rv += ")"
+        return rv
 
     @classmethod
     def from_json(cls, entry, context):
@@ -38,7 +48,7 @@ class MicroBenchmark(DataPoint):
         file = open(filename)
         try:
             benchmark = json.load(file)
-        except:
+        except json.JSONDecodeError:
             print(f"Error loading json file {filename}")
             shutil.move(filename, f"{dirname(filename)}/broken/{basename(filename)}")
             return None
@@ -51,293 +61,26 @@ class MicroBenchmark(DataPoint):
         for m in rv:
             setattr(m, "source_data_file", filename)
         return rv
-    
-#===============================================================================
 
-class Variable:
-    pass
-
-class InputVariable(Variable):
-    pass
-
-class OutputVariable(Variable):
-    pass
-
-#===============================================================================
-
-@plottable_attribute
-class cpu_time(OutputVariable):
-    name = "Time"
-    value_type = float
-
-
-@plottable_attribute
-class real_time(OutputVariable):
-    name = "Time"
-    value_type = float
-    
-@plottable_attribute
-class repeats(InputVariable):
-    name = "Repeats"
[…]

### 2019-09-01 latex/machine_table.tex
Adding machine table

\begin{tabular}{|l|l|l|}
\hline
Machine | Hardware | Compilers \\
\hline
Apollo & NVIDIA TitanV & CUDA 10.1 (nvcc),GCC 5.3.0 \\
Astra & Cavium Thunder-X2 CN99xx & GCC 8.2 (ARM) \\ 
Blake & Intel(R) Xeon(R) Platinum 8160 & Intel 19.3.199, Intel 18.5.0 \\
Mutrino & Intel(R) Xeon(R) CPU E5-2698 v3 & Intel 18.5.0 \\
\hline
\end{tabular}

### 2019-09-02 latex/layout_policy_table.tex
Design access policy subsection copy-edits.

@@
     \texttt{e} & An instance of \texttt{E}. \\
-    \texttt{i...} and \texttt{j...} & Multidimensional indices in the multidimensional index space described by \texttt{e} \\
+    \texttt{i...} and \texttt{j...} & Multidimensional indices in the multidimensional index space described by \texttt{e}. \\
     \texttt{m.extents()} & The extents object \texttt{e} representing the multidimensional index domain of the mapping.  \\
@@
     \texttt{m.is\_unique()} & \texttt{true} only if for every \texttt{i... != j...}, \texttt{m(i...) != m(j...)}. \\
-    \texttt{m.is\_contiguous()} & \texttt{true} only if the set defined by all \texttt{m(i...)} equals the set $\{0, ..., \mathtt{m.required\_span\_size()} - 1\}$ \\
+    \texttt{m.is\_contiguous()} & \texttt{true} only if the set defined by all \texttt{m(i...)} equals the set $\{0, ..., \mathtt{m.required\_span\_size()} - 1\}$. \\
     \texttt{m.is\_strided()} & \texttt{true} only if $\forall r \in [0, \mathtt{e.rank()}), \exists K_r$ such that $\forall \mathtt{i...}, \mathtt{j...} \in \mathtt{e}$, if all elements of \texttt{i...} and \texttt{j...} are equal except for the $r^{\mathrm{th}}$ element, with $j_r = i_r + 1$, then $K_r$ equals \texttt{m(j...) - m(i...)}.

### 2019-09-02 latex/machine_table.tex
added static extents figure

@@
-\begin{tabular}{|l|l|l|}
+%\rowcolors{1}{white}{black!10!white}
+\renewcommand{\arraystretch}{1.5}
+\begin{tabularx}{0.45\textwidth}{|c|>{\raggedright\arraybackslash}X|>{\raggedright\arraybackslash}X|}
 \hline
-Machine | Hardware | Compilers \\
+Machine & Hardware & Compilers \\
 \hline
-Apollo & NVIDIA TitanV & CUDA 10.1 (nvcc),GCC 5.3.0 \\
-Astra & Cavium Thunder-X2 CN99xx & GCC 8.2 (ARM) \\ 
-Blake & Intel(R) Xeon(R) Platinum 8160 & Intel 19.3.199, Intel 18.5.0 \\
-Mutrino & Intel(R) Xeon(R) CPU E5-2698 v3 & Intel 18.5.0 \\
+    Apollo & NVIDIA TitanV & CUDA~10.1~(nvcc),\newline GCC~5.3.0 \\
+Astra & Cavium Thunder-X2 CN9975-2000 & GCC~8.2~(ARM) \\ 
+Blake & Intel(R) Xeon(R) Platinum 8160 & Intel~19.3.199, Intel~18.0.5 \\
+Mutrino & Intel(R) Xeon(R) CPU E5-2698 v3 & Intel~18.0.5, GCC~8.2.0, Clang~9~(develop) \\
 \hline
-\end{tabular}
+\end{tabularx}
+\renewcommand{\arraystretch}{1.0}

### 2019-09-02 latex/make_pdf.sh
added bibtex stuff and bibtex pass

@@ -14,4 +14,6 @@ section_to_tex conclusions
 section_to_tex introduction
 cd $mydir
 pdflatex $mydir/paper.tex
+bibtex paper
+pdflatex $mydir/paper.tex
 pdflatex $mydir/paper.tex

### 2019-09-05 latex/machine_table.tex
added IBM Power 9 numbers

@@
 Mutrino & Intel(R) Xeon(R) CPU E5-2698 v3 & Intel~18.0.5, GCC~8.2.0, Clang~9~(develop) \\
+Waterman & IBM Power9 10C 80HT & GCC~7.3.0 \\
 \hline

### 2020-01-20 EduPar2020/src/1.Introduction.tex
Initial version for review.

\section{Introduction}\label{chap:introduction}

Parallel and distributed computing moved into focus at universities and teaching institutions. The interest is two-fold. Firstly, given their role, educators adapt to trends in research and industry and include parallel computing in their curricula, and secondly, parallel computing is an attractive field for department research and industry collaborations.

While industry-driven collaborations often predefine technological frameworks such as programming environments and targeted architectures, academic teaching leaves this mostly undefined. In this case, educators are exposed to a variety of solutions, approaches and trends in parallel computing. Which particular set to choose for teaching in class is not trivial. While decision factors differ, maximizing the subject relevance by teaching generic principles of parallel programming generically is among the most important ones.

Consequently, generic parallel programming models are needed that target mainstream programming languages and allow a vendor- and architecture independent, performance-portable expression of concurrency. In this way, a parallel program written in a popular language becomes capable of running efficiently on parallel computer architectures that conform to a common parallel machine model specification. We would like to compare this to the support of Von Neumann architectures in programming languages today. We define a abstract machine model later in this work. The lack of such parallel programming models motivated us to develop Kokkos.

Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS_PAPER_HERE}, is a performance-portable, parallel programming model for C++ developed at Sandia National Laboratories (SNL). It consists of a programming model specification and a library. The specification defines execution- and memory model and an application programming interface (library API). The runtime library exposes programming primitives, namely, patterns and types and maps them to the underlying abstract machine model. The machine model creates a generic view of compute hardware and memories. The Kokkos library mapps execution primitives to vendor-specific programming models, libraries and memory layouts. Kokkos has been successfully used in many large-scale projects at SNL, LLNL and other research centers~\cite{CITEKOKKOSUSECASES}. We list its key properties as follows.

\begin{itemize}
\item Parallel programming model for C++
\item Targets an abstract machine model
\item Implements a parallel-patterns and tasking paradigms
\item Offers abstract data types for performance portability
\item Pure C++ implementation through template meta programming
\item Aligns with to C++ standardization efforts for parallel programming
\item Vendor- and architecture independent and open-source
\item Includes exercises and teaching material
\end{itemize}

The objective of this paper is to present Kokkos from two perspectives. Firstly, we would like to present our thoughts that accompanied the design and development of the Kokkos parallel programming model. This insights may be considered useful in parallel programming classes that introduce programming models and reason about design choices.

Further, we would like to present Kokkos to the educational community as one representative parallel programming model for a trend towards architecture-independent and performance-portable programming of CPUs and accelerators. We show a set of types that implement patterns and encapsulate data with enough semantic information to produce performance-portable, parallel code on current architectures. Further, we highlight its alignment and contributions to standardization efforts towards supporting parallel programming in C++.

Secondly, we would like to promote Kokkos to students as a platform to understand and experiment with parallel patterns and types. Type specialization through templating allows to experiment with memory layouts and investigate their performance implications. Further, its modular code organization and open access under the BSD license allow students to peak behind the scenes and explore the mappings of abstractions to a variety of vendor-specific implementations. We believe that this makes Kokkos a valuable programming model for students to reason about vendor-provided programming models and to see them as particular implementations of higher abstractions in a parallel programming landscape.

This rest of the paper is structured as follows. The next Chapter gives an overview on parallel programming concepts and present the notion of \emph{semantic capture}. Chapter~\ref{chap:kokkosMM} presents the abstract machine model used to derive required semantic information for performance portability in Kokkos. Chapters~\ref{chap:kokkosPM} and ~\ref{chap:kokkosBackend} present the Kokkos programming model and its back-end support. Chapter~\ref{chap:C++} shows the alignment to C++ standardization efforts and lastly Chapter~\ref{chap:conclusion} concludes this work.

### 2020-01-20 EduPar2020-2/Test_App/src/helper.hpp

@@ -1,0 +1,6 @@
+#ifndef HELPER_HPP
+#define HELPER_HPP
+typedef struct  {
+  int N, M, S, nrepeat;
+} Args;
+#endif //HELPER_HPP

### 2020-01-20 EduPar2020-2/Test_App/src/main.cpp

@@ -1,0 +1,36 @@
+#include <limits>
+#include <cstdio>
+#include <cstdlib>
+#include <cstring>
+#include <sys/time.h>
+
+#include "helper.hpp"
+
+Args parse_args(int argc, char* argv[]);
+
+void checkSizes( int &N, int &M, int &S, int &nrepeat );
+double run(int argc, char* argv[], Args args);
+
+int main( int argc, char* argv[] )
+{
+  Args args  = parse_args (argc, argv);
+  
+  int N = args.N;
+  int M = args.M;
+  int nrepeat = args.nrepeat;
+
+  double time = run(argc, argv, args);
+
+  // Calculate bandwidth.
+  // Each matrix A row (each of length M) is read once.
+  // The x vector (of length M) is read N times.
+  // The y vector (of length N) is read once.
+  // double Gbytes = 1.0e-9 * double( sizeof(double) * ( 2 * M * N + N ) );
+  double Gbytes = 1.0e-9 * double( sizeof(double) * ( M + M * N + N ) );
+
+  // Print results (problem size, time and bandwidth in GB/s).
+  printf( "  N( %d ) M( %d ) nrepeat ( %d ) problem( %g MB ) time( %g s ) bandwidth( %g GB/s )\n",
+          N, M, nrepeat, Gbytes * 1000, time, Gbytes * nrepeat / time );
+
+  return 0;
+}

### 2020-01-20 EduPar2020-2/Test_App/src/matvec_kokkos.cpp

@@ -1,0 +1,93 @@
+#include <Kokkos_Core.hpp>
+#include "helper.hpp"
+
+double run(int argc, char* argv[], Args args)
+{
+  int N = args.N;
+  int M = args.M;
+  int nrepeat = args.nrepeat;
+  double time;
+
+  Kokkos::initialize( argc, argv );
+  {
+    // typedef Kokkos::Serial   ExecSpace;
+    // typedef Kokkos::Threads  ExecSpace;
+    // typedef Kokkos::OpenMP   ExecSpace;
+    typedef Kokkos::Cuda     ExecSpace;
+
+    // typedef Kokkos::HostSpace     MemSpace;
+    // typedef Kokkos::OpenMP        MemSpace;
+    typedef Kokkos::CudaSpace     MemSpace;
+    // typedef Kokkos::CudaUVMSpace  MemSpace;
+
+    typedef Kokkos::LayoutLeft   Layout;
+    //typedef Kokkos::LayoutRight  Layout;
+    typedef Kokkos::RangePolicy<ExecSpace>  range_policy;
+
+    // Allocate y, x vectors and Matrix A on device.
+    typedef Kokkos::View<double*, Layout, MemSpace>   ViewVectorType;
+    typedef Kokkos::View<double**, Layout, MemSpace>  ViewMatrixType;
+    ViewVectorType y( "y", N );
+    ViewVectorType x( "x", M );
+    ViewMatrixType A( "A", N, M );
+
+    // Create host mirrors of device views.
+    ViewVectorType::HostMirror h_y = Kokkos::create_mirror_view( y );
+    ViewVectorType::HostMirror h_x = Kokkos::create_mirror_view( x );
+    ViewMatrixType::HostMirror h_A = Kokkos::create_mirror_view( A );
+
+    // Initialize y vector on host.
+    for ( int i = 0; i < N; ++i ) {
+      h_y( i ) = 1;
+    }
+
+    // Initialize x vector on host.
+    for ( int i = 0; i < M; ++i ) {
+      h_x( i ) = 1;
+    }
+
+    // Initialize A matrix on host.
+    for ( int j = 0; j < N; ++j ) {
+      for ( int i = 0; i < M; ++i ) {
+        h_A( j, i ) = 1;
+      }
+    }
+
+    // Deep copy host views to device views.
+    Kokkos::deep_copy( y, h_y );
+    Kokkos::deep_copy( x, h_x );
[…]

### 2020-01-20 EduPar2020-2/Test_App/src/matvec_openmp_device.cpp

@@ -1,0 +1,86 @@
+#include <Kokkos_Core.hpp>
+#include <assert.h>
+#include "helper.hpp"
+
+
+template <class T>
+struct vec_t
+{
+  T * v;
+  vec_t(size_t m){v = (T*) malloc (m * sizeof(T)); assert(v!=NULL);}
+  T & operator () (const size_t i) {return v[i];}
+};
+
+template <class T>
+struct mtrx_t
+{
+  T * v;
+  size_t dim;
+  mtrx_t(size_t n, size_t m) {v = (T*) malloc (n * m * sizeof(T)); dim = m; assert(v!=NULL); }
+  T & operator () (const size_t i, const size_t j) {return v[j*dim+i];}
+};
+
+double run(int argc, char* argv[], Args args)
+{
+  int N = args.N;
+  int M = args.M;
+  int nrepeat = args.nrepeat;
+  double time;
+  {
+    typedef vec_t<double> vec_double_t;
+    typedef mtrx_t<double> mtrx_double_t;
+
+    vec_double_t y( N );
+    vec_double_t x( M );
+    mtrx_double_t A( N, M );
+
+        // Initialize y vector on host.
+    for ( int i = 0; i < N; ++i ) {
+      y( i ) = 1;
+    }
+
+    // Initialize x vector on host.
+    for ( int i = 0; i < M; ++i ) {
+      x( i ) = 1;
+    }
+
+    // Initialize A matrix on host.
+    for ( int j = 0; j < N; ++j ) {
+      for ( int i = 0; i < M; ++i ) {
+        A( j, i ) = 1;
+      }
+    }
+
+    // Timer products.
+    Kokkos::Timer timer;
+
+    for ( int repeat = 0; repeat < nrepeat; repeat++ ) {
+      // Application: <y,Ax> = y^T*A*x
+      double result = 0;
+      # pragma omp target data map (A, x, y, A.v, x.v, y.v)
+      # pragma omp teams distribute reduction (+:result)
+      for ( int j = 0; j < N; ++j ) {
+        double temp = 0;
+        #pragma omp parallel for
+        for ( int i = 0; i < M; ++i ) {
+          temp += A( j, i ) * x( i );
+        }
+        result += y( j ) * temp;
+      }
+
+      // Output result.
+      if ( repeat == ( nrepeat - 1 ) ) {
+        printf( "  Computed result for %d x %d is %lf\n", N, M, result );
+      }
+    
+      const double solution = (double) N * (double) M;
+
[…]

### 2020-01-20 EduPar2020-2/code/for_loop.cpp

@@ -1,0 +1,37 @@
+
+#define NUMQPs 8
+
+void f()
+{
+  int element, qp, numElements = 100, NUMQPs = 8;
+  int *data_in[NUMQPs];
+
+  # pragma omp target data map (...)
+  # pragma omp teams num_teams (...) num_threads (...) private (...)
+  # pragma omp distribute
+  for ( element = 0; element < numElements ; ++ element ) {
+  int total = 0;
+  # pragma omp parallel for
+  for (qp = 0; qp < numQPs ; ++ qp)
+  total += dot ( left [ element ][ qp], right [ element ][ qp ]);
+  elementValues [ element ] = total ;
+  }
+  //Option 2: OpenACC
+  # pragma acc parallel copy (...) num_gangs (...) vector_length (...)
+  # pragma acc loop gang vector
+  for ( element = 0; element < numElements ; ++ element ) {
+  total = 0;
+  for (qp = 0; qp < numQPs ; ++ qp)
+  total += dot ( left [ element ][ qp], right [ element ][ qp ]);
+  elementValues
+
+  //Option 2: OpenCL
+  # pragma acc parallel copy (...) num_gangs (...) vector_length (...)
+  # pragma acc loop gang vector
+  for ( element = 0; element < numElements ; ++ element ) {
+  total = 0;
+  for (qp = 0; qp < numQPs ; ++ qp)
+  total += dot ( left [ element ][ qp], right [ element ][ qp ]);
+  elementValues
+
+}

### 2020-01-20 EduPar2020/Test_App/src/helper.cpp

@@ -1,0 +1,95 @@
+#include <limits>
+#include <cstdio>
+#include <cstdlib>
+#include <cmath>
+#include <cstring>
+#include <sys/time.h>
+
+#include "helper.hpp"
+
+void checkSizes( int &N, int &M, int &S, int &nrepeat );
+
+Args parse_args(int argc, char* argv[])
+{
+  Args args;
+  args.N = -1;         // number of rows 2^12
+  args.M = -1;         // number of columns 2^10
+  args.S = -1;         // total size 2^22
+  args.nrepeat = 100;  // number of repeats of the test
+
+  // Read command line arguments.
+  for ( int i = 0; i < argc; i++ ) {
+    if ( ( strcmp( argv[ i ], "-N" ) == 0 ) || ( strcmp( argv[ i ], "-Rows" ) == 0 ) ) {
+      args.N = pow( 2, atoi( argv[ ++i ] ) );
+      printf( "  User N is %d\n", args.N );
+    }
+    else if ( ( strcmp( argv[ i ], "-M" ) == 0 ) || ( strcmp( argv[ i ], "-Columns" ) == 0 ) ) {
+      args.M = pow( 2, atof( argv[ ++i ] ) );
+      printf( "  User M is %d\n", args.M );
+    }
+    else if ( ( strcmp( argv[ i ], "-S" ) == 0 ) || ( strcmp( argv[ i ], "-Size" ) == 0 ) ) {
+      args.S = pow( 2, atof( argv[ ++i ] ) );
+      printf( "  User S is %d\n", args.S );
+    }
+    else if ( strcmp( argv[ i ], "-nrepeat" ) == 0 ) {
+      args.nrepeat = atoi( argv[ ++i ] );
+    }
+    else if ( ( strcmp( argv[ i ], "-h" ) == 0 ) || ( strcmp( argv[ i ], "-help" ) == 0 ) ) {
+      printf( "  y^T*A*x Options:\n" );
+      printf( "  -Rows (-N) <int>:      exponent num, determines number of rows 2^num (default: 2^12 = 4096)\n" );
+      printf( "  -Columns (-M) <int>:   exponent num, determines number of columns 2^num (default: 2^10 = 1024)\n" );
+      printf( "  -Size (-S) <int>:      exponent num, determines total matrix size 2^num (default: 2^22 = 4096*1024 )\n" );
[…]

### 2020-01-20 EduPar2020/Test_App/src/matvec_openacc.cpp

@@ -1,0 +1,92 @@
+#include <Kokkos_Core.hpp>
+#include <assert.h>
+#include "helper.hpp"
+
+
+template <class T>
+struct vec_t
+{
+  T * v;
+  vec_t(size_t m){v = (T*) malloc (m * sizeof(T)); assert(v!=NULL);}
+  T & operator () (const size_t i) {return v[i];}
+};
+
+template <class T>
+struct mtrx_t
+{
+  T * v;
+  size_t dim;
+  mtrx_t(size_t n, size_t m) {v = (T*) malloc (n * m * sizeof(T)); dim = m; assert(v!=NULL); }
+  T & operator () (const size_t i, const size_t j) {return v[j*dim+i];}
+};
+
+double run(int argc, char* argv[], Args args)
+{
+  int N = args.N;
+  int M = args.M;
+  int nrepeat = args.nrepeat;
+  double time;
+  {
+    typedef vec_t<double> vec_double_t;
+    typedef mtrx_t<double> mtrx_double_t;
+
+    vec_double_t y( N );
+    vec_double_t x( M );
+    mtrx_double_t A( N, M );
+
+
+  
+    // Initialize y vector on host.
+    for ( int i = 0; i < N; ++i ) {
+      y( i ) = 1;
+    }
+
+    // Initialize x vector on host.
+    for ( int i = 0; i < M; ++i ) {
+      x( i ) = 1;
+    }
+
+    // Initialize A matrix on host.
+    for ( int j = 0; j < N; ++j ) {
+      for ( int i = 0; i < M; ++i ) {
+        A( j, i ) = 1;
+      }
+    }
+
+    #pragma acc data copy(y, x, A, y.v[0:N], x.v[0:M], A.v[0:N*M])
+    {
+
+      // Timer products.
+      Kokkos::Timer timer;
+
+      for ( int repeat = 0; repeat < nrepeat; repeat++ ) {
+        // Application: <y,Ax> = y^T*A*x
+        double result = 0;
+      
+        #pragma acc kernels loop independent reduction(+:result)
+        for ( int j = 0; j < N; ++j ) {
+          double temp = 0;
+          #pragma parallel loop
+          for ( int i = 0; i < M; ++i ) {
+            temp += A( j, i ) * x( i );
+          }
+          result += y( j ) * temp;
+        }
+
+        // Output result.
+        if ( repeat == ( nrepeat - 1 ) ) {
+          printf( "  Computed result for %d x %d is %lf\n", N, M, result );
+        }
+      
[…]

### 2020-01-20 EduPar2020/Test_App/src/matvec_serial.cpp

@@ -1,0 +1,83 @@
+#include <Kokkos_Core.hpp>
+#include <assert.h>
+#include "helper.hpp"
+
+
+template <class T>
+struct vec_t
+{
+  T * v;
+  vec_t(size_t m){v = (T*) malloc (m * sizeof(T)); assert(v!=NULL);}
+  T & operator () (const size_t i) {return v[i];}
+};
+
+template <class T>
+struct mtrx_t
+{
+  T * v;
+  size_t dim;
+  mtrx_t(size_t n, size_t m) {v = (T*) malloc (n * m * sizeof(T)); dim = m; assert(v!=NULL); }
+  T & operator () (const size_t i, const size_t j) {return v[j*dim+i];}
+};
+
+double run(int argc, char* argv[], Args args)
+{
+  int N = args.N;
+  int M = args.M;
+  int nrepeat = args.nrepeat;
+  double time;
+  {
+    typedef vec_t<double> vec_double_t;
+    typedef mtrx_t<double> mtrx_double_t;
+
+    vec_double_t y( N );
+    vec_double_t x( M );
+    mtrx_double_t A( N, M );
+
+        // Initialize y vector on host.
+    for ( int i = 0; i < N; ++i ) {
+      y( i ) = 1;
+    }
+
+    // Initialize x vector on host.
+    for ( int i = 0; i < M; ++i ) {
+      x( i ) = 1;
+    }
+
+    // Initialize A matrix on host.
+    for ( int j = 0; j < N; ++j ) {
+      for ( int i = 0; i < M; ++i ) {
+        A( j, i ) = 1;
+      }
+    }
+
+    // Timer products.
+    Kokkos::Timer timer;
+
+    for ( int repeat = 0; repeat < nrepeat; repeat++ ) {
+      // Application: <y,Ax> = y^T*A*x
+      double result = 0;
+      for ( int j = 0; j < N; ++j ) {
+        double temp = 0;
+        for ( int i = 0; i < M; ++i ) {
+          temp += A( j, i ) * x( i );
+        }
+        result += y( j ) * temp;
+      }
+
+      // Output result.
+      if ( repeat == ( nrepeat - 1 ) ) {
+        printf( "  Computed result for %d x %d is %lf\n", N, M, result );
+      }
+    
+      const double solution = (double) N * (double) M;
+
+      if ( result != solution ) {
+        printf( "  Error: result( %lf ) != solution( %lf )\n", result, solution );
+      }
+    }
[…]

### 2020-01-20 EduPar2020-2/Test_App/src/helper.cpp
Removed old dirs.

@@ -1,95 +1,0 @@
-#include <limits>
-#include <cstdio>
-#include <cstdlib>
-#include <cmath>
-#include <cstring>
-#include <sys/time.h>
-
-#include "helper.hpp"
-
-void checkSizes( int &N, int &M, int &S, int &nrepeat );
-
-Args parse_args(int argc, char* argv[])
-{
-  Args args;
-  args.N = -1;         // number of rows 2^12
-  args.M = -1;         // number of columns 2^10
-  args.S = -1;         // total size 2^22
-  args.nrepeat = 100;  // number of repeats of the test
-
-  // Read command line arguments.
-  for ( int i = 0; i < argc; i++ ) {
-    if ( ( strcmp( argv[ i ], "-N" ) == 0 ) || ( strcmp( argv[ i ], "-Rows" ) == 0 ) ) {
-      args.N = pow( 2, atoi( argv[ ++i ] ) );
-      printf( "  User N is %d\n", args.N );
-    }
-    else if ( ( strcmp( argv[ i ], "-M" ) == 0 ) || ( strcmp( argv[ i ], "-Columns" ) == 0 ) ) {
-      args.M = pow( 2, atof( argv[ ++i ] ) );
-      printf( "  User M is %d\n", args.M );
-    }
-    else if ( ( strcmp( argv[ i ], "-S" ) == 0 ) || ( strcmp( argv[ i ], "-Size" ) == 0 ) ) {
-      args.S = pow( 2, atof( argv[ ++i ] ) );
-      printf( "  User S is %d\n", args.S );
-    }
-    else if ( strcmp( argv[ i ], "-nrepeat" ) == 0 ) {
-      args.nrepeat = atoi( argv[ ++i ] );
-    }
-    else if ( ( strcmp( argv[ i ], "-h" ) == 0 ) || ( strcmp( argv[ i ], "-help" ) == 0 ) ) {
-      printf( "  y^T*A*x Options:\n" );
-      printf( "  -Rows (-N) <int>:      exponent num, determines number of rows 2^num (default: 2^12 = 4096)\n" );
-      printf( "  -Columns (-M) <int>:   exponent num, determines number of columns 2^num (default: 2^10 = 1024)\n" );
-      printf( "  -Size (-S) <int>:      exponent num, determines total matrix size 2^num (default: 2^22 = 4096*1024 )\n" );
[…]

### 2020-01-20 EduPar2020-2/Test_App/src/helper.hpp

@@ -1,6 +1,0 @@
-#ifndef HELPER_HPP
-#define HELPER_HPP
-typedef struct  {
-  int N, M, S, nrepeat;
-} Args;
-#endif //HELPER_HPP

### 2020-01-20 EduPar2020-2/Test_App/src/main.cpp

@@ -1,36 +1,0 @@
-#include <limits>
-#include <cstdio>
-#include <cstdlib>
-#include <cstring>
-#include <sys/time.h>
-
-#include "helper.hpp"
-
-Args parse_args(int argc, char* argv[]);
-
-void checkSizes( int &N, int &M, int &S, int &nrepeat );
-double run(int argc, char* argv[], Args args);
-
-int main( int argc, char* argv[] )
-{
-  Args args  = parse_args (argc, argv);
-  
-  int N = args.N;
-  int M = args.M;
-  int nrepeat = args.nrepeat;
-
-  double time = run(argc, argv, args);
-
-  // Calculate bandwidth.
-  // Each matrix A row (each of length M) is read once.
-  // The x vector (of length M) is read N times.
-  // The y vector (of length N) is read once.
-  // double Gbytes = 1.0e-9 * double( sizeof(double) * ( 2 * M * N + N ) );
-  double Gbytes = 1.0e-9 * double( sizeof(double) * ( M + M * N + N ) );
-
-  // Print results (problem size, time and bandwidth in GB/s).
-  printf( "  N( %d ) M( %d ) nrepeat ( %d ) problem( %g MB ) time( %g s ) bandwidth( %g GB/s )\n",
-          N, M, nrepeat, Gbytes * 1000, time, Gbytes * nrepeat / time );
-
-  return 0;
-}

### 2020-01-20 EduPar2020-2/Test_App/src/matvec_kokkos.cpp

@@ -1,93 +1,0 @@
-#include <Kokkos_Core.hpp>
-#include "helper.hpp"
-
-double run(int argc, char* argv[], Args args)
-{
-  int N = args.N;
-  int M = args.M;
-  int nrepeat = args.nrepeat;
-  double time;
-
-  Kokkos::initialize( argc, argv );
-  {
-    // typedef Kokkos::Serial   ExecSpace;
-    // typedef Kokkos::Threads  ExecSpace;
-    // typedef Kokkos::OpenMP   ExecSpace;
-    typedef Kokkos::Cuda     ExecSpace;
-
-    // typedef Kokkos::HostSpace     MemSpace;
-    // typedef Kokkos::OpenMP        MemSpace;
-    typedef Kokkos::CudaSpace     MemSpace;
-    // typedef Kokkos::CudaUVMSpace  MemSpace;
-
-    typedef Kokkos::LayoutLeft   Layout;
-    //typedef Kokkos::LayoutRight  Layout;
-    typedef Kokkos::RangePolicy<ExecSpace>  range_policy;
-
-    // Allocate y, x vectors and Matrix A on device.
-    typedef Kokkos::View<double*, Layout, MemSpace>   ViewVectorType;
-    typedef Kokkos::View<double**, Layout, MemSpace>  ViewMatrixType;
-    ViewVectorType y( "y", N );
-    ViewVectorType x( "x", M );
-    ViewMatrixType A( "A", N, M );
-
-    // Create host mirrors of device views.
-    ViewVectorType::HostMirror h_y = Kokkos::create_mirror_view( y );
-    ViewVectorType::HostMirror h_x = Kokkos::create_mirror_view( x );
-    ViewMatrixType::HostMirror h_A = Kokkos::create_mirror_view( A );
-
-    // Initialize y vector on host.
-    for ( int i = 0; i < N; ++i ) {
-      h_y( i ) = 1;
-    }
-
-    // Initialize x vector on host.
-    for ( int i = 0; i < M; ++i ) {
-      h_x( i ) = 1;
-    }
-
-    // Initialize A matrix on host.
-    for ( int j = 0; j < N; ++j ) {
-      for ( int i = 0; i < M; ++i ) {
-        h_A( j, i ) = 1;
-      }
-    }
-
-    // Deep copy host views to device views.
-    Kokkos::deep_copy( y, h_y );
-    Kokkos::deep_copy( x, h_x );
[…]

### 2020-01-20 EduPar2020-2/Test_App/src/matvec_openacc.cpp

@@ -1,92 +1,0 @@
-#include <Kokkos_Core.hpp>
-#include <assert.h>
-#include "helper.hpp"
-
-
-template <class T>
-struct vec_t
-{
-  T * v;
-  vec_t(size_t m){v = (T*) malloc (m * sizeof(T)); assert(v!=NULL);}
-  T & operator () (const size_t i) {return v[i];}
-};
-
-template <class T>
-struct mtrx_t
-{
-  T * v;
-  size_t dim;
-  mtrx_t(size_t n, size_t m) {v = (T*) malloc (n * m * sizeof(T)); dim = m; assert(v!=NULL); }
-  T & operator () (const size_t i, const size_t j) {return v[j*dim+i];}
-};
-
-double run(int argc, char* argv[], Args args)
-{
-  int N = args.N;
-  int M = args.M;
-  int nrepeat = args.nrepeat;
-  double time;
-  {
-    typedef vec_t<double> vec_double_t;
-    typedef mtrx_t<double> mtrx_double_t;
-
-    vec_double_t y( N );
-    vec_double_t x( M );
-    mtrx_double_t A( N, M );
-
-
-  
-    // Initialize y vector on host.
-    for ( int i = 0; i < N; ++i ) {
-      y( i ) = 1;
-    }
-
-    // Initialize x vector on host.
-    for ( int i = 0; i < M; ++i ) {
-      x( i ) = 1;
-    }
-
-    // Initialize A matrix on host.
-    for ( int j = 0; j < N; ++j ) {
-      for ( int i = 0; i < M; ++i ) {
-        A( j, i ) = 1;
-      }
-    }
-
-    #pragma acc data copy(y, x, A, y.v[0:N], x.v[0:M], A.v[0:N*M])
-    {
-
-      // Timer products.
-      Kokkos::Timer timer;
-
-      for ( int repeat = 0; repeat < nrepeat; repeat++ ) {
-        // Application: <y,Ax> = y^T*A*x
-        double result = 0;
-      
-        #pragma acc kernels loop independent reduction(+:result)
-        for ( int j = 0; j < N; ++j ) {
-          double temp = 0;
-          #pragma parallel loop
-          for ( int i = 0; i < M; ++i ) {
-            temp += A( j, i ) * x( i );
-          }
-          result += y( j ) * temp;
-        }
-
-        // Output result.
-        if ( repeat == ( nrepeat - 1 ) ) {
-          printf( "  Computed result for %d x %d is %lf\n", N, M, result );
-        }
-      
[…]

### 2020-01-20 EduPar2020-2/Test_App/src/matvec_openmp_device.cpp

@@ -1,86 +1,0 @@
-#include <Kokkos_Core.hpp>
-#include <assert.h>
-#include "helper.hpp"
-
-
-template <class T>
-struct vec_t
-{
-  T * v;
-  vec_t(size_t m){v = (T*) malloc (m * sizeof(T)); assert(v!=NULL);}
-  T & operator () (const size_t i) {return v[i];}
-};
-
-template <class T>
-struct mtrx_t
-{
-  T * v;
-  size_t dim;
-  mtrx_t(size_t n, size_t m) {v = (T*) malloc (n * m * sizeof(T)); dim = m; assert(v!=NULL); }
-  T & operator () (const size_t i, const size_t j) {return v[j*dim+i];}
-};
-
-double run(int argc, char* argv[], Args args)
-{
-  int N = args.N;
-  int M = args.M;
-  int nrepeat = args.nrepeat;
-  double time;
-  {
-    typedef vec_t<double> vec_double_t;
-    typedef mtrx_t<double> mtrx_double_t;
-
-    vec_double_t y( N );
-    vec_double_t x( M );
-    mtrx_double_t A( N, M );
-
-        // Initialize y vector on host.
-    for ( int i = 0; i < N; ++i ) {
-      y( i ) = 1;
-    }
-
-    // Initialize x vector on host.
-    for ( int i = 0; i < M; ++i ) {
-      x( i ) = 1;
-    }
-
-    // Initialize A matrix on host.
-    for ( int j = 0; j < N; ++j ) {
-      for ( int i = 0; i < M; ++i ) {
-        A( j, i ) = 1;
-      }
-    }
-
-    // Timer products.
-    Kokkos::Timer timer;
-
-    for ( int repeat = 0; repeat < nrepeat; repeat++ ) {
-      // Application: <y,Ax> = y^T*A*x
-      double result = 0;
-      # pragma omp target data map (A, x, y, A.v, x.v, y.v)
-      # pragma omp teams distribute reduction (+:result)
-      for ( int j = 0; j < N; ++j ) {
-        double temp = 0;
-        #pragma omp parallel for
-        for ( int i = 0; i < M; ++i ) {
-          temp += A( j, i ) * x( i );
-        }
-        result += y( j ) * temp;
-      }
-
-      // Output result.
-      if ( repeat == ( nrepeat - 1 ) ) {
-        printf( "  Computed result for %d x %d is %lf\n", N, M, result );
-      }
-    
-      const double solution = (double) N * (double) M;
-
[…]

### 2020-01-20 EduPar2020-2/Test_App/src/matvec_serial.cpp

@@ -1,83 +1,0 @@
-#include <Kokkos_Core.hpp>
-#include <assert.h>
-#include "helper.hpp"
-
-
-template <class T>
-struct vec_t
-{
-  T * v;
-  vec_t(size_t m){v = (T*) malloc (m * sizeof(T)); assert(v!=NULL);}
-  T & operator () (const size_t i) {return v[i];}
-};
-
-template <class T>
-struct mtrx_t
-{
-  T * v;
-  size_t dim;
-  mtrx_t(size_t n, size_t m) {v = (T*) malloc (n * m * sizeof(T)); dim = m; assert(v!=NULL); }
-  T & operator () (const size_t i, const size_t j) {return v[j*dim+i];}
-};
-
-double run(int argc, char* argv[], Args args)
-{
-  int N = args.N;
-  int M = args.M;
-  int nrepeat = args.nrepeat;
-  double time;
-  {
-    typedef vec_t<double> vec_double_t;
-    typedef mtrx_t<double> mtrx_double_t;
-
-    vec_double_t y( N );
-    vec_double_t x( M );
-    mtrx_double_t A( N, M );
-
-        // Initialize y vector on host.
-    for ( int i = 0; i < N; ++i ) {
-      y( i ) = 1;
-    }
-
-    // Initialize x vector on host.
-    for ( int i = 0; i < M; ++i ) {
-      x( i ) = 1;
-    }
-
-    // Initialize A matrix on host.
-    for ( int j = 0; j < N; ++j ) {
-      for ( int i = 0; i < M; ++i ) {
-        A( j, i ) = 1;
-      }
-    }
-
-    // Timer products.
-    Kokkos::Timer timer;
-
-    for ( int repeat = 0; repeat < nrepeat; repeat++ ) {
-      // Application: <y,Ax> = y^T*A*x
-      double result = 0;
-      for ( int j = 0; j < N; ++j ) {
-        double temp = 0;
-        for ( int i = 0; i < M; ++i ) {
-          temp += A( j, i ) * x( i );
-        }
-        result += y( j ) * temp;
-      }
-
-      // Output result.
-      if ( repeat == ( nrepeat - 1 ) ) {
-        printf( "  Computed result for %d x %d is %lf\n", N, M, result );
-      }
-    
-      const double solution = (double) N * (double) M;
-
-      if ( result != solution ) {
-        printf( "  Error: result( %lf ) != solution( %lf )\n", result, solution );
-      }
-    }
[…]

### 2020-01-20 EduPar2020-2/code/for_loop.cpp

@@ -1,37 +1,0 @@
-
-#define NUMQPs 8
-
-void f()
-{
-  int element, qp, numElements = 100, NUMQPs = 8;
-  int *data_in[NUMQPs];
-
-  # pragma omp target data map (...)
-  # pragma omp teams num_teams (...) num_threads (...) private (...)
-  # pragma omp distribute
-  for ( element = 0; element < numElements ; ++ element ) {
-  int total = 0;
-  # pragma omp parallel for
-  for (qp = 0; qp < numQPs ; ++ qp)
-  total += dot ( left [ element ][ qp], right [ element ][ qp ]);
-  elementValues [ element ] = total ;
-  }
-  //Option 2: OpenACC
-  # pragma acc parallel copy (...) num_gangs (...) vector_length (...)
-  # pragma acc loop gang vector
-  for ( element = 0; element < numElements ; ++ element ) {
-  total = 0;
-  for (qp = 0; qp < numQPs ; ++ qp)
-  total += dot ( left [ element ][ qp], right [ element ][ qp ]);
-  elementValues
-
-  //Option 2: OpenCL
-  # pragma acc parallel copy (...) num_gangs (...) vector_length (...)
-  # pragma acc loop gang vector
-  for ( element = 0; element < numElements ; ++ element ) {
-  total = 0;
-  for (qp = 0; qp < numQPs ; ++ qp)
-  total += dot ( left [ element ][ qp], right [ element ][ qp ]);
-  elementValues
-
-}

### 2020-01-20 main.tex
Places all files into root.

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{fancyvrb} 
\usepackage{subcaption}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

\title{Towards Performance-portable Parallel Programming in Computer Science Education}

\author{\IEEEauthorblockN{1\textsuperscript{st} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{3\textsuperscript{rd} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
}

\maketitle

\input{src/0.Abstract}

\begin{IEEEkeywords}
Parallel programming, accelerator programming, C++
\end{IEEEkeywords}

\input{src/1.Introduction} %label: chap:introduction
\input{src/2.Kokkos}       %label: chap:kokkos
\input{src/3.Exercises}    %label: chap:tutorial
\input{src/4.C++}          %label: chap:c++
\input{src/5.Related}      %label: chap:related
\input{src/6.Conclusion}   %label: chap:conclusion

\begin{thebibliography}{00}
\bibitem{b1} G. Eason, B. Noble, and I. N. Sneddon, ``On certain integrals of Lipschitz-Hankel type involving products of Bessel functions,'' Phil. Trans. Roy. Soc. London, vol. A247, pp. 529--551, April 1955.
\end{thebibliography}

\end{document}

### 2020-01-20 src/0.Abstract.tex
Places all files into root.

\begin{abstract}\label{chap:abstract}
Kokkos is a generic, performance-portable parallel programming model for host- and accelerator architectures in C++. It consists of a pure C++ interface, a specification and a programming library. The programming library exposes patterns and types and maps them to an underlying abstract machine model. 
This machine model offers a generic view of parallel hardware by abstracting from architecture- and vendor-specific programming models. In academia, this can contribute to a generic eduction of performance-portable parallel programming. In this work we present Kokkos from the perspective of educators, highlight key concepts and give an instruction to examples suited for incremental teaching of parallel programming in class.
\end{abstract}

%is aligned with C++ standardization efforts and trends towards parallel programming support in base languages.

### 2020-01-20 src/2.Kokkos.tex
Places all files into root.

\section{Towards Generic Parallel Programming}\label{chap:background}

A programming model is an abstract representation of a computer. It defines a paradigm and a programming language. The programming language is used by the programmer to write source code, a human readable, textual representation of instructions and data. Source code is a sequential listing of statements evaluated from the top to bottom, one line at a time. While such ordering seems natural to describe an algorithms as it corresponds to sequential processing (both, reading code and progressing the program counter follow a time line), it does not include the notion of concurrency.

Parallel programming models add the notion of concurrency to a sequence of instruction of a program. This is commonly achieved either by adding semantic value to the base programming language, often referred to as programming with \emph{pragma} annotations, through library calls or new languages. All three approaches are conceptually suitable to represent programming paradigms such as tasking, parallel patterns and others. Well known representatives are the OpenMP programming model~\cite{CITEOPENMP} or threading libraries respectively. In general it is up to the parallel programming model to expose language constructs or library interfaces (APIs) such that they are convenient to the programmer and capture enough semantic information to allow the programming model to map the program execution to the parallel computer hardware efficiently and correctly. Questions of which semantic information is needed, what paradigm to present and by what syntactical means span a design space worth taking a closer look at. Figure~\ref{figSemCapture} shows the three areas that constitute a parallel programming model and its design. We call it the semantic capture.

\begin{figure}[h]
\begin{Verbatim}[frame=leftline]
Semantic information
Paradigm 
Language
\end{Verbatim}
\caption{Semantic capture - a triplet of language, paradigm and semantic information defined by a parallel programming model allow the programmer to express concurrency and the compiler and runtime system to understand and run applications on a parallel machine.}
\label{figSemCapture}
\end{figure}

Semantic information provided by the programmer to the parallel programming model can be grouped into information to express intent and properties. They address the question of \emph{what}, \emph{where} and \emph{how}, which, in the context of parallel programming, corresponds to defining which code portion to parallelize, where to run and access and how to run that parallel code. Synchronization primitives may be considered as part of the what and information on the execution properties such as data placement or memory access type as the how. Semantics are expressed following a paradigm and a programming language.

The parallel programming paradigm is an abstract ~\emph{representation} with the purpose of facilitating the programmer's understanding of programming rules and program behavior. Which programming paradigm to chose depends on several considerations. Parallel-patterns allow to express concurrency for commonly occurring programming patterns such as loop constructs. Tasking is a paradigm that support the expression of concurrent loops as well as irregular algorithms. Distributed and correctness-oriented programming models may implement actor-based programming. In this programming model, each unit of execution represent an actor who communicates over predefined communication interfaces. This paradigm eliminates accesses to shared state and align well with message passing programming (MPI). The execution model and memory model of a programming model defines behavior. That is, it defines the relationship between abstract concepts and program behavior on the given architecture.

Lastly, a language defines the syntax of the parallel semantic. Declarative languages add semantic information to the base language through pragma annotations while imperative languages rely on programming interfaces to capture information. Their semantic is defined in the API specification. Figures~\ref{figOMPLike} and~\ref{figKokkosLike} show examples of two sample applications using the aforementioned parallel programming language types. While both programming models implement the same programming paradigm (parallel patterns), Figures~\ref{figOMPLike} extends the base programming language through pragma annotations. Both, pragma annotations and an API call as shown in Figure~\ref{figKokkosLike} have a similar semantic of describing a parallel loop construct. It is up to the programming model specification to define the execution and memory model in both cases.

Both examples provide the semantic information on what to parallelize, however, they do not provide information on how to map the parallel execution on modern parallel architectures. In order to define what further information is needed for a generic support on concurrency on modern computer architectures, it is important to define an abstract machine model first.

\begin{figure}
\begin{Verbatim}[frame=leftline]
# pragma model parallel for
for ( size_t i = 0; i < N; ++i) {
 /* loop body */
}
\end{Verbatim}
\caption{Declarative parallel programming provides uses pragma annotations to capture semantic information.}
\label{figOMPLike}
\end{figure}

\begin{figure}
\begin{Verbatim}[frame=leftline]
model::parallel_for (N, [=] ( const size_t i) {
  /* loop body */
});
\end{Verbatim}
\caption{Imperative parallel programming uses a base language and relies on programming interfaces to capture information and a documentation that describe their semantics.}
\label{figKokkosLike}
\end{figure}

\section{The Kokkos Abstract Machine Model}\label{chap:kokkosMM}

The Kokkos machine model defines abstractions that represent hardware capabilities for processing and data access. The Kokkos parallel programming model exposes these abstractions to the programmer through C++ and the Kokkos API. By abstracting from physical hardware, the machine model ensures that applications written in programming models targeting this machine model are generic and performance-portable on current and future hardware. The Kokkos parallel programming model is one particular instance of a programming model that builds on top of that machine model. We would like to point out that it is this conceptual differentiation that allows scenarios where the underlying machine model is instantiated in other languages beyond C++, yet the algorithmic specification as well as performance characteristic and portability remain the same.

The Kokkos abstract machine model targets a design of a future shared-memory computing architectures. The design is shown in Figure~\ref{fig:execspace}. It is characterized by multiple latency-optimized execution units and off-die bandwidth-optimized accelerators. Both compute device types have disjoint memory address spaces and unique performance properties. In such an architecture,   execution units might be grouped into larger units with multiple levels of achievable parallelism with different memory access characteristics and coherence properties of caches. In order to ensure performance portability on such a wide range of configurations, an abstraction of the compute engines and available memories are required. For this purpose we introduce \emph{spaces}.

An instance of an \emph{execution space} is a specific instantiation of an execution space to which a programmer can target parallel work. For example, an execution space can be used to describe a multi-core processor. In this case, the execution space contains several homogeneous cores organized into arbitrary logical groups. In a parallel programming model that implements this machine model, an instance of such an execution space would be made available to the programmer to run kernels. Adding more logical groups or accelerators simply increases the number of available execution spaces to the programmer.

Memory and memory types are exposed through \emph{memory spaces}. Each memory space provides a finite storage capacity at which data structures can be allocated and accessed. Different memory space types have different characteristics with respect to accessibility from execution spaces and performance.

An instance of a memory space provides a concrete method for the application programmer to request data storage allocations. Following the architecture as discussed earlier, the multi-core processor may contain multiple memory spaces that abstract on-package memory, slower DRAM and non-volatile memories. Accelerators can provide an additional memory space through its local on-package memory. The programmer is free to decide where each data structure is allocated by requesting the corresponding memory space by instantiating the according memory space. Kokkos provides the appropriate abstraction of memory allocation and data management. We believe that an abstract representation of execution- and memory devices is a key property towards performance-portable parallel programming.

How a machine model is exposed to the programmer and what are the design considerations towards defining an appropriate semantic capture are discussed in the next section.

%Ang, J.A., et. al., Abstract Machine Models and Proxy Architectures for Exascale Computing, 2014, Sandia National Laboratories and Lawrence Berkeley National Laboratory, DOE Computer Architecture Laboratories Project

\begin{figure}
\begin{subfigure}[b]{0.45\textwidth}
\centering
\includegraphics[width=\textwidth]{img/ExecSpaces.png}
\caption{Execution space is an instantiation of a space that represents a processing device to which the programmer can target parallel code.}
\label{fig:execspace}
\end{subfigure}%
\hfill \break
\begin{subfigure}[b]{0.45\textwidth}

\centering
\includegraphics[width=\textwidth]{img/MemSpaces.png}
\caption{A memory space is an instantiation of a space that represents a memory location on which the programmer can allocate data.}
\label{fig:memspace}
\end{subfigure}%
\caption{Spaces represent conceptual building blocks of the abstract machine model in Kokkos.}
\label{fig:spaces}
\end{figure}

\section{The Kokkos Programming Model}\label{chap:kokkosPM}
%====================

%The Kokkos programming model defines a programming language, paradigm and semantic information. Its design has been motivated by the desire to develop a generic, performance portable programming model that would be convenient enough to the programmer and allow a runtime library to map programming concepts and execution to the underlying abstract machine model at the same time. We take a closer look as follows.
The Kokkos programming model specifies a language, programming paradigm and an API that captures intents and properties of execution. This completes the semantic capture, discussed in Chapter~\ref{chap:background}, which allows the expression of concurrency and efficient program execution on the underlying hardware. We detail all three comments as follows.

Kokkos defines a pure interface for C++ and uses C++ as the base programming language. This has multiple reasons. Firstly, application developers of scientific and high-performance computing applications at Sandia National Laboratories expressed the demand for C++ and a generic programming model in order to support their codes. Following their estimates, code porting to changing vendor APIs represents an anacceptable burden and the possibility to

Other reasons exist. From the implementation persoecte, C++ allows meta programming which is well suited to implement generic APIs and libraries. In this case, class specialization and teplating allows to compiler-generate optimized intefaces libraries for a given underlying hardware architecture. This effectively allows for compile time transformation of algorithms to allow for adaptions of varying degrees of hardware parallelism as well as of the memory hierarchy.
Consequently, while Kokkos targets C++, we believe that any programming language can be used to express the paradigms and semantic information to efficenly support the abstravt machine model.

Kokkos supports the parallel patterns and task paradigms. Parallel patters covers \emph{for}, \emph{ scan} and \emph{reduction}. This allows the expression on concurrency over iterative, for-loop computable algorithms. To cover the class of while-loop computable algorithms (irregular algorithms), Kokkos implements the tasking paradigm. Tasks encapsulate work into units that may be executed in parallel to other tasks or sections of the program.

These  concepts allow the formulation of generic algorithms and data structures which can then be mapped to different types of architectures. Effectively, they allow for compile time transformation of algorithms to allow for adaptions of varying degrees of hardware parallelism as well as of the memory hierarchy.

C++ because popular (perf critical code in industry)
Use-case (SNL)
Language features are present and first-class citizen. 
Allow to make the PM as part of the language -> type information | overloading ops
No Source2source compiler needed
Template mechanisms / type / meta info: to write abstractions 
template meta programming privde a library to implement the progamming mdoel
- pure way to suppport the machine model
- not tide to a specific implementation of underlying hardware
- No additional semantics through pragma annotations

\section{Kokkos Abstractions}

To capture semantic information, that is the~\emph{What}, the ~\emph{Where} and the ~\emph{How}, Kokkos introduces  six abstractions: \emph{execution spaces}, \emph{execution patterns}, \emph{execution policies}, \emph{memory spaces}, \emph{memory layout} and \emph{memory traits}. These abstractions specify semantic information and enable the runtime to efficiently map the program to any underlying hardware architecture represented by abstract machine model. We take a closer look as follows.

An execution space is a place where code can be executed. On current hardware architectures this correspond to accelerators and CPUs. In the future this can include Processing in Memory (PIM) modules or different core types on a heterogeneous CPU. This model allows to include computing elements In principle, execution spaces can include remote execution spaces to support e.g., the capability of sending work to a different node. Execution Spaces thus give an application developer the means to target different parts of a heterogeneous hardware architecture. This corresponds directly to the previously described machine model.

Execution patterns are the fundamental parallel algorithms in which an application has to be expressed. Examples are the parallel\_for: execute a function in undetermined order a specified amount of times,
parallel\_reduce: which combines a parallel\_for execution with a reduction operation,
parallel\_scan: which combines a parallel\_for operation with a prefix or postfix scan on output values of each operation, and
task: which executes a single function with dependencies on other functions.
Expressing an application in these patterns allows the underlying implementation or the used compiler to reason about valid transformations. For example all parallel\_*** patterns allow unspecified execution order and only promise deterministic results of the reductions themselves. This enables different mapping patterns on different hardware such as assignment of iterations to threads or vector lanes.

An execution policy determines, together with an execution pattern, how a function is executed. Some policies can be nested in others. The most simple form of execution policies are range policies. They are used to execute an operation once for each element in a range. There are no prescriptions of order of execution or concurrency, which means that it is not legal to synchronize different iterations.

Team policies are used to implement hierarchical parallelism. For that purpose Kokkos groups threads into teams. A thread team is a collection of one or more parallel "threads" of execution. Kokkos allows an arbitrary number of teams - the league size. Hardware constrains the number of threads in a team - the team size. All threads in a team are guaranteed to run concurrently.

Memory spaces are the places where data resides. They specify physical location of data as well as access characteristics. Different physical locations correspond to different device types such as high bandwidth memories, on-die scratch memories or non-volatile bulk storage. Different logical memory spaces allow for concepts such as UVM memory in the CUDA programming model, which is accessible from the host and the CUDA accelerator. Memory Spaces, similarly to execution spaces, conceptually support remote memory locations in distributed-memory scenarios. Furthermore, they encapsulate functionality such as consistency control and persistence scopes.

Layouts express the mapping from logical (algorithmic) indices to address offset for a data allocation. By adopting appropriate layouts for memory structures, an application can optimize data access patterns in a given algorithm. If an implementation provides polymorphic layouts (i.e. a data structure can be instantiated at compile or runtime with different layouts), an architecture dependent optimization can be performed.

Memory traits specify how a data structure is accessed in an algorithm. Traits express usage scenarios such as atomic access, random access and streaming loads or stores. By putting such attributes on data structures, an implementation of the programming model can insert optimal load and store operations. If a compiler implements the programming model, it could reason about the access modes and use that to inform code transformations.

These concepts allow the generic formulation of algorithms and data structures that can be mapped to the any hardware architecture that can be abstracted through the machine model.. Effectively, they allow for compile time transformation of algorithms to allow for adaption of varying degrees of hardware parallelism as well as of the memory hierarchy.

%\begin{figure}
%\begin{Verbatim}[frame=leftline]
%T main(int argc, char** argv)
%{
% agent<T> a, b;
% a.SetInput(b.GetOutput());
% a.start();
% b.start();
% agent::wait(&a);
% return a.out();
%}
%\end{Verbatim}
%\caption{The actor-based programming paradigm builds upon the idea of isolation and message-passing which is considered a contribution towards helping developers to write correct code, to express unstructured parallelism and to support the execution on distributed memory systems.}
%\label{figAgentsLike}
%\end{figure}

\begin{figure}[h]
\begin{Verbatim}[frame=leftline]
Semantic information: Patterns (intent), Spaces, Layouts, 
Policies and Traits
Paradigm: Parallel patterns and tasking
Language: C++
\end{Verbatim}
\caption{Semantic capture defining the Kokkos programming model.}
\label{figSemCapture}
\end{figure}

\begin{figure}
\begin{Verbatim}[frame=leftline]
parallel_for <ExecutionSpace<>>
parallel_scan <>
parallel_reduce <>
task <>
\end{Verbatim}
\caption{The Kokkos programming model uses template meta programming to specialize types for  exposes concurrency through parallel patterns and tasking in the C++ base language.}
\label{fig}
\end{figure}

%\begin{center}
%\begin{tabular}{ |c|c|c| } 
% \hline
% cell1 & cell2 & cell3 \\ 
% cell4 & cell5 & cell6 \\ 
% cell7 & cell8 & cell9 \\ 
% \hline
%\end{tabular}
%\end{center}

%\begin{figure}
%\begin{Verbatim}[frame=leftline]
%double f(){
% typedef Kokkos::Cuda ES;
% typedef Kokkos::CudaSpace MS;
% typedef Kokkos::LayoutLeft LL;
% typedef Kokkos::RangePolicy<ES> Range_policy;
% typedef Kokkos::View<double*, LL, MS> V
% V A( "A", N);
% V::HostMirror h_A = Kokkos::create_mirror_view( A );
% for ( int i = 0; i < N; ++i ) //Initialize on host
%   h_A( i ) = 1;
% Kokkos::deep_copy( A, h_A );
% double result = 0;
% Kokkos::parallel_reduce( "yAx", range_policy( 0, N ), 
%  KOKKOS_LAMBDA ( int i, double &update ) {
%   update += A( i );
% }, result );
% return result;
%}
%\end{Verbatim}
%\caption{Declarative parallel programming provides uses pragma annotations to capture semantic information.}
%\label{figOMPLike}
%\end{figure}

%machine model

%\begin{figure}
%\centerline{\includegraphics[width=0.4\textwidth]{img/Stack.png}}
%\caption{Stack}
%\label{fig}
%\end{figure}

%\begin{figure}
%\centerline{\includegraphics[width=0.4\textwidth]{img/Abstractions.png}}
%\caption{Building workflow}
%\label{fig}
%\end{figure}

\begin{figure}
\centerline{\includegraphics[width=0.4\textwidth]{img/Stack.png}}
\caption{Landscaoe}
\label{fig:stack}
\end{figure}

The Kokkos semantic capture des

 While design choices regarding the language and programming paradigm address requirements posed by developers in scientific computing, the set semantic information required is defined by the need to support the abstract machine model.

Figure\ref{fig:stack} shows abstraction layers in Kokkos and its fit in a parallel programming landscape.
The next Chapter shows an example application written in Kokkos and provides an insight into the runtime library.

\section{Back-end Support}\label{chap:kokkosBackend}

The important consideration here is that the method of compiling code for different execution spaces and the dispatch of kernels to instances is abstracted by the Kokkos model. This unburdens application programmers from writing algorithms in hardware specific languages.

\begin{figure}
\centerline{\includegraphics[width=0.4\textwidth]{img/Build.png}}
\caption{Building workflow}
\label{fig}
\end{figure}

### 2020-01-20 src/6.Conclusion.tex
Places all files into root.

\section{Conclusion}\label{chap:conclusion}
...
Consequently, we emphasize the importance and welcome new development and educational use of parallel programming models that align to this trend. We believe that abstracting from vendor or architecture-specific programming models not only simplifies parallel programming in science and industry but also would support educators in teaching parallel programming.
We would also like to express gratitude to educators for taking on this important role. Further materials, including a detailed description of the programming- and machine model can be found by following the reference on the Kokkos programming mode. Finnaly, we would like to mention that interested developers and student can reach the development team on Slack by following the invititation link as mentioned in the reference section.

### 2020-01-21 src/2.Kokkos.tex
Updated Chapter 4.

@@
 %====================
-
-%The Kokkos programming model defines a programming language, paradigm and semantic information.
-Its design has been motivated by the desire to develop a generic, performance portable programming model that would be convenient enough to the programmer and allow a runtime library to map programming concepts and execution to the underlying abstract machine model at the same time.
-We take a closer look as follows.
-The Kokkos programming model specifies a language, programming paradigm and an API that captures intents and properties of execution.
-This completes the semantic capture, discussed in Chapter~\ref{chap:background}, which allows the expression of concurrency and efficient program execution on the underlying hardware.
-We detail all three comments as follows.
+The Kokkos programming model specifies a language, programming paradigm and an API.
+The API exposes the abstract machine model to the developer and defines a set of pattens and types to captures intents and properties of execution.
+This completes the semantic capture, discussed in Chapter~\ref{chap:background}, We detail all three components as follows.
 
@@
 This has multiple reasons.
-Firstly, application developers of scientific and high-performance computing applications at Sandia National Laboratories expressed the demand for C++ and a generic programming model in order to support their codes.
-Following their estimates, code porting to changing vendor APIs represents an anacceptable burden and the possibility to
+Foremost, application developers of scientific and high-performance computing applications at Sandia National Laboratories expressed the demand for C++ and a generic programming model to support their codes.
+Following their estimates, porting efforts towards adapting to vendors, programming models, APIs and software releases represents a burden worth addressing.
+Further, the possibility to maintain pure C++ codes is appealing from the code development and debugging perspective.
+Lastly, from the programming model perspective, C++ offers template meta programming which is well suited to implement generic APIs and libraries.
+In this case, class specialization and templating allow for compile-type generated types and optimizations for a given hardware architecture.
+It is worth to mention that Kokkos avoids the use of pragma annotations for the sake of treating concurrency as a first-class concept in C++.
 
-Other reasons exist.
-From the implementation persoecte, C++ allows meta programming which is well suited to implement generic APIs and libraries.
-In this case, class specialization and teplating allows to compiler-generate optimized intefaces libraries for a given underlying hardware architecture.
-This effectively allows for compile time transformation of algorithms to allow for adaptions of varying degrees of hardware parallelism as well as of the memory hierarchy.
-Consequently, while Kokkos targets C++, we believe that any programming language can be used to express the paradigms and semantic information to efficenly support the abstravt machine model.
-
+Using the programming language, a parallel programming paradigm can be applied to simplify the process of of thought.
 Kokkos supports the parallel patterns and task paradigms.
@@
 Tasks encapsulate work into units that may be executed in parallel to other tasks or sections of the program.
+Equipped with a language and set of paradigms, a particular set of abstractions can be defined that exposes the abstract machine model and captures the intent of the programmer.
 
-These  concepts allow the formulation of generic algorithms and data structures which can then be mapped to different types of architectures.
-Effectively, they allow for compile time transformation of algorithms to allow for adaptions of varying degrees of hardware parallelism as well as of the memory hierarchy.
-
-C++ because popular (perf critical code in industry)
-Use-case (SNL)
-Language features are present and first-class citizen. 
-Allow to make the PM as part of the language -> type information | overloading ops
-No Source2source compiler needed
-Template mechanisms / type / meta info: to write abstractions 
-template meta programming privde a library to implement the progamming mdoel
-- pure way to suppport the machine model
-- not tide to a specific implementation of underlying hardware
-- No additional semantics through pragma annotations
-
-\section{Kokkos Abstractions}
+\subsection{Kokkos Abstractions}
 
 To capture semantic information, that is the~\emph{What}, the ~\emph{Where} and the ~\emph{How}, Kokkos introduces six abstractions: \emph{execution spaces}, \emph{execution patterns}, \emph{execution policies}, \emph{memory spaces}, \emph{memory layout} and \emph{memory traits}.
-These abstractions specify semantic information and enable the runtime to efficiently map the program to any underlying hardware architecture represented by abstract machine model.
-We take a closer look as follows.
-
-An execution space is a place where code can be executed.
-On current hardware architectures this correspond to accelerators and CPUs.
-In the future this can include Processing in Memory (PIM) modules or different core types on a heterogeneous CPU.
-This model allows to include computing elements In principle, execution spaces can include remote execution spaces to support e.g., the capability of sending work to a different node.
-Execution Spaces thus give an application developer the means to target different parts of a heterogeneous hardware architecture.
-This corresponds directly to the previously described machine model.
-
-Execution patterns are the fundamental parallel algorithms in which an application has to be expressed.
-Examples are the parallel\_for: execute a function in undetermined order a specified amount of times,
-parallel\_reduce: which combines a parallel\_for execution with a reduction operation,
-parallel\_scan: which combines a parallel\_for operation with a prefix or postfix scan on output values of each operation, and
-task: which executes a single function with dependencies on other functions.
-Expressing an application in these patterns allows the underlying implementation or the used compiler to reason about valid transformations.
-For example all parallel\_*** patterns allow unspecified execution order and only promise deterministic results of the reductions themselves.
-This enables different mapping patterns on different hardware such as assignment of iterations to threads or vector lanes.
-
-An execution policy determines, together with an execution pattern, how a function is executed.
-Some policies can be nested in others.
-The most simple form of execution policies are range policies.
-They are used to execute an operation once for each element in a range.
-There are no prescriptions of order of execution or concurrency, which means that it is not legal to synchronize different iterations.
-
-Team policies are used to implement hierarchical parallelism.
-For that purpose Kokkos groups threads into teams.
-A thread team is a collection of one or more parallel "threads" of execution.
-Kokkos allows an arbitrary number of teams - the league size.
-Hardware constrains the number of threads in a team - the team size.
-All threads in a team are guaranteed to run concurrently.
-
-Memory spaces are the places where data resides.
-They specify physical location of data as well as access characteristics.
+These abstractions specify semantic information that enable to capture the programmer's intent and the runtime to efficiently map the program to any underlying hardware architecture.
+We list them as follows:
+\begin{itemize}
+	\item  An execution space is a place where code can be executed.
+On current hardware architectures this correspond to accelerators and CPUs and can include any compute device in the future.
+This abstraction supports remote compute devices in distributed memory scenarios as remote executiuon spaces.
+	\item Execution patterns expose the parallel programming paradigm.
+Supported patterns are \emph{parallel\_for} loop that executes the loop body in any order a specified amount of times, the \emph{parallel\_reduce} which combines a parallel\_for with a reduction operation, \emph{parallel\_scan} which combines a parallel\_for operation with a prefix or postfix scan, and \emph{task} which executes a single function potentially in parallel in respect to other tasks or code sections.
+	\item Execution policy shape the iteration space of a loop pattern. A simple execution policy is a range policy. It specifies that the loop body is execution once for each element in a range. 
+	\item Memory spaces are the places where data resides.
+They specify physical locations of data as well as access characteristics.
 Different physical locations correspond to different device types such as high bandwidth memories, on-die scratch memories or non-volatile bulk storage.
 Different logical memory spaces allow for concepts such as UVM memory in the CUDA programming model, which is accessible from the host and the CUDA accelerator.
-Memory Spaces, similarly to execution spaces, conceptually support remote memory locations in distributed-memory scenarios.
+Memory spaces, similarly to execution spaces, conceptually support remote memory locations in distributed-memory scenarios.
 Furthermore, they encapsulate functionality such as consistency control and persistence scopes.
-
-Layouts express the mapping from logical (algorithmic) indices to address offset for a data allocation.
+	\item Layouts express the mapping from loop indices to address offsets.
 By adopting appropriate layouts for memory structures, an application can optimize data access patterns in a given algorithm.
 If an implementation provides polymorphic layouts (i.e.
-a data structure can be instantiated at compile or runtime with different layouts), an architecture dependent optimization can be performed.
+a data structure can be instantiated at compile or runtime with different layouts), architecture-dependent optimizations can be performed.
+	\item Memory traits specify how a data structure is accessed.
+Traits express usage scenarios such as atomic access, random access and streaming loads or stores.
+This allows the programming model to  optimize load and store operations.
 
-Memory traits specify how a data structure is accessed in an algorithm.
-Traits express usage scenarios such as atomic access, random access and streaming loads or stores.
-By putting such attributes on data structures, an implementation of the programming model can insert optimal load and store operations.
-If a compiler implements the programming model, it could reason about the access modes and use that to inform code transformations.
-
-These concepts allow the generic formulation of algorithms and data structures that can be mapped to the any hardware architecture that can be abstracted through the machine model..
-Effectively, they allow for compile time transformation of algorithms to allow for adaption of varying degrees of hardware parallelism as well as of the memory hierarchy.
-
-%\begin{figure}
-%\begin{Verbatim}[frame=leftline]
-%T main(int argc, char** argv)
-%{
-% agent<T> a, b;
-% a.SetInput(b.GetOutput());
-% a.start();
-% b.start();
-% agent::wait(&a);
-% return a.out();
-%}
-%\end{Verbatim}
-%\caption{The actor-based programming paradigm builds upon the idea of isolation and message-passing which is considered a contribution towards helping developers to write correct code, to express unstructured parallelism and to support the execution on distributed memory systems.}
-%\label{figAgentsLike}
-%\end{figure}
+\end{itemize}
+Expressing an application in these types and patterns allows the compiler and the runtime library to generate optimized transformations.
+In the case of parallel patterns, the execution is unspecified and only promise deterministic results for the scan and reduction operations.
+This gives the freedom to the underlying abstraction layers to implement different mapping patterns on different hardware such as assignment of iterations to threads or vector lanes.
 
@@
 \begin{Verbatim}[frame=leftline]
-parallel_for <ExecutionSpace<>>
-parallel_scan <>
-parallel_reduce <>
-task <>
+template <class ExecPolicy, class FunctorType>
+Kokkos::parallel_for(
+    const std::string& name, 
+    const ExecPolicy& policy, 
+    const FunctorType& functor);
 \end{Verbatim}
-\caption{The Kokkos programming model uses template meta programming to specialize types for  exposes concurrency through parallel patterns and tasking in the C++ base language.}
+\caption{The Kokkos parallel for loop class shows how template meta programming allows to specialize types.
+Its descriptive semantic offers the freedom to optimize execution on modern architectures with multiple degrees of concurreny where an Execution policy shapes the iteration space accordingly.}
 \label{fig}
@@
 
-The Kokkos semantic capture des
-
- While design choices regarding the language and programming paradigm address requirements posed by developers in scientific computing, the set semantic information required is defined by the need to support the abstract machine model.
-
 Figure\ref{fig:stack} shows abstraction layers in Kokkos and its fit in a parallel programming landscape.
 The next Chapter shows an example application written in Kokkos and provides an insight into the runtime library.
+For the interested reader, the complete programming model specification can be accessed on-line\cite{pub:KOKKOS}.
 
-\section{Back-end Support}\label{chap:kokkosBackend}
+\subsection{Back-end Support}\label{chap:kokkosBackend}
 

### 2020-01-21 src/2.Kokkos.tex
Minor change to figure.
Changes to Chapter 4.

@@
 The Kokkos programming model specifies a language, programming paradigm and an API.
-The API exposes the abstract machine model to the developer and defines a set of pattens and types to captures intents and properties of execution.
+The API exposes the abstract machine model to the developer and defines a set of pattens and types that capture intents and properties of execution.
 This completes the semantic capture, discussed in Chapter~\ref{chap:background}, We detail all three components as follows.
@@
 This has multiple reasons.
-Foremost, application developers of scientific and high-performance computing applications at Sandia National Laboratories expressed the demand for C++ and a generic programming model to support their codes.
+It is our experience that application developers of scientific and high-performance computing applications at Sandia National Laboratories express the demand for C++ and a generic parallel programming model to support their codes.
 Following their estimates, porting efforts towards adapting to vendors, programming models, APIs and software releases represents a burden worth addressing.
@@
 Tasks encapsulate work into units that may be executed in parallel to other tasks or sections of the program.
-Equipped with a language and set of paradigms, a particular set of abstractions can be defined that exposes the abstract machine model and captures the intent of the programmer.
+Equipped with a language and set of paradigms, a particular set of abstractions can be defined to expose the abstract machine model and captures the intent of the programmer.
 
@@
 This allows the programming model to  optimize load and store operations.
-
 \end{itemize}
-Expressing an application in these types and patterns allows the compiler and the runtime library to generate optimized transformations.
-In the case of parallel patterns, the execution is unspecified and only promise deterministic results for the scan and reduction operations.
-This gives the freedom to the underlying abstraction layers to implement different mapping patterns on different hardware such as assignment of iterations to threads or vector lanes.
 
@@
 \caption{Semantic capture defining the Kokkos programming model.}
-\label{figSemCapture}
+\label{fig:SemCaptureKokkos}
 \end{figure}
+
+Expressing concurrency with patterns and abstractions allows the compiler to generate optimized transformations.
+Further, a runtime library to map the execution to the parallel hardware.
+In the case of parallel patterns, the execution is unspecified and only promise deterministic results for the reduction and scan operations.
+This gives the freedom to the underlying abstraction layers to implement different mapping patterns on different hardware such as assignment of iterations to threads or vector lanes.
+We believe that these abstraction and their descriptive semantic represent key programming primitives and the way forward for native language support for parallel programming.
+Figure~\ref{fig:SemCaptureKokkos} shows the semantic capture definition of the Kokkos programming model. 
+Figure~\ref{fig:stack} shows abstraction layers in Kokkos and its fit in a parallel programming landscape.
+The next section shows an example application written in Kokkos and provides an insight into the runtime library.
+For the interested reader, the complete programming model specification can be accessed on-line\cite{pub:KOKKOS}.
 
@@
 \caption{The Kokkos parallel for loop class shows how template meta programming allows to specialize types.
-Its descriptive semantic offers the freedom to optimize execution on modern architectures with multiple degrees of concurreny where an Execution policy shapes the iteration space accordingly.}
-\label{fig}
+Its descriptive semantic offers the freedom to optimize execution on modern architectures with multiple degrees of concurrency where an Execution policy shapes the iteration space accordingly.}
+\label{fig:parallelForKokkos}
 \end{figure}
@@
 
-Figure\ref{fig:stack} shows abstraction layers in Kokkos and its fit in a parallel programming landscape.
-The next Chapter shows an example application written in Kokkos and provides an insight into the runtime library.
-For the interested reader, the complete programming model specification can be accessed on-line\cite{pub:KOKKOS}.
-
 \subsection{Back-end Support}\label{chap:kokkosBackend}

### 2020-01-21 src/1.Introduction.tex
Fix typos in Intro chapter

--- 2020-01-20 EduPar2020/src/1.Introduction.tex
@@
 We would like to compare this to the support of Von Neumann architectures in programming languages today.
-We define a abstract machine model later in this work.
+We define an abstract machine model later in this work.
 The lack of such parallel programming models motivated us to develop Kokkos.
@@
 It consists of a programming model specification and a library.
-The specification defines execution- and memory model and an application programming interface (library API).
+The specification defines execution and memory model and an application programming interface (library API).
 The runtime library exposes programming primitives, namely, patterns and types and maps them to the underlying abstract machine model.
 The machine model creates a generic view of compute hardware and memories.
-The Kokkos library mapps execution primitives to vendor-specific programming models, libraries and memory layouts.
+The Kokkos library maps execution primitives to vendor-specific programming models, libraries, and memory layouts.
 Kokkos has been successfully used in many large-scale projects at SNL, LLNL and other research centers~\cite{CITEKOKKOSUSECASES}.
@@
 \item Targets an abstract machine model
-\item Implements a parallel-patterns and tasking paradigms
+\item Implements parallel patterns and tasking paradigms
 \item Offers abstract data types for performance portability
-\item Pure C++ implementation through template meta programming
-\item Aligns with to C++ standardization efforts for parallel programming
+\item Pure C++ implementation through template metaprogramming
+\item Aligns with C++ standardization efforts for parallel programming
 \item Vendor- and architecture independent and open-source
@@
 Firstly, we would like to present our thoughts that accompanied the design and development of the Kokkos parallel programming model.
-This insights may be considered useful in parallel programming classes that introduce programming models and reason about design choices.
+These insights may be considered useful in parallel programming classes that introduce programming models and reason about design choices.
 
@@
 
-This rest of the paper is structured as follows.
-The next Chapter gives an overview on parallel programming concepts and present the notion of \emph{semantic capture}.
+The rest of this paper is structured as follows.
+The next Chapter gives an overview on parallel programming concepts and presents the notion of \emph{semantic capture}.
 Chapter~\ref{chap:kokkosMM} presents the abstract machine model used to derive required semantic information for performance portability in Kokkos.

### 2020-01-21 src/1.Introduction.tex
Typo in 1.Introduction

@@
 Type specialization through templating allows to experiment with memory layouts and investigate their performance implications.
-Further, its modular code organization and open access under the BSD license allow students to peak behind the scenes and explore the mappings of abstractions to a variety of vendor-specific implementations.
+Further, its modular code organization and open access under the BSD license allow students to peek behind the scenes and explore the mappings of abstractions to a variety of vendor-specific implementations.
 We believe that this makes Kokkos a valuable programming model for students to reason about vendor-provided programming models and to see them as particular implementations of higher abstractions in a parallel programming landscape.

### 2020-01-21 src/2.Kokkos.tex
Updated images.
Small wording updates

@@
 Source code is a sequential listing of statements evaluated from the top to bottom, one line at a time.
-While such ordering seems natural to describe an algorithms as it corresponds to sequential processing (both, reading code and progressing the program counter follow a time line), it does not include the notion of concurrency.
+While such ordering seems natural to describe an algorithm as it corresponds to sequential processing (both, reading code and progressing the program counter following a time line), it does not include the notion of concurrency.
 
@@
 Semantic information provided by the programmer to the parallel programming model can be grouped into information to express intent and properties.
-They address the question of \emph{what}, \emph{where} and \emph{how}, which, in the context of parallel programming, corresponds to defining which code portion to parallelize, where to run and access and how to run that parallel code.
+They address the question of \emph{what}, \emph{where} and \emph{how}.
+In the context of parallel programming, these correspond to defining which code portion to parallelize, where to run and access and how to run that parallel code.
 Synchronization primitives may be considered as part of the what and information on the execution properties such as data placement or memory access type as the how.
@@
 
-An instance of an \emph{execution space} is a specific instantiation of an execution space to which a programmer can target parallel work.
+An instance of an \emph{execution space} is an abstraction over an execution resource to which a programmer can target parallel work.
 For example, an execution space can be used to describe a multi-core processor.
@@
 
-%machine model
-
-%\begin{figure}
-%\centerline{\includegraphics[width=0.4\textwidth]{img/Stack.png}}
-%\caption{Stack}
-%\label{fig}
-%\end{figure}
-
-%\begin{figure}
-%\centerline{\includegraphics[width=0.4\textwidth]{img/Abstractions.png}}
-%\caption{Building workflow}
-%\label{fig}
-%\end{figure}
-
 \begin{figure}
-\centerline{\includegraphics[width=0.4\textwidth]{img/Stack.png}}
-\caption{Landscaoe}
+\centerline{\includegraphics[width=0.3\textwidth]{img/Stack.png}}
+\caption{Landscape}
 \label{fig:stack}
@@
 \begin{figure}
-\centerline{\includegraphics[width=0.4\textwidth]{img/Build.png}}
+\centerline{\includegraphics[width=0.3\textwidth]{img/Build.png}}
 \caption{Building workflow}

### 2020-01-21 src/2.Kokkos.tex
Fix typos in Kokkos chapter

--- 2020-01-21 src/2.Kokkos.tex
@@
 It defines a paradigm and a programming language.
-The programming language is used by the programmer to write source code, a human readable, textual representation of instructions and data.
+The programming language is used by the programmer to write source code, a human-readable textual representation of instructions and data.
 Source code is a sequential listing of statements evaluated from the top to bottom, one line at a time.
-While such ordering seems natural to describe an algorithms as it corresponds to sequential processing (both, reading code and progressing the program counter follow a time line), it does not include the notion of concurrency.
+While such ordering seems natural to describe an algorithm as it corresponds to sequential processing (both, reading code and progressing the program counter follow a time line), it does not include the notion of concurrency.
 
@@
 Well known representatives are the OpenMP programming model~\cite{CITEOPENMP} or threading libraries respectively.
-In general it is up to the parallel programming model to expose language constructs or library interfaces (APIs) such that they are convenient to the programmer and capture enough semantic information to allow the programming model to map the program execution to the parallel computer hardware efficiently and correctly.
+In general, it is up to the parallel programming model to expose language constructs or library interfaces (APIs) such that they are convenient to the programmer and capture enough semantic information to allow the programming model to map the program execution to the parallel computer hardware efficiently and correctly.
 Questions of which semantic information is needed, what paradigm to present and by what syntactical means span a design space worth taking a closer look at.
@@
 Further, the possibility to maintain pure C++ codes is appealing from the code development and debugging perspective.
-Lastly, from the programming model perspective, C++ offers template meta programming which is well suited to implement generic APIs and libraries.
+Lastly, from the programming model perspective, C++ offers template metaprogramming which is well suited to implement generic APIs and libraries.
 In this case, class specialization and templating allow for compile-type generated types and optimizations for a given hardware architecture.
@@
 
-Using the programming language, a parallel programming paradigm can be applied to simplify the process of of thought.
+Using the programming language, a parallel programming paradigm can be applied to simplify the process of thought.
 Kokkos supports the parallel patterns and task paradigms.
 Parallel patters covers \emph{for}, \emph{ scan} and \emph{reduction}.
-This allows the expression on concurrency over iterative, for-loop computable algorithms.
+This allows the expression of concurrency over iterative, for-loop computable algorithms.
 To cover the class of while-loop computable algorithms (irregular algorithms), Kokkos implements the tasking paradigm.
@@
 On current hardware architectures this correspond to accelerators and CPUs and can include any compute device in the future.
-This abstraction supports remote compute devices in distributed memory scenarios as remote executiuon spaces.
+This abstraction supports remote compute devices in distributed memory scenarios as remote execution spaces.
 	\item Execution patterns expose the parallel programming paradigm.
 Supported patterns are \emph{parallel\_for} loop that executes the loop body in any order a specified amount of times, the \emph{parallel\_reduce} which combines a parallel\_for with a reduction operation, \emph{parallel\_scan} which combines a parallel\_for operation with a prefix or postfix scan, and \emph{task} which executes a single function potentially in parallel in respect to other tasks or code sections.
-	\item Execution policy shape the iteration space of a loop pattern. A simple execution policy is a range policy. It specifies that the loop body is execution once for each element in a range. 
+	\item Execution policy shape the iteration space of a loop pattern. A simple execution policy is a range policy. It specifies that the loop body is executed once for each element in a range.
 	\item Memory spaces are the places where data resides.
@@
 \end{Verbatim}
-\caption{The Kokkos parallel for loop class shows how template meta programming allows to specialize types.
+\caption{The Kokkos parallel for loop class shows how template metaprogramming allows to specialize types.
 Its descriptive semantic offers the freedom to optimize execution on modern architectures with multiple degrees of concurrency where an Execution policy shapes the iteration space accordingly.}

### 2020-01-21 src/2.Kokkos.tex
UVM memory -> unified memory

@@
 Different physical locations correspond to different device types such as high bandwidth memories, on-die scratch memories or non-volatile bulk storage.
-Different logical memory spaces allow for concepts such as UVM memory in the CUDA programming model, which is accessible from the host and the CUDA accelerator.
+Different logical memory spaces allow for concepts such as memory in the CUDA programming model, which is accessible from the host and the CUDA accelerator.
 Memory spaces, similarly to execution spaces, conceptually support remote memory locations in distributed-memory scenarios.

### 2020-01-21 src/2.Kokkos.tex

@@
 It defines a paradigm and a programming language.
-The programming language is used by the programmer to write source code, a human-readable textual representation of instructions and data.
+The programming language is used by the programmer to write source code, a human readable, textual representation of instructions and data.
 Source code is a sequential listing of statements evaluated from the top to bottom, one line at a time.
-While such ordering seems natural to describe an algorithm as it corresponds to sequential processing (both, reading code and progressing the program counter follow a time line), it does not include the notion of concurrency.
+While such ordering seems natural to describe an algorithm as it corresponds to sequential processing (both, reading code and progressing the program counter following a time line), it does not include the notion of concurrency.
 

### 2020-01-22 src/1.Introduction.tex
Minor fixes.

@@
 Further, its modular code organization and open access under the BSD license allow students to peek behind the scenes and explore the mappings of abstractions to a variety of vendor-specific implementations.
-We believe that this makes Kokkos a valuable programming model for students to reason about vendor-provided programming models and to see them as particular implementations of higher abstractions in a parallel programming landscape.
+We believe that this makes Kokkos a valuable programming model for students which allows them to reason about vendor-provided programming models and to see them as particular implementations of higher abstractions in a parallel programming landscape.
 
@@
 Chapters~\ref{chap:kokkosPM} and ~\ref{chap:kokkosBackend} present the Kokkos programming model and its back-end support.
-Chapter~\ref{chap:C++} shows the alignment to C++ standardization efforts and lastly Chapter~\ref{chap:conclusion} concludes this work.
+Chapters~\ref{chap:c++} and~\ref{chap:related} show the alignment to C++ standardization efforts and dicuss related work and lastly Chapter~\ref{chap:conclusion} concludes this work.

### 2020-01-22 src/2.Kokkos.tex
Moved Kokkos backend chapter to separate file.
Minor fixes.

@@
 
-To capture semantic information, that is the~\emph{What}, the ~\emph{Where} and the ~\emph{How}, Kokkos introduces six abstractions: \emph{execution spaces}, \emph{execution patterns}, \emph{execution policies}, \emph{memory spaces}, \emph{memory layout} and \emph{memory traits}.
-These abstractions specify semantic information that enable to capture the programmer's intent and the runtime to efficiently map the program to any underlying hardware architecture.
+\begin{figure}
+\begin{Verbatim}[frame=leftline]
+template <class ExecPolicy, class FunctorType>
+Kokkos::parallel_for(
+    const std::string& name, 
+    const ExecPolicy& policy, 
+    const FunctorType& functor);
+\end{Verbatim}
+\caption{The Kokkos parallel for loop class shows how template metaprogramming allows to specialize types.
+Its descriptive semantic offers the freedom to optimize execution on modern architectures with multiple degrees of concurrency where an Execution policy shapes the iteration space accordingly.}
+\label{fig:parallelForKokkos}
+\end{figure}
+
+To capture semantic information, that is the~\emph{What}, the~\emph{Where} and the~\emph{How}, Kokkos introduces six abstractions:~\emph{execution spaces},~\emph{execution patterns},~\emph{execution policies},~\emph{memory spaces},~\emph{memory layout} and~\emph{memory traits}.
+These abstractions specify semantic information that enable to capture the programmer's intent and allow the runtime to efficiently map the program to any underlying hardware architecture.
 We list them as follows:
@@
 \end{Verbatim}
-\caption{Semantic capture defining the Kokkos programming model.}
+\caption{Semantic capture defined by the Kokkos programming model.}
 \label{fig:SemCaptureKokkos}
 \end{figure}
+
+Figure~\ref{fig:SemCaptureKokkos}
 
@@
 We believe that these abstraction and their descriptive semantic represent key programming primitives and the way forward for native language support for parallel programming.
-Figure~\ref{fig:SemCaptureKokkos} shows the semantic capture definition of the Kokkos programming model. 
+Figure~\ref{fig:SemCaptureKokkos} shows the semantic capture defined by the Kokkos programming model. 
 Figure~\ref{fig:stack} shows abstraction layers in Kokkos and its fit in a parallel programming landscape.
@@
 For the interested reader, the complete programming model specification can be accessed on-line\cite{pub:KOKKOS}.
-
-\begin{figure}
-\begin{Verbatim}[frame=leftline]
-template <class ExecPolicy, class FunctorType>
-Kokkos::parallel_for(
-    const std::string& name, 
-    const ExecPolicy& policy, 
-    const FunctorType& functor);
-\end{Verbatim}
-\caption{The Kokkos parallel for loop class shows how template metaprogramming allows to specialize types.
-Its descriptive semantic offers the freedom to optimize execution on modern architectures with multiple degrees of concurrency where an Execution policy shapes the iteration space accordingly.}
-\label{fig:parallelForKokkos}
-\end{figure}
 
@@
 \end{figure}
-
-\subsection{Back-end Support}\label{chap:kokkosBackend}
-
-The important consideration here is that the method of compiling code for different execution spaces and the dispatch of kernels to instances is abstracted by the Kokkos model.
-This unburdens application programmers from writing algorithms in hardware specific languages.
-
-\begin{figure}
-\centerline{\includegraphics[width=0.3\textwidth]{img/Build.png}}
-\caption{Building workflow}
-\label{fig}
-\end{figure}

### 2020-01-22 src/2.Kokkos.tex
Minor changes.

@@
 Both examples provide the semantic information on what to parallelize, however, they do not provide information on how to map the parallel execution on modern parallel architectures.
-In order to define what further information is needed for a generic support on concurrency on modern computer architectures, it is important to define an abstract machine model first.
+In order to define what further information is needed to express concurrency on modern computer architectures, it is important to define an abstract machine model first.
 
@@
 Accelerators can provide an additional memory space through its local on-package memory.
-The programmer is free to decide where each data structure is allocated by requesting the corresponding memory space by instantiating the according memory space.
+An architectural layout is shown in Figure~\ref{fig:memspace}.
+The programmer is free to decide where each data structure is allocated by requesting the corresponding memory space.
+Programmatically this is achieved by instantiating the according memory space.
 Kokkos provides the appropriate abstraction of memory allocation and data management.
@@
 
-Kokkos defines a pure interface for C++ and uses C++ as the base programming language.
+Kokkos defines a pure interface for C++ and uses C++ as a programming language.
 This has multiple reasons.

### 2020-01-22 src/2.Kokkos.tex
Minor changes.

@@
 \begin{figure}
-\begin{subfigure}[b]{0.45\textwidth}
+\begin{subfigure}[b]{0.41\textwidth}
 \centering
@@
 \label{fig:execspace}
-\end{subfigure}%
+\end{subfigure}
 \hfill \break
-\begin{subfigure}[b]{0.45\textwidth}
-
+\begin{subfigure}[b]{0.41\textwidth}
 \centering
@@
 \label{fig:memspace}
-\end{subfigure}%
+\end{subfigure}
 \caption{Spaces represent conceptual building blocks of the abstract machine model in Kokkos.}
@@
 \section{The Kokkos Programming Model}\label{chap:kokkosPM}
-%====================
+
 The Kokkos programming model specifies a language, programming paradigm and an API.
@@
 
-To capture semantic information, that is the~\emph{What}, the~\emph{Where} and the~\emph{How}, Kokkos introduces six abstractions:~\emph{execution spaces},~\emph{execution patterns},~\emph{execution policies},~\emph{memory spaces},~\emph{memory layout} and~\emph{memory traits}.
+To capture semantic information, that is the~\emph{what}, the~\emph{where} and the~\emph{how}, Kokkos introduces six abstractions:~\emph{execution spaces},~\emph{execution patterns},~\emph{execution policies},~\emph{memory spaces},~\emph{memory layout} and~\emph{memory traits}.
 These abstractions specify semantic information that enable to capture the programmer's intent and allow the runtime to efficiently map the program to any underlying hardware architecture.
@@
 
-Figure~\ref{fig:SemCaptureKokkos}
-
 Expressing concurrency with patterns and abstractions allows the compiler to generate optimized transformations.
@@
 We believe that these abstraction and their descriptive semantic represent key programming primitives and the way forward for native language support for parallel programming.
+
 Figure~\ref{fig:SemCaptureKokkos} shows the semantic capture defined by the Kokkos programming model. 
-Figure~\ref{fig:stack} shows abstraction layers in Kokkos and its fit in a parallel programming landscape.
+Figure~\ref{fig:stack} shows abstraction layers in Kokkos and and how it fits in a parallel programming landscape.
 The next section shows an example application written in Kokkos and provides an insight into the runtime library.

### 2020-01-22 src/2.Kokkos.tex
Commiting latest changes for review purposes.

@@
 
-\begin{figure}
-\begin{Verbatim}[frame=leftline]
-template <class ExecPolicy, class FunctorType>
-Kokkos::parallel_for(
-    const std::string& name, 
-    const ExecPolicy& policy, 
-    const FunctorType& functor);
-\end{Verbatim}
-\caption{The Kokkos parallel for loop class shows how template metaprogramming allows to specialize types.
-Its descriptive semantic offers the freedom to optimize execution on modern architectures with multiple degrees of concurrency where an Execution policy shapes the iteration space accordingly.}
-\label{fig:parallelForKokkos}
-\end{figure}
-
 To capture semantic information, that is the~\emph{what}, the~\emph{where} and the~\emph{how}, Kokkos introduces six abstractions:~\emph{execution spaces},~\emph{execution patterns},~\emph{execution policies},~\emph{memory spaces},~\emph{memory layout} and~\emph{memory traits}.
@@
 
+\begin{figure}
+\begin{Verbatim}[frame=leftline]
+template <class ExecPolicy, class FunctorType>
+Kokkos::parallel_for(const std::string& name, 
+ const ExecPolicy& policy, 
+ const FunctorType& functor);
+template <class ExecPolicy, class FunctorType, class ReducerVarNonConst>
+Kokkos::parallel_reduce(const std::string& name, 
+ const ExecPolicy& policy, 
+ const FunctorType& functor, 
+ ReducerVarNonConst& reducer);
+\end{Verbatim}
+\caption{The Kokkos~\emph{parallel_for} and \emph{parallel_reduce} class show how template metaprogramming allows to specialize types.
+Its descriptive semantic offers the freedom to optimize execution on modern architectures with multiple degrees of concurrency where an Execution policy shapes the iteration space accordingly.}
+\label{fig:parallelForAndReduceKokkos}
+\end{figure}
+
+WIP: Figure~\ref{fig:parallelForAndReduceKokkos} shows definitions of the~\emph{parallel for} and~\emph{parallel reduction} interfaces.
+Both interfaces take a set of template parameters and function arguments.
+\emph{ExecPolicy} shapes the iteration space for a particular device with varying degrees of hierarchical parallelism.
+A valid template parameter for ExecPolicy is a \emph{RangePolicy}.
+This type defines a 1-dimensional iteration range that is templated.
+This allows to consider the distribtion of iterations into  hierarchical orghanization of parallel processing units.
+In this case, ExecPolicy is   and~\emph{FunctorType} and a set of function arguments.
+
 Expressing concurrency with patterns and abstractions allows the compiler to generate optimized transformations.
@@
 For the interested reader, the complete programming model specification can be accessed on-line\cite{pub:KOKKOS}.
+
+\begin{figure}
+\centerline{\includegraphics[width=0.3\textwidth]{img/Stack.png}}
+\caption{Landscape}
+\label{fig:stack}
+\end{figure}
 
@@
 %\end{figure}
-
-\begin{figure}
-\centerline{\includegraphics[width=0.3\textwidth]{img/Stack.png}}
-\caption{Landscape}
-\label{fig:stack}
-\end{figure}

### 2020-01-22 main.tex
Updated main.tex
Slit text into multiple files.

@@
 
-\title{Towards Performance-portable Parallel Programming in Computer Science Education}
+\title{Performance-portable Parallel Programming in Computer Science Education with Kokkos and C++}
 
@@
 \begin{IEEEkeywords}
-Parallel programming, accelerator programming, C++
+Parallel programming, accelerator programming, Kokkos, C++
 \end{IEEEkeywords}
@@
 \input{src/2.Kokkos}              %label: chap:kokkos
+\input{src/2.1.KokkosBackend.tex} %label: chap:kokkosBackend
 \input{src/3.Exercises}           %label: chap:tutorial

### 2020-01-22 src/2.GenericProgramming.tex
Slit text into multiple files.

--- 2020-01-22 src/2.Kokkos.tex
@@
 \end{figure}
-
-\section{The Kokkos Abstract Machine Model}\label{chap:kokkosMM}
-
-The Kokkos machine model defines abstractions that represent hardware capabilities for processing and data access.
-The Kokkos parallel programming model exposes these abstractions to the programmer through C++ and the Kokkos API.
-By abstracting from physical hardware, the machine model ensures that applications written in programming models targeting this machine model are generic and performance-portable on current and future hardware.
-The Kokkos parallel programming model is one particular instance of a programming model that builds on top of that machine model.
-We would like to point out that it is this conceptual differentiation that allows scenarios where the underlying machine model is instantiated in other languages beyond C++, yet the algorithmic specification as well as performance characteristic and portability remain the same.
-
-The Kokkos abstract machine model targets a design of a future shared-memory computing architectures.
-The design is shown in Figure~\ref{fig:execspace}.
-It is characterized by multiple latency-optimized execution units and off-die bandwidth-optimized accelerators.
-Both compute device types have disjoint memory address spaces and unique performance properties.
-In such an architecture,   execution units might be grouped into larger units with multiple levels of achievable parallelism with different memory access characteristics and coherence properties of caches.
-In order to ensure performance portability on such a wide range of configurations, an abstraction of the compute engines and available memories are required.
-For this purpose we introduce \emph{spaces}.
-
-An instance of an \emph{execution space} is an abstraction over an execution resource to which a programmer can target parallel work.
-For example, an execution space can be used to describe a multi-core processor.
-In this case, the execution space contains several homogeneous cores organized into arbitrary logical groups.
-In a parallel programming model that implements this machine model, an instance of such an execution space would be made available to the programmer to run kernels.
-Adding more logical groups or accelerators simply increases the number of available execution spaces to the programmer.
-
-Memory and memory types are exposed through \emph{memory spaces}.
-Each memory space provides a finite storage capacity at which data structures can be allocated and accessed.
-Different memory space types have different characteristics with respect to accessibility from execution spaces and performance.
-
-An instance of a memory space provides a concrete method for the application programmer to request data storage allocations.
-Following the architecture as discussed earlier, the multi-core processor may contain multiple memory spaces that abstract on-package memory, slower DRAM and non-volatile memories.
-Accelerators can provide an additional memory space through its local on-package memory.
-An architectural layout is shown in Figure~\ref{fig:memspace}.
-The programmer is free to decide where each data structure is allocated by requesting the corresponding memory space.
-Programmatically this is achieved by instantiating the according memory space.
-Kokkos provides the appropriate abstraction of memory allocation and data management.
-We believe that an abstract representation of execution- and memory devices is a key property towards performance-portable parallel programming.
-
-How a machine model is exposed to the programmer and what are the design considerations towards defining an appropriate semantic capture are discussed in the next section.
-
-%Ang, J.A., et.
-al., Abstract Machine Models and Proxy Architectures for Exascale Computing, 2014, Sandia National Laboratories and Lawrence Berkeley National Laboratory, DOE Computer Architecture Laboratories Project
-
-\begin{figure}
-\begin{subfigure}[b]{0.41\textwidth}
-\centering
-\includegraphics[width=\textwidth]{img/ExecSpaces.png}
-\caption{Execution space is an instantiation of a space that represents a processing device to which the programmer can target parallel code.}
-\label{fig:execspace}
-\end{subfigure}
-\hfill \break
-\begin{subfigure}[b]{0.41\textwidth}
-\centering
-\includegraphics[width=\textwidth]{img/MemSpaces.png}
-\caption{A memory space is an instantiation of a space that represents a memory location on which the programmer can allocate data.}
-\label{fig:memspace}
-\end{subfigure}
-\caption{Spaces represent conceptual building blocks of the abstract machine model in Kokkos.}
-\label{fig:spaces}
-\end{figure}
-
-\section{The Kokkos Programming Model}\label{chap:kokkosPM}
-
-The Kokkos programming model specifies a language, programming paradigm and an API.
-The API exposes the abstract machine model to the developer and defines a set of pattens and types that capture intents and properties of execution.
-This completes the semantic capture, discussed in Chapter~\ref{chap:background}, We detail all three components as follows.
-
-Kokkos defines a pure interface for C++ and uses C++ as a programming language.
-This has multiple reasons.
-It is our experience that application developers of scientific and high-performance computing applications at Sandia National Laboratories express the demand for C++ and a generic parallel programming model to support their codes.
-Following their estimates, porting efforts towards adapting to vendors, programming models, APIs and software releases represents a burden worth addressing.
-Further, the possibility to maintain pure C++ codes is appealing from the code development and debugging perspective.
-Lastly, from the programming model perspective, C++ offers template metaprogramming which is well suited to implement generic APIs and libraries.
-In this case, class specialization and templating allow for compile-type generated types and optimizations for a given hardware architecture.
-It is worth to mention that Kokkos avoids the use of pragma annotations for the sake of treating concurrency as a first-class concept in C++.
-
-Using the programming language, a parallel programming paradigm can be applied to simplify the process of thought.
-Kokkos supports the parallel patterns and task paradigms.
-Parallel patters covers \emph{for}, \emph{ scan} and \emph{reduction}.
-This allows the expression of concurrency over iterative, for-loop computable algorithms.
-To cover the class of while-loop computable algorithms (irregular algorithms), Kokkos implements the tasking paradigm.
-Tasks encapsulate work into units that may be executed in parallel to other tasks or sections of the program.
-Equipped with a language and set of paradigms, a particular set of abstractions can be defined to expose the abstract machine model and captures the intent of the programmer.
-
-\subsection{Kokkos Abstractions}
-
-\begin{figure}
-\begin{Verbatim}[frame=leftline]
-template <class ExecPolicy, class FunctorType>
-Kokkos::parallel_for(
-    const std::string& name, 
-    const ExecPolicy& policy, 
-    const FunctorType& functor);
-\end{Verbatim}
-\caption{The Kokkos parallel for loop class shows how template metaprogramming allows to specialize types.
-Its descriptive semantic offers the freedom to optimize execution on modern architectures with multiple degrees of concurrency where an Execution policy shapes the iteration space accordingly.}
-\label{fig:parallelForKokkos}
-\end{figure}
-
-To capture semantic information, that is the~\emph{what}, the~\emph{where} and the~\emph{how}, Kokkos introduces six abstractions:~\emph{execution spaces},~\emph{execution patterns},~\emph{execution policies},~\emph{memory spaces},~\emph{memory layout} and~\emph{memory traits}.
-These abstractions specify semantic information that enable to capture the programmer's intent and allow the runtime to efficiently map the program to any underlying hardware architecture.
-We list them as follows:
-\begin{itemize}
-	\item  An execution space is a place where code can be executed.
-On current hardware architectures this correspond to accelerators and CPUs and can include any compute device in the future.
-This abstraction supports remote compute devices in distributed memory scenarios as remote execution spaces.
-	\item Execution patterns expose the parallel programming paradigm.
-Supported patterns are \emph{parallel\_for} loop that executes the loop body in any order a specified amount of times, the \emph{parallel\_reduce} which combines a parallel\_for with a reduction operation, \emph{parallel\_scan} which combines a parallel\_for operation with a prefix or postfix scan, and \emph{task} which executes a single function potentially in parallel in respect to other tasks or code sections.
-	\item Execution policy shape the iteration space of a loop pattern. A simple execution policy is a range policy. It specifies that the loop body is executed once for each element in a range.
-	\item Memory spaces are the places where data resides.
-They specify physical locations of data as well as access characteristics.
-Different physical locations correspond to different device types such as high bandwidth memories, on-die scratch memories or non-volatile bulk storage.
-Different logical memory spaces allow for concepts such as memory in the CUDA programming model, which is accessible from the host and the CUDA accelerator.
-Memory spaces, similarly to execution spaces, conceptually support remote memory locations in distributed-memory scenarios.
-Furthermore, they encapsulate functionality such as consistency control and persistence scopes.
-	\item Layouts express the mapping from loop indices to address offsets.
-By adopting appropriate layouts for memory structures, an application can optimize data access patterns in a given algorithm.
-If an implementation provides polymorphic layouts (i.e.
-a data structure can be instantiated at compile or runtime with different layouts), architecture-dependent optimizations can be performed.
-	\item Memory traits specify how a data structure is accessed.
-Traits express usage scenarios such as atomic access, random access and streaming loads or stores.
-This allows the programming model to  optimize load and store operations.
-\end{itemize}
-
-\begin{figure}[h]
-\begin{Verbatim}[frame=leftline]
-Semantic information: Patterns (intent), Spaces, Layouts, 
-Policies and Traits
-Paradigm: Parallel patterns and tasking
-Language: C++
-\end{Verbatim}
-\caption{Semantic capture defined by the Kokkos programming model.}
-\label{fig:SemCaptureKokkos}
-\end{figure}
-
-Expressing concurrency with patterns and abstractions allows the compiler to generate optimized transformations.
-Further, a runtime library to map the execution to the parallel hardware.
-In the case of parallel patterns, the execution is unspecified and only promise deterministic results for the reduction and scan operations.
-This gives the freedom to the underlying abstraction layers to implement different mapping patterns on different hardware such as assignment of iterations to threads or vector lanes.
-We believe that these abstraction and their descriptive semantic represent key programming primitives and the way forward for native language support for parallel programming.
-
-Figure~\ref{fig:SemCaptureKokkos} shows the semantic capture defined by the Kokkos programming model. 
-Figure~\ref{fig:stack} shows abstraction layers in Kokkos and and how it fits in a parallel programming landscape.
-The next section shows an example application written in Kokkos and provides an insight into the runtime library.
-For the interested reader, the complete programming model specification can be accessed on-line\cite{pub:KOKKOS}.
-
-%\begin{center}
-%\begin{tabular}{ |c|c|c| } 
-% \hline
-% cell1 & cell2 & cell3 \\ 
-% cell4 & cell5 & cell6 \\ 
-% cell7 & cell8 & cell9 \\ 
-% \hline
-%\end{tabular}
-%\end{center}
-
-%\begin{figure}
-%\begin{Verbatim}[frame=leftline]
-%double f(){
-% typedef Kokkos::Cuda ES;
-% typedef Kokkos::CudaSpace MS;
-% typedef Kokkos::LayoutLeft LL;
-% typedef Kokkos::RangePolicy<ES> Range_policy;
-% typedef Kokkos::View<double*, LL, MS> V
-% V A( "A", N);
-% V::HostMirror h_A = Kokkos::create_mirror_view( A );
-% for ( int i = 0; i < N; ++i ) //Initialize on host
-%   h_A( i ) = 1;
-% Kokkos::deep_copy( A, h_A );
-% double result = 0;
-% Kokkos::parallel_reduce( "yAx", range_policy( 0, N ), 
-%  KOKKOS_LAMBDA ( int i, double &update ) {
-%   update += A( i );
-% }, result );
-% return result;
-%}
-%\end{Verbatim}
-%\caption{Declarative parallel programming provides uses pragma annotations to capture semantic information.}
-%\label{figOMPLike}
-%\end{figure}
-
-\begin{figure}
-\centerline{\includegraphics[width=0.3\textwidth]{img/Stack.png}}
-\caption{Landscape}
-\label{fig:stack}
-\end{figure}

### 2020-01-22 src/4.ExecutionModel.tex
Slit text into multiple files.

--- 2020-01-22 src/2.Kokkos.tex
@@
-\section{Towards Generic Parallel Programming}\label{chap:background}
-
-A programming model is an abstract representation of a computer.
-It defines a paradigm and a programming language.
-The programming language is used by the programmer to write source code, a human readable, textual representation of instructions and data.
-Source code is a sequential listing of statements evaluated from the top to bottom, one line at a time.
-While such ordering seems natural to describe an algorithm as it corresponds to sequential processing (both, reading code and progressing the program counter following a time line), it does not include the notion of concurrency.
-
-Parallel programming models add the notion of concurrency to a sequence of instruction of a program.
-This is commonly achieved either by adding semantic value to the base programming language, often referred to as programming with \emph{pragma} annotations, through library calls or new languages.
-All three approaches are conceptually suitable to represent programming paradigms such as tasking, parallel patterns and others.
-Well known representatives are the OpenMP programming model~\cite{CITEOPENMP} or threading libraries respectively.
-In general, it is up to the parallel programming model to expose language constructs or library interfaces (APIs) such that they are convenient to the programmer and capture enough semantic information to allow the programming model to map the program execution to the parallel computer hardware efficiently and correctly.
-Questions of which semantic information is needed, what paradigm to present and by what syntactical means span a design space worth taking a closer look at.
-Figure~\ref{figSemCapture} shows the three areas that constitute a parallel programming model and its design.
-We call it the semantic capture.
-
-\begin{figure}[h]
-\begin{Verbatim}[frame=leftline]
-Semantic information
-Paradigm 
-Language
-\end{Verbatim}
-\caption{Semantic capture - a triplet of language, paradigm and semantic information defined by a parallel programming model allow the programmer to express concurrency and the compiler and runtime system to understand and run applications on a parallel machine.}
-\label{figSemCapture}
-\end{figure}
-
-Semantic information provided by the programmer to the parallel programming model can be grouped into information to express intent and properties.
-They address the question of \emph{what}, \emph{where} and \emph{how}.
-In the context of parallel programming, these correspond to defining which code portion to parallelize, where to run and access and how to run that parallel code.
-Synchronization primitives may be considered as part of the what and information on the execution properties such as data placement or memory access type as the how.
-Semantics are expressed following a paradigm and a programming language.
-
-The parallel programming paradigm is an abstract ~\emph{representation} with the purpose of facilitating the programmer's understanding of programming rules and program behavior.
-Which programming paradigm to chose depends on several considerations.
-Parallel-patterns allow to express concurrency for commonly occurring programming patterns such as loop constructs.
-Tasking is a paradigm that support the expression of concurrent loops as well as irregular algorithms.
-Distributed and correctness-oriented programming models may implement actor-based programming.
-In this programming model, each unit of execution represent an actor who communicates over predefined communication interfaces.
-This paradigm eliminates accesses to shared state and align well with message passing programming (MPI).
-The execution model and memory model of a programming model defines behavior.
-That is, it defines the relationship between abstract concepts and program behavior on the given architecture.
-
-Lastly, a language defines the syntax of the parallel semantic.
-Declarative languages add semantic information to the base language through pragma annotations while imperative languages rely on programming interfaces to capture information.
-Their semantic is defined in the API specification.
-Figures~\ref{figOMPLike} and~\ref{figKokkosLike} show examples of two sample applications using the aforementioned parallel programming language types.
-While both programming models implement the same programming paradigm (parallel patterns), Figures~\ref{figOMPLike} extends the base programming language through pragma annotations.
-Both, pragma annotations and an API call as shown in Figure~\ref{figKokkosLike} have a similar semantic of describing a parallel loop construct.
-It is up to the programming model specification to define the execution and memory model in both cases.
-
-Both examples provide the semantic information on what to parallelize, however, they do not provide information on how to map the parallel execution on modern parallel architectures.
-In order to define what further information is needed to express concurrency on modern computer architectures, it is important to define an abstract machine model first.
-
-\begin{figure}
-\begin{Verbatim}[frame=leftline]
-# pragma model parallel for
-for ( size_t i = 0; i < N; ++i) {
- /* loop body */
-}
-\end{Verbatim}
-\caption{Declarative parallel programming provides uses pragma annotations to capture semantic information.}
-\label{figOMPLike}
-\end{figure}
-
-\begin{figure}
-\begin{Verbatim}[frame=leftline]
-model::parallel_for (N, [=] ( const size_t i) {
-  /* loop body */
-});
-\end{Verbatim}
-\caption{Imperative parallel programming uses a base language and relies on programming interfaces to capture information and a documentation that describe their semantics.}
-\label{figKokkosLike}
-\end{figure}
-
-\section{The Kokkos Abstract Machine Model}\label{chap:kokkosMM}
+\section{The Kokkos Execution Model Abstractions}\label{chap:kokkosEMA}
 
@@
 It is characterized by multiple latency-optimized execution units and off-die bandwidth-optimized accelerators.
-Both compute device types have disjoint memory address spaces and unique performance properties.
-In such an architecture,   execution units might be grouped into larger units with multiple levels of achievable parallelism with different memory access characteristics and coherence properties of caches.
-In order to ensure performance portability on such a wide range of configurations, an abstraction of the compute engines and available memories are required.
-For this purpose we introduce \emph{spaces}.
+Both compute device types can have disjoint memory address spaces with unique performance properties.
+In such an architecture, execution units might be hierarchically organized with multiple levels of achievable parallelism with different memory access characteristics and coherence properties of caches.
+In order to ensure performance portability on such a wide range of configurations, an abstraction of compute resources and available memories are required.
+For this purpose we introduce the abstract concept called \emph{space}.
 
@@
 Memory and memory types are exposed through \emph{memory spaces}.
-Each memory space provides a finite storage capacity at which data structures can be allocated and accessed.
+Each memory space provides storage capacity at which data structures can be allocated and accessed.
 Different memory space types have different characteristics with respect to accessibility from execution spaces and performance.
@@
 Accelerators can provide an additional memory space through its local on-package memory.
-An architectural layout is shown in Figure~\ref{fig:memspace}.
+An example of such an architectural layout is shown in Figure~\ref{fig:memspace}.
 The programmer is free to decide where each data structure is allocated by requesting the corresponding memory space.
@@
 \end{figure}
-
-\section{The Kokkos Programming Model}\label{chap:kokkosPM}
-
-The Kokkos programming model specifies a language, programming paradigm and an API.
-The API exposes the abstract machine model to the developer and defines a set of pattens and types that capture intents and properties of execution.
-This completes the semantic capture, discussed in Chapter~\ref{chap:background}, We detail all three components as follows.
-
-Kokkos defines a pure interface for C++ and uses C++ as a programming language.
-This has multiple reasons.
-It is our experience that application developers of scientific and high-performance computing applications at Sandia National Laboratories express the demand for C++ and a generic parallel programming model to support their codes.
-Following their estimates, porting efforts towards adapting to vendors, programming models, APIs and software releases represents a burden worth addressing.
-Further, the possibility to maintain pure C++ codes is appealing from the code development and debugging perspective.
-Lastly, from the programming model perspective, C++ offers template metaprogramming which is well suited to implement generic APIs and libraries.
-In this case, class specialization and templating allow for compile-type generated types and optimizations for a given hardware architecture.
-It is worth to mention that Kokkos avoids the use of pragma annotations for the sake of treating concurrency as a first-class concept in C++.
-
-Using the programming language, a parallel programming paradigm can be applied to simplify the process of thought.
-Kokkos supports the parallel patterns and task paradigms.
-Parallel patters covers \emph{for}, \emph{ scan} and \emph{reduction}.
-This allows the expression of concurrency over iterative, for-loop computable algorithms.
-To cover the class of while-loop computable algorithms (irregular algorithms), Kokkos implements the tasking paradigm.
-Tasks encapsulate work into units that may be executed in parallel to other tasks or sections of the program.
-Equipped with a language and set of paradigms, a particular set of abstractions can be defined to expose the abstract machine model and captures the intent of the programmer.
-
-\subsection{Kokkos Abstractions}
-
-\begin{figure}
-\begin{Verbatim}[frame=leftline]
-template <class ExecPolicy, class FunctorType>
-Kokkos::parallel_for(
-    const std::string& name, 
-    const ExecPolicy& policy, 
-    const FunctorType& functor);
-\end{Verbatim}
-\caption{The Kokkos parallel for loop class shows how template metaprogramming allows to specialize types.
-Its descriptive semantic offers the freedom to optimize execution on modern architectures with multiple degrees of concurrency where an Execution policy shapes the iteration space accordingly.}
-\label{fig:parallelForKokkos}
-\end{figure}
-
-To capture semantic information, that is the~\emph{what}, the~\emph{where} and the~\emph{how}, Kokkos introduces six abstractions:~\emph{execution spaces},~\emph{execution patterns},~\emph{execution policies},~\emph{memory spaces},~\emph{memory layout} and~\emph{memory traits}.
-These abstractions specify semantic information that enable to capture the programmer's intent and allow the runtime to efficiently map the program to any underlying hardware architecture.
-We list them as follows:
-\begin{itemize}
-	\item  An execution space is a place where code can be executed.
-On current hardware architectures this correspond to accelerators and CPUs and can include any compute device in the future.
-This abstraction supports remote compute devices in distributed memory scenarios as remote execution spaces.
-	\item Execution patterns expose the parallel programming paradigm.
-Supported patterns are \emph{parallel\_for} loop that executes the loop body in any order a specified amount of times, the \emph{parallel\_reduce} which combines a parallel\_for with a reduction operation, \emph{parallel\_scan} which combines a parallel\_for operation with a prefix or postfix scan, and \emph{task} which executes a single function potentially in parallel in respect to other tasks or code sections.
-	\item Execution policy shape the iteration space of a loop pattern. A simple execution policy is a range policy. It specifies that the loop body is executed once for each element in a range.
-	\item Memory spaces are the places where data resides.
-They specify physical locations of data as well as access characteristics.
-Different physical locations correspond to different device types such as high bandwidth memories, on-die scratch memories or non-volatile bulk storage.
-Different logical memory spaces allow for concepts such as memory in the CUDA programming model, which is accessible from the host and the CUDA accelerator.
-Memory spaces, similarly to execution spaces, conceptually support remote memory locations in distributed-memory scenarios.
-Furthermore, they encapsulate functionality such as consistency control and persistence scopes.
-	\item Layouts express the mapping from loop indices to address offsets.
-By adopting appropriate layouts for memory structures, an application can optimize data access patterns in a given algorithm.
-If an implementation provides polymorphic layouts (i.e.
-a data structure can be instantiated at compile or runtime with different layouts), architecture-dependent optimizations can be performed.
-	\item Memory traits specify how a data structure is accessed.
-Traits express usage scenarios such as atomic access, random access and streaming loads or stores.
-This allows the programming model to  optimize load and store operations.
-\end{itemize}
-
-\begin{figure}[h]
-\begin{Verbatim}[frame=leftline]
-Semantic information: Patterns (intent), Spaces, Layouts, 
-Policies and Traits
-Paradigm: Parallel patterns and tasking
-Language: C++
-\end{Verbatim}
-\caption{Semantic capture defined by the Kokkos programming model.}
-\label{fig:SemCaptureKokkos}
-\end{figure}
-
-Expressing concurrency with patterns and abstractions allows the compiler to generate optimized transformations.
-Further, a runtime library to map the execution to the parallel hardware.
-In the case of parallel patterns, the execution is unspecified and only promise deterministic results for the reduction and scan operations.
-This gives the freedom to the underlying abstraction layers to implement different mapping patterns on different hardware such as assignment of iterations to threads or vector lanes.
-We believe that these abstraction and their descriptive semantic represent key programming primitives and the way forward for native language support for parallel programming.
-
-Figure~\ref{fig:SemCaptureKokkos} shows the semantic capture defined by the Kokkos programming model. 
-Figure~\ref{fig:stack} shows abstraction layers in Kokkos and and how it fits in a parallel programming landscape.
-The next section shows an example application written in Kokkos and provides an insight into the runtime library.
-For the interested reader, the complete programming model specification can be accessed on-line\cite{pub:KOKKOS}.
-
-%\begin{center}
-%\begin{tabular}{ |c|c|c| } 
-% \hline
-% cell1 & cell2 & cell3 \\ 
-% cell4 & cell5 & cell6 \\ 
-% cell7 & cell8 & cell9 \\ 
-% \hline
-%\end{tabular}
-%\end{center}
-
-%\begin{figure}
-%\begin{Verbatim}[frame=leftline]
-%double f(){
-% typedef Kokkos::Cuda ES;
-% typedef Kokkos::CudaSpace MS;
-% typedef Kokkos::LayoutLeft LL;
-% typedef Kokkos::RangePolicy<ES> Range_policy;
-% typedef Kokkos::View<double*, LL, MS> V
-% V A( "A", N);
-% V::HostMirror h_A = Kokkos::create_mirror_view( A );
-% for ( int i = 0; i < N; ++i ) //Initialize on host
-%   h_A( i ) = 1;
-% Kokkos::deep_copy( A, h_A );
-% double result = 0;
-% Kokkos::parallel_reduce( "yAx", range_policy( 0, N ), 
-%  KOKKOS_LAMBDA ( int i, double &update ) {
-%   update += A( i );
-% }, result );
-% return result;
-%}
-%\end{Verbatim}
-%\caption{Declarative parallel programming provides uses pragma annotations to capture semantic information.}
-%\label{figOMPLike}
-%\end{figure}
-
-\begin{figure}
-\centerline{\includegraphics[width=0.3\textwidth]{img/Stack.png}}
-\caption{Landscape}
-\label{fig:stack}
-\end{figure}

### 2020-01-22 src/9.Conclusion.tex
Slit text into multiple files.

\section{Conclusion}\label{chap:conclusion}
%...
%Consequently, we emphasize the importance and welcome new development and educational use of parallel programming models that align to this trend. We believe that abstracting from vendor or architecture-specific programming models not only simplifies parallel programming in science and industry but also would support educators in teaching parallel programming.
%We would also like to express gratitude to educators for taking on this important role. Further materials, including a detailed description of the programming- and machine model can be found by following the reference on the Kokkos programming mode. Finnaly, we would like to mention that interested developers and student can reach the development team on Slack by following the invititation link as mentioned in the reference section.

### 2020-01-22 src/4.ProgammingModel.tex
Further renaming.

--- 2020-01-22 src/2.Kokkos.tex
@@
-\section{Towards Generic Parallel Programming}\label{chap:background}
-
-A programming model is an abstract representation of a computer.
-It defines a paradigm and a programming language.
-The programming language is used by the programmer to write source code, a human readable, textual representation of instructions and data.
-Source code is a sequential listing of statements evaluated from the top to bottom, one line at a time.
-While such ordering seems natural to describe an algorithm as it corresponds to sequential processing (both, reading code and progressing the program counter following a time line), it does not include the notion of concurrency.
-
-Parallel programming models add the notion of concurrency to a sequence of instruction of a program.
-This is commonly achieved either by adding semantic value to the base programming language, often referred to as programming with \emph{pragma} annotations, through library calls or new languages.
-All three approaches are conceptually suitable to represent programming paradigms such as tasking, parallel patterns and others.
-Well known representatives are the OpenMP programming model~\cite{CITEOPENMP} or threading libraries respectively.
-In general, it is up to the parallel programming model to expose language constructs or library interfaces (APIs) such that they are convenient to the programmer and capture enough semantic information to allow the programming model to map the program execution to the parallel computer hardware efficiently and correctly.
-Questions of which semantic information is needed, what paradigm to present and by what syntactical means span a design space worth taking a closer look at.
-Figure~\ref{figSemCapture} shows the three areas that constitute a parallel programming model and its design.
-We call it the semantic capture.
-
-\begin{figure}[h]
-\begin{Verbatim}[frame=leftline]
-Semantic information
-Paradigm 
-Language
-\end{Verbatim}
-\caption{Semantic capture - a triplet of language, paradigm and semantic information defined by a parallel programming model allow the programmer to express concurrency and the compiler and runtime system to understand and run applications on a parallel machine.}
-\label{figSemCapture}
-\end{figure}
-
-Semantic information provided by the programmer to the parallel programming model can be grouped into information to express intent and properties.
-They address the question of \emph{what}, \emph{where} and \emph{how}.
-In the context of parallel programming, these correspond to defining which code portion to parallelize, where to run and access and how to run that parallel code.
-Synchronization primitives may be considered as part of the what and information on the execution properties such as data placement or memory access type as the how.
-Semantics are expressed following a paradigm and a programming language.
-
-The parallel programming paradigm is an abstract ~\emph{representation} with the purpose of facilitating the programmer's understanding of programming rules and program behavior.
-Which programming paradigm to chose depends on several considerations.
-Parallel-patterns allow to express concurrency for commonly occurring programming patterns such as loop constructs.
-Tasking is a paradigm that support the expression of concurrent loops as well as irregular algorithms.
-Distributed and correctness-oriented programming models may implement actor-based programming.
-In this programming model, each unit of execution represent an actor who communicates over predefined communication interfaces.
-This paradigm eliminates accesses to shared state and align well with message passing programming (MPI).
-The execution model and memory model of a programming model defines behavior.
-That is, it defines the relationship between abstract concepts and program behavior on the given architecture.
-
-Lastly, a language defines the syntax of the parallel semantic.
-Declarative languages add semantic information to the base language through pragma annotations while imperative languages rely on programming interfaces to capture information.
-Their semantic is defined in the API specification.
-Figures~\ref{figOMPLike} and~\ref{figKokkosLike} show examples of two sample applications using the aforementioned parallel programming language types.
-While both programming models implement the same programming paradigm (parallel patterns), Figures~\ref{figOMPLike} extends the base programming language through pragma annotations.
-Both, pragma annotations and an API call as shown in Figure~\ref{figKokkosLike} have a similar semantic of describing a parallel loop construct.
-It is up to the programming model specification to define the execution and memory model in both cases.
-
-Both examples provide the semantic information on what to parallelize, however, they do not provide information on how to map the parallel execution on modern parallel architectures.
-In order to define what further information is needed to express concurrency on modern computer architectures, it is important to define an abstract machine model first.
-
-\begin{figure}
-\begin{Verbatim}[frame=leftline]
-# pragma model parallel for
-for ( size_t i = 0; i < N; ++i) {
- /* loop body */
-}
-\end{Verbatim}
-\caption{Declarative parallel programming provides uses pragma annotations to capture semantic information.}
-\label{figOMPLike}
-\end{figure}
-
-\begin{figure}
-\begin{Verbatim}[frame=leftline]
-model::parallel_for (N, [=] ( const size_t i) {
-  /* loop body */
-});
-\end{Verbatim}
-\caption{Imperative parallel programming uses a base language and relies on programming interfaces to capture information and a documentation that describe their semantics.}
-\label{figKokkosLike}
-\end{figure}
-
-\section{The Kokkos Abstract Machine Model}\label{chap:kokkosMM}
-
-The Kokkos machine model defines abstractions that represent hardware capabilities for processing and data access.
-The Kokkos parallel programming model exposes these abstractions to the programmer through C++ and the Kokkos API.
-By abstracting from physical hardware, the machine model ensures that applications written in programming models targeting this machine model are generic and performance-portable on current and future hardware.
-The Kokkos parallel programming model is one particular instance of a programming model that builds on top of that machine model.
-We would like to point out that it is this conceptual differentiation that allows scenarios where the underlying machine model is instantiated in other languages beyond C++, yet the algorithmic specification as well as performance characteristic and portability remain the same.
-
-The Kokkos abstract machine model targets a design of a future shared-memory computing architectures.
-The design is shown in Figure~\ref{fig:execspace}.
-It is characterized by multiple latency-optimized execution units and off-die bandwidth-optimized accelerators.
-Both compute device types have disjoint memory address spaces and unique performance properties.
-In such an architecture,   execution units might be grouped into larger units with multiple levels of achievable parallelism with different memory access characteristics and coherence properties of caches.
-In order to ensure performance portability on such a wide range of configurations, an abstraction of the compute engines and available memories are required.
-For this purpose we introduce \emph{spaces}.
-
-An instance of an \emph{execution space} is an abstraction over an execution resource to which a programmer can target parallel work.
-For example, an execution space can be used to describe a multi-core processor.
-In this case, the execution space contains several homogeneous cores organized into arbitrary logical groups.
-In a parallel programming model that implements this machine model, an instance of such an execution space would be made available to the programmer to run kernels.
-Adding more logical groups or accelerators simply increases the number of available execution spaces to the programmer.
-
-Memory and memory types are exposed through \emph{memory spaces}.
-Each memory space provides a finite storage capacity at which data structures can be allocated and accessed.
-Different memory space types have different characteristics with respect to accessibility from execution spaces and performance.
-
-An instance of a memory space provides a concrete method for the application programmer to request data storage allocations.
-Following the architecture as discussed earlier, the multi-core processor may contain multiple memory spaces that abstract on-package memory, slower DRAM and non-volatile memories.
-Accelerators can provide an additional memory space through its local on-package memory.
-An architectural layout is shown in Figure~\ref{fig:memspace}.
-The programmer is free to decide where each data structure is allocated by requesting the corresponding memory space.
-Programmatically this is achieved by instantiating the according memory space.
-Kokkos provides the appropriate abstraction of memory allocation and data management.
-We believe that an abstract representation of execution- and memory devices is a key property towards performance-portable parallel programming.
-
-How a machine model is exposed to the programmer and what are the design considerations towards defining an appropriate semantic capture are discussed in the next section.
-
-%Ang, J.A., et.
-al., Abstract Machine Models and Proxy Architectures for Exascale Computing, 2014, Sandia National Laboratories and Lawrence Berkeley National Laboratory, DOE Computer Architecture Laboratories Project
-
-\begin{figure}
-\begin{subfigure}[b]{0.41\textwidth}
-\centering
-\includegraphics[width=\textwidth]{img/ExecSpaces.png}
-\caption{Execution space is an instantiation of a space that represents a processing device to which the programmer can target parallel code.}
-\label{fig:execspace}
-\end{subfigure}
-\hfill \break
-\begin{subfigure}[b]{0.41\textwidth}
-\centering
-\includegraphics[width=\textwidth]{img/MemSpaces.png}
-\caption{A memory space is an instantiation of a space that represents a memory location on which the programmer can allocate data.}
-\label{fig:memspace}
-\end{subfigure}
-\caption{Spaces represent conceptual building blocks of the abstract machine model in Kokkos.}
-\label{fig:spaces}
-\end{figure}
-
 \section{The Kokkos Programming Model}\label{chap:kokkosPM}
@@
 \end{Verbatim}
-\caption{The Kokkos~\emph{parallel_for} and \emph{parallel_reduce} class show how template metaprogramming allows to specialize types.
+\caption{The Kokkos~\emph{parallel\_for} and \emph{parallel\_reduce} class show how template metaprogramming allows to specialize types.
 Its descriptive semantic offers the freedom to optimize execution on modern architectures with multiple degrees of concurrency where an Execution policy shapes the iteration space accordingly.}

### 2020-01-23 main.tex
Updated main.tex to include new tex files.
Some more file name edits.
Further renaming.
Updated authors

@@
 
-\title{Performance-portable Parallel Programming in Computer Science Education with Kokkos and C++}
+\title{Performance-portable Parallel Programming in Computer Science Education with Kokkos}
 
-\author{\IEEEauthorblockN{1\textsuperscript{st} Given Name Surname}
-\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-\textit{name of organization (of Aff.)}\\
-City, Country \\
-email address or ORCID}
+\author{
+\IEEEauthorblockN{Jan Ciesko}
+\IEEEauthorblockA{\textit{Computer Science Reseach Center} \\
+\textit{Sandia National Laboratories}\\
+Albuquerque, NM, USA\\
+jciesko@sandia.gov}
 \and
-\IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
-\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-\textit{name of organization (of Aff.)}\\
-City, Country \\
-email address or ORCID}
+\IEEEauthorblockN{David Poliakoff}
+\IEEEauthorblockA{\textit{Computer Science Reseach Center} \\
+\textit{Sandia National Laboratories}\\
+Albuquerque, NM, USA\\
+dzpolia@sandia.gov}
 \and
-\IEEEauthorblockN{3\textsuperscript{rd} Given Name Surname}
-\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-\textit{name of organization (of Aff.)}\\
-City, Country \\
-email address or ORCID}
+\IEEEauthorblockN{David S. Hollman}
+\IEEEauthorblockA{\textit{Computer Science Reseach Center} \\
+\textit{Sandia National Laboratories}\\
+Livermore, CA, USA\\
+dshallm@sandia.gov}
+\and
+\IEEEauthorblockN{Author Name}
+\IEEEauthorblockA{\textit{Computer Science Reseach Center} \\
+\textit{Sandia National Laboratories}\\
+Albuquerque, NM, USA\\
+aname@sandia.gov}
+\and
+\IEEEauthorblockN{Author Name}
+\IEEEauthorblockA{\textit{Computer Science Reseach Center} \\
+\textit{Sandia National Laboratories}\\
+Albuquerque, NM, USA\\
+aname@sandia.gov}
+\and
+\IEEEauthorblockN{Author Name}
+\IEEEauthorblockA{\textit{Computer Science Reseach Center} \\
+\textit{Sandia National Laboratories}\\
+Albuquerque, NM, USA\\
+aname@sandia.gov}
 }
@@
 \input{src/1.Introduction}           %label: chap:introduction
-\input{src/2.Kokkos}              %label: chap:kokkos
-\input{src/2.1.KokkosBackend.tex} %label: chap:kokkosBackend
-\input{src/3.Exercises}           %label: chap:tutorial
-\input{src/4.C++}                 %label: chap:c++
-\input{src/5.Related}             %label: chap:related
-\input{src/6.Conclusion}          %label: chap:conclusion
+\input{src/2.GenericProgramming.tex} %label: chap:background
+\input{src/3.MachineModel.tex}             %label: chap:kokkosPM
+\input{src/4.ProgammingModel.tex}     %label: chap:kokkosEMA
+\input{src/5.Backend.tex}            %label: chap:kokkosBackend
+\input{src/6.Exercises}              %label: chap:tutorial
+\input{src/7.C++}                    %label: chap:c++
+\input{src/8.Related}                %label: chap:related
+\input{src/9.Conclusion}             %label: chap:conclusion
 

### 2020-01-23 src/4.ProgammingModel.tex
Latest changes.

@@
  const FunctorType& functor);
-template <class ExecPolicy, class FunctorType, class ReducerVarNonConst>
-Kokkos::parallel_reduce(const std::string& name, 
- const ExecPolicy& policy, 
- const FunctorType& functor, 
- ReducerVarNonConst& reducer);
 \end{Verbatim}
-\caption{The Kokkos~\emph{parallel\_for} and \emph{parallel\_reduce} class show how template metaprogramming allows to specialize types.
-Its descriptive semantic offers the freedom to optimize execution on modern architectures with multiple degrees of concurrency where an Execution policy shapes the iteration space accordingly.}
-\label{fig:parallelForAndReduceKokkos}
+\caption{The Kokkos~\emph{parallel\_for} class shows how template metaprogramming allows to specialize types.
+Its descriptive semantic offers the freedom to optimize execution on modern architectures with multiple degrees of concurrency where an execution policy (\emph{ExecPolicy}) shapes the iteration space accordingly.}
+\label{fig:parallelFor}
 \end{figure}
 
-WIP: Figure~\ref{fig:parallelForAndReduceKokkos} shows definitions of the~\emph{parallel for} and~\emph{parallel reduction} interfaces.
-Both interfaces take a set of template parameters and function arguments.
-\emph{ExecPolicy} shapes the iteration space for a particular device with varying degrees of hierarchical parallelism.
-A valid template parameter for ExecPolicy is a \emph{RangePolicy}.
-This type defines a 1-dimensional iteration range that is templated.
-This allows to consider the distribtion of iterations into  hierarchical orghanization of parallel processing units.
-In this case, ExecPolicy is   and~\emph{FunctorType} and a set of function arguments.
+Figure~\ref{fig:parallelFor} shows the definition of the Kokkos~\emph{parallel\_for} interface.
+The interface accepts two template parameters and three function arguments.
+\emph{ExecPolicy} shapes the iteration space for a particular execution space.
+Valid template parameter are the~\emph{RangePolicy},~\emph{TeamPolicy},~\emph{MDRangePolicy} and others.
+As the name suggests, these policies express how an iteration space is mapped to the underlying hardware.
+In particular, a range policy defines a one dimensional iteration range.
+Team policies allow to map an iteration space to architectures that expose hierarchical parallelism..
+Policies commonly accept template parameters such as~\emph{execution space},~\emph{schedule} or~\emph{index type}.
+For the interested reader, the complete programming model specification can be accessed on-line\cite{pub:KOKKOS}.
 
-Expressing concurrency with patterns and abstractions allows the compiler to generate optimized transformations.
-Further, a runtime library to map the execution to the parallel hardware.
-In the case of parallel patterns, the execution is unspecified and only promise deterministic results for the reduction and scan operations.
+Programming with patterns and abstractions allows the compiler to generate optimized transformations.
+In case of parallel patterns, the execution is unspecified and only promises deterministic results for the reduction and scan operations.
 This gives the freedom to the underlying abstraction layers to implement different mapping patterns on different hardware such as assignment of iterations to threads or vector lanes.
@@
 Figure~\ref{fig:SemCaptureKokkos} shows the semantic capture defined by the Kokkos programming model. 
-Figure~\ref{fig:stack} shows abstraction layers in Kokkos and and how it fits in a parallel programming landscape.
-The next section shows an example application written in Kokkos and provides an insight into the runtime library.
-For the interested reader, the complete programming model specification can be accessed on-line\cite{pub:KOKKOS}.
+%Figure~\ref{fig:stack} shows abstraction layers in Kokkos and and how it fits in a parallel programming landscape.
+The next section shows how the Kokkos parallel\_for class is mapped to the the underlying back-end programming model.
 

### 2020-01-23 src/1.Introduction.tex
Section 1, para 1-2

@@
 Firstly, given their role, educators adapt to trends in research and industry and include parallel computing in their curricula, and secondly, parallel computing is an attractive field for department research and industry collaborations.
+At Sandia, we develop a performance portability library called Kokkos that is ideal for educators looking to teach parallel computing.
 
-While industry-driven collaborations often predefine technological frameworks such as programming environments and targeted architectures, academic teaching leaves this mostly undefined.
-In this case, educators are exposed to a variety of solutions, approaches and trends in parallel computing.
-Which particular set to choose for teaching in class is not trivial.
-While decision factors differ, maximizing the subject relevance by teaching generic principles of parallel programming generically is among the most important ones.
+In teaching a class on parallel programming, educators must balance considerations of architecture, programming model, and theoretical concepts.
+Unfortunately, not all architectures support all programming models, nor do many programming models cleanly express all parallel programming constructs.
+Many vendor models compound this, requiring students have access to specific hardware, and limiting the concepts of parallelism to those supported by that architecture.
+This makes the choice of programming model important and difficult.
 

### 2020-01-23 src/1.Introduction.tex
Fluff re: Kokkos enabling cool stuff

@@
 
-Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS_PAPER_HERE}, is a performance-portable, parallel programming model for C++ developed at Sandia National Laboratories (SNL).
+Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS_PAPER_HERE}, is a performance-portable, parallel programming model for C++ developed at Sandia National Laboratories (SNL) for developing million line scientific software on the world\'s biggest supercomputers.
 It consists of a programming model specification and a library.
 The specification defines execution and memory model and an application programming interface (library API).
-The runtime library exposes programming primitives, namely, patterns and types and maps them to the underlying abstract machine model.
-The machine model creates a generic view of compute hardware and memories.
+The runtime library exposes functions representing parallel patterns and types for data management and maps them to the underlying abstract execution model.
+This execution model creates a generic view of compute hardware and memories.
 The Kokkos library maps execution primitives to vendor-specific programming models, libraries, and memory layouts.
 Kokkos has been successfully used in many large-scale projects at SNL, LLNL and other research centers~\cite{CITEKOKKOSUSECASES}.
-We list its key properties as follows.
+Its key properties are
 
@@
 Further, we would like to present Kokkos to the educational community as one representative parallel programming model for a trend towards architecture-independent and performance-portable programming of CPUs and accelerators.
-We show a set of types that implement patterns and encapsulate data with enough semantic information to produce performance-portable, parallel code on current architectures.
+We show an abstraction with enough semantic information to produce performance-portable, parallel code on current architectures.
 Further, we highlight its alignment and contributions to standardization efforts towards supporting parallel programming in C++.

### 2020-01-23 src/1.Introduction.tex
Turned Hollman's beautiful one liner into a bulleted list and an ugly paragraph.
If brevity is the soul of wit does verbosity make me witless or soulless?

@@
 
-Secondly, we would like to promote Kokkos to students as a platform to understand and experiment with parallel patterns and types.
-Type specialization through templating allows to experiment with memory layouts and investigate their performance implications.
-Further, its modular code organization and open access under the BSD license allow students to peek behind the scenes and explore the mappings of abstractions to a variety of vendor-specific implementations.
-We believe that this makes Kokkos a valuable programming model for students which allows them to reason about vendor-provided programming models and to see them as particular implementations of higher abstractions in a parallel programming landscape.
+Secondly, we would like to promote Kokkos to students and educators as a platform to understand and experiment with parallel patterns in a fashion that is
+
+\begin{itemize}
+\item Generic
+\item Vendor-agnostic
+\item Declarative
+\end{itemize}
+
+First, Kokkos has semantics that are generic.
+This means that students are learning parallelism detached from any architectural concerns.
+Kokkos won\'t require them to know the specifics of the \"symmetric multiprocessors\" of a GPU or the cache hiearchy on a CPU\footnote{Though nothing precludes an educator from teaching these independently of the model}.
+Finally, Kokkos is vendor agnostic.
+If you have a computer lab with processors from one vendor and your students have processors from another, this presents no problem, Kokkos likely works on both.
+Kokkos is declarative, students aren\'t saying \"map these work items to this set of threads\" but \"do a reduction over this set of data.\" This is critical, much of software engineering is moving to declarative models and languages, learning to operate in such a mindset is useful outside of parallel programming.
 

### 2020-01-23 src/2.GenericProgramming.tex
Modified my part of Chapter 2. Expanded OpenMP-like to be more imperative

@@
 
-A programming model is an abstract representation of a computer.
-It defines a paradigm and a programming language.
-The programming language is used by the programmer to write source code, a human readable, textual representation of instructions and data.
-Source code is a sequential listing of statements evaluated from the top to bottom, one line at a time.
-While such ordering seems natural to describe an algorithm as it corresponds to sequential processing (both, reading code and progressing the program counter following a time line), it does not include the notion of concurrency.
-
-Parallel programming models add the notion of concurrency to a sequence of instruction of a program.
-This is commonly achieved either by adding semantic value to the base programming language, often referred to as programming with \emph{pragma} annotations, through library calls or new languages.
+A programming model exposes constrained semantics to the programmer which that programmer uses to express intent.
+Absent an embedded programming model, the semantics of C-like languages can lead to sequential thinking and execution.
+Parallel programming models expose parallel semantics to enable developers to create performance portable code.
+This is commonly achieved by adding semantic value to the base programming language, often referred to as programming with \emph{pragma} annotations.
+Other methods expose such semantics through library calls or new languages.
 All three approaches are conceptually suitable to represent programming paradigms such as tasking, parallel patterns and others.
-Well known representatives are the OpenMP programming model~\cite{CITEOPENMP} or threading libraries respectively.
-In general, it is up to the parallel programming model to expose language constructs or library interfaces (APIs) such that they are convenient to the programmer and capture enough semantic information to allow the programming model to map the program execution to the parallel computer hardware efficiently and correctly.
-Questions of which semantic information is needed, what paradigm to present and by what syntactical means span a design space worth taking a closer look at.
+Some representatives are the OpenMP programming model~\cite{CITEOPENMP}, threading libraries, or languages like Erlang in which parallelism is a first-class citizen.
+In general, it is up to the parallel programming model to expose language constructs or library interfaces (APIs) such that they balance convenience to the programmer with semantics sufficient to map the program execution to the parallel computer hardware efficiently and correctly.
+Questions of what semantic information is needed, what paradigms to present and by what syntactical means span a design space worth taking a closer look at.
 Figure~\ref{figSemCapture} shows the three areas that constitute a parallel programming model and its design.
@@
 
-Lastly, a language defines the syntax of the parallel semantic.
-Declarative languages add semantic information to the base language through pragma annotations while imperative languages rely on programming interfaces to capture information.
-Their semantic is defined in the API specification.
-Figures~\ref{figOMPLike} and~\ref{figKokkosLike} show examples of two sample applications using the aforementioned parallel programming language types.
-While both programming models implement the same programming paradigm (parallel patterns), Figures~\ref{figOMPLike} extends the base programming language through pragma annotations.
-Both, pragma annotations and an API call as shown in Figure~\ref{figKokkosLike} have a similar semantic of describing a parallel loop construct.
-It is up to the programming model specification to define the execution and memory model in both cases.
+Lastly, a model must decide between declarative and imperative semantics.
+Declarative semantics require a developer to express intent, implementation is the responsibility of the programming model and underlying toolchain.
+Imperative semantics require the developer to specify the mechanisms by which their intent will be implemented.
+In an imperative model, developer success is tied to their knowledge of implementation mechanisms.
+In a declarative model that success is tied to the developer\'s ability to accurately declare their requirements, and the toolchain\'s ability to turn that into an implementation.
+Figures~\ref{figOMPLike} and~\ref{figKokkosLike} show examples of two sample applications using the aforementioned types of semantics.
+While both programming models implement the same programming paradigm (parallel patterns), Figure~\ref{figOMPLike} required the developer to express how to parallelize the loop, while in~\ref{figKokkosLike} the developer just declared their parallel semantics, and relied on the programming model to pick an implementation.
 
@@
 \begin{Verbatim}[frame=leftline]
-# pragma model parallel for
+# pragma model parallel for simd vectorize tile unroll fuse
 for ( size_t i = 0; i < N; ++i) {

### 2020-01-23 src/2.GenericProgramming.tex
Touched up final paragraph

@@
 
-Both examples provide the semantic information on what to parallelize, however, they do not provide information on how to map the parallel execution on modern parallel architectures.
-In order to define what further information is needed to express concurrency on modern computer architectures, it is important to define an abstract machine model first.
+Kokkos takes the declarative approach.
+For the developers of our large scientific applications, this is appealing because it allows them to express their algorithms in one way, and rely on Kokkos to map that intent effectively to a variety of architectures.
+For students this is appealing because they\'re unlikely to have expertise in these architectures.
+For an educator looking for an implementation vehicle with which to teach students about parallelism, the declarative model eliminates the burden of reimplementing exercises as computer labs and dominant architectures change.
+For a declarative model to make sense, the educator should know something about the abstract execution model being used to ensure performance portability.
+Depending on the nature of the course, this could be mostly hidden from students, or discussed in detail.
 

### 2020-01-23 src/8.Related.tex
First draft of the Related Work chapter

\section{Related Work}\label{chap:related}
\subsection{RAJA}
Like Kokkos, RAJA is a C++ framework for porting massive DOE codes to a variety of architectures. In both frameworks, a developer first expresses the operations their kernel performs using simple C++ functional programming mechanisms. The difference is that Kokkos is declarative, while RAJA is imperative. In Kokkos, the framework figures out how to map that kernel to different architectures. In RAJA, functionality is provided to make it easy for a developer to try different strategies to map those kernels to architectures, but the developer still decides how that mapping happens. Which model makes more sense in a given circumstance depends on how much exposure to the underlying programming model a developer wants. RAJA would be ideal for a class in which the educator wants to expose students to the semantics of different programming models without exposing them to the detailed syntax of those models.
\subsection{DPC++}
DPC++ is a model developed by Intel(TM) for expressing parallelism across their architectures. In addition to sharing a declarative philosophy with Kokkos, it even share a number of patterns. Unlike Kokkos, at time of publication it targets only Intel architectures (it isn\'t vendor-agnostic) For a course targeting Intel(TM) platforms, you\'re likely to see good tool support, DPC++ could be a valid option.
\subsection{Parallel Constructs in the C++ Standard}
C++ increasingly supports parallel constructs. We discuss the contributions of Kokkos to C++ in~\ref{chap:c++}, it\'s increasingly possible to teach parallelism with nothing but a C++ compiler. A problem is that at time of publication C++ doesn\'t support accelerators, or many of the complexities of parallel computing as it\'s used in practice. A positive is that instead of requiring libraries, you just need a modern C++ compiler. For a course in which topics like accelerators aren\'t discussed, this can be a valid option.
%\end{itemize}

### 2020-01-23 src/4.ProgammingModel.tex
Added parallel_for example.

@@
 
-Figure~\ref{fig:parallelFor} shows the definition of the Kokkos~\emph{parallel\_for} interface.
-The interface accepts two template parameters and three function arguments.
+Figure~\ref{fig:parallelFor} shows one definition of the Kokkos~\emph{parallel\_for} interface.
+This interfac accepts two template parameters and three function arguments.
 \emph{ExecPolicy} shapes the iteration space for a particular execution space.
-Valid template parameter are the~\emph{RangePolicy},~\emph{TeamPolicy},~\emph{MDRangePolicy} and others.
+Valid template parameter are the~\emph{RangePolicy},~\emph{TeamPolicy},~\emph{MDRangePolicy},~\emph{TeamThreadRange} and~\{ThreadVectorRange}.
 As the name suggests, these policies express how an iteration space is mapped to the underlying hardware.
-In particular, a range policy defines a one dimensional iteration range.
-Team policies allow to map an iteration space to architectures that expose hierarchical parallelism..
-Policies commonly accept template parameters such as~\emph{execution space},~\emph{schedule} or~\emph{index type}.
+For example, a range policy defines a one dimensional iteration range.
+A team policy defines a 1D iteration range, each of which is assigned to a thread team.
+This policy allows the expression of hierarchical parallelism.
 For the interested reader, the complete programming model specification can be accessed on-line\cite{pub:KOKKOS}.
+Figure~\ref{fig:abstractions} lists patterns and abstractions and structures them by information type in the semantic capture.
 
@@
 This gives the freedom to the underlying abstraction layers to implement different mapping patterns on different hardware such as assignment of iterations to threads or vector lanes.
-We believe that these abstraction and their descriptive semantic represent key programming primitives and the way forward for native language support for parallel programming.
-
-Figure~\ref{fig:SemCaptureKokkos} shows the semantic capture defined by the Kokkos programming model. 
+We believe that these abstraction and their descriptive semantic represent key programming primitives and the way forward for generic support for parallel programming.
+In conclusion, Figure~\ref{fig:SemCaptureKokkos} shows the semantic capture defined by the Kokkos programming model.
 %Figure~\ref{fig:stack} shows abstraction layers in Kokkos and and how it fits in a parallel programming landscape.
@@
 \label{fig:stack}
+\end{figure}
+
+\begin{figure}
+\centerline{\includegraphics[width=0.3\textwidth]{img/Abstractions.png}}
+\caption{Landscape}
+\label{fig:abstractions}
 \end{figure}

### 2020-01-23 src/4.ProgammingModel.tex
Minor changes.

@@
 Figure~\ref{fig:parallelFor} shows one definition of the Kokkos~\emph{parallel\_for} interface.
-This interfac accepts two template parameters and three function arguments.
+This interface accepts two template parameters and three function arguments.
 \emph{ExecPolicy} shapes the iteration space for a particular execution space.
@@
 This policy allows the expression of hierarchical parallelism.
+The second template parameter,~\emph{FuntorType}, represent a functor that implements the~\{()-operator} with a matching signature for the given execution policy.
+The functor can be defined using a C++ class, struct or lambda expression.
+We show an example of how a parallel for loop is expressed and mapping to the vendor-specific back-end in the next chapter.
 For the interested reader, the complete programming model specification can be accessed on-line\cite{pub:KOKKOS}.

### 2020-01-23 src/4.ProgammingModel.tex
Removed subsection-headline.
Bug fix.

@@
 Equipped with a language and set of paradigms, a particular set of abstractions can be defined to expose the abstract machine model and captures the intent of the programmer.
-
-\subsection{Kokkos Abstractions}
 
@@
 \emph{ExecPolicy} shapes the iteration space for a particular execution space.
-Valid template parameter are the~\emph{RangePolicy},~\emph{TeamPolicy},~\emph{MDRangePolicy},~\emph{TeamThreadRange} and~\{ThreadVectorRange}.
+Valid template parameter are the~\emph{RangePolicy},~\emph{TeamPolicy},~\emph{MDRangePolicy},~\emph{TeamThreadRange} and~\emph{ThreadVectorRange}.
 As the name suggests, these policies express how an iteration space is mapped to the underlying hardware.
@@
 This policy allows the expression of hierarchical parallelism.
-The second template parameter,~\emph{FuntorType}, represent a functor that implements the~\{()-operator} with a matching signature for the given execution policy.
+The second template parameter,~\emph{FuntorType}, represent a functor that implements the~\emph{()-operator} with a matching signature for the given execution policy.
 The functor can be defined using a C++ class, struct or lambda expression.

### 2020-01-23 src/8.Related.tex
Lerned 2 LaTeX

@@
 In addition to sharing a declarative philosophy with Kokkos, it even share a number of patterns.
-Unlike Kokkos, at time of publication it targets only Intel architectures (it isn\'t vendor-agnostic) For a course targeting Intel(TM) platforms, you\'re likely to see good tool support, DPC++ could be a valid option.
+Unlike Kokkos, at time of publication it targets only Intel architectures (it isn't vendor-agnostic) For a course targeting Intel(TM) platforms, you\'re likely to see good tool support, DPC++ could be a valid option.
 \subsection{Parallel Constructs in the C++ Standard}
 C++ increasingly supports parallel constructs.
-We discuss the contributions of Kokkos to C++ in~\ref{chap:c++}, it\'s increasingly possible to teach parallelism with nothing but a C++ compiler.
-A problem is that at time of publication C++ doesn\'t support accelerators, or many of the complexities of parallel computing as it\'s used in practice.
+We discuss the contributions of Kokkos to C++ in~\ref{chap:c++}, it's increasingly possible to teach parallelism with nothing but a C++ compiler.
+A problem is that at time of publication C++ doesn't support accelerators, or many of the complexities of parallel computing as it\'s used in practice.
 A positive is that instead of requiring libraries, you just need a modern C++ compiler.

### 2020-01-23 src/2.GenericProgramming.tex
Added Hollman's pic.
More typos

@@
 In an imperative model, developer success is tied to their knowledge of implementation mechanisms.
-In a declarative model that success is tied to the developer\'s ability to accurately declare their requirements, and the toolchain\'s ability to turn that into an implementation.
+In a declarative model that success is tied to the developer's ability to accurately declare their requirements, and the toolchain's ability to turn that into an implementation.
 Figures~\ref{figOMPLike} and~\ref{figKokkosLike} show examples of two sample applications using the aforementioned types of semantics.
@@
 For the developers of our large scientific applications, this is appealing because it allows them to express their algorithms in one way, and rely on Kokkos to map that intent effectively to a variety of architectures.
-For students this is appealing because they\'re unlikely to have expertise in these architectures.
+For students this is appealing because they're unlikely to have expertise in these architectures.
 For an educator looking for an implementation vehicle with which to teach students about parallelism, the declarative model eliminates the burden of reimplementing exercises as computer labs and dominant architectures change.

### 2020-01-23 src/8.Related.tex
More typos

@@
 In addition to sharing a declarative philosophy with Kokkos, it even share a number of patterns.
-Unlike Kokkos, at time of publication it targets only Intel architectures (it isn't vendor-agnostic) For a course targeting Intel(TM) platforms, you\'re likely to see good tool support, DPC++ could be a valid option.
+Unlike Kokkos, at time of publication it targets only Intel architectures (it isn't vendor-agnostic) For a course targeting Intel(TM) platforms, you're likely to see good tool support, DPC++ could be a valid option.
 \subsection{Parallel Constructs in the C++ Standard}
@@
 We discuss the contributions of Kokkos to C++ in~\ref{chap:c++}, it's increasingly possible to teach parallelism with nothing but a C++ compiler.
-A problem is that at time of publication C++ doesn't support accelerators, or many of the complexities of parallel computing as it\'s used in practice.
+A problem is that at time of publication C++ doesn't support accelerators, or many of the complexities of parallel computing as it's used in practice.
 A positive is that instead of requiring libraries, you just need a modern C++ compiler.
-For a course in which topics like accelerators aren\'t discussed, this can be a valid option.
+For a course in which topics like accelerators aren't discussed, this can be a valid option.
 %\end{itemize}

### 2020-01-23 src/4.ProgammingModel.tex
Added description on ret semantic of the parallel_for.

@@
 The functor can be defined using a C++ class, struct or lambda expression.
-We show an example of how a parallel for loop is expressed and mapping to the vendor-specific back-end in the next chapter.
-For the interested reader, the complete programming model specification can be accessed on-line\cite{pub:KOKKOS}.
+We would like to highlight that the return semantic of the parallel_for is defined as \emph{potentially asynchronous}.
+This shows the descriptive nature of the Kokkos programming model.
+In this case, it depends on the specified execution space and weather it implements a synchronization point.
+To the interested reader, the complete programming model specification can be accessed on-line\cite{pub:KOKKOS}.
 Figure~\ref{fig:abstractions} lists patterns and abstractions and structures them by information type in the semantic capture.

### 2020-01-23 src/1.Introduction.tex
The David Poliakoff Center For Kids Who Want To Write Good

@@
 
-Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS_PAPER_HERE}, is a performance-portable, parallel programming model for C++ developed at Sandia National Laboratories (SNL) for developing million line scientific software on the world\'s biggest supercomputers.
+Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS_PAPER_HERE}, is a performance-portable, parallel programming model for C++ developed at Sandia National Laboratories (SNL) for developing million line scientific software on the world's biggest supercomputers.
 It consists of a programming model specification and a library.
@@
 
-First, Kokkos has semantics that are generic.
+Kokkos has semantics that are generic.
 This means that students are learning parallelism detached from any architectural concerns.
-Kokkos won\'t require them to know the specifics of the \"symmetric multiprocessors\" of a GPU or the cache hiearchy on a CPU\footnote{Though nothing precludes an educator from teaching these independently of the model}.
-Finally, Kokkos is vendor agnostic.
+Kokkos won't require them to know the specifics of the "symmetric multiprocessors" of a GPU or the cache hiearchy on a CPU\footnote{Though nothing precludes an educator from teaching these independently of the model}.
+Kokkos is vendor agnostic.
 If you have a computer lab with processors from one vendor and your students have processors from another, this presents no problem, Kokkos likely works on both.
-Kokkos is declarative, students aren\'t saying \"map these work items to this set of threads\" but \"do a reduction over this set of data.\" This is critical, much of software engineering is moving to declarative models and languages, learning to operate in such a mindset is useful outside of parallel programming.
+Kokkos is declarative, students aren't saying "map these work items to this set of threads" but "do a reduction over this set of data." This is critical, much of software engineering is moving to declarative models and languages, learning to operate in such a mindset is useful outside of parallel programming.
+All of these are desirable in a programming model for education.
 

### 2020-01-23 src/5.Backend.tex

--- 2020-01-22 src/2.Kokkos.tex
@@
-\section{Towards Generic Parallel Programming}\label{chap:background}
+\subsection{Back-end Support}\label{chap:kokkosBackend}
 
-A programming model is an abstract representation of a computer.
-It defines a paradigm and a programming language.
-The programming language is used by the programmer to write source code, a human readable, textual representation of instructions and data.
-Source code is a sequential listing of statements evaluated from the top to bottom, one line at a time.
-While such ordering seems natural to describe an algorithm as it corresponds to sequential processing (both, reading code and progressing the program counter following a time line), it does not include the notion of concurrency.
-
-Parallel programming models add the notion of concurrency to a sequence of instruction of a program.
-This is commonly achieved either by adding semantic value to the base programming language, often referred to as programming with \emph{pragma} annotations, through library calls or new languages.
-All three approaches are conceptually suitable to represent programming paradigms such as tasking, parallel patterns and others.
-Well known representatives are the OpenMP programming model~\cite{CITEOPENMP} or threading libraries respectively.
-In general, it is up to the parallel programming model to expose language constructs or library interfaces (APIs) such that they are convenient to the programmer and capture enough semantic information to allow the programming model to map the program execution to the parallel computer hardware efficiently and correctly.
-Questions of which semantic information is needed, what paradigm to present and by what syntactical means span a design space worth taking a closer look at.
-Figure~\ref{figSemCapture} shows the three areas that constitute a parallel programming model and its design.
-We call it the semantic capture.
-
-\begin{figure}[h]
-\begin{Verbatim}[frame=leftline]
-Semantic information
-Paradigm 
-Language
-\end{Verbatim}
-\caption{Semantic capture - a triplet of language, paradigm and semantic information defined by a parallel programming model allow the programmer to express concurrency and the compiler and runtime system to understand and run applications on a parallel machine.}
-\label{figSemCapture}
-\end{figure}
-
-Semantic information provided by the programmer to the parallel programming model can be grouped into information to express intent and properties.
-They address the question of \emph{what}, \emph{where} and \emph{how}.
-In the context of parallel programming, these correspond to defining which code portion to parallelize, where to run and access and how to run that parallel code.
-Synchronization primitives may be considered as part of the what and information on the execution properties such as data placement or memory access type as the how.
-Semantics are expressed following a paradigm and a programming language.
-
-The parallel programming paradigm is an abstract ~\emph{representation} with the purpose of facilitating the programmer's understanding of programming rules and program behavior.
-Which programming paradigm to chose depends on several considerations.
-Parallel-patterns allow to express concurrency for commonly occurring programming patterns such as loop constructs.
-Tasking is a paradigm that support the expression of concurrent loops as well as irregular algorithms.
-Distributed and correctness-oriented programming models may implement actor-based programming.
-In this programming model, each unit of execution represent an actor who communicates over predefined communication interfaces.
-This paradigm eliminates accesses to shared state and align well with message passing programming (MPI).
-The execution model and memory model of a programming model defines behavior.
-That is, it defines the relationship between abstract concepts and program behavior on the given architecture.
-
-Lastly, a language defines the syntax of the parallel semantic.
-Declarative languages add semantic information to the base language through pragma annotations while imperative languages rely on programming interfaces to capture information.
-Their semantic is defined in the API specification.
-Figures~\ref{figOMPLike} and~\ref{figKokkosLike} show examples of two sample applications using the aforementioned parallel programming language types.
-While both programming models implement the same programming paradigm (parallel patterns), Figures~\ref{figOMPLike} extends the base programming language through pragma annotations.
-Both, pragma annotations and an API call as shown in Figure~\ref{figKokkosLike} have a similar semantic of describing a parallel loop construct.
-It is up to the programming model specification to define the execution and memory model in both cases.
-
-Both examples provide the semantic information on what to parallelize, however, they do not provide information on how to map the parallel execution on modern parallel architectures.
-In order to define what further information is needed to express concurrency on modern computer architectures, it is important to define an abstract machine model first.
+The important consideration here is that the method of compiling code for different execution spaces and the dispatch of kernels to instances is abstracted by the Kokkos model.
+This unburdens application programmers from writing algorithms in hardware specific languages.
 
@@
 \begin{Verbatim}[frame=leftline]
-# pragma model parallel for
-for ( size_t i = 0; i < N; ++i) {
- /* loop body */
+
+int main(int argc, char* argv[]) {
+   Kokkos::initialize(argc,argv);
+
+   int N = atoi(argv[1]);
+
+   Foo foo;
+
+   Kokkos::parallel_for(Kokkos::TeamPolicy<Tag1>(N,Kokkos::AUTO), foo);
+   Kokkos::parallel_for("Loop2", Kokkos::TeamPolicy<Tag2>(N,Kokkos::AUTO), foo);
+
+   Kokkos::finalize();
 }
+
 \end{Verbatim}
@@
 \end{figure}
-
-\begin{figure}
-\begin{Verbatim}[frame=leftline]
-model::parallel_for (N, [=] ( const size_t i) {
-  /* loop body */
-});
-\end{Verbatim}
-\caption{Imperative parallel programming uses a base language and relies on programming interfaces to capture information and a documentation that describe their semantics.}
-\label{figKokkosLike}
-\end{figure}
-
-\section{The Kokkos Abstract Machine Model}\label{chap:kokkosMM}
-
-The Kokkos machine model defines abstractions that represent hardware capabilities for processing and data access.
-The Kokkos parallel programming model exposes these abstractions to the programmer through C++ and the Kokkos API.
-By abstracting from physical hardware, the machine model ensures that applications written in programming models targeting this machine model are generic and performance-portable on current and future hardware.
-The Kokkos parallel programming model is one particular instance of a programming model that builds on top of that machine model.
-We would like to point out that it is this conceptual differentiation that allows scenarios where the underlying machine model is instantiated in other languages beyond C++, yet the algorithmic specification as well as performance characteristic and portability remain the same.
-
-The Kokkos abstract machine model targets a design of a future shared-memory computing architectures.
-The design is shown in Figure~\ref{fig:execspace}.
-It is characterized by multiple latency-optimized execution units and off-die bandwidth-optimized accelerators.
-Both compute device types have disjoint memory address spaces and unique performance properties.
-In such an architecture,   execution units might be grouped into larger units with multiple levels of achievable parallelism with different memory access characteristics and coherence properties of caches.
-In order to ensure performance portability on such a wide range of configurations, an abstraction of the compute engines and available memories are required.
-For this purpose we introduce \emph{spaces}.
-
-An instance of an \emph{execution space} is an abstraction over an execution resource to which a programmer can target parallel work.
-For example, an execution space can be used to describe a multi-core processor.
-In this case, the execution space contains several homogeneous cores organized into arbitrary logical groups.
-In a parallel programming model that implements this machine model, an instance of such an execution space would be made available to the programmer to run kernels.
-Adding more logical groups or accelerators simply increases the number of available execution spaces to the programmer.
-
-Memory and memory types are exposed through \emph{memory spaces}.
-Each memory space provides a finite storage capacity at which data structures can be allocated and accessed.
-Different memory space types have different characteristics with respect to accessibility from execution spaces and performance.
-
-An instance of a memory space provides a concrete method for the application programmer to request data storage allocations.
-Following the architecture as discussed earlier, the multi-core processor may contain multiple memory spaces that abstract on-package memory, slower DRAM and non-volatile memories.
-Accelerators can provide an additional memory space through its local on-package memory.
-An architectural layout is shown in Figure~\ref{fig:memspace}.
-The programmer is free to decide where each data structure is allocated by requesting the corresponding memory space.
-Programmatically this is achieved by instantiating the according memory space.
-Kokkos provides the appropriate abstraction of memory allocation and data management.
-We believe that an abstract representation of execution- and memory devices is a key property towards performance-portable parallel programming.
-
-How a machine model is exposed to the programmer and what are the design considerations towards defining an appropriate semantic capture are discussed in the next section.
-
-%Ang, J.A., et.
-al., Abstract Machine Models and Proxy Architectures for Exascale Computing, 2014, Sandia National Laboratories and Lawrence Berkeley National Laboratory, DOE Computer Architecture Laboratories Project
-
-\begin{figure}
-\begin{subfigure}[b]{0.41\textwidth}
-\centering
-\includegraphics[width=\textwidth]{img/ExecSpaces.png}
-\caption{Execution space is an instantiation of a space that represents a processing device to which the programmer can target parallel code.}
-\label{fig:execspace}
-\end{subfigure}
-\hfill \break
-\begin{subfigure}[b]{0.41\textwidth}
-\centering
-\includegraphics[width=\textwidth]{img/MemSpaces.png}
-\caption{A memory space is an instantiation of a space that represents a memory location on which the programmer can allocate data.}
-\label{fig:memspace}
-\end{subfigure}
-\caption{Spaces represent conceptual building blocks of the abstract machine model in Kokkos.}
-\label{fig:spaces}
-\end{figure}
-
-\section{The Kokkos Programming Model}\label{chap:kokkosPM}
-
-The Kokkos programming model specifies a language, programming paradigm and an API.
-The API exposes the abstract machine model to the developer and defines a set of pattens and types that capture intents and properties of execution.
-This completes the semantic capture, discussed in Chapter~\ref{chap:background}, We detail all three components as follows.
-
-Kokkos defines a pure interface for C++ and uses C++ as a programming language.
-This has multiple reasons.
-It is our experience that application developers of scientific and high-performance computing applications at Sandia National Laboratories express the demand for C++ and a generic parallel programming model to support their codes.
-Following their estimates, porting efforts towards adapting to vendors, programming models, APIs and software releases represents a burden worth addressing.
-Further, the possibility to maintain pure C++ codes is appealing from the code development and debugging perspective.
-Lastly, from the programming model perspective, C++ offers template metaprogramming which is well suited to implement generic APIs and libraries.
-In this case, class specialization and templating allow for compile-type generated types and optimizations for a given hardware architecture.
-It is worth to mention that Kokkos avoids the use of pragma annotations for the sake of treating concurrency as a first-class concept in C++.
-
-Using the programming language, a parallel programming paradigm can be applied to simplify the process of thought.
-Kokkos supports the parallel patterns and task paradigms.
-Parallel patters covers \emph{for}, \emph{ scan} and \emph{reduction}.
-This allows the expression of concurrency over iterative, for-loop computable algorithms.
-To cover the class of while-loop computable algorithms (irregular algorithms), Kokkos implements the tasking paradigm.
-Tasks encapsulate work into units that may be executed in parallel to other tasks or sections of the program.
-Equipped with a language and set of paradigms, a particular set of abstractions can be defined to expose the abstract machine model and captures the intent of the programmer.
-
-\subsection{Kokkos Abstractions}
-
-\begin{figure}
-\begin{Verbatim}[frame=leftline]
-template <class ExecPolicy, class FunctorType>
-Kokkos::parallel_for(
-    const std::string& name, 
-    const ExecPolicy& policy, 
-    const FunctorType& functor);
-\end{Verbatim}
-\caption{The Kokkos parallel for loop class shows how template metaprogramming allows to specialize types.
-Its descriptive semantic offers the freedom to optimize execution on modern architectures with multiple degrees of concurrency where an Execution policy shapes the iteration space accordingly.}
-\label{fig:parallelForKokkos}
-\end{figure}
-
-To capture semantic information, that is the~\emph{what}, the~\emph{where} and the~\emph{how}, Kokkos introduces six abstractions:~\emph{execution spaces},~\emph{execution patterns},~\emph{execution policies},~\emph{memory spaces},~\emph{memory layout} and~\emph{memory traits}.
-These abstractions specify semantic information that enable to capture the programmer's intent and allow the runtime to efficiently map the program to any underlying hardware architecture.
-We list them as follows:
-\begin{itemize}
-	\item  An execution space is a place where code can be executed.
-On current hardware architectures this correspond to accelerators and CPUs and can include any compute device in the future.
-This abstraction supports remote compute devices in distributed memory scenarios as remote execution spaces.
-	\item Execution patterns expose the parallel programming paradigm.
-Supported patterns are \emph{parallel\_for} loop that executes the loop body in any order a specified amount of times, the \emph{parallel\_reduce} which combines a parallel\_for with a reduction operation, \emph{parallel\_scan} which combines a parallel\_for operation with a prefix or postfix scan, and \emph{task} which executes a single function potentially in parallel in respect to other tasks or code sections.
-	\item Execution policy shape the iteration space of a loop pattern. A simple execution policy is a range policy. It specifies that the loop body is executed once for each element in a range.
-	\item Memory spaces are the places where data resides.
-They specify physical locations of data as well as access characteristics.
-Different physical locations correspond to different device types such as high bandwidth memories, on-die scratch memories or non-volatile bulk storage.
-Different logical memory spaces allow for concepts such as memory in the CUDA programming model, which is accessible from the host and the CUDA accelerator.
-Memory spaces, similarly to execution spaces, conceptually support remote memory locations in distributed-memory scenarios.
-Furthermore, they encapsulate functionality such as consistency control and persistence scopes.
-	\item Layouts express the mapping from loop indices to address offsets.
-By adopting appropriate layouts for memory structures, an application can optimize data access patterns in a given algorithm.
-If an implementation provides polymorphic layouts (i.e.
-a data structure can be instantiated at compile or runtime with different layouts), architecture-dependent optimizations can be performed.
-	\item Memory traits specify how a data structure is accessed.
-Traits express usage scenarios such as atomic access, random access and streaming loads or stores.
-This allows the programming model to  optimize load and store operations.
-\end{itemize}
-
-\begin{figure}[h]
-\begin{Verbatim}[frame=leftline]
-Semantic information: Patterns (intent), Spaces, Layouts, 
-Policies and Traits
-Paradigm: Parallel patterns and tasking
-Language: C++
-\end{Verbatim}
-\caption{Semantic capture defined by the Kokkos programming model.}
-\label{fig:SemCaptureKokkos}
-\end{figure}
-
-Expressing concurrency with patterns and abstractions allows the compiler to generate optimized transformations.
-Further, a runtime library to map the execution to the parallel hardware.
-In the case of parallel patterns, the execution is unspecified and only promise deterministic results for the reduction and scan operations.
-This gives the freedom to the underlying abstraction layers to implement different mapping patterns on different hardware such as assignment of iterations to threads or vector lanes.
-We believe that these abstraction and their descriptive semantic represent key programming primitives and the way forward for native language support for parallel programming.
-
-Figure~\ref{fig:SemCaptureKokkos} shows the semantic capture defined by the Kokkos programming model. 
-Figure~\ref{fig:stack} shows abstraction layers in Kokkos and and how it fits in a parallel programming landscape.
-The next section shows an example application written in Kokkos and provides an insight into the runtime library.
-For the interested reader, the complete programming model specification can be accessed on-line\cite{pub:KOKKOS}.
-
-%\begin{center}
-%\begin{tabular}{ |c|c|c| } 
-% \hline
-% cell1 & cell2 & cell3 \\ 
-% cell4 & cell5 & cell6 \\ 
-% cell7 & cell8 & cell9 \\ 
-% \hline
-%\end{tabular}
-%\end{center}
 
@@
 \begin{figure}
-\centerline{\includegraphics[width=0.3\textwidth]{img/Stack.png}}
-\caption{Landscape}
-\label{fig:stack}
+\centerline{\includegraphics[width=0.3\textwidth]{img/Build.png}}
+\caption{Building workflow}
+\label{fig}
 \end{figure}

### 2020-01-23 src/4.ProgammingModel.tex
Bug fix.
Improved description on parallel_for semantics.

@@
 The functor can be defined using a C++ class, struct or lambda expression.
-We would like to highlight that the return semantic of the parallel_for is defined as \emph{potentially asynchronous}.
-This shows the descriptive nature of the Kokkos programming model.
-In this case, it depends on the specified execution space and weather it implements a synchronization point.
+We would like to highlight that the return semantic of the parallel\_for is defined as \emph{potentially asynchronous}.
+To guarantee a kernel has finished, a developer should call the fence of the execution space on which the kernel is being executed.
+Otherwise it depends on the specified execution space and weather this execution space implements a barrier.
+The semantic of the parallel_for class in Kokkos demonstrates the descriptive nature of the programming model.
 To the interested reader, the complete programming model specification can be accessed on-line\cite{pub:KOKKOS}.
-Figure~\ref{fig:abstractions} lists patterns and abstractions and structures them by information type in the semantic capture.
+Figure~\ref{fig:abstractions} summarizes patterns and abstractions and structures them by information type in the semantic capture.
 

### 2020-01-23 src/4.ProgammingModel.tex
Bug fix. Added sample code.

@@
 To guarantee a kernel has finished, a developer should call the fence of the execution space on which the kernel is being executed.
-Otherwise it depends on the specified execution space and weather this execution space implements a barrier.
-The semantic of the parallel_for class in Kokkos demonstrates the descriptive nature of the programming model.
+Otherwise it depends on the execution space where the loop executes and weather this execution space implements a barrier.
+The semantic of the parallel\_for class in Kokkos demonstrates the descriptive nature of the programming model.
 To the interested reader, the complete programming model specification can be accessed on-line\cite{pub:KOKKOS}.

### 2020-01-23 src/5.Backend.tex
Bug fix. Added sample code.

@@
 \begin{Verbatim}[frame=leftline]
-
 int main(int argc, char* argv[]) {
  Kokkos::initialize(argc,argv);
-
+ {
+  Foo foo;
   int N = atoi(argv[1]);
-
-   Foo foo;
-
-   Kokkos::parallel_for(Kokkos::TeamPolicy<Tag1>(N,Kokkos::AUTO), foo);
-   Kokkos::parallel_for("Loop2", Kokkos::TeamPolicy<Tag2>(N,Kokkos::AUTO), foo);
-
+  Kokkos::View<int*> a("A",N);  
+  Kokkos::parallel_for(Kokkos::RangePolicy<OpenMP>(N), 
+     KOKKOS_LAMBDA (const int & i)
+  {
+  	a(i)=i;
+  }); 
+ }
  Kokkos::finalize();
 }
-
 \end{Verbatim}

### 2020-01-23 src/2.GenericProgramming.tex
Language edits, caption edits

@@
 We call it the semantic capture.
+
+\begin{figure}
+\centerline{\includegraphics[width=0.45\textwidth]{img/Hollman.png}}
+\caption{The Hollman pic :) - placeholder}
+\label{fig}
+\end{figure}
 
@@
 \end{Verbatim}
-\caption{Declarative parallel programming provides uses pragma annotations to capture semantic information.}
+\caption{Imperative model tells the compiler to use simd and vectorize and tile and unroll and fuse. It lists specific optimizations that should be made}
 \label{figOMPLike}
@@
 \end{Verbatim}
-\caption{Imperative parallel programming uses a base language and relies on programming interfaces to capture information and a documentation that describe their semantics.}
+\caption{Declarative model expresses parallelism and relies on the model to map it effectively to the architecture}
 \label{figKokkosLike}

### 2020-01-23 src/2.GenericProgramming.tex
First draft of Hollman-Jan bridge

@@
 Some representatives are the OpenMP programming model~\cite{CITEOPENMP}, threading libraries, or languages like Erlang in which parallelism is a first-class citizen.
-In general, it is up to the parallel programming model to expose language constructs or library interfaces (APIs) such that they balance convenience to the programmer with semantics sufficient to map the program execution to the parallel computer hardware efficiently and correctly.
-Questions of what semantic information is needed, what paradigms to present and by what syntactical means span a design space worth taking a closer look at.
-Figure~\ref{figSemCapture} shows the three areas that constitute a parallel programming model and its design.
-We call it the semantic capture.
+In general, it is up to the parallel programming model to expose language constructs or library interfaces (APIs) such that they combine convenience to the programmer with semantics sufficient to map the program execution to the parallel computer hardware efficiently and correctly.
+
+In designing Kokkos with this mentality, we looked at a broad variety of current and future architectures on which software would need to run.
+We also looked at our codes, which express many different algorithmic property.
+We conceived of Kokkos as a generic programming system that could cover the space of architectures we cared about, while exposing semantics in which the algorithms of our codes could be easily expressed.
+In this way, a generic programming system embedded in the language became the glue between the semantics of a problem domain and the capabilities of an architecture domain.
 
@@
 
+Earlier we discuss how a programming model constrains semantics to allow the expression of intent in such a way that a developer achieves their goals.
+Putting together an effective programming model requires constraining semantics such that they enable the necessary parallel paradigms, and expose to users only those aspects of the underlying hardware that they need to think about.
+We call the combination of constrained semantics, parallel paradigms, and division of responsibility between developer and framework the Semantic Capture.
+
 \begin{figure}[h]
 \begin{Verbatim}[frame=leftline]
-Semantic information
-Paradigm 
-Language
+Constrained Semantics
+Parallel Paradigms 
+Division of Responsibility
 \end{Verbatim}
-\caption{Semantic capture - a triplet of language, paradigm and semantic information defined by a parallel programming model allow the programmer to express concurrency and the compiler and runtime system to understand and run applications on a parallel machine.}
+\caption{Semantic capture - a triplet of constrained semantics, parallel paradigms, and division of responsibilities defined by a parallel programming model to allow developers to run performance-portable code on a set of architectures}
 \label{figSemCapture}

### 2020-01-23 src/5.Backend.tex
Added examples in Backend.

--- 2020-01-23 src/5.Backend.tex
@@
 \subsection{Back-end Support}\label{chap:kokkosBackend}
+
+Explain fence guard, KOKKOS\_LAMBDA,
 
@@
 \begin{figure}
+\begin{small}
+\begin{Verbatim}[frame=leftline]
+int main(int argc, char* argv[]) {
+  Kokkos::initialize(argc,argv);
+  {
+    int N = atoi(argv[1]);
+    Kokkos::View<int*> a("A", N);  
+    Kokkos::parallel_for(
+    Kokkos::RangePolicy<Kokkos::OpenMP>(N,
+    KOKKOS_LAMBDA (const int & i)
+    {
+      a(i)=i;
+    }); 
+  }
+  Kokkos::finalize();
+}
+\end{Verbatim}
+\end{small}
+\caption{An application written in Kokkos shows the use of a~\emph{view}-type and parallel for loop.}
+\label{fig:KokkosExample}
+\end{figure}
+
+\begin{figure}
+\begin{small}
 \begin{Verbatim}[frame=leftline]
 
-int main(int argc, char* argv[]) {
-   Kokkos::initialize(argc,argv);
+template <class FunctorType, class... Traits>
+class ParallelFor<FunctorType, 
+    Kokkos::RangePolicy<Traits...>, 
+    Kokkos::OpenMP> 
+{
+  ...
+private:
+  ... exec_range(const FunctorType& functor, 
+  const Member i_beg,
+  const Member i_end) 
+  {
+    for (Member i = i_beg; iwork < i_end; ++i){
+      functor(t, i);
+    }
+  }
 
-   int N = atoi(argv[1]);
-
-   Foo foo;
-
-   Kokkos::parallel_for(Kokkos::TeamPolicy<Tag1>(N,Kokkos::AUTO), foo);
-   Kokkos::parallel_for("Loop2", Kokkos::TeamPolicy<Tag2>(N,Kokkos::AUTO), foo);
-
-   Kokkos::finalize();
+public:
+  ...
+  ... execute() const {(
+  #pragma omp parallel num_threads(OpenMP::getPoolSize())
+  {
+    HostThreadTeamData& data=*(i->get_thread_data());
+    ...
+    range = data.get_work_partition();
+    range(m_functor, range.first + m_policy.begin(),
+    range.second + m_policy.begin());
+  }
 }
@@
 \end{Verbatim}
-\caption{Declarative parallel programming provides uses pragma annotations to capture semantic information.}
-\label{figOMPLike}
+\end{small}
+\caption{An application written in Kokkos shows a paralle for loop that .}
+\label{fig:KokkosExample}
 \end{figure}

### 2020-01-23 src/1.Introduction.tex
Further further further (this one was my fault)

@@
 
-Parallel and distributed computing moved into focus at universities and teaching institutions.
+Parallel and distributed computing is a focus at universities and teaching institutions.
 The interest is two-fold.
-Firstly, given their role, educators adapt to trends in research and industry and include parallel computing in their curricula, and secondly, parallel computing is an attractive field for department research and industry collaborations.
+Firstly, given their role, educators adapt to trends in research and industry and include parallel computing in their curricula.
+Secondly, parallel computing is an attractive field for department research and industry collaborations.
 At Sandia, we develop a performance portability library called Kokkos that is ideal for educators looking to teach parallel computing.
@@
 Unfortunately, not all architectures support all programming models, nor do many programming models cleanly express all parallel programming constructs.
-Many vendor models compound this, requiring students have access to specific hardware, and limiting the concepts of parallelism to those supported by that architecture.
+Many vendor models compound this, requiring students to have access to specific hardware, and limiting the concepts of parallelism to those supported by that hardware.
 This makes the choice of programming model important and difficult.
@@
 Consequently, generic parallel programming models are needed that target mainstream programming languages and allow a vendor- and architecture independent, performance-portable expression of concurrency.
-In this way, a parallel program written in a popular language becomes capable of running efficiently on parallel computer architectures that conform to a common parallel machine model specification.
-We would like to compare this to the support of Von Neumann architectures in programming languages today.
-We define an abstract machine model later in this work.
+In this way, a programming exercise for students becomes capable of running efficiently on parallel computer architectures that conform to a common parallel execution model.
+This compares to the support of Von Neumann architectures in programming languages today.
 The lack of such parallel programming models motivated us to develop Kokkos.
@@
 It consists of a programming model specification and a library.
-The specification defines execution and memory model and an application programming interface (library API).
+The specification defines an execution and memory model and an application programming interface (library API).
 The runtime library exposes functions representing parallel patterns and types for data management and maps them to the underlying abstract execution model.
-This execution model creates a generic view of compute hardware and memories.
-The Kokkos library maps execution primitives to vendor-specific programming models, libraries, and memory layouts.
+The Kokkos library maps those patterns to vendor-specific programming models, libraries, and memory layouts.
 Kokkos has been successfully used in many large-scale projects at SNL, LLNL and other research centers~\cite{CITEKOKKOSUSECASES}.
@@
 \item Parallel programming model for C++
-\item Targets an abstract machine model
+\item Targets an abstract execution model
 \item Implements parallel patterns and tasking paradigms
@@
 We show an abstraction with enough semantic information to produce performance-portable, parallel code on current architectures.
-Further, we highlight its alignment and contributions to standardization efforts towards supporting parallel programming in C++.
+We highlight its alignment and contributions to standardization efforts towards supporting parallel programming in C++.
 

### 2020-01-23 src/5.Backend.tex
First draft of Chapter 5.

@@
 
-Explain fence guard, KOKKOS\_LAMBDA,
+This section provides a practical insight into programming with Kokkos and the inner workings of the programming model library.
+Figure~\ref{fig:KokkosExample} shows am example application that algorithmically initializes an array to integers and includes several points of interests.
+Taking a closer look, the reader can observe the initialization and finalization of the Kokkos runtime library through the~\emph{Kokkos::initialize} and~\emph{Kokkos::finalize} method calls.
+Further the example shows the use of the~\emph{view} abstraction that represents a one-dimensional integer array of size~\emph{N}.
+The object carries the name~\emph{A} internally to facilitate tracing and debugging.
+In this example, a parallel\_for pattern object receives a range policy that is templated to the OpenMP execution space and takes~\emph{N} and a lambda expression as parameters.
+Vies behave as pointers and there a copy-semantic in the lambda expression is used in this case.
+Within the loop-body, the view object is accessed through the~\emph{()-operator}.
+It is up to the OpenMP execution space to provide performance implementations of the view, an efficient internal representation of the array and for the access operator for the underlying hardware architecture.
 
-The important consideration here is that the method of compiling code for different execution spaces and the dispatch of kernels to instances is abstracted by the Kokkos model.
-This unburdens application programmers from writing algorithms in hardware specific languages.
+At compile time, the compiler generates the particular types, where in this case, the parallel\_for is types for the OpenMP execution space.
+At link time, the linker links this method to the corresponding parallel\_for implementation in the Kokkos library.
+
+The implementation of the parallel\_for specific to the OpenMP execution space is shown in Figure~\ref{fig:KokkosExampleOMPBackEnd}.
+It shows the template argument of that class predefined to that execution space and shows how the~\emph{execute()} method calls the functor by calling ~\emph{exec\_range}.
+In can be seen that the Kokkos library relies on the back-end compiler to generate platform optimized, parallel code.
 
@@
     Kokkos::RangePolicy<Kokkos::OpenMP>(N,
-    KOKKOS_LAMBDA (const int & i)
+    [=](const int & i)
     {
-      a(i)=i;
+      a(i)=i*i;
     }); 
@@
 \begin{Verbatim}[frame=leftline]
-
 template <class FunctorType, class... Traits>
 class ParallelFor<FunctorType, 
-    Kokkos::RangePolicy<Traits...>, 
-    Kokkos::OpenMP> 
-{
+    ..., 
+    Kokkos::OpenMP> {
   ...
 private:
-  ... exec_range(const FunctorType& functor, 
+  ... cconst FunctorType& functor, 
   const Member i_beg,
@@
 \end{small}
-\caption{An application written in Kokkos shows a paralle for loop that .}
-\label{fig:KokkosExample}
+\caption{Library code giving an  insight into the impelemtnation of the OpenMP back-end shows ParallelFor class, its~\emph{execute} method and the internal use of the OpenMP pragma annotation.
+internally gives an insight into the m maos how .}
+\label{fig:KokkosExampleOMPBackEnd}
 \end{figure}
-
-%\begin{figure}
-%\begin{Verbatim}[frame=leftline]
-%double f(){
-% typedef Kokkos::Cuda ES;
-% typedef Kokkos::CudaSpace MS;
-% typedef Kokkos::LayoutLeft LL;
-% typedef Kokkos::RangePolicy<ES> Range_policy;
-% typedef Kokkos::View<double*, LL, MS> V
-% V A( "A", N);
-% V::HostMirror h_A = Kokkos::create_mirror_view( A );
-% for ( int i = 0; i < N; ++i ) //Initialize on host
-%   h_A( i ) = 1;
-% Kokkos::deep_copy( A, h_A );
-% double result = 0;
-% Kokkos::parallel_reduce( "yAx", range_policy( 0, N ), 
-%  KOKKOS_LAMBDA ( int i, double &update ) {
-%   update += A( i );
-% }, result );
-% return result;
-%}
-%\end{Verbatim}
-%\caption{Declarative parallel programming provides uses pragma annotations to capture semantic information.}
-%\label{figOMPLike}
-%\end{figure}
 
@@
 \centerline{\includegraphics[width=0.3\textwidth]{img/Build.png}}
-\caption{Building workflow}
-\label{fig}
+\caption{The compilation workflow includes the invocation of the platform compiler that is in charge to generate parallel   Building workflow}
+\label{fig:workflow}
 \end{figure}

### 2020-01-23 src/5.Backend.tex
Chapter on Backend support done.

@@
-\subsection{Back-end Support}\label{chap:kokkosBackend}
+\section{Back-end Support}\label{chap:kokkosBackend}
 
@@
 The implementation of the parallel\_for specific to the OpenMP execution space is shown in Figure~\ref{fig:KokkosExampleOMPBackEnd}.
-It shows the template argument of that class predefined to that execution space and shows how the~\emph{execute()} method calls the functor by calling ~\emph{exec\_range}.
+It shows the template argument of that class predefined to that execution space and shows how the~\emph{execute()} method calls the functor through ~\emph{exec\_range}.
 In can be seen that the Kokkos library relies on the back-end compiler to generate platform optimized, parallel code.
+For the OpenMP back-end implementation of the parallel for loop, OpenMP pragma annotations are used.
+Since the parallel for implies a barrier at the end of the parallel region, the return semantic of Kokkos::parallel\_for call is synchronous.
 
@@
     Kokkos::RangePolicy<Kokkos::OpenMP>(N,
-    [=](const int & i)
-    {
+    [=](const int & i) {
       a(i)=i*i;
@@
 \end{small}
-\caption{An application written in Kokkos shows the use of a~\emph{view}-type and parallel for loop.}
+\caption{This example application written in Kokkos shows the use of a~\emph{view}-type and a parallel for loop. Both Kokkos types are templated to the OpenMP execution space.}
 \label{fig:KokkosExample}
@@
   ... execute() const {(
-  #pragma omp parallel num_threads(OpenMP::getPoolSize())
+#pragma omp parallel num_threads(OpenMP::gPoolSize())
   {
-    HostThreadTeamData& data=*(i->get_thread_data());
+    HostThreadTeamData& data=*(i->get_thr_data());
     ...
@@
 \end{small}
-\caption{Library code giving an  insight into the impelemtnation of the OpenMP back-end shows ParallelFor class, its~\emph{execute} method and the internal use of the OpenMP pragma annotation.
-internally gives an insight into the m maos how .}
+\caption{This code shows the back-end implementation of a parallel loop in the Kokkos programing model.
+It shows a particular implementation of that generic pattern as a templated class to the OpenMP execution space.
+Also it shows the use of OpenMP pragma annotations that put platform compilers in charge to generate parallel code for the underlying hardware architecture.}
 \label{fig:KokkosExampleOMPBackEnd}
@@
 
-\begin{figure}
-\centerline{\includegraphics[width=0.3\textwidth]{img/Build.png}}
-\caption{The compilation workflow includes the invocation of the platform compiler that is in charge to generate parallel   Building workflow}
-\label{fig:workflow}
-\end{figure}
+%\begin{figure}
+%\centerline{\includegraphics[width=0.3\textwidth]{img/Build.png}}
+%\caption{The compilation workflow includes the invocation of the platform compiler that is in charge to generate parallel   Building workflow}
+%\label{fig:workflow}  
+%\end{figure}

### 2020-01-23 src/6.Tutorials.tex
Renamed files for consistency.

\section{Tutorials}\label{chap:tutorial}

Describing here the tutorial app and the example apps that are included either in the repo or in the Kokkos distribution.

### 2020-01-23 src/2.GenericProgramming.tex
Stab at point one of the "semantic capture" point

@@
 
-Semantic information provided by the programmer to the parallel programming model can be grouped into information to express intent and properties.
+%Semantic information provided by the programmer to the parallel programming model can be grouped into information to express intent and properties.
 They address the question of \emph{what}, \emph{where} and \emph{how}.
@@
 Semantics are expressed following a paradigm and a programming language.
+
+To enable developers to achieve performance portability, we needed to constrain what intent they could specify to Kokkos.
+We settled on Kokkos supporting semantics for offloading loops, the allocation and freeing of data, and data movement.
+Critically, we decided against exposing machine details to the developer, we aim to increase portability of code by only exposing an abstract execution model (described later) to the users.
+Our semantics map naturally to C-like languages, with malloc/new calls replaced with the construction and destruction of abstract data types, and many loop types replaced with corresponding parallel patterns.
+This clean mapping between traditional C-like semantics and our model came from requirements from our physics teams, but also makes Kokkos a good choice for students as it will more closely track with what they've seen before.
+
+%TODO DZP: keep going here
 

### 2020-01-23 src/2.GenericProgramming.tex
Finished out the "triad" text

@@
 
-The parallel programming paradigm is an abstract ~\emph{representation} with the purpose of facilitating the programmer's understanding of programming rules and program behavior.
+%The parallel programming paradigm is an abstract ~\emph{representation} with the purpose of facilitating the programmer's understanding of programming rules and program behavior.
 Which programming paradigm to chose depends on several considerations.
@@
 
-Lastly, a model must decide between declarative and imperative semantics.
+The paradigm most central to Kokkos are standard parallel patterns.
+The three main primitives provided to developers are \emph{map}, \emph{reduce}, and \emph{scan}, with some ability for composition.
+We found that these could pretty easily be mapped to a wide variety of architectures, and that many algorithms can be expressed in terms of these building blocks.
+Many software engineering curricula already cover patterns, for many students this will resonate immediately.
+For those that don't, this provides an excuse to discuss another important topic in the field.
+
+Lastly, a model must divide responsibility between.
+In practice, this is largely a choicebetween declarative and imperative semantics.
 Declarative semantics require a developer to express intent, implementation is the responsibility of the programming model and underlying toolchain.

### 2020-01-23 src/2.GenericProgramming.tex
With edits via Jan

@@
 \begin{Verbatim}[frame=leftline]
-Constrained Semantics
-Parallel Paradigms 
-Division of Responsibility
+Constrained Semantics: memory allocations, memory movement, parallel kernels
+Parallel Paradigms: Patterns, Tasks
+Declarative vs Imperative Semantics: Declarative
 \end{Verbatim}
@@
 
-Lastly, a model must divide responsibility between.
-In practice, this is largely a choicebetween declarative and imperative semantics.
+Lastly, a model must divide responsibility between the developer and the model.
+In practice, this is largely a choice between declarative and imperative semantics.
 Declarative semantics require a developer to express intent, implementation is the responsibility of the programming model and underlying toolchain.

### 2020-01-23 src/6.Tutorials.tex
Updates to Tutorials section.

@@
-\section{Tutorials}\label{chap:tutorial}
+%\section{Tutorials}\label{chap:tutorial}
 
-Describing here the tutorial app and the example apps that are included either in the repo or in the Kokkos distribution.
+The Kokkos programming model distribution includes many examples and benchmark applications as well as tutorial and presentation material.
+Example applications demonstrate the use of each abstraction as well supported paradigms.
+Examples cover the use of unified virtual memories on CUDA devices and object polymorphism in Kokkos (virtual functions).
+The tutorial inclemently introduces the interested reader to Kokkos abstractions and patterns.
+For the purpose of education each incremental step includes two code version.
+One code version represents the programming assignment and the other is the solution.
+The assignment includes instructions and marked code sections that require completion.
+Benchmark applications provide further insight into performance implications and allows for easy experimentation.
+Further resources are online.
+This includes presentation material as well as tutorials on application profiling and the Kokkos Kernels numerical library.

### 2020-01-23 src/6.Tutorials.tex
Small changes.

@@
 
-The Kokkos programming model distribution includes many examples and benchmark applications as well as tutorial and presentation material.
+The Kokkos programming model distribution~\cite{CITEKOKKOSGIT} includes many examples and benchmark applications as well as tutorial and presentation material.
 Example applications demonstrate the use of each abstraction as well supported paradigms.
@@
 Benchmark applications provide further insight into performance implications and allows for easy experimentation.
-Further resources are online.
-This includes presentation material as well as tutorials on application profiling and the Kokkos Kernels numerical library.
+Further resources are online\cite{CITEKOKKOSTUTORIALS}.
+They include presentation material as well as tutorials on application profiling and the Kokkos Kernels numerical library.

### 2020-01-23 src/8.Related.tex
Small changes.

@@
 \section{Related Work}\label{chap:related}
-\subsection{RAJA}
-Like Kokkos, RAJA is a C++ framework for porting massive DOE codes to a variety of architectures.
+Like Kokkos, RAJA~\cite{RAJA} is a C++ framework for porting massive DOE codes to a variety of architectures.
 In both frameworks, a developer first expresses the operations their kernel performs using simple C++ functional programming mechanisms.
@@
 Which model makes more sense in a given circumstance depends on how much exposure to the underlying programming model a developer wants.
-RAJA would be ideal for a class in which the educator wants to expose students to the semantics of different programming models without exposing them to the detailed syntax of those models.
-\subsection{DPC++}
-DPC++ is a model developed by Intel(TM) for expressing parallelism across their architectures.
-In addition to sharing a declarative philosophy with Kokkos, it even share a number of patterns.
-Unlike Kokkos, at time of publication it targets only Intel architectures (it isn't vendor-agnostic) For a course targeting Intel(TM) platforms, you're likely to see good tool support, DPC++ could be a valid option.
-\subsection{Parallel Constructs in the C++ Standard}
-C++ increasingly supports parallel constructs.
-We discuss the contributions of Kokkos to C++ in~\ref{chap:c++}, it's increasingly possible to teach parallelism with nothing but a C++ compiler.
-A problem is that at time of publication C++ doesn't support accelerators, or many of the complexities of parallel computing as it's used in practice.
-A positive is that instead of requiring libraries, you just need a modern C++ compiler.
-For a course in which topics like accelerators aren't discussed, this can be a valid option.
+RAJA is a viable option for class education in which the educator wants to expose students to the semantics of different programming models without exposing them to the detailed syntax of those models.
+
+DPC++\cite{DPCPP} is a model developed by Intel\textsuperscript{\textregistered} for expressing parallelism across Intel architectures.
+In addition to sharing a declarative philosophy with Kokkos, it also shares parallel patterns.
+Unlike Kokkos, at time of publication it targets only Intel architectures and is not vendor-agnostic.
+A course targeting Intel architectures, DPC++ could be a valid option.
+
+C++ increasingly supports parallel constructs\cite{CPP}.
+We have highlighted contributions of Kokkos to C++ in~\ref{chap:c++}.
+It is becoming increasingly possible to teach parallelism in C++ and relying on modern C++ compilers.
+However, at time of publication, C++ does not support accelerators, or many of the complexities of parallel computing related to accelerator programming.
+In courses that do not target accelerators, this can be a relevant option.
 %\end{itemize}

### 2020-01-24 src/1.Introduction.tex
Fused DZP and JC changes for Chapters 1 and 2.

--- 2020-01-23 src/1.Introduction.tex
@@
 
-Parallel and distributed computing moved into focus at universities and teaching institutions.
+Parallel and distributed computing is a focus at universities and teaching institutions.
 The interest is two-fold.
 Firstly, given their role, educators adapt to trends in research and industry and include parallel computing in their curricula, and secondly, parallel computing is an attractive field for department research and industry collaborations.
-At Sandia, we develop a performance portability library called Kokkos that is ideal for educators looking to teach parallel computing.
 
-In teaching a class on parallel programming, educators must balance considerations of architecture, programming model, and theoretical concepts.
-Unfortunately, not all architectures support all programming models, nor do many programming models cleanly express all parallel programming constructs.
-Many vendor models compound this, requiring students have access to specific hardware, and limiting the concepts of parallelism to those supported by that architecture.
-This makes the choice of programming model important and difficult.
+While industry-driven collaborations often predefine technological frameworks such as programming environments and targeted architectures, academic teaching leaves this mostly undefined.
+In this case, educators are exposed to a variety of solutions, approaches and trends in parallel computing.
+Which particular set to choose for teaching in class is not trivial.
+While decision factors differ, maximizing the subject relevance by teaching generic principles of parallel programming generically is among the most important ones.
+
+Teaching generic parallel programming today is difficult.
+Educators strive to balance considerations of architecture, programming model, and theoretical concepts.
+Unfortunately, not all architectures support all programming models, nor do programming models offer enough abstractions to support modern hardware architectures sufficiently .
+Many vendor programming models compound this, requiring students to have access to specific hardware, and limiting the concepts of parallelism to those supported by that hardware.
 
 Consequently, generic parallel programming models are needed that target mainstream programming languages and allow a vendor- and architecture independent, performance-portable expression of concurrency.
-In this way, a parallel program written in a popular language becomes capable of running efficiently on parallel computer architectures that conform to a common parallel machine model specification.
+In this way, a parallel program written in a popular language becomes capable of running efficiently on parallel computer architectures that conform to a common parallel machine specification.
 We would like to compare this to the support of Von Neumann architectures in programming languages today.
@@
 
-Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS_PAPER_HERE}, is a performance-portable, parallel programming model for C++ developed at Sandia National Laboratories (SNL) for developing million line scientific software on the world\'s biggest supercomputers.
+Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS_PAPER_HERE}, is a performance-portable, parallel programming model for C++ developed at Sandia National Laboratories (SNL).
 It consists of a programming model specification and a library.
 The specification defines execution and memory model and an application programming interface (library API).
-The runtime library exposes functions representing parallel patterns and types for data management and maps them to the underlying abstract execution model.
-This execution model creates a generic view of compute hardware and memories.
+The runtime library exposes programming primitives, namely, patterns and types and maps them to the underlying abstract machine model.
+The machine model creates a generic view of compute hardware and memories.
 The Kokkos library maps execution primitives to vendor-specific programming models, libraries, and memory layouts.
 Kokkos has been successfully used in many large-scale projects at SNL, LLNL and other research centers~\cite{CITEKOKKOSUSECASES}.
-Its key properties are
+We list its key properties as follows.
 
@@
 Further, we would like to present Kokkos to the educational community as one representative parallel programming model for a trend towards architecture-independent and performance-portable programming of CPUs and accelerators.
-We show an abstraction with enough semantic information to produce performance-portable, parallel code on current architectures.
+We show a set of types that implement patterns and encapsulate data with enough semantic information to produce performance-portable, parallel code on current architectures.
 Further, we highlight its alignment and contributions to standardization efforts towards supporting parallel programming in C++.
 
-Secondly, we would like to promote Kokkos to students and educators as a platform to understand and experiment with parallel patterns in a fashion that is
-
-\begin{itemize}
-\item Generic
-\item Vendor-agnostic
-\item Declarative
-\end{itemize}
-
-First, Kokkos has semantics that are generic.
-This means that students are learning parallelism detached from any architectural concerns.
-Kokkos won\'t require them to know the specifics of the \"symmetric multiprocessors\" of a GPU or the cache hiearchy on a CPU\footnote{Though nothing precludes an educator from teaching these independently of the model}.
-Finally, Kokkos is vendor agnostic.
-If you have a computer lab with processors from one vendor and your students have processors from another, this presents no problem, Kokkos likely works on both.
-Kokkos is declarative, students aren\'t saying \"map these work items to this set of threads\" but \"do a reduction over this set of data.\" This is critical, much of software engineering is moving to declarative models and languages, learning to operate in such a mindset is useful outside of parallel programming.
+Secondly, we would like to promote Kokkos to students and educators as a platform that helps them to understand and experiment with parallel patterns in a fashion that is generic, vendor-agnostic and declarative.
+This means that students can learn about parallelism without being exposed to many architectural concerns and that they can see benefits of such abstractions for performance portability when programming on current and upcoming hardware architectures.
 

### 2020-01-24 src/2.GenericProgramming.tex
Fused DZP and JC changes for Chapters 1 and 2.

--- 2020-01-23 src/2.GenericProgramming.tex
@@
 
-A programming model exposes constrained semantics to the programmer which that programmer uses to express intent.
+A programming model exposes constrained semantics to the programmer which the programmer uses to express intent.
 Absent an embedded programming model, the semantics of C-like languages can lead to sequential thinking and execution.
@@
 In general, it is up to the parallel programming model to expose language constructs or library interfaces (APIs) such that they combine convenience to the programmer with semantics sufficient to map the program execution to the parallel computer hardware efficiently and correctly.
+Questions of which semantic information is needed, what paradigm to present and by what syntactical means span an exciting design space worth exploring.
+From a theoritecal perspective
 
-In designing Kokkos with this mentality, we looked at a broad variety of current and future architectures on which software would need to run.
-We also looked at our codes, which express many different algorithmic property.
-We conceived of Kokkos as a generic programming system that could cover the space of architectures we cared about, while exposing semantics in which the algorithms of our codes could be easily expressed.
-In this way, a generic programming system embedded in the language became the glue between the semantics of a problem domain and the capabilities of an architecture domain.
-
-\begin{figure}
-\centerline{\includegraphics[width=0.45\textwidth]{img/Hollman.png}}
-\caption{The Hollman pic :) - placeholder}
-\label{fig}
-\end{figure}
-
-Earlier we discuss how a programming model constrains semantics to allow the expression of intent in such a way that a developer achieves their goals.
-Putting together an effective programming model requires constraining semantics such that they enable the necessary parallel paradigms, and expose to users only those aspects of the underlying hardware that they need to think about.
-We call the combination of constrained semantics, parallel paradigms, and division of responsibility between developer and framework the Semantic Capture.
+ Figure~\ref{figSemCapture} shows the three  areas that constitute a parallel programming model and its design. We call it the semantic capture.
 
@@
 \begin{Verbatim}[frame=leftline]
-Constrained Semantics
-Parallel Paradigms 
-Division of Responsibility
+Constrained Semantics: memory allocations, memory movement, parallel kernels
+Parallel Paradigms: patterns, tasks
+Expression: language 
+Concern: prescriptive, descriptive
 \end{Verbatim}
-\caption{Semantic capture - a triplet of constrained semantics, parallel paradigms, and division of responsibilities defined by a parallel programming model to allow developers to run performance-portable code on a set of architectures}
+\caption{Semantic capture - a triplet of constrained semantics, parallel paradigms, and expression define by a parallel programming model to allow developers to run performance-portable code on a set of architectures}
 \label{figSemCapture}
@@
 
-%Semantic information provided by the programmer to the parallel programming model can be grouped into information to express intent and properties.
+Constrained semantics provided by the programmer to the parallel programming model can be grouped into information to express intent and properties.
 They address the question of \emph{what}, \emph{where} and \emph{how}.
@@
 Semantics are expressed following a paradigm and a programming language.
-
-To enable developers to achieve performance portability, we needed to constrain what intent they could specify to Kokkos.
-We settled on Kokkos supporting semantics for offloading loops, the allocation and freeing of data, and data movement.
-Critically, we decided against exposing machine details to the developer, we aim to increase portability of code by only exposing an abstract execution model (described later) to the users.
-Our semantics map naturally to C-like languages, with malloc/new calls replaced with the construction and destruction of abstract data types, and many loop types replaced with corresponding parallel patterns.
-This clean mapping between traditional C-like semantics and our model came from requirements from our physics teams, but also makes Kokkos a good choice for students as it will more closely track with what they've seen before.
-
-%TODO DZP: keep going here
 
@@
 
-Lastly, a model must decide between declarative and imperative semantics.
-Declarative semantics require a developer to express intent, implementation is the responsibility of the programming model and underlying toolchain.
-Imperative semantics require the developer to specify the mechanisms by which their intent will be implemented.
-In an imperative model, developer success is tied to their knowledge of implementation mechanisms.
-In a declarative model that success is tied to the developer's ability to accurately declare their requirements, and the toolchain's ability to turn that into an implementation.
-Figures~\ref{figOMPLike} and~\ref{figKokkosLike} show examples of two sample applications using the aforementioned types of semantics.
-While both programming models implement the same programming paradigm (parallel patterns), Figure~\ref{figOMPLike} required the developer to express how to parallelize the loop, while in~\ref{figKokkosLike} the developer just declared their parallel semantics, and relied on the programming model to pick an implementation.
+Lastly, a language defines the syntax of the parallel semantic.
+Annotations-based  languages add semantic information to the base language through pragmas while embedded programming models rely on the base language and offer additional programming interfaces to capture information.
+Their semantic is defined in the API specification.
 
-Kokkos takes the declarative approach.
-For the developers of our large scientific applications, this is appealing because it allows them to express their algorithms in one way, and rely on Kokkos to map that intent effectively to a variety of architectures.
-For students this is appealing because they're unlikely to have expertise in these architectures.
-For an educator looking for an implementation vehicle with which to teach students about parallelism, the declarative model eliminates the burden of reimplementing exercises as computer labs and dominant architectures change.
-For a declarative model to make sense, the educator should know something about the abstract execution model being used to ensure performance portability.
-Depending on the nature of the course, this could be mostly hidden from students, or discussed in detail.
+Lastly, a programming model must divide responsibility between the developer and the model.
+We refer to is as defining the concern.
+In practice, this is a choice between defining a descriptive and prescriptive semantics.
+Descriptive semantics require a developer to express intent.
+The implementation is the responsibility of the programming model and of the underlying toolchain.
+Prescriptive semantics require the developer to specify the mechanisms by which their intent will be implemented.
+In an prescriptive model, developer success is tied to their knowledge of implementation mechanisms.
+In a descriptive model that success is tied to the developer's ability to accurately describe their requirements and the toolchain's ability to turn that into an implementation.
+
+Figures~\ref{figOMPLike} and~\ref{figKokkosLike} show examples of two sample applications using the aforementioned parallel programming syntaxes.
+While both programming models implement the same programming paradigm (parallel patterns), Figures~\ref{figOMPLike} extends the base programming language through pragma annotations.
+Both, pragma annotations and an API call as shown in Figure~\ref{figKokkosLike} can have a descriptive or prescriptive semantic of describing a parallel loop construct.
+It is up to the programming model specification to make further definitions regarding separation of concern.
+
+Both examples provide the semantic information on what to parallelize, however, they do not expose enough abstraction primitives that would allow to map parallel execution on modern architectures.
+In order to define what further information is needed to express concurrency on modern computer architectures, it is important to define an~\emph{abstract machine representation} first.
+In the following chapters we use the words \emph{representation} and \emph{model} interchangeably.
+In this work the abstract machine model does not describe a computer in terms of automata theory but in terms of physical or functional components.
 
@@
 \begin{Verbatim}[frame=leftline]
-# pragma model parallel for simd vectorize tile unroll fuse
+# pragma model parallel loop
 for ( size_t i = 0; i < N; ++i) {
@@
 \end{Verbatim}
-\caption{Imperative model tells the compiler to use simd and vectorize and tile and unroll and fuse. It lists specific optimizations that should be made}
+\caption{Annotations-based programming models extend the semantic of the base language through pragma annotations.}
 \label{figOMPLike}
@@
 \end{Verbatim}
-\caption{Declarative model expresses parallelism and relies on the model to map it effectively to the architecture}
+\caption{Embedded parallel programming uses the base language and relies on programming interfaces to capture information.}
 \label{figKokkosLike}

### 2020-01-24 src/3.MachineModel.tex
Further renaming.
Fused DZP and JC changes for Chapters 1 and 2.

--- 2020-01-22 src/4.ExecutionModel.tex
@@
-\section{The Kokkos Execution Model Abstractions}\label{chap:kokkosEMA}
+\section{The Kokkos Machine Model}\label{chap:kokkosMM}
 
@@
 
-The Kokkos abstract machine model targets a design of a future shared-memory computing architectures.
+The Kokkos abstract machine representation targets a design of a future shared-memory computing architectures.
 The design is shown in Figure~\ref{fig:execspace}.

### 2020-01-24 src/2.GenericProgramming.tex
Changes to accodate semantic capture.

@@
 Questions of which semantic information is needed, what paradigm to present and by what syntactical means span an exciting design space worth exploring.
-From a theoritecal perspective
 
- Figure~\ref{figSemCapture} shows the three  areas that constitute a parallel programming model and its design. We call it the semantic capture.
+ Figure~\ref{figSemCapture} shows the four properties that constitute a parallel programming model and its design. We call it the semantic capture.
 
@@
 \end{Verbatim}
-\caption{Semantic capture - a triplet of constrained semantics, parallel paradigms, and expression define by a parallel programming model to allow developers to run performance-portable code on a set of architectures}
+\caption{Semantic capture - a set of constrained semantics, parallel paradigms, expression and concern define key properties a parallel programming model.}
 \label{figSemCapture}

### 2020-01-24 src/4.ProgammingModel.tex
Fused DZP and JC changes for Chapters 1 and 2.
Changes to accodate semantic capture.

@@
 The Kokkos programming model specifies a language, programming paradigm and an API.
-The API exposes the abstract machine model to the developer and defines a set of pattens and types that capture intents and properties of execution.
+The API exposes the machine representation to the developer and defines a set of pattens and types that capture intents and properties of execution.
 This completes the semantic capture, discussed in Chapter~\ref{chap:background}, We detail all three components as follows.
@@
 \begin{Verbatim}[frame=leftline]
-Semantic information: Patterns (intent), Spaces, Layouts, 
+Constrained Semantics: Patterns (intent), Spaces, Layouts, 
 Policies and Traits
-Paradigm: Parallel patterns and tasking
-Language: C++
+Parallel Paradigms: Parallel patterns and tasking
+Expression: Embedded, C++
+Concern: descriptive
 \end{Verbatim}

### 2020-01-24 src/2.GenericProgramming.tex
Formatting change.

@@
 \begin{Verbatim}[frame=leftline]
-Constrained Semantics: memory allocations, memory movement, parallel kernels
+Constrained Semantics: memory allocations, 
+memory movement, execution parameters
 Parallel Paradigms: patterns, tasks

### 2020-01-24 src/2.GenericProgramming.tex
first three paragraph edits

--- 2020-01-24 src/2.GenericProgramming.tex
@@
 
-A programming model exposes constrained semantics to the programmer which the programmer uses to express intent.
-Absent an embedded programming model, the semantics of C-like languages can lead to sequential thinking and execution.
-Parallel programming models expose parallel semantics to enable developers to create performance portable code.
-This is commonly achieved by adding semantic value to the base programming language, often referred to as programming with \emph{pragma} annotations.
+A programming model exposes constrained semantics that the programmer uses to express intent.
+Absent a programming model providing additional constraints, the semantics of C-like languages can lead to sequential thinking and execution.
+Parallel programming models expose parallel semantics to enable developers to create performance-portable code.
+This is commonly achieved by adding annotations to the base programming language, often achieved using \emph{pragma} clauses or similar language extensions.
 Other methods expose such semantics through library calls or new languages.
-All three approaches are conceptually suitable to represent programming paradigms such as tasking, parallel patterns and others.
+All three approaches are conceptually suitable to represent programming paradigms such as tasking, data parallelism, and other patterns.
 Some representatives are the OpenMP programming model~\cite{CITEOPENMP}, threading libraries, or languages like Erlang in which parallelism is a first-class citizen.
-In general, it is up to the parallel programming model to expose language constructs or library interfaces (APIs) such that they combine convenience to the programmer with semantics sufficient to map the program execution to the parallel computer hardware efficiently and correctly.
-Questions of which semantic information is needed, what paradigm to present and by what syntactical means span an exciting design space worth exploring.
+In general, it is up to the parallel programming model implementer to expose language constructs or library interfaces (APIs) that combine convenience to the programmer with sufficiently constrained semantics to map the program execution to the parallel computer hardware efficiently and correctly.
+Questions of which semantic information is needed, what paradigm to present, and by what syntactic means to do so span an exciting design space that is well worth exploring.
 
@@
 Expression: language 
-Concern: prescriptive, descriptive
+Control Paradigm: prescriptive, descriptive
 \end{Verbatim}
-\caption{Semantic capture - a set of constrained semantics, parallel paradigms, expression and concern define key properties a parallel programming model.}
+\caption{Semantic capture: a set of constrained semantics, parallel paradigms, expression and concern define key properties a parallel programming model.}
 \label{figSemCapture}
@@
 
-Constrained semantics provided by the programmer to the parallel programming model can be grouped into information to express intent and properties.
+\emph{Constrained semantics} provided by the programmer to the parallel programming model can be grouped into information to express intent or correctness and properties that control critical performance aspects.
 They address the question of \emph{what}, \emph{where} and \emph{how}.
-In the context of parallel programming, these correspond to defining which code portion to parallelize, where to run and access and how to run that parallel code.
-Synchronization primitives may be considered as part of the what and information on the execution properties such as data placement or memory access type as the how.
-Semantics are expressed following a paradigm and a programming language.
+In the context of parallel programming, these correspond to defining which code portion to parallelize, where to run the code, where to access the data, and how to run that parallel code.
+Synchronization primitives may be considered part of the ``what,'' and information on the execution properties such as data placement or memory access type may be considered as part of the ``how.'' Application logic is expressed following a \emph{paradigm} implemented in a \emph{programming language}.
 
-The parallel programming paradigm is an abstract ~\emph{representation} with the purpose of facilitating the programmer's understanding of programming rules and program behavior.
+A parallel programming paradigm is an abstract representation with the purpose of facilitating the programmer's understanding of programming rules and program behavior.
 Which programming paradigm to chose depends on several considerations.
-Parallel-patterns allow to express concurrency for commonly occurring programming patterns such as loop constructs.
-Tasking is a paradigm that support the expression of concurrent loops as well as irregular algorithms.
-Distributed and correctness-oriented programming models may implement actor-based programming.
-In this programming model, each unit of execution represent an actor who communicates over predefined communication interfaces.
-This paradigm eliminates accesses to shared state and align well with message passing programming (MPI).
-The execution model and memory model of a programming model defines behavior.
-That is, it defines the relationship between abstract concepts and program behavior on the given architecture.
+Parallel patterns allow the programmer to permit concurrent execution for commonly occurring programming patterns such as loop constructs.
+Tasking is a paradigm that supports the expression of interdependent work items needed for irregular algorithms.
+Distributed and correctness-oriented programming models may implement actor-based programming, where each unit of execution represents an actor who communicates over predefined communication channels.
+This paradigm eliminates accesses to shared state and aligns well with message passing programming (MPI).
+The execution model and memory model aspects of a programming model define its behavior.
+That is, it defines the relationship between abstract concepts and program execution on the given architecture.
 

### 2020-01-24 src/2.GenericProgramming.tex
Chose->choose.

@@
 A parallel programming paradigm is an abstract representation with the purpose of facilitating the programmer's understanding of programming rules and program behavior.
-Which programming paradigm to chose depends on several considerations.
+Which programming paradigm to choose depends on several considerations.
 Parallel patterns allow the programmer to permit concurrent execution for commonly occurring programming patterns such as loop constructs.

### 2020-01-24 src/2.GenericProgramming.tex
Minor changes.

@@
 In the context of parallel programming, these correspond to defining which code portion to parallelize, where to run the code, where to access the data, and how to run that parallel code.
-Synchronization primitives may be considered part of the ``what,'' and information on the execution properties such as data placement or memory access type may be considered as part of the ``how.'' Application logic is expressed following a \emph{paradigm} implemented in a \emph{programming language}.
+Synchronization primitives may be considered part of the \emph{what}, and information on the execution properties such as data placement or memory access type may be considered as part of the \emph{how}.
+Application logic is expressed following a \emph{paradigm} implemented in a \emph{programming language}.
 
@@
 
-Lastly, a language defines the syntax of the parallel semantic.
+A language defines the syntax of the parallel semantic.
 Annotations-based  languages add semantic information to the base language through pragmas while embedded programming models rely on the base language and offer additional programming interfaces to capture information.

### 2020-01-24 main.tex
Renamed files for consistency.
Minor changes.
Added bibitems

@@
 \IEEEauthorblockN{Jan Ciesko}
-\IEEEauthorblockA{\textit{Computer Science Reseach Center} \\
+\IEEEauthorblockA{\textit{Computer Science Reseach Institute} \\
 \textit{Sandia National Laboratories}\\
@@
 \IEEEauthorblockN{David Poliakoff}
-\IEEEauthorblockA{\textit{Computer Science Reseach Center} \\
+\IEEEauthorblockA{\textit{Computer Science Reseach Institute} \\
 \textit{Sandia National Laboratories}\\
@@
 \IEEEauthorblockN{David S. Hollman}
-\IEEEauthorblockA{\textit{Computer Science Reseach Center} \\
+\IEEEauthorblockA{\textit{Computer Science Reseach Institute} \\
 \textit{Sandia National Laboratories}\\
@@
 \IEEEauthorblockN{Author Name}
-\IEEEauthorblockA{\textit{Computer Science Reseach Center} \\
+\IEEEauthorblockA{\textit{Computer Science Reseach Institute} \\
 \textit{Sandia National Laboratories}\\
@@
 \input{src/1.Introduction}           %label: chap:introduction
-\input{src/2.GenericProgramming.tex} %label: chap:background
-\input{src/3.MachineModel.tex}             %label: chap:kokkosPM
-\input{src/4.ProgammingModel.tex}     %label: chap:kokkosEMA
-\input{src/5.Backend.tex}            %label: chap:kokkosBackend
-\input{src/6.Exercises}              %label: chap:tutorial
+\input{src/2.GenericProgramming}     %label: chap:background
+\input{src/3.MachineModel}           %label: chap:kokkosPM
+\input{src/4.ProgammingModel}        %label: chap:kokkosEMA
+\input{src/5.Example}                %label: chap:example
+\input{src/6.Tutorials}              %label: chap:tutorial
 \input{src/7.C++}                    %label: chap:c++
@@
 \begin{thebibliography}{00}
-\bibitem{b1} G.
-Eason, B.
-Noble, and I.
-N.
-Sneddon, ``On certain integrals of Lipschitz-Hankel type involving products of Bessel functions,'' Phil.
-Trans.
-Roy.
-Soc.
-London, vol.
-A247, pp.
-529--551, April 1955.
+\bibitem{RAJA} Hornung, Rich, Jones, Holger, Keasler, Jeff, Neely, Rob, Pearce, Olga, Hammond, Si, Trott, Christian, Lin, Paul, Vaughan, Courtenay, Cook, Jeanine, Hoekstra, Rob, Bergen, Ben, Payne, Josh, and Womeldorff, Geoff., ASC Tri-lab Co-design Level 2 Milestone Report 2015, United States, 2015
+\bibitem{KOKKOS} H. Carter Edwards and Christian R. Trott and Daniel Sunderland, 
+Kokkos: Enabling manycore performance portability through polymorphic memory access patterns, Journal of Parallel and Distributed Computing, Vol. 74, p. 3202-3216, 2014
+\bibitem{NALU} Stefan Domino, Shreyas Ananthan, Robert Knaus, and Alan Williams, 
+Deploy Nalu/Kokkos algorithmic infrastructure with performance benchmarking, ECP Milestone Report, United States, 2017
+\bibitem{KOKKOSUSECASE1} Holmen, John K.
+and Humphrey, Alan and Sunderland, Daniel and Berzins, Martin, Improving Uintah’s Scalability Through the Use of Portable Kokkos-Based Data Parallel Tasks, Proceedings of the Practice and Experience in Advanced Research Computing 2017 on Sustainability, ACM, United States, 2017
+\bibitem{KOKKOSUSECASE2} Stan Moore, LAMMPS KOKKOS Package: The quest for performance portable MD, LAMMPS Workshop, United States, 2019
+\bibitem{OPENMP} https://www.openmp.org/specifications/
+\bibitem{ERLANG} Cesarini, Francesco and Thompson, Simon, ERLANG Programming, O’Reilly Media, Inc., 2009
+\bibitem{KOKKOS_REPO} https://github.com/kokkos/
+\bibitem{KOKKOS_TUTORIAL} https://github.com/kokkos/kokkos-tutorials/
 \end{thebibliography}

### 2020-01-24 src/1.Introduction.tex
Added bibitems

@@
 
-Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS_PAPER_HERE}, is a performance-portable, parallel programming model for C++ developed at Sandia National Laboratories (SNL).
+Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS}, is a performance-portable, parallel programming model for C++ developed at Sandia National Laboratories (SNL).
 It consists of a programming model specification and a library.
@@
 The Kokkos library maps execution primitives to vendor-specific programming models, libraries, and memory layouts.
-Kokkos has been successfully used in many large-scale projects at SNL, LLNL and other research centers~\cite{CITEKOKKOSUSECASES}.
+Kokkos has been successfully used in many large-scale projects at SNL, LLNL and other research centers~\cite{KOKKOSUSECASE2}.
 We list its key properties as follows.
@@
 
-Further, we would like to present Kokkos to the educational community as one representative parallel programming model for a trend towards architecture-independent and performance-portable programming of CPUs and accelerators.
+Further, we would like to present Kokkos to the educational community as one representative parallel programming model for a trend towards portability abstractions.
 We show a set of types that implement patterns and encapsulate data with enough semantic information to produce performance-portable, parallel code on current architectures.
@@
 Secondly, we would like to promote Kokkos to students and educators as a platform that helps them to understand and experiment with parallel patterns in a fashion that is generic, vendor-agnostic and declarative.
-This means that students can learn about parallelism without being exposed to many architectural concerns and that they can see benefits of such abstractions for performance portability when programming on current and upcoming hardware architectures.
+This means that students can learn about parallelism without being exposed to many architectural concerns and that they can see productivity benefits of such abstractions for performance portability when programming on current and upcoming hardware architectures.
 

### 2020-01-24 src/2.GenericProgramming.tex
Added bibitems

@@
 All three approaches are conceptually suitable to represent programming paradigms such as tasking, data parallelism, and other patterns.
-Some representatives are the OpenMP programming model~\cite{CITEOPENMP}, threading libraries, or languages like Erlang in which parallelism is a first-class citizen.
+Some representatives are the OpenMP programming model~\cite{OPENMP}, threading libraries, or languages like Erlang\cite{ERLANG} in which parallelism is a first-class citizen.
 In general, it is up to the parallel programming model implementer to expose language constructs or library interfaces (APIs) that combine convenience to the programmer with sufficiently constrained semantics to map the program execution to the parallel computer hardware efficiently and correctly.

### 2020-01-24 src/4.ProgammingModel.tex
Added bibitems

@@
 \begin{Verbatim}[frame=leftline]
-Constrained Semantics: Patterns (intent), Spaces, Layouts, 
-Policies and Traits
+Constrained Semantics: Patterns (intent), Spaces, 
+Layouts, Policies and Traits
 Parallel Paradigms: Parallel patterns and tasking
 Expression: Embedded, C++
-Concern: descriptive
+Control Paradigm: descriptive
 \end{Verbatim}

### 2020-01-24 src/6.Tutorials.tex
Added bibitems

@@
 
-The Kokkos programming model distribution~\cite{CITEKOKKOSGIT} includes many examples and benchmark applications as well as tutorial and presentation material.
+The Kokkos programming model distribution~\cite{KOKKOS_REPO} includes many examples and benchmark applications as well as tutorial and presentation material.
 Example applications demonstrate the use of each abstraction as well supported paradigms.
@@
 Benchmark applications provide further insight into performance implications and allows for easy experimentation.
-Further resources are online\cite{CITEKOKKOSTUTORIALS}.
+Further resources are online\cite{KOKKOS_TUTORIAL}.
 They include presentation material as well as tutorials on application profiling and the Kokkos Kernels numerical library.

### 2020-01-24 src/1.Introduction.tex
Changes after review with Christian

@@
 
-Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS}, is a performance-portable, parallel programming model for C++ developed at Sandia National Laboratories (SNL).
+Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS}, is a performance-portable, parallel programming model for C++ developed through a multi-institutional collaboration between US national laboratories as part of the US Department of Energy Exascale Project\cite{ECP}.
 It consists of a programming model specification and a library.
@@
 The Kokkos library maps execution primitives to vendor-specific programming models, libraries, and memory layouts.
-Kokkos has been successfully used in many large-scale projects at SNL, LLNL and other research centers~\cite{KOKKOSUSECASE2}.
+Kokkos has been successfully used in many large-scale projects at SNL, Utah University, ORNL, LANL and other research centers~\cite{KOKKOSUSECASE2}.
 We list its key properties as follows.
@@
 \item Offers abstract data types for performance portability
-\item Pure C++ implementation through template metaprogramming
+\item Pure C++ interface through template metaprogramming
 \item Aligns with C++ standardization efforts for parallel programming
@@
 Firstly, we would like to present our thoughts that accompanied the design and development of the Kokkos parallel programming model.
-These insights may be considered useful in parallel programming classes that introduce programming models and reason about design choices.
+These insights may be  useful in parallel programming classes that introduce programming models and reason about design choices.
 
-Further, we would like to present Kokkos to the educational community as one representative parallel programming model for a trend towards portability abstractions.
+Further, we would like to promote Kokkos to students and educators as a platform that helps them to understand and experiment with parallel patterns in a fashion that is generic, vendor-agnostic and declarative.
+This means that students can learn about parallelism without being exposed to many architectural concerns and that they can see productivity benefits of such abstractions for performance portability when programming on current and upcoming hardware architectures.
+
+Finally, we would like to present Kokkos to the educational community as one representative parallel programming model for a trend towards portability abstractions.
 We show a set of types that implement patterns and encapsulate data with enough semantic information to produce performance-portable, parallel code on current architectures.
 Further, we highlight its alignment and contributions to standardization efforts towards supporting parallel programming in C++.
-
-Secondly, we would like to promote Kokkos to students and educators as a platform that helps them to understand and experiment with parallel patterns in a fashion that is generic, vendor-agnostic and declarative.
-This means that students can learn about parallelism without being exposed to many architectural concerns and that they can see productivity benefits of such abstractions for performance portability when programming on current and upcoming hardware architectures.
 
@@
 Chapters~\ref{chap:kokkosPM} and ~\ref{chap:kokkosBackend} present the Kokkos programming model and its back-end support.
-Chapters~\ref{chap:c++} and~\ref{chap:related} show the alignment to C++ standardization efforts and dicuss related work and lastly Chapter~\ref{chap:conclusion} concludes this work.
+Chapters~\ref{chap:c++} and~\ref{chap:related} show the alignment to C++ standardization efforts and discuss related work and lastly Chapter~\ref{chap:conclusion} concludes this work.

### 2020-01-24 src/2.GenericProgramming.tex
Changes after review with Christian

@@
 
-\emph{Constrained semantics} provided by the programmer to the parallel programming model can be grouped into information to express intent or correctness and properties that control critical performance aspects.
+\emph{Constrained Semantics} provided by the programmer to the parallel programming model can be grouped into information to express intent or correctness and properties that control critical performance aspects.
 They address the question of \emph{what}, \emph{where} and \emph{how}.
@@
 
-A parallel programming paradigm is an abstract representation with the purpose of facilitating the programmer's understanding of programming rules and program behavior.
+A \emph{Parallel Paradigm} is an abstract representation with the purpose of facilitating the programmer's understanding of programming rules and program behavior.
 Which programming paradigm to choose depends on several considerations.
@@
 Lastly, a programming model must divide responsibility between the developer and the model.
-We refer to is as defining the concern.
+We refer to is as defining the \emph{Control Paradigm}.
 In practice, this is a choice between defining a descriptive and prescriptive semantics.
@@
 Prescriptive semantics require the developer to specify the mechanisms by which their intent will be implemented.
-In an prescriptive model, developer success is tied to their knowledge of implementation mechanisms.
+In a prescriptive model, developer success is tied to their knowledge of implementation mechanisms.
 In a descriptive model that success is tied to the developer's ability to accurately describe their requirements and the toolchain's ability to turn that into an implementation.
@@
 
-Both examples provide the semantic information on what to parallelize, however, they do not expose enough abstraction primitives that would allow to map parallel execution on modern architectures.
+Both examples provide the semantic information on what to parallelize, however, they do not expose enough abstraction primitives that would allow to map parallel execution on modern architectures efficiently.
 In order to define what further information is needed to express concurrency on modern computer architectures, it is important to define an~\emph{abstract machine representation} first.

### 2020-01-24 src/4.ProgammingModel.tex
Changes after review with Christian

@@
 This has multiple reasons.
-It is our experience that application developers of scientific and high-performance computing applications at Sandia National Laboratories express the demand for C++ and a generic parallel programming model to support their codes.
-Following their estimates, porting efforts towards adapting to vendors, programming models, APIs and software releases represents a burden worth addressing.
-Further, the possibility to maintain pure C++ codes is appealing from the code development and debugging perspective.
-Lastly, from the programming model perspective, C++ offers template metaprogramming which is well suited to implement generic APIs and libraries.
+It is our experience that application developers of scientific and high-performance computing applications in the high-performance computing community and in industry in general express the demand for C++ and a generic parallel programming model to support their codes.
+Further, following their estimates, porting efforts towards adapting to vendors, programming models, APIs and software releases represents a burden worth addressing.
+Lastly, the possibility to maintain pure C++ codes is appealing from the code development and debugging perspective.
+
+From the programming model perspective, C++ offers template metaprogramming which is well suited to implement generic APIs and libraries.
 In this case, class specialization and templating allow for compile-type generated types and optimizations for a given hardware architecture.
-It is worth to mention that Kokkos avoids the use of pragma annotations for the sake of treating concurrency as a first-class concept in C++.
 
 Using the programming language, a parallel programming paradigm can be applied to simplify the process of thought.
-Kokkos supports the parallel patterns and task paradigms.
-Parallel patters covers \emph{for}, \emph{ scan} and \emph{reduction}.
+Kokkos supports parallel patterns and tasking.
+Parallel patterns covers \emph{for}, \emph{scan} and \emph{reduction}.
 This allows the expression of concurrency over iterative, for-loop computable algorithms.
@@
 Tasks encapsulate work into units that may be executed in parallel to other tasks or sections of the program.
-Equipped with a language and set of paradigms, a particular set of abstractions can be defined to expose the abstract machine model and captures the intent of the programmer.
+Equipped with a language and set of paradigms, a particular set of abstractions can be defined to expose the abstract machine model and capture the intent of the programmer.
 
@@
 Furthermore, they encapsulate functionality such as consistency control and persistence scopes.
-	\item Layouts express the mapping from loop indices to address offsets.
+	\item Layouts express the mapping from array indices to address offsets.
 By adopting appropriate layouts for memory structures, an application can optimize data access patterns in a given algorithm.
@@
 \emph{ExecPolicy} shapes the iteration space for a particular execution space.
-Valid template parameter are the~\emph{RangePolicy},~\emph{TeamPolicy},~\emph{MDRangePolicy},~\emph{TeamThreadRange} and~\emph{ThreadVectorRange}.
-As the name suggests, these policies express how an iteration space is mapped to the underlying hardware.
+Examples of execution policies are the~\emph{RangePolicy}, ~\emph{MDRangePolicy} and the~\emph{TeamPolicy}.
+As the name suggests, these policies express iteration spaces.
 For example, a range policy defines a one dimensional iteration range.
-A team policy defines a 1D iteration range, each of which is assigned to a thread team.
+The MDRangePolicy represents a multi-dimensional iteration space and is used to express tightly nested loop patterns.
+A team policy defines a 1D iteration range, each of which is assigned to a team of threads.
 This policy allows the expression of hierarchical parallelism.
@@
 Otherwise it depends on the execution space where the loop executes and weather this execution space implements a barrier.
+
 The semantic of the parallel\_for class in Kokkos demonstrates the descriptive nature of the programming model.
-To the interested reader, the complete programming model specification can be accessed on-line\cite{pub:KOKKOS}.
-Figure~\ref{fig:abstractions} summarizes patterns and abstractions and structures them by information type in the semantic capture.
-
 Programming with patterns and abstractions allows the compiler to generate optimized transformations.
-In case of parallel patterns, the execution is unspecified and only promises deterministic results for the reduction and scan operations.
+Generally parallel patterns in Kokkos do not guarantee iteration ordering nor degree of concurrency during their execution.
 This gives the freedom to the underlying abstraction layers to implement different mapping patterns on different hardware such as assignment of iterations to threads or vector lanes.
@@
 %Figure~\ref{fig:stack} shows abstraction layers in Kokkos and and how it fits in a parallel programming landscape.
-The next section shows how the Kokkos parallel\_for class is mapped to the the underlying back-end programming model.
+To the interested reader, the complete programming model specification can be accessed on-line\cite{pub:KOKKOS}.
+Figure~\ref{fig:abstractions} summarizes patterns and abstractions and structures them by information type in the semantic capture.
+The next section shows an example of the Kokkos parallel\_for and explains how its mapped to the the underlying back-end programming model.
 

### 2020-01-24 src/8.Related.tex
Fused DZP and JC changes for Chapters 1 and 2.
Changes after review with Christian

@@
 In both frameworks, a developer first expresses the operations their kernel performs using simple C++ functional programming mechanisms.
-The difference is that Kokkos is declarative, while RAJA is imperative.
+The difference is that Kokkos is declarative, while RAJA is descriptive.
 In Kokkos, the framework figures out how to map that kernel to different architectures.
@@
 DPC++\cite{DPCPP} is a model developed by Intel\textsuperscript{\textregistered} for expressing parallelism across Intel architectures.
-In addition to sharing a declarative philosophy with Kokkos, it also shares parallel patterns.
+In addition to sharing a descriptive philosophy with Kokkos, it also shares parallel patterns.
 Unlike Kokkos, at time of publication it targets only Intel architectures and is not vendor-agnostic.
@@
 In courses that do not target accelerators, this can be a relevant option.
-%\end{itemize}
+
+% Alpaka Dresden
+% NV Thrust

### 2020-01-24 src/7.C++.tex
First swing at the C++ section

\section{Towards Parallel Programming in C++}\label{chap:c++}

Another important reason that Kokkos is an ideal choice for an eduactional setting is the feedback cycle it participates in with the ISO-C++ language standard itself.  Over the past decade or so, work on the Kokkos programming model has contributed to a number of current or near-future ISO-C++ features, including \code{atomic_ref}, \code{mdspan}, and executors.  This interplay between Kokkos and the C++ standard means that students using the model will be exposed to paradigms that will likely be part of the broader standard C++ ecosystem, thus preparing them for a much wider range of application areas.

In particular, the design of C++ executors, which are quite similar to Kokkos execution spaces, was heavily influenced by Kokkos.  As the demand to extract performance from increasingly deep and increasingly asynchronous software stacks across a wide variety of domains increases, many C++ experts expect executors to become a central abstraction in any performance-oriented software stack.  This is already evident with the introduction of parallel algorithms with execution policies in C++17 (also influenced by Kokkos), and the likely extensions of these algorithms to executors and asynchrony in the coming years continues to bear many similarities to the Kokkos programming model.  Thus, teaching Kokkos as a parallel programming model not only prepares students for HPC programming environments; it also prepares them for any of a number of performance sensitive fields, from embedded devices to gaming to data science and machine learning.

### 2020-01-24 src/1.Introduction.tex
Adding other example and back-end example.

@@
 
-Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS}, is a performance-portable, parallel programming model for C++ developed through a multi-institutional collaboration between US national laboratories as part of the US Department of Energy Exascale Project\cite{ECP}.
+Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS}, is a performance-portable, parallel programming model for C++ developed collaboratively between national laboratories as part of the US Department of Energy Exascale Project\cite{ECP}.
 It consists of a programming model specification and a library.
@@
 The Kokkos library maps execution primitives to vendor-specific programming models, libraries, and memory layouts.
-Kokkos has been successfully used in many large-scale projects at SNL, Utah University, ORNL, LANL and other research centers~\cite{KOKKOSUSECASE2}.
+Kokkos has been successfully used in many large-scale projects at SNL, University of Utah, Oak-Ridge National Laboratories, Los Alamos National Laboratories and others~\cite{KOKKOSUSECASE2}.
 We list its key properties as follows.

### 2020-01-24 src/5.Example.tex
Renamed files for consistency.
Small changes.
Adding other example and back-end example.

--- 2020-01-23 src/5.Backend.tex
@@
-\section{Back-end Support}\label{chap:kokkosBackend}
+\section{Example}\label{chap:kokkosExample}
 
 This section provides a practical insight into programming with Kokkos and the inner workings of the programming model library.
-Figure~\ref{fig:KokkosExample} shows am example application that algorithmically initializes an array to integers and includes several points of interests.
+Figure~\ref{fig:KokkosExample} shows am example application that algorithmically initializes an array to integers.
 Taking a closer look, the reader can observe the initialization and finalization of the Kokkos runtime library through the~\emph{Kokkos::initialize} and~\emph{Kokkos::finalize} method calls.
@@
   {
-    int N = atoi(argv[1]);
-    Kokkos::View<int*> a("A", N);  
-    Kokkos::parallel_for(
-    Kokkos::RangePolicy<Kokkos::OpenMP>(N,
-    [=](const int & i) {
-      a(i)=i*i;
+    ...
+    Kokkos::View<double*> a("A", dN);  
+    Kokkos::View<double*> r("A", N);
+    Kokkos::View<double**> M("A", dM, dN);
+
+    Kokkos::parallel_for("outer", TeamPolicy<>(M,AUTO),
+    [=](const member_type &teamID)
+    {
+      const int e = teamID.league_rank();
+      Kokkos::parallel_reduce( TeamThreadRange(teamID, N),
+        [=](const int & i, double & r_tmp) {
+          r_tmp += M(e, i) * a(i);
+        }, 
+        r(e));
     }); 
@@
 \end{figure}
+
+%\begin{figure}
+%\begin{small}
+%\begin{Verbatim}[frame=leftline]
+%int main(int argc, char* argv[]) {
+%  Kokkos::initialize(argc,argv);
+%  {
+%    int N = atoi(argv[1]);
+%    Kokkos::View<int*> a("A", N);  
+%    Kokkos::parallel_for(
+%    Kokkos::RangePolicy<Kokkos::OpenMP>(N,
+%    [=](const int & i) {
+%      a(i)=i*i;
+%    }); 
+%  }
+%  Kokkos::finalize();
+%}
+%\end{Verbatim}
+%\end{small}
+%\caption{This example application written in Kokkos shows the use of a~\emph{view}-type and a parallel for loop. Both Kokkos types are templated to the OpenMP execution space.}
+%\label{fig:KokkosExample}
+%\end{figure}
 

### 2020-01-24 src/9.Conclusion.tex
Adding other example and back-end example.

@@
 %...
-%Consequently, we emphasize the importance and welcome new development and educational use of parallel programming models that align to this trend.
+In this work we gave an introduction to the Kokkos programming model. We presented a conceptual framework consi,  presented the Kokkos parallel prgroamming model. 
+Programming support for parallel progrmaming in mainstream langiages is gaining momentung
+. Programming abstraction libraries such as Kokkos, Raja and others showcase this trend. Targeting a generic parallel machine representation allows to design and efficiently use
+
+We emphasize the importance and welcome new development and educational use of parallel programming models that align to this trend.
 We believe that abstracting from vendor or architecture-specific programming models not only simplifies parallel programming in science and industry but also would support educators in teaching parallel programming.
-%We would also like to express gratitude to educators for taking on this important role.
+We would also like to express gratitude to educators for taking on this important role.
 Further materials, including a detailed description of the programming- and machine model can be found by following the reference on the Kokkos programming mode.
-Finnaly, we would like to mention that interested developers and student can reach the development team on Slack by following the invititation link as mentioned in the reference section.
+Finally, we would like to mention that interested developers and student can reach the development team on Slack by following the invitation link as mentioned in the reference section.

### 2020-01-24 src/7.C++.tex
Minor fix

@@
 Another important reason that Kokkos is an ideal choice for an eduactional setting is the feedback cycle it participates in with the ISO-C++ language standard itself.
-Over the past decade or so, work on the Kokkos programming model has contributed to a number of current or near-future ISO-C++ features, including \code{atomic_ref}, \code{mdspan}, and executors.
+Over the past decade or so, work on the Kokkos programming model has contributed to a number of current or near-future ISO-C++ features, including \emph{atomic\_ref}, \emph{mdspan}, and executors.
 This interplay between Kokkos and the C++ standard means that students using the model will be exposed to paradigms that will likely be part of the broader standard C++ ecosystem, thus preparing them for a much wider range of application areas.

### 2020-01-24 main.tex
Adding other example and back-end example.
Added Conclusion.

@@
 \and
-\IEEEauthorblockN{Author Name}
+\IEEEauthorblockN{Christian C. Trott}
 \IEEEauthorblockA{\textit{Computer Science Reseach Institute} \\
@@
 Albuquerque, NM, USA\\
-aname@sandia.gov}
+ctrott@sandia.gov}
 \and
-\IEEEauthorblockN{Author Name}
-\IEEEauthorblockA{\textit{Computer Science Reseach Institute} \\
-\textit{Sandia National Laboratories}\\
-Albuquerque, NM, USA\\
-aname@sandia.gov}
-\and
-\IEEEauthorblockN{Author Name}
-\IEEEauthorblockA{\textit{Computer Science Reseach Institute} \\
-\textit{Sandia National Laboratories}\\
-Albuquerque, NM, USA\\
-aname@sandia.gov}
+\IEEEauthorblockN{Damien Lebrun-Grandié}
+\IEEEauthorblockA{\textit{Computational Sciences and Engineering} \\
+\textit{Oak Ridge National Laboratory}\\
+Oak Ridge, TN, USA\\
+lebrungrandt@ornl.gov}
 }
@@
 \begin{thebibliography}{00}
-\bibitem{RAJA} Hornung, Rich, Jones, Holger, Keasler, Jeff, Neely, Rob, Pearce, Olga, Hammond, Si, Trott, Christian, Lin, Paul, Vaughan, Courtenay, Cook, Jeanine, Hoekstra, Rob, Bergen, Ben, Payne, Josh, and Womeldorff, Geoff., ASC Tri-lab Co-design Level 2 Milestone Report 2015, United States, 2015
 \bibitem{KOKKOS} H. Carter Edwards and Christian R. Trott and Daniel Sunderland, 
 Kokkos: Enabling manycore performance portability through polymorphic memory access patterns, Journal of Parallel and Distributed Computing, Vol. 74, p. 3202-3216, 2014
+\bibitem{RAJA} Hornung, Rich, Jones, Holger, Keasler, Jeff, Neely, Rob, Pearce, Olga, Hammond, Si, Trott, Christian, Lin, Paul, Vaughan, Courtenay, Cook, Jeanine, Hoekstra, Rob, Bergen, Ben, Payne, Josh, and Womeldorff, Geoff., ASC Tri-lab Co-design Level 2 Milestone Report 2015, United States, 2015
 \bibitem{NALU} Stefan Domino, Shreyas Ananthan, Robert Knaus, and Alan Williams, 
@@
 \bibitem{ERLANG} Cesarini, Francesco and Thompson, Simon, ERLANG Programming, O’Reilly Media, Inc., 2009
+\bibitem{DPCPP} https://software.intel.com/en-us/oneapi/base-kit\#specifications
 \bibitem{KOKKOS_REPO} https://github.com/kokkos/
 \bibitem{KOKKOS_TUTORIAL} https://github.com/kokkos/kokkos-tutorials/
+\bibitem{KOKKO_SLACK} https://kokkosteam.slack.com
 \end{thebibliography}

### 2020-01-24 src/9.Conclusion.tex
Added Conclusion.

@@
 \section{Conclusion}\label{chap:conclusion}
-%...
-In this work we gave an introduction to the Kokkos programming model. We presented a conceptual framework consi,  presented the Kokkos parallel prgroamming model. 
-Programming support for parallel progrmaming in mainstream langiages is gaining momentung
-. Programming abstraction libraries such as Kokkos, Raja and others showcase this trend. Targeting a generic parallel machine representation allows to design and efficiently use
+In this work we gave an introduction to the Kokkos parallel programming model and our guiding thoughts behind its design.
+We discussed a machine representation of a parallel hardware architecture that is characterized by a variety of processing units, memories and their hierarchical organization.
+This machine model allowed us to define a constrained semantic represented though different types and patterns.
+Further we listed supported paradigms and considerations that influenced the design of the semantic capture.
 
-We emphasize the importance and welcome new development and educational use of parallel programming models that align to this trend.
-We believe that abstracting from vendor or architecture-specific programming models not only simplifies parallel programming in science and industry but also would support educators in teaching parallel programming.
+An example application gave insight into parallel programming with Kokkos patterns and the back-end implementation.
+The back-end implementation shows the generic programming interface and its partial type-specialization towards a particular vendor programming model.
+
+To the educational community, we presented Kokkos as a representative parallel programming model of a trend towards support of parallel programming in base languages and through library abstractions.
+We believe in the importance of this trend in computer science education as such parallel programming models contribute to the generic and relevant education of parallel programming and prepare student for their programming careers.
+
+We emphasize the importance and welcome development and educational use of other parallel programming models that align to this effort. An example of this is Raja.
+
 We would also like to express gratitude to educators for taking on this important role.
 Further materials, including a detailed description of the programming- and machine model can be found by following the reference on the Kokkos programming mode.
-Finally, we would like to mention that interested developers and student can reach the development team on Slack by following the invitation link as mentioned in the reference section.
+Finally, we would like to mention that interested readers can reach the development team on Slack\cite{KOKKOSLACK}.

### 2020-01-24 src/5.Example.tex
Rewrite Exercise and Implementation Section

--- 2020-01-23 src/5.Backend.tex
@@
-\section{Back-end Support}\label{chap:kokkosBackend}
+\section{Example}\label{chap:kokkosExample}
 
-This section provides a practical insight into programming with Kokkos and the inner workings of the programming model library.
-Figure~\ref{fig:KokkosExample} shows am example application that algorithmically initializes an array to integers and includes several points of interests.
-Taking a closer look, the reader can observe the initialization and finalization of the Kokkos runtime library through the~\emph{Kokkos::initialize} and~\emph{Kokkos::finalize} method calls.
-Further the example shows the use of the~\emph{view} abstraction that represents a one-dimensional integer array of size~\emph{N}.
-The object carries the name~\emph{A} internally to facilitate tracing and debugging.
-In this example, a parallel\_for pattern object receives a range policy that is templated to the OpenMP execution space and takes~\emph{N} and a lambda expression as parameters.
-Vies behave as pointers and there a copy-semantic in the lambda expression is used in this case.
-Within the loop-body, the view object is accessed through the~\emph{()-operator}.
-It is up to the OpenMP execution space to provide performance implementations of the view, an efficient internal representation of the array and for the access operator for the underlying hardware architecture.
+In order to provide an impression of the looks and feel of Kokkos code this section discusses a small but complete example of a data parallel Kokkos application shown in Figure~\ref{fig:KokkosExample} which allocates data, potentially initalizes it on the host and then performs a matrix-vector product.
+Similar to other programming models such as MPI, Kokkos must be initialized and finalized. 
+This serves the purpose of setting up the underlying runtime and acquiring resouces.
+The fundamental data structure of Kokkos programs is the ~\emph{Kokkos::View}.
+These views represent multi-dimensional arrays of rank one to eight.
+A general philosophy of Kokkos is to have defaults for most properties, including data layouts, memory spaces and execution spaces.
+Consequently the simple example does not explicitly set which memory space the data lives in.
+Kokkos ensures that the defaults for parallel execution and data allocations match though.
 
-At compile time, the compiler generates the particular types, where in this case, the parallel\_for is types for the OpenMP execution space.
-At link time, the linker links this method to the corresponding parallel\_for implementation in the Kokkos library.
+The example then creates copies of the views accessible by the host process via the ~\emph{create\_mirror\_view} function. 
+If the original data was already accessible by the host process, this function returns a ~\emph{View} referencing the same allocation.
+In real applications this host process accessible allocation is often necessary for I/O purposes for example to initialize the data with input read from a file.
+To get the data to the memory space accessible by the parallel execution resources the function ~\emph{deep\_copy} is then invoked.
+If two ~\emph{View}s are aliases of each other this operation is a no-op.
 
-The implementation of the parallel\_for specific to the OpenMP execution space is shown in Figure~\ref{fig:KokkosExampleOMPBackEnd}.
-It shows the template argument of that class predefined to that execution space and shows how the~\emph{execute()} method calls the functor through ~\emph{exec\_range}.
-In can be seen that the Kokkos library relies on the back-end compiler to generate platform optimized, parallel code.
-For the OpenMP back-end implementation of the parallel for loop, OpenMP pragma annotations are used.
-Since the parallel for implies a barrier at the end of the parallel region, the return semantic of Kokkos::parallel\_for call is synchronous.
+Now the parallel pattern can be invoked.
+Since the matrix-vector product is an example of hierarchical work (for each row a dot product must be performed), the ~\emph{Kokkos::TeamPolicy} is used.
+For each row of M a team of threads is launched, where the size of team is left for the Kokkos runtime to choose.
+When using a ~\emph{TeamPolicy} the operator of the lambda does not simply get an index, but rather a handle to the team of threads. 
+This handle provides the team id, and is subsequently handed to the nested parallel execution policies. 
+The nested reduction requires the operator of the nested lambda to take a reference to the thread local reduction variable (called here ~\emph{y\_tmp}) in addition to the loop index.
+It then writes the result into the correct position of the output vector.
+
+After the parallel pattern the main code calls a fence to ensure that the execution of the parallel code is finished, before copying the results of y back to the host and potentially putting them out through an I/O operation.
+
+To learn more about Kokkos's capabilties and features we recommend going through the extensive Kokkos tutorial, with multiple days worth of lectures and exercises, and read the Kokkos Programming Guide as well as the Kokkos API documentation - which provide several hundred pages of text-book like material and API reference.
+The Tutorial is intended for computational scientists with minimal or even no prior knowlegde of parallel programming.
+It introduces concepts of parallel programming using Kokkos step by step, through a series of lectures and hands-on exercises which build on each other and gradually introduce more and more Kokkos constructs.
+The Programming Guide provides more in-depth information to the concepts introduced in the Tutorial, while the API reference provides straight forward documentation of the syntax and semantics of Kokkos's library interface.
 
@@
 \begin{Verbatim}[frame=leftline]
+#include<Kokkos_Core.hpp>
 int main(int argc, char* argv[]) {
@@
   {
-    int N = atoi(argv[1]);
-    Kokkos::View<int*> a("A", N);  
-    Kokkos::parallel_for(
-    Kokkos::RangePolicy<Kokkos::OpenMP>(N,
-    [=](const int & i) {
-      a(i)=i*i;
+    Kokkos::View<double*> x("x", M);  
+    Kokkos::View<double*> y("y", N);
+    Kokkos::View<double**,Kokkos::LayoutRight> A("A", N, M);
+
+    auto x_h = Kokkos::create_mirror_view(x);
+    auto y_h = Kokkos::create_mirror_view(y);
+    auto A_h = Kokkos::create_mirror_view(A);
+    //intialize_data_on_host(x_h,A_h);
+    Kokkos::deep_copy(x,x_h);
+    Kokkos::deep_copy(A,A_h);
+
+    Kokkos::parallel_for("outer", TeamPolicy<>(N,AUTO),
+    [=](const member_type &team_handle) {
+      const int e = team_handle.league_rank();
+      Kokkos::parallel_reduce( TeamThreadRange(team_handle, M),
+        [=](const int & i, double & y_tmp) {
+          y_tmp += A(e, i) * x(i);
+        }, 
+        y(e));
     }); 
+    Kokkos::fence();
+    Kokkos::deep_copy(y_h,y);
+    //output_result_on_host(y_h);
   }
@@
 \end{small}
-\caption{This example application written in Kokkos shows the use of a~\emph{view}-type and a parallel for loop. Both Kokkos types are templated to the OpenMP execution space.}
+\caption{This example application written in Kokkos shows the use of a~\emph{view}-type and a parallel for loop using a ~\emph{TeamPolicy} and a nested parallel reduction. }
 \label{fig:KokkosExample}
@@
 
+\section{Kokkos Implementation}
+
+The following section gives a short impression of Kokkos maps generic parallel programming patterns to the hardware specific programming models.
+As a template metaprogramming library, Kokkos makes heavy use of partial specialization. 
+In particular the parallel pattern functions are mapped to partial specializations of classes for those patterns for each execution policy type and backend type.
+The library specification contains concepts for execution spaces, which define the generic interface of those classes.
+To add a new backend, each of those concepts must be implemented.
+
+A simplified version of this implementation strategy is provided in Figure\ref{fig:KokkosExampleOMPBackEnd}.
+The ~\emph{parallel\_for} function instantiates a partial specialization of the ~\emph{ParallelFor} class and calls its ~\emph{execute} function.
+That function is then responsible to implement the Kokkos parallel pattern, for example via OpenMP pragmas in the case of the OpenMP execution space.
 \begin{figure}
@@
 \begin{Verbatim}[frame=leftline]
-template <class FunctorType, class... Traits>
-class ParallelFor<FunctorType, 
-    ..., 
-    Kokkos::OpenMP> {
-  ...
-private:
-  ... cconst FunctorType& functor, 
-  const Member i_beg,
-  const Member i_end) 
-  {
-    for (Member i = i_beg; iwork < i_end; ++i){
-      functor(t, i);
+template <class FunctorType>
+class ParallelFor<FunctorType, RangePolicy<OpenMP> > {
+  const FunctorType functor;
+  const RangePolicy<OpenMP> policy;
+
+public:
+  void execute() const {
+    #pragma omp parallel for
+    for(int i=policy.begin(); i<policy.end(); i++)
+      functor(i);
   }
@@
 
-public:
-  ...
-  ... execute() const {(
-#pragma omp parallel num_threads(OpenMP::gPoolSize())
-  {
-    HostThreadTeamData& data=*(i->get_thr_data());
-    ...
-    range = data.get_work_partition();
-    range(m_functor, range.first + m_policy.begin(),
-    range.second + m_policy.begin());
-  }
+template <class FunctorType, class PolicyType>
+void parallel_for(std::string label, const PolicyType& policy,
+                  const FunctorType& functor) {
+  // Call tools hook with label
+  ParallelFor<FunctorType,PolicyType> pf{functor,policy};
+  pf.execute();
 }

### 2020-01-24 src/3.MachineModel.tex
Adding other example and back-end example.
Updated img.

@@
 The Kokkos abstract machine representation targets a design of a future shared-memory computing architectures.
-The design is shown in Figure~\ref{fig:execspace}.
+The design is shown in Figure~\ref{fig:KokkosSpaces}.
 It is characterized by multiple latency-optimized execution units and off-die bandwidth-optimized accelerators.
@@
 Accelerators can provide an additional memory space through its local on-package memory.
-An example of such an architectural layout is shown in Figure~\ref{fig:memspace}.
+This is highligted in Figure~\ref{fig:KokkosSpaces} in a green color.
+shows memory spaces in .
 The programmer is free to decide where each data structure is allocated by requesting the corresponding memory space.
@@
 \begin{figure}
-\begin{subfigure}[b]{0.41\textwidth}
-\centering
-\includegraphics[width=\textwidth]{img/ExecSpaces.png}
-\caption{Execution space is an instantiation of a space that represents a processing device to which the programmer can target parallel code.}
+
+\includegraphics[width=0.5\textwidth]{img/KokkosSpaces.png}
+\caption{Spaces represent conceptual building blocks of the abstract machine model in Kokkos.}
 \label{fig:execspace}
-\end{subfigure}
-\hfill \break
-\begin{subfigure}[b]{0.41\textwidth}
-\centering
-\includegraphics[width=\textwidth]{img/MemSpaces.png}
-\caption{A memory space is an instantiation of a space that represents a memory location on which the programmer can allocate data.}
-\label{fig:memspace}
-\end{subfigure}
-\caption{Spaces represent conceptual building blocks of the abstract machine model in Kokkos.}
 \label{fig:spaces}

### 2020-01-25 src/4.ProgammingModel.tex
Adding other example and back-end example.
Added new picture.

@@
 In conclusion, Figure~\ref{fig:SemCaptureKokkos} shows the semantic capture defined by the Kokkos programming model.
-%Figure~\ref{fig:stack} shows abstraction layers in Kokkos and and how it fits in a parallel programming landscape.
-To the interested reader, the complete programming model specification can be accessed on-line\cite{pub:KOKKOS}.
 Figure~\ref{fig:abstractions} summarizes patterns and abstractions and structures them by information type in the semantic capture.
-The next section shows an example of the Kokkos parallel\_for and explains how its mapped to the the underlying back-end programming model.
+To the interested reader, the complete programming model description can be accessed on-line\cite{KOKKOS_WIKI}.
+The next section shows an example of the Kokkos parallel\_for and explains how it is mapped to the the underlying back-end programming model.
 
 \begin{figure}
-\centerline{\includegraphics[width=0.3\textwidth]{img/Stack.png}}
-\caption{Landscape}
-\label{fig:stack}
-\end{figure}
-
-\begin{figure}
-\centerline{\includegraphics[width=0.3\textwidth]{img/Abstractions.png}}
-\caption{Landscape}
+\centerline{\includegraphics[width=0.45\textwidth]{img/Abstractions.png}}
+\caption{Overview of abstractions that define the constrained semantic of the Kokkos parallel programming model.}
 \label{fig:abstractions}
 \end{figure}
-
-%\begin{center}
-%\begin{tabular}{ |c|c|c| } 
-% \hline
-% cell1 & cell2 & cell3 \\ 
-% cell4 & cell5 & cell6 \\ 
-% cell7 & cell8 & cell9 \\ 
-% \hline
-%\end{tabular}
-%\end{center}
-
-%\begin{figure}
-%\begin{Verbatim}[frame=leftline]
-%double f(){
-% typedef Kokkos::Cuda ES;
-% typedef Kokkos::CudaSpace MS;
-% typedef Kokkos::LayoutLeft LL;
-% typedef Kokkos::RangePolicy<ES> Range_policy;
-% typedef Kokkos::View<double*, LL, MS> V
-% V A( "A", N);
-% V::HostMirror h_A = Kokkos::create_mirror_view( A );
-% for ( int i = 0; i < N; ++i ) //Initialize on host
-%   h_A( i ) = 1;
-% Kokkos::deep_copy( A, h_A );
-% double result = 0;
-% Kokkos::parallel_reduce( "yAx", range_policy( 0, N ), 
-%  KOKKOS_LAMBDA ( int i, double &update ) {
-%   update += A( i );
-% }, result );
-% return result;
-%}
-%\end{Verbatim}
-%\caption{Declarative parallel programming provides uses pragma annotations to capture semantic information.}
-%\label{figOMPLike}
-%\end{figure}

### 2020-01-25 src/5.Example.tex
Improvements to Chapter on Example and Backend

@@
 
-In order to provide an impression of the looks and feel of Kokkos code this section discusses a small but complete example of a data parallel Kokkos application shown in Figure~\ref{fig:KokkosExample} which allocates data, potentially initalizes it on the host and then performs a matrix-vector product.
-Similar to other programming models such as MPI, Kokkos must be initialized and finalized. 
-This serves the purpose of setting up the underlying runtime and acquiring resouces.
-The fundamental data structure of Kokkos programs is the ~\emph{Kokkos::View}.
-These views represent multi-dimensional arrays of rank one to eight.
-A general philosophy of Kokkos is to have defaults for most properties, including data layouts, memory spaces and execution spaces.
+In this section we provide an example of an application written in the Kokkos programming model.
+The code in Figure~\ref{fig:KokkosExample} shows a parallel matrix-vector multiplication implemented as two nested loops.
+The outer loop, that is the outer parallel_for, iterates over the rows of the input matrix and of the input vector while the inner loop iterates over elements in each row.
+Since the inner loop potentially executes in parallel, the update operation of the aggregating variable~\emph{y_tmp} must be protected against data races.
+Using the parallel_reduction pattern, the programming model provides a local variable to that loop iteration.
+Once the computation finishes, the final result of the reduction operation is stored in the output vector~\emph{y} at the row index~\emph{e}.
+As a first step, a Kokkos application must call ~\emph{Kokkos::initialize}. Initialization and finalization are required and serve the purpose of setting up the underlying runtime. 
+Further, the example code shows the use of the ~\emph{Kokkos::View} abstraction.
+
+Views are abstractions that represent data.
+Naturally data can reside in host memory where it is accessible to the host execution space or on device memory where it is accessible to the device execution space.
+In this example, copies of the views are created in order to make them accessible by the host process.
+This is accomplished by calling the ~\emph{create\_mirror\_view} function.
+
+If the original data is already accessible by the host process, this function returns a ~\emph{View} referencing the same allocation.
+In real applications host-accessible allocation is necessary in order to allow I/O operations during fileystem access.
+To move data to the respective memory space that is accessible by the parallel execution resources, the function ~\emph{deep\_copy} has to be invoked.
+In case two view objects are aliased, this operation results in a~\emph{no-op}.
+
+%A general philosophy of Kokkos is to have defaults for most properties, including data layouts, memory spaces and execution spaces.
 Consequently the simple example does not explicitly set which memory space the data lives in.
@@
 
-The example then creates copies of the views accessible by the host process via the ~\emph{create\_mirror\_view} function. 
-If the original data was already accessible by the host process, this function returns a ~\emph{View} referencing the same allocation.
-In real applications this host process accessible allocation is often necessary for I/O purposes for example to initialize the data with input read from a file.
-To get the data to the memory space accessible by the parallel execution resources the function ~\emph{deep\_copy} is then invoked.
-If two ~\emph{View}s are aliases of each other this operation is a no-op.
+Taking a closer look at the loop step shows the hierarchical expression of concurrency.
+In case of the matrix-vector multiplication,This is accomplished by using the ~\emph{Kokkos::TeamPolicy}.
+For each row of~\emph{M} a team of threads is launched, where the size of the team is not further specified.
+When using a ~\emph{TeamPolicy} the operator of the lambda expression does not receive an index, but rather a handle to the team of threads.
+This handle provides the team identifier \emph{e}, and is subsequently handed to the nested parallel execution policies.
+The nested reduction requires the operator of the nested lambda to take a reference to the thread-local reduction variable (called here ~\emph{y\_tmp}) in addition to the loop index.
+It then writes the result to the correct position of the output vector.
+The parallel loops are follows by a call to \emph{Kokkos::fence}.
+This ensures that the execution of the parallel code is completed before copying the results back to the host memory space.
+This is implemented through the~\emph{Kokkos::deep_copy} call.
+Lastly, the Kokkos application calls~\emph{Kokkos::finalize}.
 
-Now the parallel pattern can be invoked.
-Since the matrix-vector product is an example of hierarchical work (for each row a dot product must be performed), the ~\emph{Kokkos::TeamPolicy} is used.
-For each row of M a team of threads is launched, where the size of team is left for the Kokkos runtime to choose.
-When using a ~\emph{TeamPolicy} the operator of the lambda does not simply get an index, but rather a handle to the team of threads. 
-This handle provides the team id, and is subsequently handed to the nested parallel execution policies. 
-The nested reduction requires the operator of the nested lambda to take a reference to the thread local reduction variable (called here ~\emph{y\_tmp}) in addition to the loop index.
-It then writes the result into the correct position of the output vector.
-
-After the parallel pattern the main code calls a fence to ensure that the execution of the parallel code is finished, before copying the results of y back to the host and potentially putting them out through an I/O operation.
-
-To learn more about Kokkos's capabilties and features we recommend going through the extensive Kokkos tutorial, with multiple days worth of lectures and exercises, and read the Kokkos Programming Guide as well as the Kokkos API documentation - which provide several hundred pages of text-book like material and API reference.
-The Tutorial is intended for computational scientists with minimal or even no prior knowlegde of parallel programming.
-It introduces concepts of parallel programming using Kokkos step by step, through a series of lectures and hands-on exercises which build on each other and gradually introduce more and more Kokkos constructs.
-The Programming Guide provides more in-depth information to the concepts introduced in the Tutorial, while the API reference provides straight forward documentation of the syntax and semantics of Kokkos's library interface.
+While this example can only provide a brief insight into Kokkos, we recommend to the interested reader to access further resources.
+Resources include example applications for individual Kokkos features, tutorials with multiple days worth of lectures and exercises, a programming guide and an API documentation.
+Source code examples as well as the tutorial are located in the Kokkos repository~\cite{KOKKOS_REPO}.
+The tutorial is intended for students with minimal or even no prior knowledge of parallel programming.
+It introduces concepts of parallel programming using Kokkos through a series of lectures and hands-on exercises which build up on each other and gradually introduce new concepts.
+The programming guide is available on-line~\cite{KOKKOS_WIKI}.
 
@@
 
-The following section gives a short impression of Kokkos maps generic parallel programming patterns to the hardware specific programming models.
-As a template metaprogramming library, Kokkos makes heavy use of partial specialization. 
-In particular the parallel pattern functions are mapped to partial specializations of classes for those patterns for each execution policy type and backend type.
-The library specification contains concepts for execution spaces, which define the generic interface of those classes.
-To add a new backend, each of those concepts must be implemented.
+In this section we peak inside the Kokkos library and show how the a generic parallel for loop is mapped to the OpenMP back-end programming model.
+As a template metaprogramming library, Kokkos makes use of partial specialization.
+In particular the~\emph{parallel_for} or ~\emph{parallel_reduce} functional calls as shown in Figure~\ref{fig:KokkosExample} are mapped to partially specialized classes.
+Such classes correspond to particular back-end where for each back-end, concepts with partial specialization must be provided.
+Today, Kokkos supports the CUDA and OpenMP back-ends.
+For each new back-end, partially implemented classes that implement necessary concepts must be provided.
 
-A simplified version of this implementation strategy is provided in Figure\ref{fig:KokkosExampleOMPBackEnd}.
+A simplified version of a partially specialized class for a parallel for loop implementation is shown in Figure\ref{fig:KokkosExampleOMPBackEnd}.
 The ~\emph{parallel\_for} function instantiates a partial specialization of the ~\emph{ParallelFor} class and calls its ~\emph{execute} function.
-That function is then responsible to implement the Kokkos parallel pattern, for example via OpenMP pragmas in the case of the OpenMP execution space.
+That function is then responsible to implement the Kokkos parallel pattern.
+In this back-end, the parallel implementation is generated by using OpenMP pragma annotations and a by invoking a compatible compiler when the library is compiled.
 \begin{figure}
@@
 \caption{This code shows the back-end implementation of a parallel loop in the Kokkos programing model.
-It shows a particular implementation of that generic pattern as a templated class to the OpenMP execution space.
+It shows a particular implementation of a generic pattern as a templated class to the OpenMP execution space.
 Also it shows the use of OpenMP pragma annotations that put platform compilers in charge to generate parallel code for the underlying hardware architecture.}

### 2020-01-25 src/5.Example.tex
Bug fixes.

@@
 The code in Figure~\ref{fig:KokkosExample} shows a parallel matrix-vector multiplication implemented as two nested loops.
-The outer loop, that is the outer parallel_for, iterates over the rows of the input matrix and of the input vector while the inner loop iterates over elements in each row.
-Since the inner loop potentially executes in parallel, the update operation of the aggregating variable~\emph{y_tmp} must be protected against data races.
-Using the parallel_reduction pattern, the programming model provides a local variable to that loop iteration.
+The outer loop, that is the outer~\emph{parallel\_for}, iterates over the rows of the input matrix and of the input vector while the inner loop iterates over elements in each row.
+Since the inner loop potentially executes in parallel, the update operation of the aggregating variable~\emph{y\_tmp} must be protected against data races.
+Using the~\emph{parallel\_reduction} pattern, the programming model provides a local variable to that loop iteration.
 Once the computation finishes, the final result of the reduction operation is stored in the output vector~\emph{y} at the row index~\emph{e}.
@@
 This ensures that the execution of the parallel code is completed before copying the results back to the host memory space.
-This is implemented through the~\emph{Kokkos::deep_copy} call.
+This is implemented through the~\emph{Kokkos::deep\_copy} call.
 Lastly, the Kokkos application calls~\emph{Kokkos::finalize}.
@@
 As a template metaprogramming library, Kokkos makes use of partial specialization.
-In particular the~\emph{parallel_for} or ~\emph{parallel_reduce} functional calls as shown in Figure~\ref{fig:KokkosExample} are mapped to partially specialized classes.
+In particular the~\emph{parallel\_for} or ~\emph{parallel\_reduce} functional calls as shown in Figure~\ref{fig:KokkosExample} are mapped to partially specialized classes.
 Such classes correspond to particular back-end where for each back-end, concepts with partial specialization must be provided.

### 2020-01-25 src/8.Related.tex
Adding other example and back-end example.
Updated Conclusion

@@
 \section{Related Work}\label{chap:related}
-Like Kokkos, RAJA~\cite{RAJA} is a C++ framework for porting massive DOE codes to a variety of architectures.
-In both frameworks, a developer first expresses the operations their kernel performs using simple C++ functional programming mechanisms.
-The difference is that Kokkos is declarative, while RAJA is descriptive.
-In Kokkos, the framework figures out how to map that kernel to different architectures.
-In RAJA, functionality is provided to make it easy for a developer to try different strategies to map those kernels to architectures, but the developer still decides how that mapping happens.
-Which model makes more sense in a given circumstance depends on how much exposure to the underlying programming model a developer wants.
-RAJA is a viable option for class education in which the educator wants to expose students to the semantics of different programming models without exposing them to the detailed syntax of those models.
+RAJA~\cite{RAJA} is a C++ abstraction library with similarities to Kokkos.
+The important different however is that Kokkos is descriptive, while RAJA is prescriptive.
+In Kokkos, the framework determines how an application is mapped to the underlying hardware.
+RAJA provides functionality that exposes hardware details and relies on the developer to try different strategies to map kernels to architectures.
+RAJA is a viable option for class education in which the educator would like to expose students to the semantics of different programming models without exposing them to the detailed syntax of those models.
 
@@
 
-C++ increasingly supports parallel constructs\cite{CPP}.
+C++ increasingly supports parallel constructs.
 We have highlighted contributions of Kokkos to C++ in~\ref{chap:c++}.

### 2020-01-25 src/9.Conclusion.tex
Minor bug.
Updated Conclusion

--- 2020-01-22 src/9.Conclusion.tex
@@
 \section{Conclusion}\label{chap:conclusion}
-%...
-%Consequently, we emphasize the importance and welcome new development and educational use of parallel programming models that align to this trend.
-We believe that abstracting from vendor or architecture-specific programming models not only simplifies parallel programming in science and industry but also would support educators in teaching parallel programming.
-%We would also like to express gratitude to educators for taking on this important role.
-Further materials, including a detailed description of the programming- and machine model can be found by following the reference on the Kokkos programming mode.
-Finnaly, we would like to mention that interested developers and student can reach the development team on Slack by following the invititation link as mentioned in the reference section.
+In this work we gave an introduction to the Kokkos parallel programming model and presented guiding thoughts behind its design.
+We discussed a machine representation of a parallel hardware architecture that is characterized by a variety of processing units, memories and their hierarchical organization.
+This machine model allowed us to define a constrained semantic represented though different types and patterns.
+Further we listed supported paradigms and considerations that influenced the design of the semantic capture.
+
+An example application gave insight into parallel programming with patterns and into the back-end implementation.
+We used them to show the generic programming interface and type specialization of the implementing classes to support a particular vendor programming model.
+
+To the educational community, we presented Kokkos as a representative programming model of a trend towards support of parallel programming in base languages and libraries.
+We believe in the importance of embracing this trend in computer science education as such parallel programming models can contribute to the generic and relevant education of parallel programming and prepare student for their programming careers.
+
+Finally we would like to express gratitude to educators for taking on this important role.
+We are open to provide further material or support and propose the interested reader to reach out to the development team on Slack\cite{KOKKOS_SLACK}.

### 2020-01-25 src/1.Introduction.tex
changes.

@@
 Chapter~\ref{chap:kokkosMM} presents the abstract machine model used to derive required semantic information for performance portability in Kokkos.
-Chapters~\ref{chap:kokkosPM} and ~\ref{chap:kokkosBackend} present the Kokkos programming model and its back-end support.
+Chapters~\ref{chap:kokkosPM} and ~\ref{chap:kokkosExample} present the Kokkos programming model and an example application.
+In Chapter~\ref{chap:backend} provides insight into the Kokkos library implementation.
 Chapters~\ref{chap:c++} and~\ref{chap:related} show the alignment to C++ standardization efforts and discuss related work and lastly Chapter~\ref{chap:conclusion} concludes this work.

### 2020-01-25 src/5.Example.tex
changes.

@@
 \end{figure}
-
-\section{Kokkos Implementation}
-
-In this section we peak inside the Kokkos library and show how the a generic parallel for loop is mapped to the OpenMP back-end programming model.
-As a template metaprogramming library, Kokkos makes use of partial specialization.
-In particular the~\emph{parallel\_for} or ~\emph{parallel\_reduce} functional calls as shown in Figure~\ref{fig:KokkosExample} are mapped to partially specialized classes.
-Such classes correspond to particular back-end where for each back-end, concepts with partial specialization must be provided.
-Today, Kokkos supports the CUDA and OpenMP back-ends.
-For each new back-end, partially implemented classes that implement necessary concepts must be provided.
-
-A simplified version of a partially specialized class for a parallel for loop implementation is shown in Figure\ref{fig:KokkosExampleOMPBackEnd}.
-The ~\emph{parallel\_for} function instantiates a partial specialization of the ~\emph{ParallelFor} class and calls its ~\emph{execute} function.
-That function is then responsible to implement the Kokkos parallel pattern.
-In this back-end, the parallel implementation is generated by using OpenMP pragma annotations and a by invoking a compatible compiler when the library is compiled.
-\begin{figure}
-\begin{small}
-\begin{Verbatim}[frame=leftline]
-template <class FunctorType>
-class ParallelFor<FunctorType, RangePolicy<OpenMP> > {
-  const FunctorType functor;
-  const RangePolicy<OpenMP> policy;
-
-public:
-  void execute() const {
-    #pragma omp parallel for
-    for(int i=policy.begin(); i<policy.end(); i++)
-      functor(i);
-  }
-}
-
-template <class FunctorType, class PolicyType>
-void parallel_for(std::string label, const PolicyType& policy,
-                  const FunctorType& functor) {
-  // Call tools hook with label
-  ParallelFor<FunctorType,PolicyType> pf{functor,policy};
-  pf.execute();
-}
-
-\end{Verbatim}
-\end{small}
-\caption{This code shows the back-end implementation of a parallel loop in the Kokkos programing model.
-It shows a particular implementation of a generic pattern as a templated class to the OpenMP execution space.
-Also it shows the use of OpenMP pragma annotations that put platform compilers in charge to generate parallel code for the underlying hardware architecture.}
-\label{fig:KokkosExampleOMPBackEnd}
-\end{figure}
-
-%\begin{figure}
-%\centerline{\includegraphics[width=0.3\textwidth]{img/Build.png}}
-%\caption{The compilation workflow includes the invocation of the platform compiler that is in charge to generate parallel   Building workflow}
-%\label{fig:workflow}  
-%\end{figure}

### 2020-01-25 src/6.Backend.tex
changes.

\section{Kokkos Implementation}\label{chap:backend}

In this section we peak inside the Kokkos library and show how the a generic parallel for loop is mapped to the OpenMP back-end programming model. As a template metaprogramming library, Kokkos makes use of partial specialization. In particular the~\emph{parallel\_for} or ~\emph{parallel\_reduce} functional calls as shown in Figure~\ref{fig:KokkosExample} are mapped to partially specialized classes. Such classes correspond to particular back-end where for each back-end, concepts with partial specialization must be provided. Today, Kokkos supports the CUDA and OpenMP back-ends. For each new back-end, partially implemented classes that implement necessary concepts must be provided.

A simplified version of a partially specialized class for a parallel for loop implementation is shown in Figure\ref{fig:KokkosExampleOMPBackEnd}. The ~\emph{parallel\_for} function instantiates a partial specialization of the ~\emph{ParallelFor} class and calls its ~\emph{execute} function. That function is then responsible to implement the Kokkos parallel pattern. In this back-end, the parallel implementation is generated by using OpenMP pragma annotations and a by invoking a compatible compiler when the library is compiled. 
\begin{figure}
\begin{small}
\begin{Verbatim}[frame=leftline]
template <class FunctorType>
class ParallelFor<FunctorType, RangePolicy<OpenMP> > {
  const FunctorType functor;
  const RangePolicy<OpenMP> policy;

public:
  void execute() const {
    #pragma omp parallel for
    for(int i=policy.begin(); i<policy.end(); i++)
      functor(i);
  }
}

template <class FunctorType, class PolicyType>
void parallel_for(std::string label, const PolicyType& policy,
                  const FunctorType& functor) {
  // Call tools hook with label
  ParallelFor<FunctorType,PolicyType> pf{functor,policy};
  pf.execute();
}

\end{Verbatim}
\end{small}
\caption{This code shows the back-end implementation of a parallel loop in the Kokkos programing model. It shows a particular implementation of a generic pattern as a templated class to the OpenMP execution space. Also it shows the use of OpenMP pragma annotations that put platform compilers in charge to generate parallel code for the underlying hardware architecture.}
\label{fig:KokkosExampleOMPBackEnd}
\end{figure}

%\begin{figure}
%\centerline{\includegraphics[width=0.3\textwidth]{img/Build.png}}
%\caption{The compilation workflow includes the invocation of the platform compiler that is in charge to generate parallel   Building workflow}
%\label{fig:workflow}  
%\end{figure}

### 2020-01-25 main.tex
Minor bug.
Rewrite Exercise and Implementation Section
Updated img.
Added new picture.
changes.
Improving bibtex

--- 2020-01-23 main.tex
@@
 \and
-\IEEEauthorblockN{Author Name}
-\IEEEauthorblockA{\textit{Computer Science Reseach Center} \\
+\IEEEauthorblockN{Christian C. Trott}
+\IEEEauthorblockA{\textit{Computer Science Reseach Institute} \\
 \textit{Sandia National Laboratories}\\
 Albuquerque, NM, USA\\
-aname@sandia.gov}
+ctrott@sandia.gov}
 \and
-\IEEEauthorblockN{Author Name}
-\IEEEauthorblockA{\textit{Computer Science Reseach Center} \\
-\textit{Sandia National Laboratories}\\
-Albuquerque, NM, USA\\
-aname@sandia.gov}
-\and
-\IEEEauthorblockN{Author Name}
-\IEEEauthorblockA{\textit{Computer Science Reseach Center} \\
-\textit{Sandia National Laboratories}\\
-Albuquerque, NM, USA\\
-aname@sandia.gov}
+\IEEEauthorblockN{Damien Lebrun-Grandié}
+\IEEEauthorblockA{\textit{Computational Sciences and Engineering} \\
+\textit{Oak Ridge National Laboratory}\\
+Oak Ridge, TN, USA\\
+lebrungrandt@ornl.gov}
 }
@@
 \input{src/1.Introduction}           %label: chap:introduction
-\input{src/2.GenericProgramming.tex} %label: chap:background
-\input{src/3.MachineModel.tex}             %label: chap:kokkosPM
-\input{src/4.ProgammingModel.tex}     %label: chap:kokkosEMA
-\input{src/5.Backend.tex}            %label: chap:kokkosBackend
-\input{src/6.Exercises}              %label: chap:tutorial
-\input{src/7.C++}                    %label: chap:c++
+\input{src/2.GenericProgramming}     %label: chap:background
+\input{src/3.MachineModel}           %label: chap:kokkosPM
+\input{src/4.ProgammingModel}        %label: chap:kokkosEMA
+\input{src/5.Example}                %label: chap:example
+\input{src/6.Backend}                %label: chap:backend
 \input{src/8.Related}                %label: chap:related
@@
 
-\begin{thebibliography}{00}
-\bibitem{b1} G.
-Eason, B.
-Noble, and I.
-N.
-Sneddon, ``On certain integrals of Lipschitz-Hankel type involving products of Bessel functions,'' Phil.
-Trans.
-Roy.
-Soc.
-London, vol.
-A247, pp.
-529--551, April 1955.
-\end{thebibliography}
-
+\bibliographystyle{IEEEtran}
+\bibliography{ref}
 \end{document}

### 2020-01-25 src/1.Introduction.tex
Improving bibtex

@@
 The Kokkos library maps execution primitives to vendor-specific programming models, libraries, and memory layouts.
-Kokkos has been successfully used in many large-scale projects at SNL, University of Utah, Oak-Ridge National Laboratories, Los Alamos National Laboratories and others~\cite{KOKKOSUSECASE2}.
+Kokkos has been successfully used in many large-scale projects at SNL, University of Utah, Oak-Ridge National Laboratories, Los Alamos National Laboratories and others~\cite{KOKKOSPROJS}.
 We list its key properties as follows.
@@
 Chapters~\ref{chap:kokkosPM} and ~\ref{chap:kokkosExample} present the Kokkos programming model and an example application.
-In Chapter~\ref{chap:backend} provides insight into the Kokkos library implementation.
-Chapters~\ref{chap:c++} and~\ref{chap:related} show the alignment to C++ standardization efforts and discuss related work and lastly Chapter~\ref{chap:conclusion} concludes this work.
+In Chapter~\ref{chap:backend} we provide insight into a Kokkos back-end implementation.
+Chapters~\ref{chap:related} discusses related work and lastly Chapter~\ref{chap:conclusion} concludes this work.

### 2020-01-25 src/8.Related.tex
Improving bibtex

@@
 \section{Related Work}\label{chap:related}
+
+C++ increasingly supports parallel constructs.
+It is becoming increasingly possible to teach parallelism in C++ and relying on modern C++ compilers.
+However, at time of publication, C++ does not support accelerators, or many of the complexities of parallel computing related to accelerator programming.
+In courses that do not target accelerators, this can be a relevant option.
+
+Over the past decade the Kokkos programming model has contributed to a number of current and near-future ISO-C++ features, including \emph{atomic\_ref}\cite{wg21_p0019}, \emph{mdspan}\cite{MDSPAN}, and executors.
+In particular, the design of C++ executors, which are similar to Kokkos execution spaces, were influenced by this programming model.
+As the demand to extract performance from increasingly deep and increasingly asynchronous software stacks across a wide variety of domains increases, many C++ experts expect executors to become a central abstraction in any performance-oriented software stack.
+
 RAJA~\cite{RAJA} is a C++ abstraction library with similarities to Kokkos.
 The important different however is that Kokkos is descriptive, while RAJA is prescriptive.
-In Kokkos, the framework determines how an application is mapped to the underlying hardware.
+In Kokkos, the programming models determines how an application is mapped to the underlying hardware.
 RAJA provides functionality that exposes hardware details and relies on the developer to try different strategies to map kernels to architectures.
@@
 In addition to sharing a descriptive philosophy with Kokkos, it also shares parallel patterns.
-Unlike Kokkos, at time of publication it targets only Intel architectures and is not vendor-agnostic.
+Unlike Kokkos, at time of publication it targets Intel architectures only and is not vendor-agnostic.
 A course targeting Intel architectures, DPC++ could be a valid option.
-
-C++ increasingly supports parallel constructs.
-We have highlighted contributions of Kokkos to C++ in~\ref{chap:c++}.
-It is becoming increasingly possible to teach parallelism in C++ and relying on modern C++ compilers.
-However, at time of publication, C++ does not support accelerators, or many of the complexities of parallel computing related to accelerator programming.
-In courses that do not target accelerators, this can be a relevant option.
 

### 2020-01-25 main.tex
improved author formatting.

@@
 \IEEEauthorblockN{David S. Hollman}
-\IEEEauthorblockA{\textit{Computer Science Reseach Institute} \\
+\IEEEauthorblockA{\textit{Computer Science Reseach Institute}\\
 \textit{Sandia National Laboratories}\\
@@
 \and
-\IEEEauthorblockN{Christian C. Trott}
-\IEEEauthorblockA{\textit{Computer Science Reseach Institute} \\
-\textit{Sandia National Laboratories}\\
-Albuquerque, NM, USA\\
-ctrott@sandia.gov}
+\IEEEauthorblockN{
+\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ Christian C. Trott }
+\IEEEauthorblockA{
+\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \textit{Computer Science Reseach Institute} \\
+\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \textit{Sandia National Laboratories}\\
+\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \  Albuquerque, NM, USA\\
+\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ ctrott@sandia.gov}
 \and
-\IEEEauthorblockN{Damien Lebrun-Grandié}
-\IEEEauthorblockA{\textit{Computational Sciences and Engineering} \\
-\textit{Oak Ridge National Laboratory}\\
-Oak Ridge, TN, USA\\
-lebrungrandt@ornl.gov}
+\IEEEauthorblockN{ Damien Lebrun-Grandié\ \ \ \ \ \ \ \ \ \ \ }
+\IEEEauthorblockA{\textit{Computational Sciences and Engineering}\ \ \ \ \ \ \ \ \ \ \ \\
+\textit{Oak Ridge National Laboratory}\ \ \ \ \ \ \ \ \ \ \ \\
+Oak Ridge, TN, USA\ \ \ \ \ \ \ \ \ \ \ \\
+lebrungrandt@ornl.gov\ \ \ \ \ \ \ \ \ \ \ }
 }

### 2020-01-25 src/1.Introduction.tex
Small edits

@@
 Educators strive to balance considerations of architecture, programming model, and theoretical concepts.
-Unfortunately, not all architectures support all programming models, nor do programming models offer enough abstractions to support modern hardware architectures sufficiently .
+Unfortunately, not all architectures support all programming models, nor do many programming models offer enough abstractions to support modern hardware architectures sufficiently.
 Many vendor programming models compound this, requiring students to have access to specific hardware, and limiting the concepts of parallelism to those supported by that hardware.
@@
 We would like to compare this to the support of Von Neumann architectures in programming languages today.
-We define an abstract machine model later in this work.
-The lack of such parallel programming models motivated us to develop Kokkos.
 
-Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS}, is a performance-portable, parallel programming model for C++ developed collaboratively between national laboratories as part of the US Department of Energy Exascale Project\cite{ECP}.
+Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS}, is a performance-portable, parallel programming model for C++ developed collaboratively by a number of national laboratories as part of the US Department of Energy Exascale Project\cite{ECP}.
 It consists of a programming model specification and a library.
 The specification defines execution and memory model and an application programming interface (library API).
-The runtime library exposes programming primitives, namely, patterns and types and maps them to the underlying abstract machine model.
+The runtime library exposes programming primitives, namelyexecution patterns and data structures which it maps to the underlying abstract machine model.
 The machine model creates a generic view of compute hardware and memories.
-The Kokkos library maps execution primitives to vendor-specific programming models, libraries, and memory layouts.
+The Kokkos library maps implements execution primitives on top of vendor-specific programming models, libraries, and memory layouts.
 Kokkos has been successfully used in many large-scale projects at SNL, University of Utah, Oak-Ridge National Laboratories, Los Alamos National Laboratories and others~\cite{KOKKOSPROJS}.
@@
 
-The objective of this paper is to present Kokkos from two perspectives.
+The objective of this paper is to present Kokkos from three perspectives.
 Firstly, we would like to present our thoughts that accompanied the design and development of the Kokkos parallel programming model.
@@
 Finally, we would like to present Kokkos to the educational community as one representative parallel programming model for a trend towards portability abstractions.
-We show a set of types that implement patterns and encapsulate data with enough semantic information to produce performance-portable, parallel code on current architectures.
+We show a set of interfaces that implement patterns and encapsulate data with enough semantic information to produce performance-portable, parallel code on current architectures.
 Further, we highlight its alignment and contributions to standardization efforts towards supporting parallel programming in C++.

### 2020-01-25 src/2.GenericProgramming.tex
Small edits

@@
 Parallel programming models expose parallel semantics to enable developers to create performance-portable code.
-This is commonly achieved by adding annotations to the base programming language, often achieved using \emph{pragma} clauses or similar language extensions.
+This is commonly achieved by adding annotations to the base programming language, often through \emph{pragma} clauses or similar language extensions.
 Other methods expose such semantics through library calls or new languages.
@@
 
-Figures~\ref{figOMPLike} and~\ref{figKokkosLike} show examples of two sample applications using the aforementioned parallel programming syntaxes.
+Figures~\ref{figOMPLike} and~\ref{figKokkosLike} show two code examples using the aforementioned parallel programming syntaxes.
 While both programming models implement the same programming paradigm (parallel patterns), Figures~\ref{figOMPLike} extends the base programming language through pragma annotations.

### 2020-01-25 src/2.GenericProgramming.tex
Syntax highlighting

--- 2020-01-24 src/2.GenericProgramming.tex
@@
 \begin{figure}[h]
-\begin{Verbatim}[frame=leftline]
+\begin{minted}[fontsize=\small]{text}
 Constrained Semantics: memory allocations, 
@@
 Control Paradigm: prescriptive, descriptive
-\end{Verbatim}
+\end{minted}
 \caption{Semantic capture: a set of constrained semantics, parallel paradigms, expression and concern define key properties a parallel programming model.}
@@
 \begin{figure}
-\begin{Verbatim}[frame=leftline]
-# pragma model parallel loop
+\begin{minted}[fontsize=\small]{c++}
+#pragma model parallel loop
 for ( size_t i = 0; i < N; ++i) {
@@
 }
-\end{Verbatim}
+\end{minted}
 \caption{Annotations-based programming models extend the semantic of the base language through pragma annotations.}
@@
 \begin{figure}
-\begin{Verbatim}[frame=leftline]
+\begin{minted}[fontsize=\small]{c++}
 model::parallel_for (N, [=] ( const size_t i) {
@@
 });
-\end{Verbatim}
+\end{minted}
 \caption{Embedded parallel programming uses the base language and relies on programming interfaces to capture information.}

### 2020-01-25 src/4.ProgammingModel.tex
Improving bibtex
Small edits
Syntax highlighting

@@
 \begin{figure}[h]
-\begin{Verbatim}[frame=leftline]
-Constrained Semantics: Patterns (intent), Spaces, 
-Layouts, Policies and Traits
-Parallel Paradigms: Parallel patterns and tasking
+\begin{minted}[fontsize=\small]{text}
+Constrained Semantics: Patterns (intent),
+Spaces, Layouts, Policies and Traits
+Parallel Paradigms: Parallel patterns and 
+tasking
 Expression: Embedded, C++
 Control Paradigm: descriptive
-\end{Verbatim}
+\end{minted}
 \caption{Semantic capture defined by the Kokkos programming model.}
@@
 \begin{figure}
-\begin{Verbatim}[frame=leftline]
+\begin{minted}[fontsize=\small]{c++}
 template <class ExecPolicy, class FunctorType>
@@
  const FunctorType& functor);
-\end{Verbatim}
+\end{minted}
 \caption{The Kokkos~\emph{parallel\_for} class shows how template metaprogramming allows to specialize types.
@@
 Figure~\ref{fig:abstractions} summarizes patterns and abstractions and structures them by information type in the semantic capture.
+
 To the interested reader, the complete programming model description can be accessed on-line\cite{KOKKOS_WIKI}.

### 2020-01-25 src/5.Example.tex
Syntax highlighting

@@
 \begin{small}
-\begin{Verbatim}[frame=leftline]
+\begin{minted}[fontsize=\small]{c++}
 #include<Kokkos_Core.hpp>
@@
     Kokkos::View<double*> y("y", N);
-    Kokkos::View<double**,Kokkos::LayoutRight> A("A", N, M);
-
+    Kokkos::View<double**,Kokkos::LayoutRight> 
+                                  A("A", N, M);  
     auto x_h = Kokkos::create_mirror_view(x);
@@
     auto A_h = Kokkos::create_mirror_view(A);
-    //intialize_data_on_host(x_h,A_h);
     Kokkos::deep_copy(x,x_h);
@@
 
-    Kokkos::parallel_for("outer", TeamPolicy<>(N,AUTO),
+    Kokkos::parallel_for("outer", 
+      TeamPolicy<>(N,AUTO),
       [=](const member_type &team_handle) {
         const int e = team_handle.league_rank();
-      Kokkos::parallel_reduce( TeamThreadRange(team_handle, M),
+        Kokkos::parallel_reduce( 
+          TeamThreadRange(team_handle, M),
           [=](const int & i, double & y_tmp) {
@@
 }
-\end{Verbatim}
+\end{minted}
 \end{small}

### 2020-01-25 src/6.Backend.tex
Improving bibtex
Syntax highlighting

@@
-\section{Kokkos Implementation}\label{chap:backend}
+\section{Back-end Support}\label{chap:backend}
 
@@
 \begin{small}
-\begin{Verbatim}[frame=leftline]
+\begin{minted}[fontsize=\small]{c++}
 template <class FunctorType>
-class ParallelFor<FunctorType, RangePolicy<OpenMP> > {
+class ParallelFor<FunctorType,
+              RangePolicy<OpenMP>> 
+{
   const FunctorType functor;
   const RangePolicy<OpenMP> policy; 
-
 public:
@@
     #pragma omp parallel for
-    for(int i=policy.begin(); i<policy.end(); i++)
+    for(int i=policy.begin();
+            i<policy.end();
+            ++i)
       functor(i);
   }
-}
-
-template <class FunctorType, class PolicyType>
-void parallel_for(std::string label, const PolicyType& policy,
+};
+template <class FunctorType,class PolicyType>
+void parallel_for(std::string label, 
+    const PolicyType& policy,
     const FunctorType& functor) {
-  // Call tools hook with label
-  ParallelFor<FunctorType,PolicyType> pf{functor,policy};
+  // Call profiling tool hooks with label
+  ParallelFor<FunctorType,PolicyType> 
+    pf{functor,policy};
   pf.execute();
 }
-
-\end{Verbatim}
+\end{minted}
 \end{small}

### 2020-01-25 src/3.MachineModel.tex
Updated Conclusion

@@
 The Kokkos abstract machine representation targets a design of a future shared-memory computing architectures.
-The design is shown in Figure~\ref{fig:KokkosSpaces}.
+The design is shown in Figure~\ref{fig:spaces}.
 It is characterized by multiple latency-optimized execution units and off-die bandwidth-optimized accelerators.
@@
 Accelerators can provide an additional memory space through its local on-package memory.
-This is highligted in Figure~\ref{fig:KokkosSpaces} in a green color.
-shows memory spaces in .
+This is also highligted in Figure~\ref{fig:spaces}.
 The programmer is free to decide where each data structure is allocated by requesting the corresponding memory space.
@@
 
-How a machine model is exposed to the programmer and what are the design considerations towards defining an appropriate semantic capture are discussed in the next section.
+How a machine model is exposed to the programmer and what the design considerations are towards defining an appropriate semantic capture are discussed in the next section.
 

### 2020-01-25 src/1.Introduction.tex
Fix a few typos

@@
 The specification defines execution and memory model and an application programming interface (library API).
-The runtime library exposes programming primitives, namelyexecution patterns and data structures which it maps to the underlying abstract machine model.
+The runtime library exposes programming primitives, namely execution patterns and data structures, which it maps to the underlying abstract machine model.
 The machine model creates a generic view of compute hardware and memories.
-The Kokkos library maps implements execution primitives on top of vendor-specific programming models, libraries, and memory layouts.
-Kokkos has been successfully used in many large-scale projects at SNL, University of Utah, Oak-Ridge National Laboratories, Los Alamos National Laboratories and others~\cite{KOKKOSPROJS}.
+The Kokkos library implements execution primitives on top of vendor-specific programming models, libraries, and memory layouts.
+Kokkos has been successfully used in many large-scale projects at Sandia National Laboratories, University of Utah, Oak Ridge National Laboratory, Los Alamos National Laboratory and others~\cite{KOKKOSPROJS}.
 We list its key properties as follows.

### 2020-01-25 src/4.ProgammingModel.tex
Fix a few typos

@@
 
-The Kokkos programming model specifies a language, programming paradigm and an API.
-The API exposes the machine representation to the developer and defines a set of pattens and types that capture intents and properties of execution.
+The Kokkos programming model specifies a programming paradigm and an API.
+The API exposes the machine representation to the developer and defines a set of patterns and types that capture intents and properties of execution.
 This completes the semantic capture, discussed in Chapter~\ref{chap:background}, We detail all three components as follows.
@@
 Supported patterns are \emph{parallel\_for} loop that executes the loop body in any order a specified amount of times, the \emph{parallel\_reduce} which combines a parallel\_for with a reduction operation, \emph{parallel\_scan} which combines a parallel\_for operation with a prefix or postfix scan, and \emph{task} which executes a single function potentially in parallel in respect to other tasks or code sections.
-	\item Execution policy shape the iteration space of a loop pattern. A simple execution policy is a range policy. It specifies that the loop body is executed once for each element in a range.
+	\item Execution policies shape the iteration space of a loop pattern. A simple execution policy is a range policy. It specifies that the loop body is executed once for each element in a range.
 	\item Memory spaces are the places where data resides.

### 2020-01-25 src/5.Example.tex
Fix a few typos

@@
 If the original data is already accessible by the host process, this function returns a ~\emph{View} referencing the same allocation.
-In real applications host-accessible allocation is necessary in order to allow I/O operations during fileystem access.
+In real applications host-accessible allocation is necessary in order to allow I/O operations during filesystem access.
 To move data to the respective memory space that is accessible by the parallel execution resources, the function ~\emph{deep\_copy} has to be invoked.
@@
 In case of the matrix-vector multiplication,This is accomplished by using the ~\emph{Kokkos::TeamPolicy}.
-For each row of~\emph{M} a team of threads is launched, where the size of the team is not further specified.
+For each row of~\emph{A} a team of threads is launched, where the size of the team is not further specified.
 When using a ~\emph{TeamPolicy} the operator of the lambda expression does not receive an index, but rather a handle to the team of threads.

### 2020-02-10 src/0.Abstract.tex
Slit text into multiple files.
Fix a few typos
Minor changes. Added pdf output.

@@
 In academia, this can contribute to a generic eduction of performance-portable parallel programming.
-In this work we present Kokkos from the perspective of educators, highlight key concepts and give an instruction to examples suited for incremental teaching of parallel programming in class.
+In this work we present Kokkos from the perspective of educators, highlight key concepts and point the reader to examples and tutorials suited for incremental teaching of parallel programming in class.
 \end{abstract}

### 2020-02-10 src/1.Introduction.tex
Minor changes. Added pdf output.

--- 2020-01-25 src/1.Introduction.tex
@@
 
-Teaching generic parallel programming today is difficult.
+Education of parallel programming today is difficult.
 Educators strive to balance considerations of architecture, programming model, and theoretical concepts.
@@
 The specification defines execution and memory model and an application programming interface (library API).
-The runtime library exposes programming primitives, namelyexecution patterns and data structures which it maps to the underlying abstract machine model.
+The runtime library exposes programming primitives, namely execution patterns and data structures which it maps to the underlying abstract machine model.
 The machine model creates a generic view of compute hardware and memories.

### 2020-08-10 main.tex
Syntax highlighting
Updated Title, Abstract.

@@
 \usepackage{graphicx}
+\usepackage{minted}
 \usepackage{textcomp}
@@
 
-\title{Performance-portable Parallel Programming in Computer Science Education with Kokkos}
+\title{Towards Generic Parallel Programming in Computer Science Education with Kokkos}
 
@@
 \begin{IEEEkeywords}
-Parallel programming, accelerator programming, Kokkos, C++
+parallel programming, C++, performance portability, programming model
 \end{IEEEkeywords}
 
-\input{src/1.Introduction}           %label: chap:introduction
-\input{src/2.GenericProgramming}     %label: chap:background
-\input{src/3.MachineModel}           %label: chap:kokkosPM
-\input{src/4.ProgammingModel}        %label: chap:kokkosEMA
-\input{src/5.Example}                %label: chap:example
-\input{src/6.Backend}                %label: chap:backend
-\input{src/8.Related}                %label: chap:related
-\input{src/9.Conclusion}             %label: chap:conclusion
+% \input{src/1.Introduction}           %label: chap:introduction
+%\input{src/2.GenericProgramming}     %label: chap:background
+%\input{src/3.MachineModel}           %label: chap:kokkosPM
+%\input{src/4.ProgammingModel}        %label: chap:kokkosEMA
+%\input{src/5.Example}                %label: chap:example
+%\input{src/6.Backend}                %label: chap:backend
+%\input{src/8.Related}                %label: chap:related
+%\input{src/9.Conclusion}             %label: chap:conclusion
 

### 2020-08-10 src/0.Abstract.tex
Updated Title, Abstract.

--- 2020-01-20 src/0.Abstract.tex
@@
 \begin{abstract}\label{chap:abstract}
-Kokkos is a generic, performance-portable parallel programming model for host- and accelerator architectures in C++.
+Parallel patterns, views and spaces are promising abstractions to capture the programmer's intent as well as the contextual information that can be used by an underlying runtime to efficiently map software to parallel hardware.
+These abstractions can be valuable in cases where an algorithm must accommodate requirements of code and performance portability across hardware architectures and vendor programming models.
+Kokkos is a parallel programing model for host- and accelerator architectures that relies on these abstractions and targets these requirements.
 It consists of a pure C++ interface, a specification and a programming library.
 The programming library exposes patterns and types and maps them to an underlying abstract machine model.
-This machine model offers a generic view of parallel hardware by abstracting from architecture- and vendor-specific programming models.
-In academia, this can contribute to a generic eduction of performance-portable parallel programming.
-In this work we present Kokkos from the perspective of educators, highlight key concepts and give an instruction to examples suited for incremental teaching of parallel programming in class.
+The abstract machine model offers a generic view of parallel hardware.
+While Kokkos is gaining popularity in large-scale HPC applications at some DOE laboratories, we believe that the implemented concepts are of interest to a broader audience including academia as they may contribute to creating a generic view on parallel programming.
+In this work we give an insight into design considerations of a representative programming model, list important abstractions, document feedback received from workshops and give pointers to resources that the reader may consider valuable for a lecture on generic parallel programming for students with preexisting knowledge on this matter.
 \end{abstract}
-
-%is aligned with C++ standardization efforts and trends towards parallel programming support in base languages.

### 2020-08-10 main.tex
Fix typos highlighted by the reviewers.

--- 2020-01-25 main.tex
@@
 \begin{IEEEkeywords}
-Parallel programming, accelerator programming, Kokkos, C++
+parallel programming, C++, performance portability, programming model
 \end{IEEEkeywords}

### 2020-08-10 src/0.Abstract.tex
Fix typos highlighted by the reviewers.

--- 2020-01-20 src/0.Abstract.tex
@@
 \begin{abstract}\label{chap:abstract}
-Kokkos is a generic, performance-portable parallel programming model for host- and accelerator architectures in C++.
-It consists of a pure C++ interface, a specification and a programming library.
+Parallel patterns, views, and spaces are promising abstractions to capture the programmer's intent as well as the contextual information that can be used by an underlying runtime to efficiently map software to parallel hardware.
+These abstractions can be valuable in cases where an algorithm must accommodate requirements of code and performance portability across hardware architectures and vendor programming models.
+Kokkos is a parallel programming model for host- and accelerator architectures that relies on these abstractions and targets these requirements.
+It consists of a pure C++ interface, a specification, and a programming library.
 The programming library exposes patterns and types and maps them to an underlying abstract machine model.
-This machine model offers a generic view of parallel hardware by abstracting from architecture- and vendor-specific programming models.
-In academia, this can contribute to a generic eduction of performance-portable parallel programming.
-In this work we present Kokkos from the perspective of educators, highlight key concepts and give an instruction to examples suited for incremental teaching of parallel programming in class.
+The abstract machine model offers a generic view of parallel hardware.
+While Kokkos is gaining popularity in large-scale HPC applications at some DOE laboratories, we believe that the implemented concepts are of interest to a broader audience including academia as they may contribute to parallel programming productivity.
+In this work, we give an insight into design considerations of a representative programming model, list important abstractions, document feedback received during workshops, and give pointers to resources that the reader may consider valuable for a lecture on generic parallel programming for students with preexisting knowledge on this matter.
 \end{abstract}
-
-%is aligned with C++ standardization efforts and trends towards parallel programming support in base languages.

### 2020-08-10 src/1.Introduction.tex
Fix typos highlighted by the reviewers.

@@
-\section{Introduction}\label{chap:introduction}
-
 Parallel and distributed computing is a focus at universities and teaching institutions.
@@
 While industry-driven collaborations often predefine technological frameworks such as programming environments and targeted architectures, academic teaching leaves this mostly undefined.
-In this case, educators are exposed to a variety of solutions, approaches and trends in parallel computing.
+In this case, educators can select from a variety of solutions, approaches, and trends in parallel computing.
 Which particular set to choose for teaching in class is not trivial.
-While decision factors differ, maximizing the subject relevance by teaching generic principles of parallel programming generically is among the most important ones.
+Complexity is further amplified by the breath of related topics ranging from parallel hardware architectures, programming models, algorithms as well as productivity in parallel software development.
+However, it is the productivity concern that is gaining importance as the number of hardware solutions, their complexity as well as demands for a quick time-to-solution increase.
 
-Education of parallel programming today is difficult.
-Educators strive to balance considerations of architecture, programming model, and theoretical concepts.
-Unfortunately, not all architectures support all programming models, nor do many programming models offer enough abstractions to support modern hardware architectures sufficiently.
-Many vendor programming models compound this, requiring students to have access to specific hardware, and limiting the concepts of parallelism to those supported by that hardware.
+While vendor programming models maintain their position in curricula today, we think that the inclusion of parallel programming models such as Kokkos is worth consideration.
+They offer an easy to understand set of concepts for a vendor- and architecture-independent, performance-portable expression of concurrency which significantly contributes to parallel programming productivity.
 
-Consequently, generic parallel programming models are needed that target mainstream programming languages and allow a vendor- and architecture independent, performance-portable expression of concurrency.
-In this way, a parallel program written in a popular language becomes capable of running efficiently on parallel computer architectures that conform to a common parallel machine specification.
-We would like to compare this to the support of Von Neumann architectures in programming languages today.
-
-Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS}, is a performance-portable, parallel programming model for C++ developed collaboratively by a number of national laboratories as part of the US Department of Energy Exascale Project\cite{ECP}.
+Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS} is a performance-portable, parallel programming model for C++ developed collaboratively by several national laboratories as part of the US Department of Energy Exascale Project\cite{ECP}.
 It consists of a programming model specification and a library.
-The specification defines execution and memory model and an application programming interface (library API).
-The runtime library exposes programming primitives, namely execution patterns and data structures which it maps to the underlying abstract machine model.
-The machine model creates a generic view of compute hardware and memories.
-The Kokkos library maps implements execution primitives on top of vendor-specific programming models, libraries, and memory layouts.
-Kokkos has been successfully used in many large-scale projects at SNL, University of Utah, Oak-Ridge National Laboratories, Los Alamos National Laboratories and others~\cite{KOKKOSPROJS}.
+The specification defines the execution and memory model and an application programming interface (library API).
+The runtime library exposes programming primitives, namely execution patterns and types, including views and execution spaces and maps them to the underlying abstract machine model.
+The machine model creates a generic view of hardware and memories.
+Finally, the Kokkos library implements the mapping of the programming model to vendor-specific programming models, libraries, and memory layouts.
+Kokkos has been successfully used in many large-scale projects at Sandia National Laboratories, Oak-Ridge National Laboratories, Los Alamos National Laboratories, University of Utah, and others~\cite{KOKKOSPROJS}.
 We list its key properties as follows.
@@
 \begin{itemize}
-\item Parallel programming model for C++
 \item Targets an abstract machine model
-\item Implements parallel patterns and tasking paradigms
-\item Offers abstract data types for performance portability
-\item Pure C++ interface through template metaprogramming
+\item Implements patterns, views, and spaces
+\item Vendor- and architecture-independent and open-source
 \item Aligns with C++ standardization efforts for parallel programming
-\item Vendor- and architecture independent and open-source
 \item Includes exercises and teaching material
@@
 Firstly, we would like to present our thoughts that accompanied the design and development of the Kokkos parallel programming model.
-These insights may be  useful in parallel programming classes that introduce programming models and reason about design choices.
+These insights may be useful in parallel programming classes that introduce programming models, reason about design choices, and cover productivity in parallel software development.
 
-Further, we would like to promote Kokkos to students and educators as a platform that helps them to understand and experiment with parallel patterns in a fashion that is generic, vendor-agnostic and declarative.
-This means that students can learn about parallelism without being exposed to many architectural concerns and that they can see productivity benefits of such abstractions for performance portability when programming on current and upcoming hardware architectures.
+Further, we would like to promote Kokkos to students and educators as a platform that helps them to understand and experiment with parallel patterns in a way that is generic, vendor-agnostic, and declarative.
+Abstractions allow students to learn about concurrency without being exposed to many architectural concerns initially and help to create awareness of their productivity benefits in terms of code- and performance portability on current and upcoming hardware architectures.
 
-Finally, we would like to present Kokkos to the educational community as one representative parallel programming model for a trend towards portability abstractions.
-We show a set of interfaces that implement patterns and encapsulate data with enough semantic information to produce performance-portable, parallel code on current architectures.
+Finally, we would like to present Kokkos to the educational community as one representative parallel programming model for a trend towards portability abstractions in parallel programming models.
+We show a set of interfaces that implement patterns, views, and spaces, and capture enough semantic information to produce performance-portable, parallel code on current architectures.
 Further, we highlight its alignment and contributions to standardization efforts towards supporting parallel programming in C++.
+For teaching insights, we document feedback received during on-line lectures.
 
 The rest of this paper is structured as follows.
-The next Chapter gives an overview on parallel programming concepts and presents the notion of \emph{semantic capture}.
-Chapter~\ref{chap:kokkosMM} presents the abstract machine model used to derive required semantic information for performance portability in Kokkos.
+The next chapter gives an overview of parallel programming concepts and presents the notion of \emph{semantic capture}.
+Chapter~\ref{chap:kokkosMM} presents the abstract machine model used to derive the required semantic information for performance portability in Kokkos.
 Chapters~\ref{chap:kokkosPM} and ~\ref{chap:kokkosExample} present the Kokkos programming model and an example application.
 In Chapter~\ref{chap:backend} we provide insight into a Kokkos back-end implementation.
-Chapters~\ref{chap:related} discusses related work and lastly Chapter~\ref{chap:conclusion} concludes this work.
+Chapters~\ref{chap:related} and ~\ref{chap:tutorials} discuss related work, tutorials and student feedback from tutorials.
+Lastly Chapter~\ref{chap:conclusion} concludes this work.

### 2020-08-10 src/2.GenericProgramming.tex
Fix typos highlighted by the reviewers.

@@
 Some representatives are the OpenMP programming model~\cite{OPENMP}, threading libraries, or languages like Erlang\cite{ERLANG} in which parallelism is a first-class citizen.
-In general, it is up to the parallel programming model implementer to expose language constructs or library interfaces (APIs) that combine convenience to the programmer with sufficiently constrained semantics to map the program execution to the parallel computer hardware efficiently and correctly.
+In general, it is up to the parallel programming model implementer to expose language constructs or library interfaces (APIs) that combine convenience to the programmer with a sufficiently constrained semantics to map the program execution to the parallel computer hardware efficiently and correctly.
 Questions of which semantic information is needed, what paradigm to present, and by what syntactic means to do so span an exciting design space that is well worth exploring.
@@
 \end{minted}
-\caption{Semantic capture: a set of constrained semantics, parallel paradigms, expression and concern define key properties a parallel programming model.}
+\caption{Semantic capture: a set of constrained semantics, parallel paradigms, expression and concern define key properties of a parallel programming model.}
 \label{figSemCapture}
@@
 
-A \emph{Parallel Paradigm} is an abstract representation with the purpose of facilitating the programmer's understanding of programming rules and program behavior.
+A \emph{Parallel Paradigm} is an abstract representation to facilitate the programmer's understanding of programming rules and program behavior.
 Which programming paradigm to choose depends on several considerations.
@@
 Distributed and correctness-oriented programming models may implement actor-based programming, where each unit of execution represents an actor who communicates over predefined communication channels.
-This paradigm eliminates accesses to shared state and aligns well with message passing programming (MPI).
+This paradigm eliminates access to a shared state and aligns well with message passing programming (MPI).
 The execution model and memory model aspects of a programming model define its behavior.
@@
 
-A language defines the syntax of the parallel semantic.
+A language defines the syntax of the parallel semantics.
 Annotations-based languages add semantic information to the base language through pragmas while embedded programming models rely on the base language and offer additional programming interfaces to capture information.
-Their semantic is defined in the API specification.
+Their semantics is defined in the API specification.
 
 Lastly, a programming model must divide responsibility between the developer and the model.
-We refer to is as defining the \emph{Control Paradigm}.
+We refer to this as defining the \emph{Control Paradigm}.
 In practice, this is a choice between defining a descriptive and prescriptive semantics.
-Descriptive semantics require a developer to express intent.
+Descriptive semantics requires a developer to express intent.
 The implementation is the responsibility of the programming model and of the underlying toolchain.
-Prescriptive semantics require the developer to specify the mechanisms by which their intent will be implemented.
+Prescriptive semantics requires the developer to specify the mechanisms by which their intent will be implemented.
 In a prescriptive model, developer success is tied to their knowledge of implementation mechanisms.
-In a descriptive model that success is tied to the developer's ability to accurately describe their requirements and the toolchain's ability to turn that into an implementation.
+In a descriptive model that success is tied to the developer's ability to accurately describe their requirements and the toolchain's ability to turn that into implementation.
 
@@
 While both programming models implement the same programming paradigm (parallel patterns), Figures~\ref{figOMPLike} extends the base programming language through pragma annotations.
-Both, pragma annotations and an API call as shown in Figure~\ref{figKokkosLike} can have a descriptive or prescriptive semantic of describing a parallel loop construct.
-It is up to the programming model specification to make further definitions regarding separation of concern.
+Both, pragma annotations and an API call as shown in Figure~\ref{figKokkosLike} can have a descriptive or prescriptive semantics of describing a parallel loop construct.
+It is up to the programming model specification to make further definitions regarding the separation of concern.
 
-Both examples provide the semantic information on what to parallelize, however, they do not expose enough abstraction primitives that would allow to map parallel execution on modern architectures efficiently.
-In order to define what further information is needed to express concurrency on modern computer architectures, it is important to define an~\emph{abstract machine representation} first.
+Both examples provide the semantic information on what to parallelize, however, they do not expose enough abstraction primitives that would allow them to map parallel execution on modern architectures efficiently.
+To define what further information is needed to express concurrency on modern computer architectures, it is important to define an~\emph{abstract machine representation} first.
 In the following chapters we use the words \emph{representation} and \emph{model} interchangeably.
-In this work the abstract machine model does not describe a computer in terms of automata theory but in terms of physical or functional components.
+In this work, the abstract machine model does not describe a computer in terms of automata theory but in terms of physical or functional components.
 

### 2020-08-10 src/3.MachineModel.tex
Fix a few typos
Fix typos highlighted by the reviewers.

@@
 The Kokkos parallel programming model is one particular instance of a programming model that builds on top of that machine model.
-We would like to point out that it is this conceptual differentiation that allows scenarios where the underlying machine model is instantiated in other languages beyond C++, yet the algorithmic specification as well as performance characteristic and portability remain the same.
+We would like to point out that it is this conceptual differentiation that allows scenarios where the underlying machine model is instantiated in other languages beyond C++, yet the algorithmic specification as well as performance characteristics and portability remain the same.
 
-The Kokkos abstract machine representation targets a design of a future shared-memory computing architectures.
+The Kokkos abstract machine representation targets a design of a future shared-memory computing architecture.
 The design is shown in Figure~\ref{fig:spaces}.
@@
 In such an architecture, execution units might be hierarchically organized with multiple levels of achievable parallelism with different memory access characteristics and coherence properties of caches.
-In order to ensure performance portability on such a wide range of configurations, an abstraction of compute resources and available memories are required.
-For this purpose we introduce the abstract concept called \emph{space}.
+In order to ensure performance portability on such a wide range of configurations, an abstraction of compute-resources and available memories is required.
+For this purpose, we introduce the abstract concept called \emph{space}.
 
@@
 An instance of a memory space provides a concrete method for the application programmer to request data storage allocations.
-Following the architecture as discussed earlier, the multi-core processor may contain multiple memory spaces that abstract on-package memory, slower DRAM and non-volatile memories.
+Following the architecture as discussed earlier, the multi-core processor may contain multiple memory spaces that abstract on-package memory, slower DRAM, and non-volatile memories.
 Accelerators can provide an additional memory space through its local on-package memory.
-This is also highligted in Figure~\ref{fig:spaces}.
+This is also highlighted in Figure~\ref{fig:spaces}.
 The programmer is free to decide where each data structure is allocated by requesting the corresponding memory space.

### 2020-08-10 src/4.ProgammingModel.tex
Fix typos highlighted by the reviewers.

@@
 This has multiple reasons.
-It is our experience that application developers of scientific and high-performance computing applications in the high-performance computing community and in industry in general express the demand for C++ and a generic parallel programming model to support their codes.
-Further, following their estimates, porting efforts towards adapting to vendors, programming models, APIs and software releases represents a burden worth addressing.
+It is our experience that application developers of scientific and high-performance computing applications in the high-performance computing community and in the industry in general express the demand for C++ and a generic parallel programming model to support their codes.
+Further, following their estimates, porting efforts towards adapting to vendors, programming models, APIs, and software releases represents a burden worth addressing.
 Lastly, the possibility to maintain pure C++ codes is appealing from the code development and debugging perspective.
@@
 From the programming model perspective, C++ offers template metaprogramming which is well suited to implement generic APIs and libraries.
-In this case, class specialization and templating allow for compile-type generated types and optimizations for a given hardware architecture.
+In this case, class specialization and templating allow for compile-time generated types and optimizations for a given hardware architecture.
 
@@
 To capture semantic information, that is the~\emph{what}, the~\emph{where} and the~\emph{how}, Kokkos introduces six abstractions:~\emph{execution spaces},~\emph{execution patterns},~\emph{execution policies},~\emph{memory spaces},~\emph{memory layout} and~\emph{memory traits}.
-These abstractions specify semantic information that enable to capture the programmer's intent and allow the runtime to efficiently map the program to any underlying hardware architecture.
+These abstractions specify semantic information that enables to capture the programmer's intent and allows the runtime to efficiently map the program to any underlying hardware architecture.
 We list them as follows:
@@
 	\item  An execution space is a place where code can be executed.
-On current hardware architectures this correspond to accelerators and CPUs and can include any compute device in the future.
+On current hardware architectures, this corresponds to accelerators and CPUs and can include other compute devices in the future.
 This abstraction supports remote compute devices in distributed memory scenarios as remote execution spaces.
@@
 	\item Memory spaces are the places where data resides.
-They specify physical locations of data as well as access characteristics.
-Different physical locations correspond to different device types such as high bandwidth memories, on-die scratch memories or non-volatile bulk storage.
+They specify the physical locations of data as well as access characteristics.
+Different physical locations correspond to different device types such as high bandwidth memories, on-die scratch memories, or non-volatile bulk storage.
 Different logical memory spaces allow for concepts such as memory in the CUDA programming model, which is accessible from the host and the CUDA accelerator.
@@
 	\item Memory traits specify how a data structure is accessed.
-Traits express usage scenarios such as atomic access, random access and streaming loads or stores.
+Traits express usage scenarios such as atomic access, random access, and streaming loads or stores.
 This allows the programming model to optimize load and store operations.
@@
 As the name suggests, these policies express iteration spaces.
-For example, a range policy defines a one dimensional iteration range.
+For example, a range policy defines a one-dimensional iteration range.
 The MDRangePolicy represents a multi-dimensional iteration space and is used to express tightly nested loop patterns.
@@
 This policy allows the expression of hierarchical parallelism.
-The second template parameter,~\emph{FuntorType}, represent a functor that implements the~\emph{()-operator} with a matching signature for the given execution policy.
+The second template parameter,~\emph{FunctorType}, represent a functor that implements the~\emph{()-operator} with a matching signature for the given execution policy.
 The functor can be defined using a C++ class, struct or lambda expression.
@@
 To guarantee a kernel has finished, a developer should call the fence of the execution space on which the kernel is being executed.
-Otherwise it depends on the execution space where the loop executes and weather this execution space implements a barrier.
+Otherwise, it depends on the execution space where the loop executes and whether this execution space implements a barrier.
 
@@
 Programming with patterns and abstractions allows the compiler to generate optimized transformations.
-Generally parallel patterns in Kokkos do not guarantee iteration ordering nor degree of concurrency during their execution.
-This gives the freedom to the underlying abstraction layers to implement different mapping patterns on different hardware such as assignment of iterations to threads or vector lanes.
-We believe that these abstraction and their descriptive semantic represent key programming primitives and the way forward for generic support for parallel programming.
+Generally, parallel patterns in Kokkos do not guarantee iteration ordering nor the degree of concurrency during their execution.
+This gives the freedom to the underlying abstraction layers to implement different mapping patterns on different hardware such as the assignment of iterations to threads or vector lanes.
+We believe that these abstractions and their descriptive semantic represent key programming primitives and the way forward for generic support for parallel programming.
 In conclusion, Figure~\ref{fig:SemCaptureKokkos} shows the semantic capture defined by the Kokkos programming model.
@@
 To the interested reader, the complete programming model description can be accessed on-line\cite{KOKKOS_WIKI}.
-The next section shows an example of the Kokkos parallel\_for and explains how it is mapped to the the underlying back-end programming model.
+The next section shows an example of the Kokkos parallel\_for and explains how it is mapped to the underlying back-end programming model.
 

### 2020-08-10 src/5.Example.tex
Fix typos highlighted by the reviewers.

@@
 Views are abstractions that represent data.
-Naturally data can reside in host memory where it is accessible to the host execution space or on device memory where it is accessible to the device execution space.
-In this example, copies of the views are created in order to make them accessible by the host process.
+Naturally, data can reside in host memory where it is accessible to the host execution space or on device memory where it is accessible to the device execution space.
+In this example, copies of the views are created to make them accessible by the host process.
 This is accomplished by calling the ~\emph{create\_mirror\_view} function.
@@
 If the original data is already accessible by the host process, this function returns a ~\emph{View} referencing the same allocation.
-In real applications host-accessible allocation is necessary in order to allow I/O operations during filesystem access.
+In real applications, a host-accessible allocation is necessary to allow I/O operations during filesystem access.
 To move data to the respective memory space that is accessible by the parallel execution resources, the function ~\emph{deep\_copy} has to be invoked.
@@
 
-%A general philosophy of Kokkos is to have defaults for most properties, including data layouts, memory spaces and execution spaces.
-Consequently the simple example does not explicitly set which memory space the data lives in.
-Kokkos ensures that the defaults for parallel execution and data allocations match though.
+Consequently, the simple example does not explicitly specify a memory space where the data lives in. However, Kokkos ensures that the defaults for parallel execution and data allocations match.
 
 Taking a closer look at the loop step shows the hierarchical expression of concurrency.
-In case of the matrix-vector multiplication,This is accomplished by using the ~\emph{Kokkos::TeamPolicy}.
+This is accomplished by using the ~\emph{Kokkos::TeamPolicy}.
 For each row of~\emph{A} a team of threads is launched, where the size of the team is not further specified.
@@
 It then writes the result to the correct position of the output vector.
-The parallel loops are follows by a call to \emph{Kokkos::fence}.
+The parallel loops are followed by a call to \emph{Kokkos::fence}.
 This ensures that the execution of the parallel code is completed before copying the results back to the host memory space.
@@
 While this example can only provide a brief insight into Kokkos, we recommend to the interested reader to access further resources.
-Resources include example applications for individual Kokkos features, tutorials with multiple days worth of lectures and exercises, a programming guide and an API documentation.
+Resources include example applications for individual Kokkos features, tutorials with multiple days worth of lectures and exercises, a programming guide, and API documentation.
 Source code examples as well as the tutorial are located in the Kokkos repository~\cite{KOKKOS_REPO}.
 The tutorial is intended for students with minimal or even no prior knowledge of parallel programming.
-It introduces concepts of parallel programming using Kokkos through a series of lectures and hands-on exercises which build up on each other and gradually introduce new concepts.
+It introduces concepts of parallel programming using Kokkos through a series of lectures and hands-on exercises that build upon each other and gradually introduce new concepts.
 The programming guide is available on-line~\cite{KOKKOS_WIKI}.

### 2020-08-10 src/6.Backend.tex
Fix a few typos
Fix typos highlighted by the reviewers.

@@
 
-In this section we peak inside the Kokkos library and show how the a generic parallel for loop is mapped to the OpenMP back-end programming model.
+In this section we peek inside the Kokkos library and show how a generic parallel for loop is mapped to the OpenMP backend programming model.
 As a template metaprogramming library, Kokkos makes use of partial specialization.
 In particular the~\emph{parallel\_for} or ~\emph{parallel\_reduce} functional calls as shown in Figure~\ref{fig:KokkosExample} are mapped to partially specialized classes.
-Such classes correspond to particular back-end where for each back-end, concepts with partial specialization must be provided.
-Today, Kokkos supports the CUDA and OpenMP back-ends.
+Such classes correspond to a particular back-end where for each back-end, concepts with partial specialization must be provided.
+Today, Kokkos supports the CUDA, OpenMP, HIP as well as one-sided MPI, SHMEM and NVSHMEM back-ends.
 For each new back-end, partially implemented classes that implement necessary concepts must be provided.
@@
 \end{small}
-\caption{This code shows the back-end implementation of a parallel loop in the Kokkos programing model.
+\caption{This code shows the back-end implementation of a parallel loop in the Kokkos programming model.
 It shows a particular implementation of a generic pattern as a templated class to the OpenMP execution space.
-Also it shows the use of OpenMP pragma annotations that put platform compilers in charge to generate parallel code for the underlying hardware architecture.}
+Also, it shows the use of OpenMP pragma annotations that put platform compilers in charge to generate parallel code for the underlying hardware architecture.}
 \label{fig:KokkosExampleOMPBackEnd}
 \end{figure}
-
-%\begin{figure}
-%\centerline{\includegraphics[width=0.3\textwidth]{img/Build.png}}
-%\caption{The compilation workflow includes the invocation of the platform compiler that is in charge to generate parallel   Building workflow}
-%\label{fig:workflow}  
-%\end{figure}

### 2020-08-10 src/8.Related.tex
Fix typos highlighted by the reviewers.

@@
 It is becoming increasingly possible to teach parallelism in C++ and relying on modern C++ compilers.
-However, at time of publication, C++ does not support accelerators, or many of the complexities of parallel computing related to accelerator programming.
+However, at the time of publication, C++ does not support accelerators or many of the complexities of parallel computing related to accelerator programming.
 In courses that do not target accelerators, this can be a relevant option.
@@
 RAJA~\cite{RAJA} is a C++ abstraction library with similarities to Kokkos.
-The important different however is that Kokkos is descriptive, while RAJA is prescriptive.
-In Kokkos, the programming models determines how an application is mapped to the underlying hardware.
+The important difference however is that Kokkos is descriptive, while RAJA is prescriptive.
+In Kokkos, the programming models determine how an application is mapped to the underlying hardware.
 RAJA provides functionality that exposes hardware details and relies on the developer to try different strategies to map kernels to architectures.
@@
 In addition to sharing a descriptive philosophy with Kokkos, it also shares parallel patterns.
-Unlike Kokkos, at time of publication it targets Intel architectures only and is not vendor-agnostic.
+Unlike Kokkos, it targets Intel architectures only and is not vendor-agnostic.
 A course targeting Intel architectures, DPC++ could be a valid option.

### 2020-08-10 src/9.Conclusion.tex
Fix typos highlighted by the reviewers.

@@
 In this work we gave an introduction to the Kokkos parallel programming model and presented guiding thoughts behind its design.
-We discussed a machine representation of a parallel hardware architecture that is characterized by a variety of processing units, memories and their hierarchical organization.
+We discussed a machine representation of a parallel hardware architecture that is characterized by a variety of processing units, memories, and their hierarchical organization.
 This machine model allowed us to define a constrained semantic represented though different types and patterns.
-Further we listed supported paradigms and considerations that influenced the design of the semantic capture.
+Further, we listed supported paradigms and considerations that influenced the design of the semantic capture.
 
@@
 
-To the educational community, we presented Kokkos as a representative programming model of a trend towards support of parallel programming in base languages and libraries.
-We believe in the importance of embracing this trend in computer science education as such parallel programming models can contribute to the generic and relevant education of parallel programming and prepare student for their programming careers.
+To the educational community, we presented Kokkos as a representative programming model of a trend towards the support of parallel programming in base languages and libraries.
+We believe in the importance of embracing this trend in computer science education as such parallel programming models can contribute to the generic and relevant education of parallel programming and prepare students for their programming careers.
 
-Finally we would like to express gratitude to educators for taking on this important role.
+Finally, we would like to express gratitude to educators for taking on this important role.
 We are open to provide further material or support and propose the interested reader to reach out to the development team on Slack\cite{KOKKOS_SLACK}.

### 2020-08-15 src/0.Abstract.tex
Updates for submission to EduHPC at SC20

@@
 The abstract machine model offers a generic view of parallel hardware.
-While Kokkos is gaining popularity in large-scale HPC applications at some DOE laboratories, we believe that the implemented concepts are of interest to a broader audience including academia as they may contribute to parallel programming productivity.
-In this work, we give an insight into design considerations of a representative programming model, list important abstractions, document feedback received during workshops, and give pointers to resources that the reader may consider valuable for a lecture on generic parallel programming for students with preexisting knowledge on this matter.
+While Kokkos is gaining popularity in large-scale HPC applications at some DOE laboratories, we believe that the implemented concepts are of interest to a broader audience including academia as they may contribute to a generic, vendor, and architecture-independent education of parallel programming.
+In this work, we give an insight into the design considerations of this programming model and list important abstractions.
+Further, we document best practices obtained from giving virtual classes on Kokkos, document attendee feedback, and give pointers to resources that the reader may consider valuable for a lecture on generic parallel programming for students with preexisting knowledge on this matter.
 \end{abstract}

### 2020-08-15 src/1.Introduction.tex
Updates for submission to EduHPC at SC20

@@
-Parallel and distributed computing is a focus at universities and teaching institutions.
+Parallel computing is an area of academic focus.
 The interest is two-fold.
@@
 
-While industry-driven collaborations often predefine technological frameworks such as programming environments and targeted architectures, academic teaching leaves this mostly undefined.
-In this case, educators can select from a variety of solutions, approaches, and trends in parallel computing.
+While industry-driven collaborations often predefine technological frameworks such as programming environments and target architectures, academic teaching leaves this mostly undefined.
+In this case, educators can select from a variety of architectures and programming models for their parallel computing curricula.
 Which particular set to choose for teaching in class is not trivial.
-Complexity is further amplified by the breath of related topics ranging from parallel hardware architectures, programming models, algorithms as well as productivity in parallel software development.
-However, it is the productivity concern that is gaining importance as the number of hardware solutions, their complexity as well as demands for a quick time-to-solution increase.
+While decision factors differ, maximizing the subject relevance by teaching generic principles of parallel programming is among the most important ones.
 
-While vendor programming models maintain their position in curricula today, we think that the inclusion of parallel programming models such as Kokkos is worth consideration.
-They offer an easy to understand set of concepts for a vendor- and architecture-independent, performance-portable expression of concurrency which significantly contributes to parallel programming productivity.
+Consequently, generic parallel programming models are desirable that target mainstream programming languages and allow a vendor- and architecture-independent, performance-portable expression of concurrency.
+In this way, a parallel program becomes capable of running efficiently on computer architectures that conform to a common parallel machine specification.
+We would like to compare this to the support of Von Neumann architectures in sequential programming languages today.
+
+While message-passing interfaces, OpenMP, OpenCL or CUDA maintain their role in teaching, the plethora of novel parallel programming models~\cite{Chapel}~\cite{UPC++}~\cite{Raja} hints towards an emphasis on productivity in parallel programming and also brings up the valid question whether this strengthens the argument for their inclusion in education.
+
+Based on our experience in supercomputing, we see that productivity is gaining importance as the number of hardware solutions, their complexity to program as well as demands for a quick time-to-solution increase.
+Consequently, we think that the inclusion of novel programming models is worth consideration in case they offer generic programming concepts for a vendor- and architecture-independent, performance-portable expression of concurrency and can contribute to parallel programming productivity.
+
+It is the productivity concern as well as the need for a generic expression of concurrency due to periodic changes of supercomputer architectures and vendors in national projects that inspired the development of Kokkos.
+The increasing adoption of Kokkos in these projects confirms its usefulness which motivates us to present Kokkos to academia in this work.
 
@@
 Further, we highlight its alignment and contributions to standardization efforts towards supporting parallel programming in C++.
-For teaching insights, we document feedback received during on-line lectures.
+For teaching insights, we document our best practices obtained from teaching Kokkos as well as feedback received during on-line lectures.
 
@@
 Chapters~\ref{chap:kokkosPM} and ~\ref{chap:kokkosExample} present the Kokkos programming model and an example application.
-In Chapter~\ref{chap:backend} we provide insight into a Kokkos back-end implementation.
-Chapters~\ref{chap:related} and ~\ref{chap:tutorials} discuss related work, tutorials and student feedback from tutorials.
+In Chapter~\ref{chap:backend} we provide insight into a Kokkos backend implementation.
+Chapters~\ref{chap:related} and ~\ref{chap:teaching} discuss related work, best practices in teaching, and student feedback from tutorials.
 Lastly Chapter~\ref{chap:conclusion} concludes this work.

### 2020-08-15 src/10.Conclusion.tex
Updates for submission to EduHPC at SC20

\section{Conclusion}\label{chap:conclusion}
In this work, we gave an introduction to the Kokkos parallel programming model and presented guiding thoughts behind its design. We discussed a machine representation of a parallel hardware architecture that is characterized by a variety of processing units, memories, and their hierarchical organization. This machine model allowed us to define a constrained semantic represented through common abstractions such as patterns, views, and spaces.

An example application gave insight into parallel programming with Kokkos and its OpenMP backend implementation. We used them to show the generic programming interface and type specialization of the implementing classes to support a particular vendor programming model.

To the educational community, we presented Kokkos as a representative programming model of a trend towards the support of parallel programming in base languages and libraries through a common vendor- and architecture-independent abstractions.

The recent shift towards online classes impacted our teaching of Kokkos as well. In this work, we presented our best-practices when conducting online classes including lecture structure, material preparation, and methodology.

Finally, we would like to express gratitude to educators of parallel programming and computer science for taking on this important and challenging role. In case of interest, we would gladly provide further material and support on Kokkos and offer to the interested reader to reach out to the development team on Slack\cite{KOKKOS_SLACK}.

### 2020-08-15 src/2.GenericProgramming.tex
Updates for submission to EduHPC at SC20

@@
 Absent a programming model providing additional constraints, the semantics of C-like languages can lead to sequential thinking and execution.
-Parallel programming models expose parallel semantics to enable developers to create performance-portable code.
+Parallel programming models expose parallel semantics to enable developers to add concurrency to their code.
 This is commonly achieved by adding annotations to the base programming language, often through \emph{pragma} clauses or similar language extensions.
@@
 Some representatives are the OpenMP programming model~\cite{OPENMP}, threading libraries, or languages like Erlang\cite{ERLANG} in which parallelism is a first-class citizen.
-In general, it is up to the parallel programming model implementer to expose language constructs or library interfaces (APIs) that combine convenience to the programmer with a sufficiently constrained semantics to map the program execution to the parallel computer hardware efficiently and correctly.
+In general, it is up to the parallel programming model implementer to expose language constructs or library interfaces (APIs) that combine convenience to the programmer with a sufficiently constrained semantics that is used to map the program execution to the parallel computer hardware efficiently and correctly.
 Questions of which semantic information is needed, what paradigm to present, and by what syntactic means to do so span an exciting design space that is well worth exploring.
@@
 The execution model and memory model aspects of a programming model define its behavior.
-That is, it defines the relationship between abstract concepts and program execution on the given architecture.
+That is, it defines the relationship between abstract concepts and program execution on a given hardware architecture.
 

### 2020-08-15 src/3.MachineModel.tex
Updates for submission to EduHPC at SC20

@@
 The Kokkos parallel programming model exposes these abstractions to the programmer through C++ and the Kokkos API.
-By abstracting from physical hardware, the machine model ensures that applications written in programming models targeting this machine model are generic and performance-portable on current and future hardware.
-The Kokkos parallel programming model is one particular instance of a programming model that builds on top of that machine model.
+By abstracting from physical hardware, the machine model ensures that applications written in any programming model that is targeting this machine model are generic and performance-portable on current and future hardware.
+Kokkos is one particular instance of a programming model that builds on top of that machine model.
 We would like to point out that it is this conceptual differentiation that allows scenarios where the underlying machine model is instantiated in other languages beyond C++, yet the algorithmic specification as well as performance characteristics and portability remain the same.
@@
 
-\includegraphics[width=0.5\textwidth]{img/KokkosSpaces.png}
+\includegraphics[width=0.49\textwidth]{img/KokkosSpaces.png}
+\centering
 \caption{Spaces represent conceptual building blocks of the abstract machine model in Kokkos.}

### 2020-08-15 src/4.ProgammingModel.tex
Updates for submission to EduHPC at SC20

@@
 The API exposes the machine representation to the developer and defines a set of patterns and types that capture intents and properties of execution.
-This completes the semantic capture, discussed in Chapter~\ref{chap:background}, We detail all three components as follows.
+This completes the semantic capture, discussed in Chapter~\ref{chap:background}.
+We detail all three components as follows.
 
@@
 This has multiple reasons.
-It is our experience that application developers of scientific and high-performance computing applications in the high-performance computing community and in the industry in general express the demand for C++ and a generic parallel programming model to support their codes.
+It is our experience that developers of scientific and high-performance computing applications express the demand for C++ and a generic parallel programming model to support their codes.
 Further, following their estimates, porting efforts towards adapting to vendors, programming models, APIs, and software releases represents a burden worth addressing.
@@
 Kokkos supports parallel patterns and tasking.
-Parallel patterns covers \emph{for}, \emph{scan} and \emph{reduction}.
+Parallel patterns cover \emph{for}, \emph{scan} and \emph{reduction}.
 This allows the expression of concurrency over iterative, for-loop computable algorithms.
@@
 To the interested reader, the complete programming model description can be accessed on-line\cite{KOKKOS_WIKI}.
-The next section shows an example of the Kokkos parallel\_for and explains how it is mapped to the underlying back-end programming model.
+The next section shows an example of the Kokkos parallel\_for and explains how it is mapped to the underlying backend programming model.
 
 \begin{figure}
-\centerline{\includegraphics[width=0.45\textwidth]{img/Abstractions.png}}
+\centerline{\includegraphics[width=0.49\textwidth]{img/Abstractions.png}}
 \caption{Overview of abstractions that define the constrained semantic of the Kokkos parallel programming model.}

### 2020-08-15 src/5.Example.tex
Updates for submission to EduHPC at SC20

@@
 In case two view objects are aliased, this operation results in a~\emph{no-op}.
-
-Consequently, the simple example does not explicitly specify a memory space where the data lives in. However, Kokkos ensures that the defaults for parallel execution and data allocations match.
+Consequently, the simple example does not explicitly specify a memory space where the data lives in.
+However, Kokkos ensures that the defaults for parallel execution and data allocations match.
 
@@
 While this example can only provide a brief insight into Kokkos, we recommend to the interested reader to access further resources.
-Resources include example applications for individual Kokkos features, tutorials with multiple days worth of lectures and exercises, a programming guide, and API documentation.
+Resources include example applications for each individual Kokkos feature, tutorials with multiple days worth of lectures and exercises, a programming guide, and API documentation.
 Source code examples as well as the tutorial are located in the Kokkos repository~\cite{KOKKOS_REPO}.
-The tutorial is intended for students with minimal or even no prior knowledge of parallel programming.
+The tutorial is intended for students with previous knowledge of one parallel programming model and basic knowledge of C++.
 It introduces concepts of parallel programming using Kokkos through a series of lectures and hands-on exercises that build upon each other and gradually introduce new concepts.
 The programming guide is available on-line~\cite{KOKKOS_WIKI}.
+In chapter~\ref{chap:teaching} we provide further details on best practices obtained from teaching Kokkos as well as list feedback received from lecture attendees.
 
@@
             y_tmp += A(e, i) * x(i);
-          }, 
-          y(e));
+          }, y(e));
     }); 
     Kokkos::fence();
-    Kokkos::deep_copy(y_h,y);
-    //output_result_on_host(y_h);
+    Kokkos::deep_copy(y_h,y); //copy to host memory space    
   }

### 2020-08-15 src/6.Backend.tex
Updates for submission to EduHPC at SC20

@@
-\section{Back-end Support}\label{chap:backend}
+\section{Backend Support}\label{chap:backend}
 
-In this section we peek inside the Kokkos library and show how a generic parallel for loop is mapped to the OpenMP backend programming model.
+In this section, we peek inside the Kokkos library and show how a generic parallel for loop is mapped to the OpenMP backend programming model.
 As a template metaprogramming library, Kokkos makes use of partial specialization.
-In particular the~\emph{parallel\_for} or ~\emph{parallel\_reduce} functional calls as shown in Figure~\ref{fig:KokkosExample} are mapped to partially specialized classes.
-Such classes correspond to a particular back-end where for each back-end, concepts with partial specialization must be provided.
-Today, Kokkos supports the CUDA, OpenMP, HIP as well as one-sided MPI, SHMEM and NVSHMEM back-ends.
-For each new back-end, partially implemented classes that implement necessary concepts must be provided.
+In particular the~\emph{parallel\_for} or ~\emph{parallel\_reduce} function calls, as shown in Figure~\ref{fig:KokkosExample}, are mapped to partially specialized classes.
+Such classes correspond to a particular backend where for each backend, concepts with partial specialization must be provided.
+Today, Kokkos supports the CUDA, OpenMP, HIP as well as one-sided MPI, SHMEM and NVSHMEM backends.
+For each new backend, partially implemented classes that implement necessary concepts must be provided.
 
-A simplified version of a partially specialized class for a parallel for loop implementation is shown in Figure\ref{fig:KokkosExampleOMPBackEnd}.
+A simplified version of a partially specialized class for a parallel for-loop implementation in the OpenMP backend is shown in Figure\ref{fig:KokkosExampleOMPBackEnd}.
 The ~\emph{parallel\_for} function instantiates a partial specialization of the ~\emph{ParallelFor} class and calls its ~\emph{execute} function.
 That function is then responsible to implement the Kokkos parallel pattern.
-In this back-end, the parallel implementation is generated by using OpenMP pragma annotations and a by invoking a compatible compiler when the library is compiled.
+In this backend, the parallel implementation is generated by using OpenMP pragma annotations and a by invoking a compatible compiler when the library is compiled.
+Figure~\ref{fig:stack} gives an architectural overview of Kokkos and lists important abstractions as well as the currently supported backends.
 \begin{figure}
@@
 class ParallelFor<FunctorType,
-              RangePolicy<OpenMP>> 
-{
+              RangePolicy<OpenMP>> {
   const FunctorType functor;
@@
 \end{small}
-\caption{This code shows the back-end implementation of a parallel loop in the Kokkos programming model.
+\caption{This code shows the backend implementation of a parallel loop in the Kokkos programming model.
 It shows a particular implementation of a generic pattern as a templated class to the OpenMP execution space.
@@
 \end{figure}
+
+\begin{figure}
+\centerline{\includegraphics[width=0.42\textwidth]{img/Stack.png}}
+\caption{Overview of the Kokkos programming model shows important abstractions as well as the inclusion of backend programming models used by the runtime to map higher level concepts to concrete hardware architectures.}
+\label{fig:stack}
+\end{figure}

### 2020-08-15 src/8.Related.tex
Updates for submission to EduHPC at SC20

@@
 In courses that do not target accelerators, this can be a relevant option.
-
 Over the past decade the Kokkos programming model has contributed to a number of current and near-future ISO-C++ features, including \emph{atomic\_ref}\cite{wg21_p0019}, \emph{mdspan}\cite{MDSPAN}, and executors.
@@
 As the demand to extract performance from increasingly deep and increasingly asynchronous software stacks across a wide variety of domains increases, many C++ experts expect executors to become a central abstraction in any performance-oriented software stack.
-
-RAJA~\cite{RAJA} is a C++ abstraction library with similarities to Kokkos.
+RAJA~\cite{Raja} is a C++ abstraction library with similarities to Kokkos.
 The important difference however is that Kokkos is descriptive, while RAJA is prescriptive.
-In Kokkos, the programming models determine how an application is mapped to the underlying hardware.
+In Kokkos, the programming model determines how an application is mapped to the underlying hardware.
 RAJA provides functionality that exposes hardware details and relies on the developer to try different strategies to map kernels to architectures.
 RAJA is a viable option for class education in which the educator would like to expose students to the semantics of different programming models without exposing them to the detailed syntax of those models.
-
-DPC++\cite{DPCPP} is a model developed by Intel\textsuperscript{\textregistered} for expressing parallelism across Intel architectures.
+DPC++\cite{DPCPP} is a model developed by Intel\textsuperscript{\textregistered} for expressing parallelism across Intel hardware architectures.
 In addition to sharing a descriptive philosophy with Kokkos, it also shares parallel patterns.
-Unlike Kokkos, it targets Intel architectures only and is not vendor-agnostic.
+Unlike Kokkos however, it targets Intel architectures only and is not vendor-agnostic.
 A course targeting Intel architectures, DPC++ could be a valid option.

### 2020-08-15 src/9.Teaching.tex
Updates for submission to EduHPC at SC20

\section{Virtual Teaching Experiences}\label{chap:teaching}

In our experience, teaching Kokkos benefits from a set of good practices that we document in this section. We exclude obvious practices such as good documentation, accessibility, or responsiveness to user questions. Our best practices can be divided into accessibility and organization of source code distribution, presentation material, and presentation methodology.

Our Kokkos source code is accessible through GitHub and supports a variety of build systems including Makefile, Cmake as well as Spack\cite{SPACK}. This variety in build system support makes the code easily accessible and executable to a wider range of users who are familiar with one of these build systems. Further, the Kokkos source code directly includes an example for each feature. The inclusion of examples in the source code distribution is beneficial for quick look-up of syntax, APIs, and usage patterns. All examples are embedded into small but complete test applications as this unburdens the interested programmer from any additional programming. For completeness, examples include codes that illustrate important performance implications of certain properties such as the definition of access patterns in views that differ between host and accelerator devices (strided versus coalesced).

Our presentation and tutorial material is divided into 8 units that gradually roll out all features and use-cases of the Kokkos programming model. Our tutorial material covers approximately 8 two-hour lectures, where each lecture summarizes the key learning points of the previous lecture at the beginning, summarizes the current lecture at the end, and gives a preview on the upcoming lecture. This creates continuity, improves the attendee's orientation in the subject, and spurs interest. Each lecture is concluded with a call for experimentation. Each lecture has a prepared hands-on exercise that can be used either in-lined during presential lectures or given out as homework for individual experimentation. During online lectures, the presenter explains the solution of the homework through screen-sharing and repeats important concepts. We would like to point out that each homework assignment contains the solution. Providing a solution lowers the barrier of experimentation and helps impatient students.

Online lectures tend to restrict possible interaction between the presenter and attendees to chats. In turns out useful to assign assistant presenters to actively participate in the chat during the presentation. This helps to resolve questions and support audience engagement. Since some chat conversations tend to fall short, it is a good practice to save the chat transcript and to publish the transcript together with refined responses on-line where they can be accessed by attendees for reference. To help interested attendees, we offer consulting hours. Consulting hours are hosted as online telephone conferences where attendees can call in and ask further questions regarding lecture content or homework.

The advantage of virtual classes is their possibility of being recording. Recordings can serve as a valuable resource for future reference with quick access links pointing to particular moments in the recording relevant to a topic.

A recent online course\cite{KOKKOS_LECTURE} has been publicly announced through social media and counted approximately 200 attendees. The majority of attendees are familiar with at least one parallel programming model and have a background in scientific computing. The listing in Figure~\ref{fig:lectureOutline} shows the structure of our on-line lecture.

Question asked by attendees during the recent online course showed that they did understand the relevance as well as key concepts. Questions were mostly related to details in behavior, differences to other programming models and best practices. We would like to direct the interested reader to the our online website where the Q\&A sessions are listed in detail.

\begin{figure}

\begin{itemize}
\item Introduction, Building and Parallel Dispatch
\item Views and Spaces
\item Data Structures + MultiDimensional Loops
\item Hierarchical Parallelism
\item Tasking, Streams and SIMD
\item Internode: MPI and PGAS
\item Tools: Profiling, Tuning and Debugging
\item Kernels: Sparse and Dense Linear Algebra
\end{itemize}
\caption{Outline of the Kokkos on-line lecture series with 8 two-hour sessions incrementally cover all concepts and features.}
\label{fig:lectureOutline}
\end{figure}

### 2020-08-24 src/0.Abstract.tex
Updates for EduHPC2020.

@@
 In this work, we give an insight into the design considerations of this programming model and list important abstractions.
-Further, we document best practices obtained from giving virtual classes on Kokkos, document attendee feedback, and give pointers to resources that the reader may consider valuable for a lecture on generic parallel programming for students with preexisting knowledge on this matter.
+Further, we document best practices obtained from giving virtual classes on Kokkos and give pointers to resources that the reader may consider valuable for a lecture on generic parallel programming for students with preexisting knowledge on this matter.
 \end{abstract}

### 2020-08-24 src/1.Introduction.tex
Updates for EduHPC2020.

@@
 
-While message-passing interfaces, OpenMP, OpenCL or CUDA maintain their role in teaching, the plethora of novel parallel programming models~\cite{Chapel}~\cite{UPC++}~\cite{Raja} hints towards an emphasis on productivity in parallel programming and also brings up the valid question whether this strengthens the argument for their inclusion in education.
+While message-passing, OpenMP, OpenCL or CUDA maintain their role in teaching, the plethora of novel parallel programming models~\cite{Chapel}~\cite{UPC++}~\cite{Raja} hints towards an emphasis on productivity in parallel programming and also brings up the valid question whether this strengthens the argument for their inclusion in education.
 
@@
 
-It is the productivity concern as well as the need for a generic expression of concurrency due to periodic changes of supercomputer architectures and vendors in national projects that inspired the development of Kokkos.
+It is the productivity concern as well as the need for a generic expression of concurrency due to periodic changes of supercomputer architectures and their vendors in national projects that inspired the development of Kokkos.
 The increasing adoption of Kokkos in these projects confirms its usefulness which motivates us to present Kokkos to academia in this work.
@@
 In Chapter~\ref{chap:backend} we provide insight into a Kokkos backend implementation.
-Chapters~\ref{chap:related} and ~\ref{chap:teaching} discuss related work, best practices in teaching, and student feedback from tutorials.
+Chapters~\ref{chap:related} and ~\ref{chap:teaching} discuss related work and best practices for teaching in online classes.
 Lastly Chapter~\ref{chap:conclusion} concludes this work.

### 2020-08-24 src/9.Teaching.tex
Updates for EduHPC2020.

@@
 The advantage of virtual classes is their possibility of being recording.
-Recordings can serve as a valuable resource for future reference with quick access links pointing to particular moments in the recording relevant to a topic.
+Recordings can serve as a valuable resource for future reference with quick access links pointing to particular timestamps in the recording relevant to a topic.
 
@@
 The majority of attendees are familiar with at least one parallel programming model and have a background in scientific computing.
-The listing in Figure~\ref{fig:lectureOutline} shows the structure of our on-line lecture.
+The listing in Figure~\ref{fig:lectureOutline} shows the structure of our online lecture.
 

### 2020-10-07 src/6.Backend.tex
EduHPC @ SC2020 Camera Ready

@@
 
-A simplified version of a partially specialized class for a parallel for-loop implementation in the OpenMP backend is shown in Figure\ref{fig:KokkosExampleOMPBackEnd}.
+A simplified version of a partially specialized class for a parallel for-loop implementation in the OpenMP backend is shown in Figure~\ref{fig:KokkosExampleOMPBackEnd}.
 The ~\emph{parallel\_for} function instantiates a partial specialization of the ~\emph{ParallelFor} class and calls its ~\emph{execute} function.
@@
 In this backend, the parallel implementation is generated by using OpenMP pragma annotations and a by invoking a compatible compiler when the library is compiled.
-Figure~\ref{fig:stack} gives an architectural overview of Kokkos and lists important abstractions as well as the currently supported backends.
+Figure ~\ref{fig:stack} gives an architectural overview of Kokkos and lists important abstractions as well as the currently supported backends.
 \begin{figure}

### 2020-10-07 src/9.Teaching.tex
EduHPC @ SC2020 Camera Ready

@@
 Our presentation and tutorial material is divided into 8 units that gradually roll out all features and use-cases of the Kokkos programming model.
-Our tutorial material covers approximately 8 two-hour lectures, where each lecture summarizes the key learning points of the previous lecture at the beginning, summarizes the current lecture at the end, and gives a preview on the upcoming lecture.
+Our tutorial material covers approximately 8 two-hour lectures, where each lecture summarizes the key learning points of the previous lecture at the beginning, summarizes the current lecture at the end, and gives a preview on the upcoming one.
 This creates continuity, improves the attendee's orientation in the subject, and spurs interest.
 Each lecture is concluded with a call for experimentation.
-Each lecture has a prepared hands-on exercise that can be used either in-lined during presential lectures or given out as homework for individual experimentation.
-During online lectures, the presenter explains the solution of the homework through screen-sharing and repeats important concepts.
-We would like to point out that each homework assignment contains the solution.
+
+Each unit has a prepared hands-on exercise that can be used either in-lined during presential classes or given out as homework for individual experimentation.
+During the online presentation, the presenter explains the solution of the homework through screen-sharing and repeats important concepts.
+We would like to point out that each homework assignment contains the solution in a separate directory.
 Providing a solution lowers the barrier of experimentation and helps impatient students.
 
-Online lectures tend to restrict possible interaction between the presenter and attendees to chats.
+Online classes tend to restrict possible interaction between the presenter and attendees to chats.
 In turns out useful to assign assistant presenters to actively participate in the chat during the presentation.
 This helps to resolve questions and support audience engagement.
-Since some chat conversations tend to fall short, it is a good practice to save the chat transcript and to publish the transcript together with refined responses on-line where they can be accessed by attendees for reference.
+Since some chat conversations tend to fall short, it is a good practice to save the chat transcript and to publish the transcript together with refined responses online where they can be accessed by attendees for reference.
 To help interested attendees, we offer consulting hours.
@@
 The majority of attendees are familiar with at least one parallel programming model and have a background in scientific computing.
-The listing in Figure~\ref{fig:lectureOutline} shows the structure of our online lecture.
+The listing in Figure~\ref{fig:lectureOutline} shows the structure of our online lectures.
 
-Question asked by attendees during the recent online course showed that they did understand the relevance as well as key concepts.
-Questions were mostly related to details in behavior, differences to other programming models and best practices.
-We would like to direct the interested reader to the our online website where the Q\&A sessions are listed in detail.
+Questions asked by attendees during the recent online course showed that they did understand the relevance as well as key concepts.
+Questions were mostly related to details in behavior, differences to other programming models, and best practices.
+We would like to direct the interested reader to our online website where the Q\&A sessions are listed in detail.
 

### 2020-10-13 src/2.GenericProgramming.tex
Cam-ready version.

@@
 
-\begin{figure}[h]
+\begin{figure}[t!]
 \begin{minted}[fontsize=\small]{text}
@@
 \end{minted}
-\caption{Semantic capture: a set of constrained semantics, parallel paradigms, expression and concern define key properties of a parallel programming model.}
+\caption{Semantic capture: a set of constrained semantics, parallel paradigms, expression, and control paradigm define key properties of a parallel programming model.}
 \label{figSemCapture}
@@
 \emph{Constrained Semantics} provided by the programmer to the parallel programming model can be grouped into information to express intent or correctness and properties that control critical performance aspects.
-They address the question of \emph{what}, \emph{where} and \emph{how}.
+They address the question of \emph{what}, \emph{where}, and \emph{how}.
 In the context of parallel programming, these correspond to defining which code portion to parallelize, where to run the code, where to access the data, and how to run that parallel code.
@@
 To define what further information is needed to express concurrency on modern computer architectures, it is important to define an~\emph{abstract machine representation} first.
-In the following chapters we use the words \emph{representation} and \emph{model} interchangeably.
+In the following chapters, we use the words \emph{representation} and \emph{model} interchangeably.
 In this work, the abstract machine model does not describe a computer in terms of automata theory but in terms of physical or functional components.
 
-\begin{figure}
+\begin{figure}[t!]
 \begin{minted}[fontsize=\small]{c++}
@@
 
-\begin{figure}
+\begin{figure}[t!]
 \begin{minted}[fontsize=\small]{c++}

### 2020-10-13 src/3.MachineModel.tex
Cam-ready version.

@@
 In such an architecture, execution units might be hierarchically organized with multiple levels of achievable parallelism with different memory access characteristics and coherence properties of caches.
-In order to ensure performance portability on such a wide range of configurations, an abstraction of compute-resources and available memories is required.
+To ensure performance portability on such a wide range of configurations, an abstraction of compute-resources and available memories is required.
 For this purpose, we introduce the abstract concept called \emph{space}.
@@
 An instance of a memory space provides a concrete method for the application programmer to request data storage allocations.
-Following the architecture as discussed earlier, the multi-core processor may contain multiple memory spaces that abstract on-package memory, slower DRAM, and non-volatile memories.
+Following the architecture, as discussed earlier, the multi-core processor may contain multiple memory spaces that abstract on-package memory, slower DRAM, and non-volatile memories.
 Accelerators can provide an additional memory space through its local on-package memory.
@@
 
-\begin{figure}
+\begin{figure}[t!]
 

### 2020-10-13 src/4.ProgammingModel.tex
EduHPC @ SC2020 Camera Ready
Cam-ready version.

@@
 It is our experience that developers of scientific and high-performance computing applications express the demand for C++ and a generic parallel programming model to support their codes.
-Further, following their estimates, porting efforts towards adapting to vendors, programming models, APIs, and software releases represents a burden worth addressing.
+Further, following their estimates, porting efforts towards adapting to vendors, programming models, APIs, and software releases represent a burden worth addressing.
 Lastly, the possibility to maintain pure C++ codes is appealing from the code development and debugging perspective.
@@
 
-To capture semantic information, that is the~\emph{what}, the~\emph{where} and the~\emph{how}, Kokkos introduces six abstractions:~\emph{execution spaces},~\emph{execution patterns},~\emph{execution policies},~\emph{memory spaces},~\emph{memory layout} and~\emph{memory traits}.
+To capture semantic information, that is the~\emph{what}, the~\emph{where}, and the~\emph{how}, Kokkos introduces six abstractions:~\emph{execution spaces},~\emph{execution patterns},~\emph{execution policies},~\emph{memory spaces},~\emph{memory layout}, and~\emph{memory traits}.
 These abstractions specify semantic information that enables to capture the programmer's intent and allows the runtime to efficiently map the program to any underlying hardware architecture.
@@
 	\item Execution patterns expose the parallel programming paradigm.
-Supported patterns are \emph{parallel\_for} loop that executes the loop body in any order a specified amount of times, the \emph{parallel\_reduce} which combines a parallel\_for with a reduction operation, \emph{parallel\_scan} which combines a parallel\_for operation with a prefix or postfix scan, and \emph{task} which executes a single function potentially in parallel in respect to other tasks or code sections.
+Supported patterns are the \emph{parallel\_for} loop that executes the loop body in any order a specified amount of times, the \emph{parallel\_reduce} which combines a parallel\_for with a reduction operation, the \emph{parallel\_scan} which combines a parallel\_for operation with a prefix or postfix scan, and \emph{task} which executes a single function potentially in parallel with respect to other tasks or code sections.
 	\item Execution policies shape the iteration space of a loop pattern. A simple execution policy is a range policy. It specifies that the loop body is executed once for each element in a range.
@@
 
-\begin{figure}[h]
-\begin{minted}[fontsize=\small]{text}
-Constrained Semantics: Patterns (intent),
-Spaces, Layouts, Policies and Traits
-Parallel Paradigms: Parallel patterns and 
-tasking
-Expression: Embedded, C++
-Control Paradigm: descriptive
-\end{minted}
-\caption{Semantic capture defined by the Kokkos programming model.}
-\label{fig:SemCaptureKokkos}
-\end{figure}
-
-\begin{figure}
+\begin{figure}[t!]
 \begin{minted}[fontsize=\small]{c++}
@@
 
+\begin{figure}[t!]
+\begin{minted}[fontsize=\small]{text}
+Constrained Semantics: Patterns (intent),
+Spaces, Layouts, Policies and Traits
+Parallel Paradigms: Parallel patterns and 
+tasking
+Expression: Embedded, C++
+Control Paradigm: descriptive
+\end{minted}
+\caption{Semantic capture defined by the Kokkos programming model.}
+\label{fig:SemCaptureKokkos}
+\end{figure}
+
 The semantic of the parallel\_for class in Kokkos demonstrates the descriptive nature of the programming model.
@@
 
-\begin{figure}
+\begin{figure}[t!]
 \centerline{\includegraphics[width=0.49\textwidth]{img/Abstractions.png}}

### 2020-10-13 src/5.Example.tex
Cam-ready version.

@@
 
-In this section we provide an example of an application written in the Kokkos programming model.
+In this section, we provide an example of an application written in the Kokkos programming model.
 The code in Figure~\ref{fig:KokkosExample} shows a parallel matrix-vector multiplication implemented as two nested loops.
@@
 While this example can only provide a brief insight into Kokkos, we recommend to the interested reader to access further resources.
-Resources include example applications for each individual Kokkos feature, tutorials with multiple days worth of lectures and exercises, a programming guide, and API documentation.
+Resources include example applications for each Kokkos feature, tutorials with multiple days worth of lectures and exercises, a programming guide, and API documentation.
 Source code examples as well as the tutorial are located in the Kokkos repository~\cite{KOKKOS_REPO}.
@@
 
-\begin{figure}
+\begin{figure}[t!]
 \begin{small}

### 2020-10-13 src/8.Related.tex
Cam-ready version.

@@
 In courses that do not target accelerators, this can be a relevant option.
-Over the past decade the Kokkos programming model has contributed to a number of current and near-future ISO-C++ features, including \emph{atomic\_ref}\cite{wg21_p0019}, \emph{mdspan}\cite{MDSPAN}, and executors.
+Over the past decade, the Kokkos programming model has contributed to a number of current and near-future ISO-C++ features, including \emph{atomic\_ref}\cite{wg21_p0019}, \emph{mdspan}\cite{MDSPAN}, and executors.
 In particular, the design of C++ executors, which are similar to Kokkos execution spaces, were influenced by this programming model.

### 2020-10-15 src/5.Example.tex
Final cam-ready version.

@@
       [=](const member_type &team_handle) {
-        const int e = team_handle.league_rank();
+        const int e=team_handle.league_rank();
         Kokkos::parallel_reduce( 
           TeamThreadRange(team_handle, M),
-          [=](const int & i, double & y_tmp) {
+          [=](const int & i, double & y_tmp){
             y_tmp += A(e, i) * x(i);
@@
     Kokkos::fence();
-    Kokkos::deep_copy(y_h,y); //copy to host memory space    
+    //copy to host memory space    
+    Kokkos::deep_copy(y_h,y); 
   }
